### 2020-11-10 docs/manuscript.tex
initial docs commit

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[fleqn,11pt]{SelfArx} % Document font size and equations flushed left

\usepackage{parskip}
\usepackage{microtype}
\usepackage{svg}
\usepackage{float}
\usepackage{placeins}
\usepackage[english]{babel} % Specify a different language here - english by default
\usepackage{listings}
\usepackage{fontspec}
\usepackage{transparent}
% \usepackage{lipsum} % Required to insert dummy text. To be removed otherwise
\setmainfont{TeX Gyre Termes}
%----------------------------------------------------------------------------------------
%	COLUMNS
%----------------------------------------------------------------------------------------

\setlength{\columnsep}{0.55cm} % Distance between the two columns of text
\setlength{\fboxrule}{1pt} % Width of the border around the abstract

%----------------------------------------------------------------------------------------
%	COLORS
%----------------------------------------------------------------------------------------

\definecolor{color1}{RGB}{124,0,0} % Color of the article title and sections
\definecolor{color2}{RGB}{150,150,150} % Color of the boxes behind the abstract and headings

%----------------------------------------------------------------------------------------
%	HYPERLINKS
%----------------------------------------------------------------------------------------

\usepackage{hyperref} % Required for hyperlinks
\hypersetup{hidelinks,colorlinks,breaklinks=true,urlcolor=color2,citecolor=color1,
  linkcolor=color1,bookmarksopen=false,pdftitle={Title},pdfauthor={Author}}

%----------------------------------------------------------------------------------------
%	ARTICLE INFORMATION
%----------------------------------------------------------------------------------------

\PaperTitle{
    Predicting HLA associated mutations with a \\ Bayesian regression model
} % Article title

\Authors{} % Authors
% \affiliation{\textsuperscript{1}\textit{Department of Biology, University of Examples, London, United Kingdom}} % Author affiliation
% \affiliation{\textsuperscript{2}\textit{Department of Chemistry, University of Examples, London, United Kingdom}} % Author affiliation
% \affiliation{*\textbf{Corresponding author}: john@smith.com} % Corresponding author

\Keywords{} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

%----------------------------------------------------------------------------------------
%	ABSTRACT
%----------------------------------------------------------------------------------------

\Abstract{

}

%----------------------------------------------------------------------------------------
\begin{document}

\flushbottom % Makes all text pages the same height

\maketitle % Print the title and abstract box

{
  \hypersetup{linkcolor=black}
  \tableofcontents
}

\thispagestyle{empty} % Removes page numbering from the first page
%----------------------------------------------------------------------------------------
%	ARTICLE CONTENTS
%----------------------------------------------------------------------------------------

\FloatBarrier

%----------------------------------------------------------------------------------------
%	REFERENCE LIST
%----------------------------------------------------------------------------------------
\phantomsection
\bibliographystyle{unsrt}

%----------------------------------------------------------------------------------------

\end{document}

### 2020-11-10 docs/manuscript.tex
progress on title page

@@
 \PaperTitle{
-    Predicting HLA associated mutations with a \\ Bayesian regression model
+    HAMFinder: Combining information to predict HLA-associated mutations with a
+    Bayesian regression model
 } % Article title
 
-\Authors{} % Authors
-% \affiliation{\textsuperscript{1}\textit{Department of Biology, University of Examples, London, United Kingdom}} % Author affiliation
+\Authors{Daniel Habermann\textsuperscript{1}, Daniel Hoffmann\textsuperscript{1}} % Authors
+\affiliation{\textsuperscript{1}\textit{Department of Biology, University of Duisburg-Essen, Essen, Germany}} % Author affiliation
 % \affiliation{\textsuperscript{2}\textit{Department of Chemistry, University of Examples, London, United Kingdom}} % Author affiliation
@@
 
-\Keywords{} % Keywords - if you don't want any simply remove all the text between the curly brackets
+\Keywords{human leucocyte anigen system, multiple sequence alignment, escape mutations, 
+  Bayesian inference, sparsity, horseshoe, epitope prediction} % Keywords - if you don't want any simply remove all the text between the curly brackets
 \newcommand{\keywordname}{Keywords} % Defines the keywords heading name
@@
 \Abstract{
+  % - what are HLA associated mutations and why identifying HLA escape is important
+  % 
+  % - current methods lack important aspects
+  % - Progress in Bayesian methods are not reflected in current HLA models
+  % - Bayesian statistics provides a coherent framework to combine information from different sources
+  % - Using this kind of imformation and new methods improve predicted performance tremendously
+  % -
+  \\
+  \textbf{Motivation} \\
+  The human leucoyte antigen system (HLA) is of paramount importance to combat viral
+  infections by presenting peptides on the cell surface via MHC I.
+  In this way, CD8+ cytotoxic T-Lymphocytes exert a strong selection pressure towards
+  virus variants that escape that immune recognition pathway, e.g. through point 
+  mutations that decrease binding of the respective peptide to MHC I. \\
 
+  Reliably identifying HLA-associated mutations is important for understanding viral
+  evolution, but experimental methods like binding assays are prohibitively expensive
+  for large-scale use and fail to recognize other mechanisms of immune escape like
+  proteasomal processing.
+
+  One step in finding these mutations is through the statistical analysis
+  of sequence data. However, existing methods are based on nullhypothesis significance 
+  testing and do not make use of all the available information and therefore have poor
+  real-world performance. \\
+
+  \textbf{Results} \\
+  Here, we present a Bayesian regression model that is easily extensible to include 
+  information from different sources (e.g. epitope prediction software) 
+  and makes use of recent advances in Bayesian inference, e.g. by using a sparsifying 
+  prior. We show that including this kind of information improves predictive performance 
+  considerably over state-of-the-art methods. \\
+
+  \textbf{Availability and Implementation} \\
+  The source code of this software is available at 
+  \url{http://gogs.uni-due.de/habermann/Escape.git} under a permissive MIT license. \\
+
+  \textbf{Supplementary information} \\
+  \url{google.de}[Supplementary data] are available at \textit{Bioinformatics} online. \\
 }
@@
 }
+
+\pagebreak
+
+\section{Introduction}
+\section{System and methods}
+\section{Algorithm}
+\section{Implementation}
+\section{Discussion}
+\section{References}
 

### 2020-11-10 docs/manuscript.tex
update abstract

@@
 \usepackage{fontspec}
-\usepackage{transparent}
+\usepackage{hyperref}
 % \usepackage{lipsum} % Required to insert dummy text. To be removed otherwise
@@
 \Authors{Daniel Habermann\textsuperscript{1}, Daniel Hoffmann\textsuperscript{1}} % Authors
-\affiliation{\textsuperscript{1}\textit{Department of Biology, University of Duisburg-Essen, Essen, Germany}} % Author affiliation
+\affiliation{\textsuperscript{1}\textit{Bioinformatics \& Computational Biophysics, Faculty of Biology, University of Duisburg-Essen, 45117 Essen, Germany}} % Author affiliation
 % \affiliation{\textsuperscript{2}\textit{Department of Chemistry, University of Examples, London, United Kingdom}} % Author affiliation
@@
   \textbf{Motivation} \\
-  The human leucoyte antigen system (HLA) is of paramount importance to combat viral
+  The human leucocyte antigen system (HLA) is of paramount importance to combat viral
   infections by presenting peptides on the cell surface via MHC I.
@@
   of sequence data. However, existing methods are based on nullhypothesis significance 
-  testing and do not make use of all the available information and therefore have poor
-  real-world performance. \\
+  testing and do not make use of all the available information and therefore have
+  unsatisfactory real-world performance. \\
 
@@
   \textbf{Supplementary information} \\
-  \url{google.de}[Supplementary data] are available at \textit{Bioinformatics} online. \\
+  \href{https://google.com}{Supplementary data} are available at \textit{Bioinformatics} online. \\
 }
@@
 \section{Introduction}
+
 \section{System and methods}

### 2020-11-11 docs/manuscript.tex
progress on introduction

@@
 \PaperTitle{
-    HAMFinder: Combining information to predict HLA-associated mutations with a
+    HAMDetector: Combining information to predict HLA-associated mutations with a
     Bayesian regression model
@@
 
-\Keywords{human leucocyte anigen system, multiple sequence alignment, escape mutations, 
-  Bayesian inference, sparsity, horseshoe, epitope prediction} % Keywords - if you don't want any simply remove all the text between the curly brackets
+\Keywords{human leucocyte antigen system, HLA, multiple sequence alignment, escape mutations, 
+  viral escape, Bayesian inference, sparsity, horseshoe, epitope prediction} % Keywords - if you don't want any simply remove all the text between the curly brackets
 \newcommand{\keywordname}{Keywords} % Defines the keywords heading name
@@
 \Abstract{
-  % - what are HLA associated mutations and why identifying HLA escape is important
-  % 
-  % - current methods lack important aspects
-  % - Progress in Bayesian methods are not reflected in current HLA models
-  % - Bayesian statistics provides a coherent framework to combine information from different sources
-  % - Using this kind of imformation and new methods improve predicted performance tremendously
-  % -
   \\
@@
 
+% - subset of peptidome presented on the cell surface via MHC I.
+% - Cytotoxic t cells are selected to not recognize self.
+% - by rule of exclusion, recognized peptides are foreign
+% - t cells induce cytolytic activity and recruit other immune cells, reference
+
+% - basic function of statistical analysis of sequence data
+% - state of the art + brumme paper, phylogeny, linkage disequilibrium and codon covariation
+% cite brumme and other paper
+
+% - problems of p value based methods and statistical significance testing
+% - highlight benefits of Bayesian approaches
+\subsection{The HLA system}
+
+One way how the human immune system is able to recognize intracellular viral infections
+is through the human leucocyte antigen system:
+In cells with active protein biosynthesis, proteins are continuously synthesized
+and also degraded by a process called proteasomal degradation, which cleaves
+proteins into linear peptides of varying length.
+A small subset of these peptides is presented on the cell surface via
+receptors called MHC class I. The genomic region encoding for MHC I is known to be
+highly polymorphic, with more than 20000 different HLA alleles described today.
+The resulting gene products differ in their binding properties, which means that
+cells from different individuals present a highly diverse set of peptides on their
+surface.
+Cytotoxic T cells are selected during maturation to only weakly bind to 
+peptide/MHC I complexes when the peptide originated from proteins of the usual proteome, 
+but might be able to strongly bind to complexes of MHC I with peptides which are 
+generated through proteosomal degradation of a viral protein.
+Upon activation, T cells induce cytolytic activity and recruit other immune cells.
+
+\subsection{HLA escape}
+In this way, the HLA system exerts a strong selection pressure towards virus variants
+that escape T cell recognition, for example through a point mutation
+that results in reduced binding of an immunogenic peptide to MHC I or through a
+set of mutations that alters the viral protein in such a way that it is cleaved into 
+different peptides that are not recognized by the host's T cell repertoire.
+
+The evolutionary events are complex and occur not only on the level of individuals, 
+where a virus adapts to specific features of the host, but also on the population level,
+because HLA alleles differ in their frequency across geographic region, as they are inherited
+according to standard Mendelian rules.
+
+Upon transmission, HLA escape mutations typically quickly revert to their wild type
+as they usually reduce viral replicative capacity (if a mutation would increase
+viral replicative capacity regardless of the presence of a given HLA allele, it would
+probably already be the wild type). Kawashima et al
+describe an escape mutation that is selected by HLA allele B\*53,
+does not strongly affect viral replicative capacity and therefore slowly enriches
+over time in Japan, where B\*53 commonly occurs.
+How quickly a given escape mutation is selected upon transmission in a host depends on the
+magnitude of the reduction in viral replicative capacity, on the strength of selection
+pressure and also on the genetic background, e.g. some escape mutations require compensatory 
+mutations which partly attenuate the negative impact on viral replicative capacity.
+
+Studying HLA escape therefore provides the unique opportunity to gain insight into
+viral evolution, on the host level but also on the population level.
+Unfortunately, identifying HLA escape mutations is difficult in practice.
+
+\subsection{Identifying HLA-escape mutations}
+
 \section{System and methods}

### 2020-11-12 docs/manuscript.tex
add 'Identifying HLA-escape mutations' section

@@
 but might be able to strongly bind to complexes of MHC I with peptides which are 
-generated through proteosomal degradation of a viral protein.
+generated from of a viral protein.
 Upon activation, T cells induce cytolytic activity and recruit other immune cells.
@@
 where a virus adapts to specific features of the host, but also on the population level,
-because HLA alleles differ in their frequency across geographic region, as they are inherited
+because HLA alleles differ in their frequency across geographic regions, as they are inherited
 according to standard Mendelian rules.
@@
 viral replicative capacity regardless of the presence of a given HLA allele, it would
-probably already be the wild type). Kawashima et al
+probably already be the wild type). Kawashima et al.
 describe an escape mutation that is selected by HLA allele B\*53,
@@
 
-Studying HLA escape therefore provides the unique opportunity to gain insight into
+Studying HLA escape therefore provides an unique opportunity to gain insight into
 viral evolution, on the host level but also on the population level.
@@
 \subsection{Identifying HLA-escape mutations}
+
+There are several experimental methods available to study HLA escape:
+Recombinant MHC-I molecules can be used in binding assays: Upon complex formation
+with a peptide, a change in conformation can be detected with conformation-specific
+antibodies. This method is relatively fast, but only measures binding affinity of a
+peptide to MHC-I and does not account for antigen processing or immunodominance, 
+which describes the observation that a peptide may be presented via MHC-I on the cell surface but does
+not induce an immune response.
+An experimental setup that resembles the conditions in-vitro more closely but is also
+more time-consuming is to measure
+CD8+ T cell responses instead. This is usually done by stimulating PBMCs with prototype
+and variant peptides and measuring the secretion of IFN-gamma by intracellular cytokine
+staining and fluorescence-assisted cell sorting.
+To analyse CD8+ T cell responses against endogenously processed antigens it is necessary
+to generate cell-lines stably expressing the antigen in question and adding antigen-specific
+CD8+ T cells. This method scales poorly as it requires transfection of cell lines and
+antigen-specific expansion of CD8+ T cells.
 

### 2020-11-14 docs/manuscript.tex
add 'Issues with using p-values as a screening tool' subsection

@@
 \PaperTitle{
-    HAMDetector: Combining information to predict HLA-associated mutations with a
+    HAMDetector: Combining information to detect HLA-associated mutations with a
     Bayesian regression model
@@
 
+\subsection{Computational methods}
+
+Because the currently available experimental methods do not scale well enough to analyze
+whole viral genomes, a useful strategy might be to use annotated sequence data to identfiy 
+candidate HLA escape mutations that can then be verified experimentally.
+
+As the selection pressure exerted by cytotoxic T cells depends on successful recognition
+of viral peptides on the cell surface, escape mutations are often HLA-allele specific
+and can therefore  be detected as HLA-allele dependent footprints in sequence
+alignments of viral proteins: At certain alignment positions, a replacement
+might be more frequently observed in sequences from hosts with a specific HLA allele than in 
+sequences from hosts without that HLA allele. By quantifying this difference
+for all replacement and HLA allele pairs it is possible to idenfify replacements that
+are enriched in sequences coming from hosts with certain HLA alleles and are therefore
+likely to be HLA escape mutations.
+
+One way of quantifying this enrichment is Fisher\'s exact test.
+For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\),
+a 2-by-2 contingency table is constructed containing the absolute counts of the
+number of sequences in the four possible categories: 
+(\(R_{i}\), \(H\)), (\(R_{i}\), \(!H\)), (\(!R_{i}\), \(H\)) and (\(!R_{i}\), \(H\)),
+where \(!R_{i}\) denotes any replacement except \(R_{i}\) and \(!H\) denotes any HLA allele
+except \(H\).
+Fisher\'s exact test is a conventional nullhypothesis significance test (NHST) that
+generates p-values. In this case, the nullhypothesis is that HLA allele \(H\) and 
+replacement \(R_{i}\) are independent, and the p-value is the probability of
+observing a deviation from independence that is at least as extreme as in the data at hand,
+under the assumption that the nullhypothesis is true.
+
+Fisher\'s exact test has the advantage of being easy to apply, but has also several 
+disadvantages that are outlined in Carlson et al.
+The most striking one is that viral sequences share a common phylogenetic history and therfore,
+treating sequences as independent and identically distributed samples may under- or
+overestimate effect sizes and in the context of hypothesis testing leads to
+increased false positive and false negative rates.
+Another issue of applying Fisher\'s exact test is that HLA class I loci are located in
+close proximity on chromosome 6 and are therefore in linkage disquilibrium, which means 
+that HLA alleles are not inherited completely independent of each other, so that
+inheritance of one HLA allele correlates with inheritance of another HLA allele.
+When using a statistical method that tests each HLA allele individually without
+considering the whole set of alleles present, spurious associations might occur
+: If HLA allele \(H_{1}\) is associated with an amino acid replacement \(R\), but
+\(H_{1}\) is in linkage disequilibrium with another allele \(H_{2}\), this also means that
+we observe an association between replacement \(R\) and \(H_{2}\), even in
+absence of any underlying escape mechanism.
+
+Correlations can not only occur between HLA alleles, but also between replacements.
+This kind of codon covariation occurs for example in compensatory mutations that
+attenuate the negative impact of immune escape mutations. For instance,
+a compensatory mutation might lead to a conformational change in such a way that
+the mutated protein resembles the original wild type more closely.
+
+Carlson et. al. developed a method called Phylogenetic Dependency Network
+that accounts for phylogenetic bias, HLA linkage disequilibrium and codon covariation.
+and is based on nullhypothesis significance tests.
+
+\subsection{Issues with using p-values as a screening tool}
+
+In addition to the aforementioned biological reasons why Fisher's exact test is 
+not suited for the analysis of sequence data, there are also more fundamental statistical
+issues that universally occur when using p-values as a screening tool.
+
+In the presence of small effect sizes and high variance between measurements, as 
+it is typically the case when working with biological data, statistically significant
+results can often be misleading and are likely to be in the wrong direction
+(a so-called type S error) or greatly overestimate an effect (a so-called type M error).
+This problem has recently been widely appreciated in the literature in the context
+of the current \glqq replication crisis\grqq{}, which describes the circumstance that
+scientific claims with seemingly strong statistical evidence fail to replicate.
+
+Another issue that occurs when using p-values as a screening tool is the problem
+of multiple comparisons. When applying a statistical test, the probability  of 
+obtaining a statistically significant result increases with
+each additional test, even in absence of any real effect. When using p-values as a filter
+it is therefore likely to obtain significant effects that are in fact not real.
+To circumvent this problem, a common strategy is to control the false discovery
+rate, which is the expected proportion of false positives.
+These adjustment procedures have the problem that, when performing many of such comparisons,
+none but the very largest effects remain.
+Instead of performing many hypothesis tests and trying to adjust for them, we instead prefer
+to fit a single, multilevel model that contains all comparisons of interest.
+In this setting, the problem of multiple comparisons can disappear entirely and also
+yield more valid estimates.
+
 \section{System and methods}
+
 \section{Algorithm}

### 2020-11-18 docs/manuscript.tex
add 'System and Methods' section

@@
 \PaperTitle{
-    HAMDetector: Combining information to detect HLA-associated mutations with a
+    HAMdetector: Combining information to detect HLA-associated mutations with a
     Bayesian regression model
@@
   The source code of this software is available at 
-  \url{http://gogs.uni-due.de/habermann/Escape.git} under a permissive MIT license. \\
+  \url{http://github.com} under a permissive MIT license. \\
 
@@
 
-% - subset of peptidome presented on the cell surface via MHC I.
-% - Cytotoxic t cells are selected to not recognize self.
-% - by rule of exclusion, recognized peptides are foreign
-% - t cells induce cytolytic activity and recruit other immune cells, reference
-
-% - basic function of statistical analysis of sequence data
-% - state of the art + brumme paper, phylogeny, linkage disequilibrium and codon covariation
-% cite brumme and other paper
-
-% - problems of p value based methods and statistical significance testing
-% - highlight benefits of Bayesian approaches
 \subsection{The HLA system}
@@
 probably already be the wild type). Kawashima et al.
-describe an escape mutation that is selected by HLA allele B\*53,
+describe an escape mutation that is selected by HLA allele B*51,
 does not strongly affect viral replicative capacity and therefore slowly enriches
-over time in Japan, where B\*53 commonly occurs.
+over time in Japan, where B*53 commonly occurs.
 How quickly a given escape mutation is selected upon transmission in a host depends on the
@@
 peptide to MHC-I and does not account for antigen processing or immunodominance, 
-which describes the observation that a peptide may be presented via MHC-I on the cell surface but does
-not induce an immune response.
+which describes the observation that a peptide may be presented via MHC-I on the cell 
+surface but does not induce an immune response.
 An experimental setup that resembles the conditions in-vitro more closely but is also
@@
 
-Fisher\'s exact test has the advantage of being easy to apply, but has also several 
+Fisher's exact test has the advantage of being easy to apply, but has also several 
 disadvantages that are outlined in Carlson et al.
@@
 increased false positive and false negative rates.
-Another issue of applying Fisher\'s exact test is that HLA class I loci are located in
+Another issue of applying Fisher's exact test is that HLA class I loci are located in
 close proximity on chromosome 6 and are therefore in linkage disquilibrium, which means 
@@
 none but the very largest effects remain.
-Instead of performing many hypothesis tests and trying to adjust for them, we instead prefer
+Instead of performing many hypothesis tests and trying to adjust for them we instead prefer
 to fit a single, multilevel model that contains all comparisons of interest.
-In this setting, the problem of multiple comparisons can disappear entirely and also
-yield more valid estimates.
+When using multilevel models, the problem of multiple comparisons can disappear entirely and 
+also yield more valid estimates.
 
 \section{System and methods}
+
+When possible, we choose to fit Bayesian models. By using prior information,
+adding problem-specific structure and partial pooling, the accuracy of estimates
+can often be noticably improved. Prior information does not necessarily mean to
+use external data, even a rough idea about the expected magnitude of estimates is often
+surprisingly effective.
+Additionally, Bayesian statistics provides an accessible way to test models:
+By comparing data generated under the model's assumptions to the actually observed data,
+it is possible to identify important aspects of the dataset that the models fails to
+capture and subsequently improve the model until it is consistent with the observed data.
+
+In the context of identifying HLA-associated mutations we propose to improve existing
+methods by the following additions, which can be broadly divided into additional information
+and model structure:
+\subsection*{Additional information}
+
+  \textit{Binding affinity prediction}
+
+    HLA-associated mutations are expected to lie more frequently in regions of known
+    epitopes. There are vast experimental binding affinity data available of different
+    peptides and MHC I molecule pairs and there are well-established computational
+    methods that use these data to extrapolate HLA binding for untested peptides.
+    We show that by including the outcome of these computational tools as input for a
+    probabilistic model, the prediction of HLA-associated mutations can be improved.
+
+  \textit{Antigen processing prediction}
+
+  Similarly, there are also HLA-allele independent effects like antigen processing
+  that influence presentation on MHC I. Second generational tools like
+  MHCFlurry 2.0 use both binding affinity and antigen processing data to improve
+  epitope prediction. For our tool HAMdetector, we use the output of MHCFlurry to
+  benefit from this binding affinity and antigen processing data for predicting
+  HLA-associated mutations.
+
+\subsection*{Model structure}
+
+  \textit{Sparsity-inducing priors}
+
+  Recent advantages in Bayesian inference include so-called sparsity-promoting priors.
+  Sparsity-promoting priors convey the apriori expectation that most coefficients in 
+  a regression model are close to 0.
+  This assumption also applies to HLA-associated mutations, because the number of epitopes
+  that are restricted by a given HLA allele is typically small compared to the number
+  of all possible epitopes. If no epitope spanning a given alignment position is presented
+  on the MHC I molecule, any association between a replacement and the respective HLA allele 
+  is likely to be due to random variation alone (or HLA linkage disequilibrium).
+
+  It has been shown that using a sparsity-promoting prior when non-zero coefficients
+  are sparse can drastically improve predictive performance because the model is
+  better able to differentiate between signal and noise.
+
+  We show that a simple logistic regression model with a sparsity-promoting prior
+  on the regression coefficients for the HLA alleles alone already performs roughly on-par 
+  with the much more elaborate Phylogenetic Dependency Network approach from Carlson et. al.
+  when using the fraction of identified escapes that lie in known HLA epitopes as a
+  benchmark.
+
+  \textit{Partial pooling of 4-digit HLA alleles}
+
+  It has recently been appreciated that binding specifities can vary drastically across 
+  HLA alleles from the same allele group, e.g. between HLA-B*51:01 and HLA-B*51:03.
+  For predicting HLA associated mutations, there has therefore been a shift to use
+  4-digit resolution data whenever possible.
+  This is not without downsides however, because overall, binding specifities
+  are more similar for alleles in the same group and therefore, treating them
+  as completely separate might unnecessarily fragment the available data.
+
+  In Bayesian statistics, it is not required to chose one extreme or the other:
+  Instead of either treating all HLA alleles from the same allele group as identical or
+  as completely separate, partial pooling of the estimates allows something inbetween:
+  By partial pooling, estimates do share some information across each other, but they are
+  also allowed to vary if necessary.
+
+  In the context of our model this means that observing a relevant association between
+  a replacement and an HLA allele also makes the model more inclined to estimate
+  relevant associations between that replacement and all other alleles from that allele group.
+  The degree of this partial pooling can be estimated from the data, so if we do observe
+  strong similarity of HLA alleles in a given allele group, the estimates are more
+  influenced by each other than if they do not.
 

### 2020-11-30 docs/manuscript.tex
add implementation talking points

--- 2020-11-12 docs/manuscript.tex
@@
 \PaperTitle{
-    HAMDetector: Combining information to predict HLA-associated mutations with a
+    HAMdetector: Combining information to detect HLA-associated mutations with a
     Bayesian regression model
@@
 
-\Authors{Daniel Habermann\textsuperscript{1}, Daniel Hoffmann\textsuperscript{1}} % Authors
+\Authors{Daniel Habermann\textsuperscript{1}, ..., Daniel Hoffmann\textsuperscript{1}} % Authors
 \affiliation{\textsuperscript{1}\textit{Bioinformatics \& Computational Biophysics, Faculty of Biology, University of Duisburg-Essen, 45117 Essen, Germany}} % Author affiliation
@@
   infections by presenting peptides on the cell surface via MHC I.
-  In this way, CD8+ cytotoxic T-Lymphocytes exert a strong selection pressure towards
+  Thus, CD8+ cytotoxic T-Lymphocytes exert a strong selection pressure towards
   virus variants that escape that immune recognition pathway, e.g. through point 
-  mutations that decrease binding of the respective peptide to MHC I. \\
+  mutations that decreases binding of the respective peptide to MHC I. \\
 
@@
 
-\pagebreak
-
 \section{Introduction}
 
-% - subset of peptidome presented on the cell surface via MHC I.
-% - Cytotoxic t cells are selected to not recognize self.
-% - by rule of exclusion, recognized peptides are foreign
-% - t cells induce cytolytic activity and recruit other immune cells, reference
-
-% - basic function of statistical analysis of sequence data
-% - state of the art + brumme paper, phylogeny, linkage disequilibrium and codon covariation
-% cite brumme and other paper
-
-% - problems of p value based methods and statistical significance testing
-% - highlight benefits of Bayesian approaches
 \subsection{The HLA system}
@@
 One way how the human immune system is able to recognize intracellular viral infections
-is through the human leucocyte antigen system:
+is through the human leucocyte antigen system\nolinebreak\cite{Germain1994}:
 In cells with active protein biosynthesis, proteins are continuously synthesized
 and also degraded by a process called proteasomal degradation, which cleaves
-proteins into linear peptides of varying length.
+proteins into linear peptides of varying length\nolinebreak\cite{Goldberg2002}.
 A small subset of these peptides is presented on the cell surface via
-receptors called MHC class I. The genomic region encoding for MHC I is known to be
-highly polymorphic, with more than 20000 different HLA alleles described today.
+receptors called MHC class I (HLA-A, HLA-B and HLA-C in humans). The genomic region encoding 
+for MHC I is known to be highly polymorphic, with more than 20000 different HLA alleles 
+described today\nolinebreak\cite{Robinson2014}.
 The resulting gene products differ in their binding properties, which means that
@@
 but might be able to strongly bind to complexes of MHC I with peptides which are 
-generated from of a viral protein.
-Upon activation, T cells induce cytolytic activity and recruit other immune cells.
+generated from of a viral protein\nolinebreak\cite{Murata2007}.
+Upon activation, T cells induce cytolytic activity and recruit other immune cells
+\nolinebreak\cite{Harty2000}.
 
 \subsection{HLA escape}
-In this way, the HLA system exerts a strong selection pressure towards virus variants
-that escape T cell recognition, for example through a point mutation
-that results in reduced binding of an immunogenic peptide to MHC I or through a
+In this way, the HLA system exerts strong selection pressure towards virus variants
+that escape T cell recognition\nolinebreak\cite{Borrow1997}, for example through a point 
+mutation that results in reduced binding of an immunogenic peptide to MHC I or through a
 set of mutations that alters the viral protein in such a way that it is cleaved into 
-different peptides that are not recognized by the host's T cell repertoire.
+different peptides that are not recognized by the host's T cell repertoire
+\nolinebreak\cite{Yewdell2002}.
 
@@
 where a virus adapts to specific features of the host, but also on the population level,
-because HLA alleles differ in their frequency across geographic regions, as they are inherited
-according to standard Mendelian rules.
+because HLA alleles differ in their frequency across geographic regions\nolinebreak\cite{Kawashima2009}.
 
-Upon transmission, HLA escape mutations typically quickly revert to their wild type
-as they usually reduce viral replicative capacity (if a mutation would increase
-viral replicative capacity regardless of the presence of a given HLA allele, it would
-probably already be the wild type). Kawashima et al.
-describe an escape mutation that is selected by HLA allele B\*53,
-does not strongly affect viral replicative capacity and therefore slowly enriches
-over time in Japan, where B\*53 commonly occurs.
+Upon transmission to a new host, HLA escape mutations can revert to their wild type,
+as HLA escape mutations are associated with a reduction in viral replicative capacity
+\nolinebreak\cite{Matthews2008}. Kawashima et al.\nolinebreak\cite{Kawashima2009}
+describe an escape mutation that is selected by HLA allele HLA-B*51,
+does not strongly affect viral replicative capacity, and therefore slowly enriches
+over time in Japan, where HLA-B*51 commonly occurs.
 How quickly a given escape mutation is selected upon transmission in a host depends on the
 magnitude of the reduction in viral replicative capacity, on the strength of selection
-pressure and also on the genetic background, e.g. some escape mutations require compensatory 
+pressure, and also on the genetic background, e.g. some escape mutations require compensatory 
 mutations which partly attenuate the negative impact on viral replicative capacity.
 
-Studying HLA escape therefore provides an unique opportunity to gain insight into
-viral evolution, on the host level but also on the population level.
+Studying HLA escape provides an unique opportunity to gain insight into
+viral evolution, on the host level, but also on the population level.
 Unfortunately, identifying HLA escape mutations is difficult in practice.
@@
 There are several experimental methods available to study HLA escape:
-Recombinant MHC-I molecules can be used in binding assays: Upon complex formation
+Recombinant MHC-I molecules can be used in binding assays. Upon complex formation
 with a peptide, a change in conformation can be detected with conformation-specific
@@
 peptide to MHC-I and does not account for antigen processing or immunodominance, 
-which describes the observation that a peptide may be presented via MHC-I on the cell surface but does
-not induce an immune response.
+which describes the observation that a peptide may be presented via MHC-I on the cell 
+surface, but does not induce an immune response.
 An experimental setup that resembles the conditions in-vitro more closely but is also
 more time-consuming is to measure
-CD8+ T cell responses instead. This is usually done by stimulating PBMCs with prototype
-and variant peptides and measuring the secretion of IFN-gamma by intracellular cytokine
+CD8+ T cell responses instead. This is usually done by stimulating peripheral blood
+mononuclear cells with prototype and variant peptides and measuring the secretion of
+IFN-\gamma{} by intracellular cytokine
 staining and fluorescence-assisted cell sorting.
-To analyse CD8+ T cell responses against endogenously processed antigens it is necessary
+To analyze CD8+ T cell responses against endogenously processed antigens, it is necessary
 to generate cell-lines stably expressing the antigen in question and adding antigen-specific
@@
 
-\section{System and methods}
-\section{Algorithm}
+\subsection{Computational methods}
+
+Because the currently available experimental methods do not scale well enough to analyze
+whole viral genomes, a useful strategy might be to use annotated sequence data to identfiy 
+candidate HLA escape mutations that can then be verified experimentally.
+
+As the selection pressure exerted by cytotoxic T cells depends on successful recognition
+of viral peptides on the cell surface, escape mutations are often HLA-allele specific
+and can therefore be detected as HLA-allele dependent footprints in sequence
+alignments of viral proteins\nolinebreak\cite{Moore2002}: At certain alignment positions, a replacement
+might be more frequently observed in sequences from hosts with a specific HLA allele than in 
+sequences from hosts without that HLA allele. By quantifying this difference
+for all replacement and HLA allele pairs, it is possible to idenfify replacements that
+are enriched in sequences coming from hosts with certain HLA alleles, and thus are
+likely to be HLA escape mutations.
+
+One way of quantifying this enrichment is Fisher's exact test\nolinebreak\cite{Fisher1922}.
+For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\),
+a 2-by-2 contingency table is constructed containing the absolute counts of the
+number of sequences in the four possible categories: 
+(\(R_{i}\), \(H\)), (\(R_{i}\), \(!H\)), (\(!R_{i}\), \(H\)) and (\(!R_{i}\), \(H\)),
+where \(!R_{i}\) denotes any replacement except \(R_{i}\), and \(!H\) denotes any HLA allele
+except \(H\).
+Fisher's exact test is a conventional nullhypothesis significance test (NHST) that
+generates p-values. In this case, the nullhypothesis is that HLA allele \(H\) and 
+replacement \(R_{i}\) are independent, and the p-value is the probability of
+observing a deviation from independence that is at least as extreme as in the data at hand
+under the assumption that the nullhypothesis is true.
+
+Fisher's exact test has the advantage of being easy to apply\nolinebreak\cite{Budeus2016}, 
+but also has several disadvantages that are outlined in Carlson et al. 
+\nolinebreak\cite{Carlson2008}.
+The most striking one is that viral sequences share a common phylogenetic history, and, 
+therefore, treating sequences as independent and identically distributed samples may under- or
+overestimate effect sizes. In the context of hypothesis testing, this leads to
+increased false positive and false negative rates.
+Another issue of applying Fisher's exact test is that HLA class I loci are located in
+close proximity on chromosome 6 and are therefore in linkage disquilibrium, which means 
+that HLA alleles are not inherited completely independent of each other, i.e.
+inheritance of one HLA allele correlates with inheritance of another HLA allele.
+When using a statistical method that tests each HLA allele individually without
+considering the whole set of alleles present, spurious associations might occur: 
+If HLA allele \(H_{1}\) is associated with an amino acid replacement \(R\), but
+\(H_{1}\) is in linkage disequilibrium with another allele \(H_{2}\), this also means that
+we observe an association between replacement \(R\) and \(H_{2}\), even in
+absence of any underlying escape mechanism.
+
+Correlations can not only occur between HLA alleles, but also between replacements.
+This kind of codon covariation occurs for example in compensatory mutations that
+attenuate the negative impact of immune escape mutations. For instance,
+a compensatory mutation might lead to a conformational change in such a way that
+the mutated protein resembles the original wild type more closely.
+
+Carlson et al. \nolinebreak\cite{Carlson2008} developed a method called 
+Phylogenetic Dependency Network that accounts for phylogenetic bias, 
+HLA linkage disequilibrium, and codon covariation and is based on 
+nullhypothesis significance testing.
+
+\subsection{Issues with using p-values as a screening tool}
+
+In addition to the aforementioned biological reasons why Fisher's exact test is 
+not suited for the analysis of sequence data, there are also more fundamental statistical
+issues that universally occur when using p-values as a screening tool\nolinebreak\cite{Amrhein2017}.
+
+In the presence of small effect sizes and high variance between measurements, as 
+it is typically the case when working with biological data, statistically significant
+results can often be misleading and are likely to be in the wrong direction
+(a so-called type S error) or greatly overestimate an effect (a so-called type M error)
+\nolinebreak\cite{Gelman2014}.
+This problem has recently been widely appreciated in the literature in the context
+of the current \glqq replication crisis\grqq{}, which describes the circumstance that
+scientific claims with seemingly strong statistical evidence fail to replicate\nolinebreak\cite{Ioannidis2005}.
+
+Another issue that occurs when using p-values as a screening tool is the problem
+of multiple comparisons. When applying a statistical test, the probability  of 
+obtaining a statistically significant result increases with
+each additional test, even in absence of any real effect. When using p-values as a filter,
+it is therefore likely to obtain significant effects that are in fact not real.
+To circumvent this problem, a common strategy is to control the false discovery
+rate, which is the expected proportion of false positives\nolinebreak\cite{Benjamini1995}.
+These adjustment procedures have the problem that, when performing many of such comparisons,
+none but the very largest effects remain.
+Instead of performing many hypothesis tests and trying to adjust for them, we instead prefer
+to fit a single, multilevel model that contains all comparisons of interest.
+When using multilevel models, the problem of multiple comparisons can disappear entirely and 
+also yield more valid estimates\nolinebreak\cite{Gelman2012}.
+
+\section{Material and methods}
+
+When possible, we choose to fit Bayesian models. By using prior information,
+adding problem-specific structure and partial pooling, the accuracy of estimates
+can often be noticably improved\nolinebreak\cite{Gelman2010}. Prior information does not necessarily mean to
+use external data, even a rough idea about the expected magnitude of estimates is often
+surprisingly effective.
+Additionally, Bayesian statistics provides an accessible way to test models:
+By comparing data generated under the model's assumptions to the actually observed data,
+it is possible to identify important aspects of the dataset that the models fails to
+capture and subsequently improve the model until it is consistent with the observed data
+\nolinebreak\cite{Gabry2019}.
+
+In the context of identifying HLA-associated mutations we propose to improve existing
+methods by the following additions, which can be broadly divided into additional information
+and model structure:
+\subsection*{Additional information}
+
+  \textit{Binding affinity prediction}
+
+    HLA-associated mutations are expected to lie more frequently in regions of known
+    epitopes. There are vast experimental binding affinity data available of different
+    peptides and MHC I molecule pairs, and there are well-established computational
+    methods that use these data to extrapolate HLA binding for untested peptides.
+    We show that, by including the outcome of these computational tools as input for a
+    probabilistic model, the prediction of HLA-associated mutations can be improved.
+
+  \textit{Antigen processing prediction}
+
+  Similarly, there are also HLA-allele independent effects like antigen processing
+  that influence presentation on MHC I. Second generational tools like
+  MHCFlurry 2.0 use both binding affinity and antigen processing data to improve
+  epitope prediction. For our tool HAMdetector, we use the output of MHCFlurry to
+  benefit from this binding affinity and antigen processing data in order to predict
+  HLA-associated mutations.
+
+\subsection*{Model structure}
+
+  \textit{Sparsity-inducing priors}
+
+  Recent advantages in Bayesian inference include so-called sparsity-promoting priors.
+  Sparsity-promoting priors convey the apriori expectation that most coefficients in 
+  a regression model are close to 0.
+  This assumption also applies to HLA-associated mutations, because the number of epitopes
+  that are restricted by a given HLA allele is typically small compared to the number
+  of all possible epitopes. If no epitope spanning a given alignment position is presented
+  on the MHC I molecule, any association between a replacement and the respective HLA allele 
+  is likely to be due to random variation alone (or HLA linkage disequilibrium).
+
+  It has been shown that using a sparsity-promoting prior when non-zero coefficients
+  are sparse can drastically improve predictive performance, because the model is
+  better able to differentiate between signal and noise.
+
+  We show that a simple logistic regression model with a sparsity-promoting prior
+  on the regression coefficients for the HLA alleles alone already performs roughly on-par 
+  with the much more elaborate Phylogenetic Dependency Network approach from Carlson et al.
+  when using the fraction of identified escapes that lie in known HLA epitopes as a
+  benchmark.
+
+  \textit{Partial pooling of 4-digit HLA alleles}
+
+  It has recently been appreciated that binding specifities can vary drastically across 
+  HLA alleles from the same allele group, e.g. between HLA-B*51:01 and HLA-B*51:03.
+  For predicting HLA associated mutations, there has consequently been a shift to use
+  4-digit resolution data whenever possible.
+  This is not without downsides however, because overall, binding specifities
+  are more similar for alleles in the same group, and, therefore, treating them
+  as completely separate might unnecessarily fragment the available data.
+
+  In Bayesian statistics, it is not required to chose one extreme or the other:
+  Instead of either treating all HLA alleles from the same allele group as identical or
+  as completely separate, partial pooling of the estimates allows something inbetween:
+  By partial pooling, estimates do share some information across each other, but they are
+  also allowed to vary if necessary.
+
+  In the context of our model this means that observing a relevant association between
+  a replacement and an HLA allele also makes the model more inclined to estimate
+  relevant associations between that replacement and all other alleles from that allele group.
+  The degree of this partial pooling can be estimated from the data, so if we do observe
+  strong similarity of HLA alleles in a given allele group the estimates are more
+  influenced by each other than if they do not.
+
 \section{Implementation}
-\section{Discussion}
-\section{References}
+
+\subsection*{Model backbone}
+\begin{itemize}
+  \item logistic regression model, because easily extensible
+  \item coefficients are interpretable as expected increase in log-odds of observing 
+    a given replacement (also: average predictive comparisons to transform estimates to
+    probability scale)
+  \item probabilistic model allows us to perform posterior predictive checks
+  \item disadvantages: none? maybe log-odds scale unintuitive. mention divide-by-4 rule
+  \item  establish notation, start simple and expand on that? Maybe confusing because
+    some things change as models get more complex 
+\end{itemize}
+
+\subsection*{Including phylogeny}
+\begin{itemize}
+  \item sequence data shares common phylogenetic history and are therefore not i.i.d,
+    cite Brumme paper
+  \item standard approach: multivariate normal intercept where covariance matrix is based on 
+    shared branch lengths
+  \item this approach is computationally too expensive because of current limitations in 
+    Stan. Would be possible by fitting the model in Julia? Probably omit that.
+  \item Our approach is close to the one in Carlson et al.
+  \item slight tweak: use tree estimation tools for tree likelihoods p(tree | sequences),
+    then bayes rule to invert that conditional probability. Optimize branch lengths
+    for each replacement, but keep tree topology fixed
+  \item advantages: faster (raxml well optimized)
+  \item disadvantages: ad-hoc, close to conventional multivariate normal approach with 
+    off-diagonal elements set to 0, therefore not as flexible. But: should be good enough,
+    phylogeny seems to be not that important (probably omit that?)
+  \item how to include p(replacement | tree) in the model: just add additional intercept
+  \item add coefficient for that information, so model can decide how to make use of
+    that information.
+  \item logit(p(replacement|tree)) used because it cancels out with the logistic
+    link function. therefore: phylogeny information as baseline in absence of any 
+    HLA effects
+  \item calibration plots to show that estimated phylogenetic probabilities are
+   reasonably well calibrated.
+  \item meaning: for sequences with an expected probability of observing the replacement
+    of x\%, we really do see that x\% of the sequences have that replacement.
+\end{itemize}
+
+\begin{figure}[!ht]
+  \includegraphics[width=1\linewidth]{plots/phylogeny_calibration.pdf}
+  \caption{Calibration plot and how it works.}
+\end{figure}
+
+\subsection*{Including HLA linkage disequilibrium and codon covariation}
+
+\begin{itemize}
+  \item HLA linkage disequilibrium is accounted for by considering all HLA alleles
+    of a sample at once.
+  \item If strong correlations between HLA alleles: marginal posterior for both of them
+    overlap with 0
+  \item Codon covariation is not accounted for because we are mainly interested in HAMs.
+  \item Model could be extended to include codon covariation by including them as additional
+    predictors. This is possible because we use the horseshoe prior.
+\end{itemize}
+
+\subsection*{Including sparsity assumptions}
+
+\begin{itemize}
+  \item one of the recent advances in Bayesian inference: development of
+    spacial priors to give model useful properties
+  \item we expect most HLA coefficients to be 0, because no epitope spanning that region
+    is presented on the cell surface, therefore no selection pressure and no association
+    with possible replacements
+  \item this is prior knowledge that should be incorporated in the model
+  \item leads to better inferences because uncertainty of the regression coefficients
+    does not propagate into observations.
+  \item works by placing most probability mass very close to 0, but with large tails
+    for non-zero coefficients.
+  \item show formula
+  \item global shrinkage parameter is very small and shrinks most coefficients to 0,
+    local shrinkage parameter is very large and allows some coefficients to escape that 
+    shrinkage
+  \item prior can be specified in terms of expected number of non-zero coefficients
+  \item in our case, degree of sparsity can be well estimated from the data because all
+    replacements share the same global shrinkage parameter.
+  \item useful addition for logistic regression models: regularized horseshoe
+    to have some regularization for non-zero coefficients. this helps to deal with
+    issues of separability and collinearity.
+  \item effect of horseshoe prior compared to standard logistic regression model:
+    refer to model testing section. In our benchmark, the horseshoe prior alone is more
+    effective than the model with phylogeny.
+  \item maybe include comparison of marginal posteriors with and without horseshoe prior
+\end{itemize}
+
+\subsection*{Including epitope prediction software}
+
+\begin{itemize}
+  \item Vast experimental data available.
+  \item HAMs are much more likely to lie in known epitopes
+  \item epitope prediction software works reasonably well and can be included into as
+    input to a probabilitic model
+  \item MHCFlurry is used for epitope prediction and Antigen processing prediction:
+    Input matrix (dimensions: \#replacements x \#HLA alleles) contains a 1 if that
+    position is either expected to be inside a predicted epitope or related to
+    antigen processing, 0 otherwise
+  \item uses some thresholds, but this is not so bad because we just use it as
+    input for a probabilistic model. Maybe play around with different options?
+    The Brumme paper uses some offsets for defining the region of known epitopes,
+    maybe this does make a difference.
+  \item including this kind of information can drastically improve predictive performance,
+    as it is relevant external data.
+  \item Bayesian inference allows us to include informationen from different sources in
+    a consistent manner.
+  \item this information may be imprecise or wrong, but the model is able to learn how much 
+    to trust this data if parameterized correctly.
+  \item epitope prediction / antigen processing escape is information about the expected
+    degree of sparsity, i.e. some coefficients are more likely to be non-zero than others.
+  \item can be included into the model by varying the local shrinkage parameter term.
+  \item higher standard deviation of the local shrinkage parameters mean that
+    it is more likely that this coefficient is non-zero.
+  \item show formula
+\end{itemize}
+
+\subsection*{Partial pooling of 4-digit HLA alleles}
+
+\begin{itemize}
+  \item (this still has to be implemented, but should be a nice feature. If this
+    is going to be too time consuming I can just leave it out)
+  \item realized through allele group specific intercepts, e.g. HLA-B\*51:01 and 
+  HLA-B\*51:03, ..., share a common intercept.
+  \item show formula
+\end{itemize}
+
+\subsection*{Full model specification}
+\begin{itemize}
+  \item show all formulas
+  \item all models were implementen in Stan. Code is available online in two versions:
+  one optimized for readability, one optimized for speed with multithreading and GPU support
+  \item data can be analyzed using a custom Julia package (tested on linux).
+\end{itemize}
+
+\subsection*{Prior justification}
+\begin{itemize}
+  \item justify priors, mostly based on the expected scales.
+\end{itemize}
+
+\section{Results and Discussion}
 
@@
 \bibliographystyle{unsrt}
-
+\bibliography{references}
 %----------------------------------------------------------------------------------------

### 2020-12-15 docs/manuscript.tex
finish rough outline for manuscript

@@
 
-\Keywords{human leucocyte antigen system, HLA, multiple sequence alignment, escape mutations, 
+\Keywords{human leukocyte antigen system, HLA, multiple sequence alignment, escape mutations, 
   viral escape, Bayesian inference, sparsity, horseshoe, epitope prediction} % Keywords - if you don't want any simply remove all the text between the curly brackets
@@
   \textbf{Motivation} \\
-  The human leucocyte antigen system (HLA) is of paramount importance to combat viral
-  infections by presenting peptides on the cell surface via MHC I.
-  Thus, CD8+ cytotoxic T-Lymphocytes exert a strong selection pressure towards
-  virus variants that escape that immune recognition pathway, e.g. through point 
-  mutations that decreases binding of the respective peptide to MHC I. \\
+  The human leukocyte antigen system (HLA) is of paramount importance to combat viral infections by presenting peptides on the cell surface via MHC I.
+Thus, CD8+ cytotoxic T-Lymphocytes exert a strong selection pressure towards virus variants that escape that immune recognition pathway, e.g.
+through point mutations that decreases binding of the respective peptide to MHC I.
+\\
 
-  Reliably identifying HLA-associated mutations is important for understanding viral
-  evolution, but experimental methods like binding assays are prohibitively expensive
-  for large-scale use and fail to recognize other mechanisms of immune escape like
-  proteasomal processing.
+  Reliably identifying HLA-associated mutations is important for understanding viral evolution, but experimental methods like binding assays are prohibitively expensive for large-scale use and fail to recognize other mechanisms of immune escape like proteasomal processing.
 
-  One step in finding these mutations is through the statistical analysis
-  of sequence data. However, existing methods are based on nullhypothesis significance 
-  testing and do not make use of all the available information and therefore have
-  unsatisfactory real-world performance. \\
+  One step in finding these mutations is through the statistical analysis of sequence data.
+However, existing methods are based on null hypothesis significance testing and do not make use of all the available information and therefore have unsatisfactory real-world performance.
+\\
 
   \textbf{Results} \\
-  Here, we present a Bayesian regression model that is easily extensible to include 
-  information from different sources (e.g. epitope prediction software) 
-  and makes use of recent advances in Bayesian inference, e.g. by using a sparsifying 
-  prior. We show that including this kind of information improves predictive performance 
-  considerably over state-of-the-art methods. \\
+  Here, we present a Bayesian regression model that is easily extensible to include information from different sources (e.g.
+epitope prediction software) and makes use of recent advances in Bayesian inference, e.g.
+by using a sparsifying prior.
+We show that including this kind of information improves predictive performance considerably over state-of-the-art methods.
+\\
 
   \textbf{Availability and Implementation} \\
-  The source code of this software is available at 
-  \url{http://github.com} under a permissive MIT license. \\
+  The source code of this software is available at \url{http://github.com} under a permissive MIT license. \\
 
@@
 
-One way how the human immune system is able to recognize intracellular viral infections
-is through the human leucocyte antigen system\nolinebreak\cite{Germain1994}:
-In cells with active protein biosynthesis, proteins are continuously synthesized
-and also degraded by a process called proteasomal degradation, which cleaves
-proteins into linear peptides of varying length\nolinebreak\cite{Goldberg2002}.
-A small subset of these peptides is presented on the cell surface via
-receptors called MHC class I (HLA-A, HLA-B and HLA-C in humans). The genomic region encoding 
-for MHC I is known to be highly polymorphic, with more than 20000 different HLA alleles 
-described today\nolinebreak\cite{Robinson2014}.
-The resulting gene products differ in their binding properties, which means that
-cells from different individuals present a highly diverse set of peptides on their
-surface.
-Cytotoxic T cells are selected during maturation to only weakly bind to 
-peptide/MHC I complexes when the peptide originated from proteins of the usual proteome, 
-but might be able to strongly bind to complexes of MHC I with peptides which are 
-generated from of a viral protein\nolinebreak\cite{Murata2007}.
-Upon activation, T cells induce cytolytic activity and recruit other immune cells
-\nolinebreak\cite{Harty2000}.
+One way how the human immune system is able to recognize intracellular viral infections is through the human leukocyte antigen system\nolinebreak\cite{Germain1994}: In cells with active protein biosynthesis, proteins are continuously synthesized and also degraded by a process called proteasomal degradation, which cleaves proteins into linear peptides of varying length\nolinebreak\cite{Goldberg2002}.
+
+A small subset of these peptides is presented on the cell surface via receptors called MHC class I (HLA-A, HLA-B and HLA-C in humans).
+The genomic region encoding for MHC I is known to be highly polymorphic, with more than 20000 different HLA alleles described today\nolinebreak\cite{Robinson2014}.
+The resulting gene products differ in their binding properties, which means that cells from different individuals present a highly diverse set of peptides on their surface.
+
+Cytotoxic T cells are selected during maturation to only weakly bind to peptide/MHC I complexes when the peptide originated from proteins of the usual proteome, but might be able to strongly bind to complexes of MHC I with peptides which are  generated from of a viral protein\nolinebreak\cite{Murata2007}.
+Upon activation, T cells induce cytolytic activity and recruit other immune cells \nolinebreak\cite{Harty2000}.
 
 \subsection{HLA escape}
-In this way, the HLA system exerts strong selection pressure towards virus variants
-that escape T cell recognition\nolinebreak\cite{Borrow1997}, for example through a point 
-mutation that results in reduced binding of an immunogenic peptide to MHC I or through a
-set of mutations that alters the viral protein in such a way that it is cleaved into 
-different peptides that are not recognized by the host's T cell repertoire
-\nolinebreak\cite{Yewdell2002}.
+In this way, the HLA system exerts strong selection pressure towards virus variants that escape T cell recognition\nolinebreak\cite{Borrow1997}, for example through a point mutation that results in reduced binding of an immunogenic peptide to MHC I or through a set of mutations that alters the viral protein in such a way that it is cleaved into different peptides that are not recognized by the host's T cell repertoire\nolinebreak\cite{Yewdell2002}.
 
-The evolutionary events are complex and occur not only on the level of individuals, 
-where a virus adapts to specific features of the host, but also on the population level,
-because HLA alleles differ in their frequency across geographic regions\nolinebreak\cite{Kawashima2009}.
+The evolutionary events are complex and occur not only on the level of individuals, where a virus adapts to specific features of the host, but also on the population level, because HLA alleles differ in their frequency across geographic regions\nolinebreak\cite{Kawashima2009}.
+Upon transmission to a new host, HLA escape mutations can revert to their wild type, as HLA escape mutations are associated with a reduction in viral replicative capacity\nolinebreak\cite{Matthews2008}.
+Kawashima et al.
+\nolinebreak\cite{Kawashima2009} describe an escape mutation that is selected by HLA allele HLA-B*51, does not strongly affect viral replicative capacity, and therefore slowly enriches over time in Japan, where HLA-B*51 commonly occurs.
 
-Upon transmission to a new host, HLA escape mutations can revert to their wild type,
-as HLA escape mutations are associated with a reduction in viral replicative capacity
-\nolinebreak\cite{Matthews2008}. Kawashima et al.\nolinebreak\cite{Kawashima2009}
-describe an escape mutation that is selected by HLA allele HLA-B*51,
-does not strongly affect viral replicative capacity, and therefore slowly enriches
-over time in Japan, where HLA-B*51 commonly occurs.
-How quickly a given escape mutation is selected upon transmission in a host depends on the
-magnitude of the reduction in viral replicative capacity, on the strength of selection
-pressure, and also on the genetic background, e.g. some escape mutations require compensatory 
-mutations which partly attenuate the negative impact on viral replicative capacity.
+How quickly a given escape mutation is selected upon transmission in a host depends on the magnitude of the reduction in viral replicative capacity, on the strength of selection pressure, and also on the genetic background, e.g.
+some escape mutations require compensatory mutations which partly attenuate the negative impact on viral replicative capacity.
 
-Studying HLA escape provides an unique opportunity to gain insight into
-viral evolution, on the host level, but also on the population level.
+Studying HLA escape provides a unique opportunity to gain insight into viral evolution, on the host level, but also on the population level.
 Unfortunately, identifying HLA escape mutations is difficult in practice.
@@
 
-There are several experimental methods available to study HLA escape:
-Recombinant MHC-I molecules can be used in binding assays. Upon complex formation
-with a peptide, a change in conformation can be detected with conformation-specific
-antibodies. This method is relatively fast, but only measures binding affinity of a
-peptide to MHC-I and does not account for antigen processing or immunodominance, 
-which describes the observation that a peptide may be presented via MHC-I on the cell 
-surface, but does not induce an immune response.
-An experimental setup that resembles the conditions in-vitro more closely but is also
-more time-consuming is to measure
-CD8+ T cell responses instead. This is usually done by stimulating peripheral blood
-mononuclear cells with prototype and variant peptides and measuring the secretion of
-IFN-\gamma{} by intracellular cytokine
-staining and fluorescence-assisted cell sorting.
-To analyze CD8+ T cell responses against endogenously processed antigens, it is necessary
-to generate cell-lines stably expressing the antigen in question and adding antigen-specific
-CD8+ T cells. This method scales poorly as it requires transfection of cell lines and
-antigen-specific expansion of CD8+ T cells.
+There are several experimental methods available to study HLA escape: Recombinant MHC-I molecules can be used in binding assays.
+Upon complex formation with a peptide, a change in conformation can be detected with conformation-specific antibodies.
+This method is relatively fast, but only measures binding affinity of a peptide to MHC-I and does not account for antigen processing or immunodominance, which describes the observation that a peptide may be presented via MHC-I on the cell  surface, but does not induce an immune response.
+
+An experimental setup that resembles the conditions in-vitro more closely but is also more time-consuming is to measure CD8+ T cell responses instead.
+This is usually done by stimulating peripheral blood mononuclear cells with prototype and variant peptides and measuring the secretion of IFN-\gamma{} by intracellular cytokine staining and fluorescence-assisted cell sorting.
+
+To analyze CD8+ T cell responses against endogenously processed antigens, it is necessary to generate cell-lines stably expressing the antigen in question and adding antigen-specific CD8+ T cells.
+This method scales poorly as it requires transfection of cell lines and antigen-specific expansion of CD8+ T cells.
 
@@
 
-Because the currently available experimental methods do not scale well enough to analyze
-whole viral genomes, a useful strategy might be to use annotated sequence data to identfiy 
-candidate HLA escape mutations that can then be verified experimentally.
+Because the currently available experimental methods do not scale well enough to analyze whole viral genomes, a useful strategy might be to use annotated sequence data to identify candidate HLA escape mutations that can then be verified experimentally.
 
-As the selection pressure exerted by cytotoxic T cells depends on successful recognition
-of viral peptides on the cell surface, escape mutations are often HLA-allele specific
-and can therefore be detected as HLA-allele dependent footprints in sequence
-alignments of viral proteins\nolinebreak\cite{Moore2002}: At certain alignment positions, a replacement
-might be more frequently observed in sequences from hosts with a specific HLA allele than in 
-sequences from hosts without that HLA allele. By quantifying this difference
-for all replacement and HLA allele pairs, it is possible to idenfify replacements that
-are enriched in sequences coming from hosts with certain HLA alleles, and thus are
-likely to be HLA escape mutations.
+As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface, escape mutations are often HLA-allele specific and can therefore be detected as HLA-allele dependent footprints in sequence alignments of viral proteins\nolinebreak\cite{Moore2002}: At certain alignment positions, a replacement might be more frequently observed in sequences from hosts with a specific HLA allele than in  sequences from hosts without that HLA allele.
+By quantifying this difference for all replacement and HLA allele pairs, it is possible to identify replacements that are enriched in sequences coming from hosts with certain HLA alleles, and thus are likely to be HLA escape mutations.
 
-One way of quantifying this enrichment is Fisher's exact test\nolinebreak\cite{Fisher1922}.
-For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\),
-a 2-by-2 contingency table is constructed containing the absolute counts of the
-number of sequences in the four possible categories: 
-(\(R_{i}\), \(H\)), (\(R_{i}\), \(!H\)), (\(!R_{i}\), \(H\)) and (\(!R_{i}\), \(H\)),
-where \(!R_{i}\) denotes any replacement except \(R_{i}\), and \(!H\) denotes any HLA allele
-except \(H\).
-Fisher's exact test is a conventional nullhypothesis significance test (NHST) that
-generates p-values. In this case, the nullhypothesis is that HLA allele \(H\) and 
-replacement \(R_{i}\) are independent, and the p-value is the probability of
-observing a deviation from independence that is at least as extreme as in the data at hand
-under the assumption that the nullhypothesis is true.
+One way of quantifying this enrichment is Fisher's exact test\nolinebreak\cite{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(!H\)), (\(!R_{i}\), \(H\)) and (\(!R_{i}\), \(H\)), where \(!R_{i}\) denotes any replacement except \(R_{i}\), and \(!H\) denotes any HLA allele except \(H\).
 
-Fisher's exact test has the advantage of being easy to apply\nolinebreak\cite{Budeus2016}, 
-but also has several disadvantages that are outlined in Carlson et al. 
-\nolinebreak\cite{Carlson2008}.
-The most striking one is that viral sequences share a common phylogenetic history, and, 
-therefore, treating sequences as independent and identically distributed samples may under- or
-overestimate effect sizes. In the context of hypothesis testing, this leads to
-increased false positive and false negative rates.
-Another issue of applying Fisher's exact test is that HLA class I loci are located in
-close proximity on chromosome 6 and are therefore in linkage disquilibrium, which means 
-that HLA alleles are not inherited completely independent of each other, i.e.
+Fisher's exact test is a conventional null hypothesis significance test (NHST) that generates p-values.
+In this case, the null hypothesis is that HLA allele \(H\) and  replacement \(R_{i}\) are independent, and the p-value is the probability of observing a deviation from independence that is at least as extreme as in the data at hand under the assumption that the null hypothesis is true.
+
+Fisher's exact test has the advantage of being easy to apply\nolinebreak\cite{Budeus2016},  but also has several disadvantages that are outlined in Carlson et al.\nolinebreak\cite{Carlson2008}.
+The most striking one is that viral sequences share a common phylogenetic history, and,  therefore, treating sequences as independent and identically distributed samples may under- or overestimate effect sizes.
+In the context of hypothesis testing, this leads to increased false positive and false negative rates.
+
+Another issue of applying Fisher's exact test is that HLA class I loci are located in proximity on chromosome 6 and are therefore in linkage disequilibrium, which means that HLA alleles are not inherited completely independent of each other, i.e.
 inheritance of one HLA allele correlates with inheritance of another HLA allele.
-When using a statistical method that tests each HLA allele individually without
-considering the whole set of alleles present, spurious associations might occur: 
-If HLA allele \(H_{1}\) is associated with an amino acid replacement \(R\), but
-\(H_{1}\) is in linkage disequilibrium with another allele \(H_{2}\), this also means that
-we observe an association between replacement \(R\) and \(H_{2}\), even in
-absence of any underlying escape mechanism.
+When using a statistical method that tests each HLA allele individually without considering the whole set of alleles present, spurious associations might occur: If HLA allele \(H_{1}\) is associated with an amino acid replacement \(R\), but \(H_{1}\) is in linkage disequilibrium with another allele \(H_{2}\), this also means that we observe an association between replacement \(R\) and \(H_{2}\), even in absence of any underlying escape mechanism.
 
 Correlations can not only occur between HLA alleles, but also between replacements.
-This kind of codon covariation occurs for example in compensatory mutations that
-attenuate the negative impact of immune escape mutations. For instance,
-a compensatory mutation might lead to a conformational change in such a way that
-the mutated protein resembles the original wild type more closely.
+This kind of codon covariation occurs for example in compensatory mutations that attenuate the negative impact of immune escape mutations.
+For instance, a compensatory mutation might lead to a conformational change in such a way that the mutated protein resembles the original wild type more closely.
 
-Carlson et al. \nolinebreak\cite{Carlson2008} developed a method called 
-Phylogenetic Dependency Network that accounts for phylogenetic bias, 
-HLA linkage disequilibrium, and codon covariation and is based on 
-nullhypothesis significance testing.
+Carlson et al.
+\nolinebreak\cite{Carlson2008} developed a method called  Phylogenetic Dependency Network that accounts for phylogenetic bias,  HLA linkage disequilibrium, and codon covariation and is based on null hypothesis significance testing.
 
@@
 
-In addition to the aforementioned biological reasons why Fisher's exact test is 
-not suited for the analysis of sequence data, there are also more fundamental statistical
-issues that universally occur when using p-values as a screening tool\nolinebreak\cite{Amrhein2017}.
+In addition to the aforementioned biological reasons why Fisher's exact test is  not suited for the analysis of sequence data, there are also more fundamental statistical issues that universally occur when using p-values as a screening tool\nolinebreak\cite{Amrhein2017}:
 
-In the presence of small effect sizes and high variance between measurements, as 
-it is typically the case when working with biological data, statistically significant
-results can often be misleading and are likely to be in the wrong direction
-(a so-called type S error) or greatly overestimate an effect (a so-called type M error)
-\nolinebreak\cite{Gelman2014}.
-This problem has recently been widely appreciated in the literature in the context
-of the current \glqq replication crisis\grqq{}, which describes the circumstance that
-scientific claims with seemingly strong statistical evidence fail to replicate\nolinebreak\cite{Ioannidis2005}.
+In the presence of small effect sizes and high variance between measurements, as  it is typically the case when working with biological data, statistically significant results can often be misleading and are likely to be in the wrong direction (a so-called type S error) or greatly overestimate an effect (a so-called type M error)\nolinebreak\cite{Gelman2014}.
+This problem has recently been widely appreciated in the literature in the context of the current \glqq replication crisis\grqq{}, which describes the circumstance that scientific claims with seemingly strong statistical evidence fail to replicate\nolinebreak\cite{Ioannidis2005}.
 
-Another issue that occurs when using p-values as a screening tool is the problem
-of multiple comparisons. When applying a statistical test, the probability  of 
-obtaining a statistically significant result increases with
-each additional test, even in absence of any real effect. When using p-values as a filter,
-it is therefore likely to obtain significant effects that are in fact not real.
-To circumvent this problem, a common strategy is to control the false discovery
-rate, which is the expected proportion of false positives\nolinebreak\cite{Benjamini1995}.
-These adjustment procedures have the problem that, when performing many of such comparisons,
-none but the very largest effects remain.
-Instead of performing many hypothesis tests and trying to adjust for them, we instead prefer
-to fit a single, multilevel model that contains all comparisons of interest.
-When using multilevel models, the problem of multiple comparisons can disappear entirely and 
-also yield more valid estimates\nolinebreak\cite{Gelman2012}.
+Another issue that occurs when using p-values as a screening tool is the problem of multiple comparisons.
+When applying a statistical test, the probability  of  obtaining a statistically significant result increases with each additional test, even in absence of any real effect.
+When using p-values as a filter, it is therefore likely to obtain significant effects that are in fact not real.
+To circumvent this problem, a common strategy is to control the false discovery rate, which is the expected proportion of false positives\nolinebreak\cite{Benjamini1995}.
+These adjustment procedures have the problem that, when performing many of such comparisons, none but the very largest effects remain.
+
+Instead of performing many hypothesis tests and trying to adjust for them, we instead prefer to fit a single, multilevel model that contains all comparisons of interest.
+When using multilevel models, the problem of multiple comparisons can disappear entirely and  also yield more valid estimates\nolinebreak\cite{Gelman2012}.
 
@@
 
-When possible, we choose to fit Bayesian models. By using prior information,
-adding problem-specific structure and partial pooling, the accuracy of estimates
-can often be noticably improved\nolinebreak\cite{Gelman2010}. Prior information does not necessarily mean to
-use external data, even a rough idea about the expected magnitude of estimates is often
-surprisingly effective.
-Additionally, Bayesian statistics provides an accessible way to test models:
-By comparing data generated under the model's assumptions to the actually observed data,
-it is possible to identify important aspects of the dataset that the models fails to
-capture and subsequently improve the model until it is consistent with the observed data
-\nolinebreak\cite{Gabry2019}.
+When possible, we choose to fit Bayesian models.
+By using prior information, adding problem-specific structure and partial pooling, the accuracy of estimates can often be noticeably improved\nolinebreak\cite{Gelman2010}.
+Prior information does not necessarily mean to use external data, even a rough idea about the expected magnitude of estimates is often surprisingly effective.
 
-In the context of identifying HLA-associated mutations we propose to improve existing
-methods by the following additions, which can be broadly divided into additional information
-and model structure:
+Additionally, Bayesian statistics provides an accessible way to test models: By comparing data generated under the model's assumptions to the actually observed data, it is possible to identify important aspects of the dataset that the models fails to capture and subsequently improve the model until it is consistent with the observed data\nolinebreak\cite{Gabry2019}.
+
+In the context of identifying HLA-associated mutations we propose to improve existing methods by the following additions, which can be broadly divided into  additional information and model structure:
 \subsection*{Additional information}
@@
 
-    HLA-associated mutations are expected to lie more frequently in regions of known
-    epitopes. There are vast experimental binding affinity data available of different
-    peptides and MHC I molecule pairs, and there are well-established computational
-    methods that use these data to extrapolate HLA binding for untested peptides.
-    We show that, by including the outcome of these computational tools as input for a
-    probabilistic model, the prediction of HLA-associated mutations can be improved.
+    HLA-associated mutations are expected to lie more frequently in regions of known epitopes.
+There are vast experimental binding affinity data available of different peptides and MHC I molecule pairs, and there are well-established computational methods that use these data to extrapolate HLA binding for untested peptides.
+We show that, by including the outcome of these computational tools as input for a probabilistic model, the prediction of HLA-associated mutations can be improved.
 
@@
 
-  Similarly, there are also HLA-allele independent effects like antigen processing
-  that influence presentation on MHC I. Second generational tools like
-  MHCFlurry 2.0 use both binding affinity and antigen processing data to improve
-  epitope prediction. For our tool HAMdetector, we use the output of MHCFlurry to
-  benefit from this binding affinity and antigen processing data in order to predict
-  HLA-associated mutations.
+    Similarly, there are also HLA-allele independent effects like antigen processing that influence presentation on MHC I.
+Second generational tools like MHCFlurry 2.0 use both binding affinity and antigen processing data to improve epitope prediction.
+For our tool HAMdetector, we use the output of MHCFlurry to benefit from this binding affinity and antigen processing data in order to predict HLA-associated mutations.
 
@@
   Recent advantages in Bayesian inference include so-called sparsity-promoting priors.
-  Sparsity-promoting priors convey the apriori expectation that most coefficients in 
-  a regression model are close to 0.
-  This assumption also applies to HLA-associated mutations, because the number of epitopes
-  that are restricted by a given HLA allele is typically small compared to the number
-  of all possible epitopes. If no epitope spanning a given alignment position is presented
-  on the MHC I molecule, any association between a replacement and the respective HLA allele 
-  is likely to be due to random variation alone (or HLA linkage disequilibrium).
+Sparsity-promoting priors convey the a priori expectation that most coefficients in a regression model are close to 0.
+This assumption also applies to HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes.
+If no epitope spanning a given alignment position is presented on the MHC I molecule, any association between a replacement and the respective HLA allele  is likely to be due to random variation alone (or HLA linkage disequilibrium).
 
-  It has been shown that using a sparsity-promoting prior when non-zero coefficients
-  are sparse can drastically improve predictive performance, because the model is
-  better able to differentiate between signal and noise.
+  It has been shown that using a sparsity-promoting prior when non-zero coefficients are sparse can drastically improve predictive performance, because the model is better able to differentiate between signal and noise.
 
-  We show that a simple logistic regression model with a sparsity-promoting prior
-  on the regression coefficients for the HLA alleles alone already performs roughly on-par 
-  with the much more elaborate Phylogenetic Dependency Network approach from Carlson et al.
-  when using the fraction of identified escapes that lie in known HLA epitopes as a
-  benchmark.
+  We show that a simple logistic regression model with a sparsity-promoting prior on the regression coefficients for the HLA alleles alone already performs roughly on-par  with the much more elaborate Phylogenetic Dependency Network approach from Carlson et al.
+when using the fraction of identified escapes that lie in known HLA epitopes as a benchmark.
 
@@
 
-  It has recently been appreciated that binding specifities can vary drastically across 
-  HLA alleles from the same allele group, e.g. between HLA-B*51:01 and HLA-B*51:03.
-  For predicting HLA associated mutations, there has consequently been a shift to use
-  4-digit resolution data whenever possible.
-  This is not without downsides however, because overall, binding specifities
-  are more similar for alleles in the same group, and, therefore, treating them
-  as completely separate might unnecessarily fragment the available data.
+  It has recently been appreciated that binding specificities can vary drastically across  HLA alleles from the same allele group, e.g.
+between HLA-B*51:01 and HLA-B*51:03.
+For predicting HLA associated mutations, there has consequently been a shift to use 4-digit resolution data whenever available.
+This is not without downsides however, because overall, binding specificities are more similar for alleles in the same group, and, therefore, treating them as completely separate might unnecessarily fragment the available data.
 
-  In Bayesian statistics, it is not required to chose one extreme or the other:
-  Instead of either treating all HLA alleles from the same allele group as identical or
-  as completely separate, partial pooling of the estimates allows something inbetween:
-  By partial pooling, estimates do share some information across each other, but they are
-  also allowed to vary if necessary.
+  In Bayesian statistics, it is not required to chose one extreme or the other: Instead of either treating all HLA alleles from the same allele group as identical or as completely separate, partial pooling of the estimates allows something in between: By partial pooling, estimates do share some information across each other, but they are also allowed to vary if necessary.
 
-  In the context of our model this means that observing a relevant association between
-  a replacement and an HLA allele also makes the model more inclined to estimate
-  relevant associations between that replacement and all other alleles from that allele group.
-  The degree of this partial pooling can be estimated from the data, so if we do observe
-  strong similarity of HLA alleles in a given allele group the estimates are more
-  influenced by each other than if they do not.
+  In the context of our model this means that observing a relevant association between a replacement and an HLA allele also makes the model more inclined to estimate relevant associations between that replacement and all other alleles from that allele group.
+The degree of this partial pooling can be estimated from the data, so if we do observe strong similarity of HLA alleles in a given allele group the estimates are more influenced by each other than if they do not.
 
@@
   \item logistic regression model, because easily extensible
-  \item coefficients are interpretable as expected increase in log-odds of observing 
-    a given replacement (also: average predictive comparisons to transform estimates to
-    probability scale)
+  \item coefficients are interpretable as expected increase in log-odds of observing  a given replacement (also: average predictive comparisons to transform estimates to probability scale)
   \item probabilistic model allows us to perform posterior predictive checks
   \item disadvantages: none? maybe log-odds scale unintuitive. mention divide-by-4 rule
-  \item  establish notation, start simple and expand on that? Maybe confusing because
-    some things change as models get more complex 
+  \item  establish notation, start simple and expand on that? Maybe confusing because some things change as models get more complex 
 \end{itemize}
@@
 \begin{itemize}
-  \item sequence data shares common phylogenetic history and are therefore not i.i.d,
-    cite Brumme paper
-  \item standard approach: multivariate normal intercept where covariance matrix is based on 
-    shared branch lengths
-  \item this approach is computationally too expensive because of current limitations in 
-    Stan. Would be possible by fitting the model in Julia? Probably omit that.
+  \item sequence data shares common phylogenetic history and are therefore not i.i.d, cite Brumme paper
+  \item standard approach: multivariate normal intercept where covariance matrix is based on shared branch lengths
+  \item this approach is computationally too expensive because of current limitations in Stan.
+Would be possible by better inferencealgorithms, see this recently published paper: \sloppy\url{https://proceedings.neurips.cc/paper/2020/file/673de96b04fa3adcae1aacda704217ef-Paper.pdf} but probably omit that?
   \item Our approach is close to the one in Carlson et al.
-  \item slight tweak: use tree estimation tools for tree likelihoods p(tree | sequences),
-    then bayes rule to invert that conditional probability. Optimize branch lengths
-    for each replacement, but keep tree topology fixed
+  \item slight tweak: use phylogeny estimation tools for tree likelihoods p(tree | sequences), then bayes rule to invert that conditional probability.
+Optimize branch lengths for each replacement, but keep tree topology fixed
   \item advantages: faster (raxml well optimized)
-  \item disadvantages: ad-hoc, close to conventional multivariate normal approach with 
-    off-diagonal elements set to 0, therefore not as flexible. But: should be good enough,
-    phylogeny seems to be not that important (probably omit that?)
+  \item disadvantages: ad-hoc, close to conventional multivariate normal    approach with off-diagonal elements set to 0, therefore not as flexible.
+But: should be good enough, phylogeny seems to be not that important (probably omit that?)
   \item how to include p(replacement | tree) in the model: just add additional intercept
-  \item add coefficient for that information, so model can decide how to make use of
-    that information.
+  \item add coefficient for that information, so model can decide how to make use of that information.
   \item logit(p(replacement|tree)) used because it cancels out with the logistic
-    link function. therefore: phylogeny information as baseline in absence of any 
-    HLA effects
-  \item calibration plots to show that estimated phylogenetic probabilities are
-   reasonably well calibrated.
-  \item meaning: for sequences with an expected probability of observing the replacement
-    of x\%, we really do see that x\% of the sequences have that replacement.
+    link function. therefore: phylogeny information as baseline in absence of any HLA effects
+  \item calibration plots to show that estimated phylogenetic probabilities are reasonably well calibrated.
+  \item meaning: for sequences with an expected probability of observing the replacement of x\%, we really do see that x\% of the sequences have that replacement.
 \end{itemize}
@@
 \begin{itemize}
-  \item HLA linkage disequilibrium is accounted for by considering all HLA alleles
-    of a sample at once.
-  \item If strong correlations between HLA alleles: marginal posterior for both of them
-    overlap with 0
+  \item HLA linkage disequilibrium is accounted for by considering all HLA alleles of a sample at once.
+  \item If strong correlations between HLA alleles: marginal posterior for both of them overlap with 0
   \item Codon covariation is not accounted for because we are mainly interested in HAMs.
-  \item Model could be extended to include codon covariation by including them as additional
-    predictors. This is possible because we use the horseshoe prior.
+  \item Model could be extended to include codon covariation by including them as additional predictors. This is possible because we use the horseshoe prior.
 \end{itemize}
@@
 \begin{itemize}
-  \item one of the recent advances in Bayesian inference: development of
-    spacial priors to give model useful properties
-  \item we expect most HLA coefficients to be 0, because no epitope spanning that region
-    is presented on the cell surface, therefore no selection pressure and no association
-    with possible replacements
+  \item one of the recent advances in Bayesian inference: development of spacial priors to give model useful properties
+  \item we expect most HLA coefficients to be 0, because no epitope spanning that region is presented on the cell surface, therefore no selection pressure and no association with possible replacements
   \item this is prior knowledge that should be incorporated in the model
-  \item leads to better inferences because uncertainty of the regression coefficients
-    does not propagate into observations.
-  \item works by placing most probability mass very close to 0, but with large tails
-    for non-zero coefficients.
+  \item leads to better inferences because uncertainty of the regression coefficients does not propagate into observations.
+  \item works by placing most probability mass very close to 0, but with large tails for non-zero coefficients.
   \item show formula
-  \item global shrinkage parameter is very small and shrinks most coefficients to 0,
-    local shrinkage parameter is very large and allows some coefficients to escape that 
-    shrinkage
+  \item global shrinkage parameter is very small and shrinks most coefficients to 0, local shrinkage parameter is very large and allows some coefficients to escape that shrinkage
   \item prior can be specified in terms of expected number of non-zero coefficients
-  \item in our case, degree of sparsity can be well estimated from the data because all
-    replacements share the same global shrinkage parameter.
-  \item useful addition for logistic regression models: regularized horseshoe
-    to have some regularization for non-zero coefficients. this helps to deal with
-    issues of separability and collinearity.
-  \item effect of horseshoe prior compared to standard logistic regression model:
-    refer to model testing section. In our benchmark, the horseshoe prior alone is more
-    effective than the model with phylogeny.
+  \item in our case, degree of sparsity can be well estimated from the data because all replacements share the same global shrinkage parameter.
+  \item useful addition for logistic regression models: regularized horseshoe to have some regularization for non-zero coefficients. this helps to deal with issues of separability and colinearity.
+  \item effect of horseshoe prior compared to standard logistic regression model: refer to model testing section.
+In our benchmark, the horseshoe prior alone is more effective than the model with phylogeny.
   \item maybe include comparison of marginal posteriors with and without horseshoe prior
@@
     input to a probabilitic model
-  \item MHCFlurry is used for epitope prediction and Antigen processing prediction:
-    Input matrix (dimensions: \#replacements x \#HLA alleles) contains a 1 if that
-    position is either expected to be inside a predicted epitope or related to
-    antigen processing, 0 otherwise
-  \item uses some thresholds, but this is not so bad because we just use it as
-    input for a probabilistic model. Maybe play around with different options?
-    The Brumme paper uses some offsets for defining the region of known epitopes,
-    maybe this does make a difference.
-  \item including this kind of information can drastically improve predictive performance,
-    as it is relevant external data.
-  \item Bayesian inference allows us to include informationen from different sources in
-    a consistent manner.
-  \item this information may be imprecise or wrong, but the model is able to learn how much 
-    to trust this data if parameterized correctly.
-  \item epitope prediction / antigen processing escape is information about the expected
-    degree of sparsity, i.e. some coefficients are more likely to be non-zero than others.
+  \item MHCFlurry is used for epitope prediction and Antigen processing prediction: Input matrix (dimensions: \#replacements x \#HLA alleles) contains a 1 if that position is either expected to be inside a predicted epitope or related to antigen processing, 0 otherwise
+  \item uses some thresholds, but this is not so bad because we just use it as input for a probabilistic model.
+Maybe play around with different options?
+The Brumme paper uses some offsets for defining the region of known epitopes, maybe this does make a difference.
+  \item including this kind of information can drastically improve predictive performance, as it is relevant external data.
+  \item Bayesian inference allows us to include informationen from different sources in a consistent manner.
+  \item this information may be imprecise or wrong, but the model is able to learn how much to trust this data if parameterized correctly.
+  \item epitope prediction / antigen processing escape is information about the expected degree of sparsity, i.e. some coefficients are more likely to be non-zero than others.
   \item can be included into the model by varying the local shrinkage parameter term.
-  \item higher standard deviation of the local shrinkage parameters mean that
-    it is more likely that this coefficient is non-zero.
+  \item larger standard deviation of the local shrinkage parameters means that it is more likely that this coefficient is non-zero.
   \item show formula
@@
 \begin{itemize}
-  \item (this still has to be implemented, but should be a nice feature. If this
-    is going to be too time consuming I can just leave it out)
-  \item realized through allele group specific intercepts, e.g. HLA-B\*51:01 and 
-  HLA-B\*51:03, ..., share a common intercept.
+  \item (this still has to be implemented, but should be a nice feature. If this is going to be too time-consuming I can just leave it out)
+  \item realized through allele group specific intercepts, e.g. HLA-B\*51:01 and  HLA-B\*51:03, ..., share a common intercept.
   \item show formula
@@
   \item show all formulas
-  \item all models were implementen in Stan. Code is available online in two versions:
-  one optimized for readability, one optimized for speed with multithreading and GPU support
+  \item all models were implemented in Stan. Code is available online in two versions: one optimized for readability, one optimized for speed with multithreading and GPU support
   \item data can be analyzed using a custom Julia package (tested on linux).
@@
 \begin{itemize}
-  \item justify priors, mostly based on the expected scales.
+  \item justify priors, mostly based on the expected scales. Compare to some 
+  Betancourt/Gelman papers to see how it is done there.
 \end{itemize}
 
-\section{Results and Discussion}
+\section{Results}
 
+\begin{itemize}
+  \item to show that model works well in several real-world scenarios, versions of the model are run on different datasets: large Brumme HIV, smaller Arevir HIV, larger Düsseldorf HBV, HDV by Michael Roggendorf, Rongge small HIV
+  \item MCMC convergence diagnostics reveal no sampling issues
+  \item general posterior predictive checks are applied to show that model explains data well
+  \item two forms of model checking: external testing through comparisons of known epitopes and leave-one-out cross-validation.
+\end{itemize}
+
+\subsubsection*{Convergence diagnostics}
+
+\begin{itemize}
+  \item inference was done with MCMC, using Stan.
+  \item One step in Bayesian workflow: checking inference with convergence diagnostics.
+  \item Stan code is available online.
+  \item All model fits show no signs of inference issues, all Rhats less than 1.01,
+    effective sample size for all model parameters greater than 300.
+\end{itemize}
+
+\subsubsection*{Posterior predictive checks}
+
+\begin{itemize}
+  \item Bayesian modeling allows a simple yet effective strategy to test models: By simulating data under the model's assumptions and comparing simulated and observed data, model misspecifications can be identified and show possible model improvements.
+  \item One way to perform posterior predictive checks for models with binary outcomes: calibration plots.
+  \item cite Gelman dog shock PPC paper
+  \item show calibration plot
+  \item calibration plots are similar to the one showed in phylogeny, but this time test the predictions of the whole model, and not just the phylogeny component
+  \item show example plot, PPCs for all the models are shown in the supplementary
+\end{itemize}
+
+\subsection*{Comparison to list of known epitopes}
+
+\begin{itemize}
+  \item When testing models, we prefer to evaluate them on real-world data whenever possible
+  \item comparing against real-world data is not straight forward for testing HAMs, because: 1.) if a tool identifies a HLA-associated mutation which is not yet documented, this could be either because it is a false positive, or a yet unknown HAM that has not been documented before.
+2.) Viewing every not identified HAM that is listed in the literature as a false negative is not correct either, because what is documented are escape mutations, which do not necessarily have to show in a sequence alignment.
+3.) What is and is not a HAM is not a binary decision, so the true positive / false negative framework does not work as well.
+  \item To circumvent these issues and still compare against real-work data, we chose the following strategy:
+  \item for each model, all evaluated replacements are ranked by decreasing confidence of being an HLA associated mutations.
+  \item Then, a list of known epitopes is used to create a plot of the cumulative number of replacements in known epitopes vs the corresponding rank.
+The assumption for this kind of model check is that we expect to see an
+  enrichment of replacements that are in known epitopes and that this enrichment is stronger for better performing models.
+  \item We compare several different models on all data sets available. 
+  \item Observations:
+  \item The logistic regression model without any additional information performs about as well as Fisher's exact test.
+  \item The horseshoe prior alone is a drastic improvement over Fisher's exact test and the logistic regression model with non-sparsifying priors, even though it does not include any external information.
+The benefit comes from providing the model the information that most coefficients are expected to be very close to 0.
+  \item Models improve as we add additional information
+  \item Adding epitope prediction does improve predictive performance considerably.
+Note that the model only uses epitope prediction software and does not utilize any information of experimentally confirmed epitopes, which we only use for model evaluation.
+Therefore, our method even works in absence of any additional epitope information.
+  \item add plot
+\end{itemize}
+
+\subsection{Leave-one-out cross-validation}
+
+\begin{itemize}
+  \item In addition to testing the model against real-world epitopes, we also perform leave-one-out cross-validation
+  \item unlike classical application of cross-validation, we are not that interested in the absolute magnitude of the prediction errors, i.e.
+how well we can estimate the occurrence of a replacement at a given alignment position, because the main objective of the model is to quantify the contribution of the HLA alleles.
+  \item But, LOO gives useful insight into how the accuracy of predictions change as we include more information in the model. 
+  \item Performing LOO has historically been challenging to apply for Bayesian models, because fitting the model N times is infeasible because of computational constrains.
+  \item Recently, Pareto-smoothed importance sampling has gained popularity.
+It approximates the LOO posterior from the full posterior and has the advantage that it allows to perform LOO-CV by only fitting the model once has good diagnostics to see if that approximation is unreliable.
+  \item cite PSIS paper
+  \item For Bayesian models, it is possible to use elpd as quantification of model performance. elpd is defined as -formula here-, i.e. the predictive density at the observed data point.
+  \item This has the advantage over other performance measures like classification accuracy that it not only takes the location of the predictive distribution into account (i.e.
+the number of correct predictions), but also the width (i.e.
+how confident the model is in its predictions).
+  \item Table shows LOO results for one dataset of serveral different models: A classical logistic regression model, similar to the one first used by Moore et al, a logistic regression model with horseshoe prior, a logistic regression model with horseshoe prior and phylogeny, and the complete model with horseshoe prior, phylogeny and epitope information.
+  \item Results are as expected, models with more information have higher predictive accuracy.
+Note that this is not due to overfitting, because LOO approximates model performance on unseen data.
+Additionally, the results are consistent across all datasets (see appendix).
+  \item Note how the model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any additional external data.
+This is because including the sparsity assumptions allows the model to better separate signal from noise and the uncertainty of the close-to-zero coefficients does not propagate into uncertainty of the estimated y.
+  \item Including phylogeny further improved model performance, as the assumption of i.i.d data does not hold for sequences that share a common phylogenetic history.
+  \item Perhaps surprisingly, the model with epitope prediction performs roughly as well as the model without.
+This is true from a perspective of predictive accuracy.
+However, in determining which HLA alleles are associated with the replacement, adding epitope prediction is highly useful, as shown in the previous section.
+This can be explained by the fact that a.) a given HLA allele is only restrincting a subset of all position in a sequence alignment and b.) the benefit of accurately identifying a relevant HLA allele is only relevant for the samples which do have that allele, which is only a small subset of all available samples.
+Therefore, looking at predictive accuracy alone is misleading in that regard, because we are more interested in the predictors than in the predictions itself.
+\end{itemize}
+
+\subsection{A real-world example}
+
+\begin{itemize}
+  \item We had the opportunity to test our model against a set of false positives that were identified by Fisher's exact test.
+On the HDV dataset, Fisher's exact test was applied to a set of HLA annotated sequences and x statistically significant results were reported.
+The identified positions were experimentally validated and y positions turned out to be false-positives, where Fisher's exact test showed extremely low p-values but experimental validation failed (how?).
+  \item This kind of data is rare because negative results are usually not published in the literature.
+  \item We ran HAMdetector on the same dataset, the posterior probabilities are shown in table x.
+  \item Note that a larger posterior probability denotes that the model is more confident in that HLA alleles being a HAM.
+  \item For most of the false positives we observe low posterior probabilities, which means that the model correctly identifies these alleles as not being relevant.
+  \item X of the false positives remain, which means that the model is not perfect, but a strong improvement over fisher's exact test.
+  \item these results were published previously using a preliminary version of HAMdetector, cite Roggendorf manuscript.
+\end{itemize}
+
+\subsection{Rongge dataset}
+\begin{itemize}
+  \item We ran HLAdetector on a set of X HIVSequences sampled in China. X associations have a posterior mass greater than Y.
+  \item Maybe show a plot of marginal posteriors.
+\end{itemize}
+
+\section{Discussion}
+
+\begin{itemize}
+  \item We present HAMdetector, a Bayesian model to identify HLA-associated mutations in a sequence alignment
+  \item We propose to use models that include as much information as possible and have shown that including sparsity and epitope prediction software can achieve better performance than existing methods.
+  \item The main advantage of using a logistic regression backbone is that it is easy to modify and the coefficients are interpretable as changes on the log-odds scale, like with a standard logistic regression model.
+  \item We can believe the general model structure can also be used in other contexts, for example in identifying associations in Sequence data and other features.
+  \item Maybe some additional points?
+\end{itemize}
 \thispagestyle{empty} % Removes page numbering from the first page

### 2020-12-16 docs/manuscript.tex
add Model backbone text

@@
 \usepackage{hyperref}
+\usepackage{amsmath}
 % \usepackage{lipsum} % Required to insert dummy text. To be removed otherwise
@@
 \subsection*{Model backbone}
-\begin{itemize}
-  \item logistic regression model, because easily extensible
-  \item coefficients are interpretable as expected increase in log-odds of observing  a given replacement (also: average predictive comparisons to transform estimates to probability scale)
-  \item probabilistic model allows us to perform posterior predictive checks
-  \item disadvantages: none? maybe log-odds scale unintuitive. mention divide-by-4 rule
-  \item  establish notation, start simple and expand on that? Maybe confusing because some things change as models get more complex 
-\end{itemize}
+
+We chose a logistic regression model as a backbone for two reasons: First, because it is easily extensible and secondly, because the coefficients are interpretable as expected increases on the log-odds scale, identical to the familiar interpretation of standard logistic regression models.
+We also chose to fit the model in a Bayesian fashion using Stan, mostly because it allows us to make use of the hierarchical structure in the data and the possibility to use posterior predictive checks as a way of model checking.
+Consider the following notation:
+
+\begin{align}
+  y_{ij} & \sim \text{bernoulli}(\theta_{ij}) \\
+  \theta_{ij} & = \text{inv\_logit}(\beta_{0_{j}} + X_{i}\beta_{\text{HLA}_{j}})
+\end{align},
+
+where \(y_{ij}\) is the binary encoded observation at sequence \(i\) for replacement \(j\), \(\theta_{ij}\) is the estimated probability that we observe replacement \(j\) in sequence \(i\), \(X_{i}\) is the binary encoded (row) vector of HLA annotations for sequence \(i\), and \(\beta_{\text{HLA}_{j}}\) is the (column) vector of HLA regression coefficients for replacement \(j\) and  \(\text{inv\_logit}\) is the logistic link function.
+Note that, if we observe N different amino acid states at a certain alignment position, each of those are considered in turn to be the "replacement".
+
+The regression coefficients \(\beta_{\text{HLA}_{j}}\) are the main focus of interest, as they quantify the association between the occurrence of replacement \(j\) and each of the observed HLA alleles.
+They are in units of log-odds, so a regression coefficient of somewhere around 2 means that when comparing sequences with and without that particular HLA allele, the log-odds of observing replacement \(j\) in presence of that HLA allele is increased by 2.
+
+Reasoning on the log-odds can sometimes be unintuitive.
+A useful approximation to interpret logistic regression coefficients on the probability scale is the so-called divide-by-4 rule, which means that a regression coefficient of 2 corresponds to an expected increase on the probability scale of about 2/4 = 50\%.
 

### 2020-12-16 docs/manuscript.tex
add phylogeny section

@@
 
-Reasoning on the log-odds can sometimes be unintuitive.
+Reasoning about coefficients on the log-odds scale can sometimes be unintuitive.
 A useful approximation to interpret logistic regression coefficients on the probability scale is the so-called divide-by-4 rule, which means that a regression coefficient of 2 corresponds to an expected increase on the probability scale of about 2/4 = 50\%.
@@
 \subsection*{Including phylogeny}
-\begin{itemize}
-  \item sequence data shares common phylogenetic history and are therefore not i.i.d, cite Brumme paper
-  \item standard approach: multivariate normal intercept where covariance matrix is based on shared branch lengths
-  \item this approach is computationally too expensive because of current limitations in Stan.
-Would be possible by better inferencealgorithms, see this recently published paper: \sloppy\url{https://proceedings.neurips.cc/paper/2020/file/673de96b04fa3adcae1aacda704217ef-Paper.pdf} but probably omit that?
-  \item Our approach is close to the one in Carlson et al.
-  \item slight tweak: use phylogeny estimation tools for tree likelihoods p(tree | sequences), then bayes rule to invert that conditional probability.
-Optimize branch lengths for each replacement, but keep tree topology fixed
-  \item advantages: faster (raxml well optimized)
-  \item disadvantages: ad-hoc, close to conventional multivariate normal    approach with off-diagonal elements set to 0, therefore not as flexible.
-But: should be good enough, phylogeny seems to be not that important (probably omit that?)
-  \item how to include p(replacement | tree) in the model: just add additional intercept
-  \item add coefficient for that information, so model can decide how to make use of that information.
-  \item logit(p(replacement|tree)) used because it cancels out with the logistic
-    link function. therefore: phylogeny information as baseline in absence of any HLA effects
-  \item calibration plots to show that estimated phylogenetic probabilities are reasonably well calibrated.
-  \item meaning: for sequences with an expected probability of observing the replacement of x\%, we really do see that x\% of the sequences have that replacement.
-\end{itemize}
+
+One problem that occurs when analyzing sequence data is that species share a common phylogenetic history.
+Standard statistical methods usually assume samples to be independent and identically distributed, which may lead to wrong conclusions when this assumption is strongly violated.
+In the context of identifying HLA associations, Bhattacharya et al. demonstrated the importance of correcting for the phylogenetic structure.
+There are many different approaches described in the literature for phylogenetic regression for binary dependent variables (see \cite{Ives2014}), where the most popular approach is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree \cite{Ives2009}.
+
+This approach showed to be too computationally expensive to include in our model, so we chose a strategy that is similar to the one in Carlson et al.
+\cite{Carlson2008} for the Phylogenetic Dependency Network:
+
+Consider a phylogenetic tree \(\Psi\) obtained from standard maximum likelihood methods for a given multiple sequence alignment.
+We are interested in estimating \(P(y_{ij}=1|\Psi)\), that is, the probability of observing the replacement \(j\) in sequence \(i\) based on the underlying phylogenetic model.
+A quantity that can be readily computed using phylogenetic software like RAxML \cite{Stamatakis2014} is \(P(\Psi|y_{ij}=1)\).
+For this, we keep the tree topology fixed, annotate the tree with the binary observations \(y\) at its leaves and optimize the branch lengths.
+\(P(\Psi|y_{ij}=1)\) is then the likelihood of the annotated phylogenetic tree.
+We can then use Bayes' rule to invert that conditional probability by additionally computing \(P(\Psi|y_{ij}=0)\), which is done by altering the annotation for sequence \(i\) and keeping all other observations constant.
+
+In order to check if the inferred probabilities \(P(y_{ij}=1|\Psi)\) faithfully reflect the observed data we use calibration plots (see fig.~\ref{fig:phylogeny-calibration}).
+All observations are binned according to their inferred probability \(P(y_{ij}=1|\Psi)\).
+If these probabilities are calibrated correctly, we expect that in a group of observations with an inferred probability \(P(y_{ij}=1|\Psi)\) of around x\% we really do observe \(y_{ij}=1\) around x\% percent of the time.
 
@@
   \includegraphics[width=1\linewidth]{plots/phylogeny_calibration.pdf}
-  \caption{Calibration plot and how it works.}
+  \caption{Calibration plot for the HBV dataset. For a description of all datasets used in this study see section~\ref{sec:results}. Calibration plots for the other datasets are shown in the appendix.
+  All observations are first sorted by increasing estimated probability \(P(y_{ij}=1|\Psi)\) and then grouped into \(n\) bins.
+  For each bin, the fraction of observations with \(y_{ij}=1\) (observed event percentage) is compared to the midpoint of each bin (the value in the center between the lowest and highest probability).
+The error bars show the cutpoints for each bin.
+If the probabilities are calibrated correctly, each dot is supposed to scatter closely around the diagonal line.}
+  \label{fig:phylogeny-calibration}
 \end{figure}
+
+The estimated probabilities are then included in the model as additional intercepts:
+
+\begin{equation}
+\begin{aligned}
+  y_{ij} & \sim \text{bernoulli}(\theta_{ij}) \\
+  \theta_{ij} & = \text{inv\_logit}(\beta_{0_{j}} + \gamma~\text{logit}(P(y_{ij}=1|\Psi)) \\ + & X_{i}\beta_{\text{HLA}_{j}})
+\end{aligned}
+\end{equation}
+
+The logit transform is used because is cancels out with the logistic link function.
+This means that the phylogeny information acts as a baseline in absence of any HLA effects.
+Note that we also include an additional parameter \(\gamma\), that is constrained to be positive.
+This helps because the inferred probabilities \(P(y_{ij}=1|\Psi)\) might not necessarily reflect the true underlying phylogenetic signal, for example because the phylogenetic tree does not match the observed data well enough.
+
+A straight-forward extension of our model would be to also account for these sources of uncertainty, for example by using a Bayesian method to estimate a posterior distribution over possible tree topologies.
+The uncertainty over the tree topology and the underlying parameters of the phylogenetic model would then propagate into uncertainty of the estimated probabilities \(P(y_{ij}=1|\Psi)\).
+However, in order to not increase the runtime of the model further we  use standard maximum likelihood estimates and then include these in a way that allows for measurement error.
 
@@
 
-\section{Results}
+\section{Results} \label{sec:results}
 

### 2020-12-17 docs/manuscript.tex
add text on linkage disequilibrium and sparsity assumption

@@
 
-\begin{itemize}
-  \item HLA linkage disequilibrium is accounted for by considering all HLA alleles of a sample at once.
-  \item If strong correlations between HLA alleles: marginal posterior for both of them overlap with 0
-  \item Codon covariation is not accounted for because we are mainly interested in HAMs.
-  \item Model could be extended to include codon covariation by including them as additional predictors. This is possible because we use the horseshoe prior.
-\end{itemize}
+The other two important confounding effects that are recognized in Carlson et al.\cite{Carlson2008} are HLA linkage disequilibrium and codon covariation.
+HLA linkage disequilibrium describes the observation that HLA alleles that lie in proximity to each other on the chromosome are also more likely to be inherited together.
+When using a statistical method that evaluates each HLA allele individually, this may lead to spurious associations: If some HLA allele \(X\) is associated with replacement \(j\), and HLA allele \(Y\) often co-occurs with HLA allele \(X\), this means that we also identify an association between HLA allele \(Y\) and replacement \(j\).
 
-\subsection*{Including sparsity assumptions}
+This association is correct in a statistical sense, but if the purpose of the model is to identify sites under HLA-mediated selection pressure, these associations due to linkage disequilibrium are most likely unwanted.
 
-\begin{itemize}
-  \item one of the recent advances in Bayesian inference: development of spacial priors to give model useful properties
-  \item we expect most HLA coefficients to be 0, because no epitope spanning that region is presented on the cell surface, therefore no selection pressure and no association with possible replacements
-  \item this is prior knowledge that should be incorporated in the model
-  \item leads to better inferences because uncertainty of the regression coefficients does not propagate into observations.
-  \item works by placing most probability mass very close to 0, but with large tails for non-zero coefficients.
-  \item show formula
-  \item global shrinkage parameter is very small and shrinks most coefficients to 0, local shrinkage parameter is very large and allows some coefficients to escape that shrinkage
-  \item prior can be specified in terms of expected number of non-zero coefficients
-  \item in our case, degree of sparsity can be well estimated from the data because all replacements share the same global shrinkage parameter.
-  \item useful addition for logistic regression models: regularized horseshoe to have some regularization for non-zero coefficients. this helps to deal with issues of separability and colinearity.
-  \item effect of horseshoe prior compared to standard logistic regression model: refer to model testing section.
-In our benchmark, the horseshoe prior alone is more effective than the model with phylogeny.
-  \item maybe include comparison of marginal posteriors with and without horseshoe prior
-\end{itemize}
+Note that these spurious associations only occur when testing each HLA allele individually.
+In a regression model where all HLA alleles of a sample are evaluated at once, the marginal poster distributions for correlated predictors become broader to accurately reflect that the model is unable to discern which of these predictors explain the observed association with a replacement.
+
+We do not currently account for codon covariation in our model, as we are interested in identifying all HLA associated mutations (which include for example compensatory mutations).
+The general structure of the model, in particular the sparsity assumption introduced in section~\ref{sec:sparsity},would allow adding codon covariation by adding all other codons as additional predictors.
+This would generally not be possible in a frequentist setting because the predictor matrix would become singular and the inverse does not exist.
+
+\subsection*{Including sparsity assumptions} \label{sec:sparsity}
+
+One of the recent advances in Bayesian inference is the development of special prior distributions which can be used to give models additional useful properties.
+An example of these special prior distributions are the so-called sparsity-inducing priors, which can be used to convey the assumption that most of the predictors in a regression model are not associated with the outcome and are therefore 0.
+
+In the context of identifying HLA associations we expect most of the HLA coefficients to be very close to 0, as a given alignment position is typically only restricted by none or few HLA alleles.
+This is important prior knowledge that should be incorporated in the model.
+Adding this sparsity assumption can lead to better inferences, because the uncertainty of the non-relevant HLA regression coefficients does not propagate further and therefore the model is able to better discern signal from noise.
+
+There are a variety of different sparsity-promoting priors with slightly different properties.
+They share the common structure of placing most probability mass very close to 0, with large to tails to accommodate the non-zero coefficients.
+For our model, we use a so-called horse prior~\cite{Carvalho2010}, which is defined as a scale-mixture of gaussians:
+
+\begin{equation}
+  \begin{aligned}
+    \beta_{j} &\sim \text{Normal}(0, \tau\lambda_{j}) \\
+    \lambda_{j} &\sim \text{Cauchy}^{+}(0, 1) \\
+    \tau &\sim \text{Cauchy}^{+}(0, \tau_{0})
+  \end{aligned}
+\end{equation},
+
+where \(\beta_{j}\) are the regression coefficients, \(\tau\) is the so-called global shrinkage parameter, \(lambda_{j}\) are the so-called local shrinkage parameters and \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution.
+
+Intuitively speaking, the global shrinkage parameter \(\tau\) is typically very small and shrinks most of the regression coefficients to 0, whereas the local shrinkage parameters can occasionally be very large to allow some coefficients to escape that shrinkage.
+
+A useful addition to the horseshoe prior is given in Piironen et al.\cite{Piironen2017}, which describes a way to choose the parameter governing the overall degree of sparsity \(\tau_{0}\) based on the expected number of non-zero coefficients.
+
+For our HLA model, we utilize an extension to the horseshoe prior called the regularized horseshoe prior~\cite{Piironen2017}, which also provides some regularization to the non-zero coefficients.
+This is particularly useful for logistic regression models, as some shrinkage helps to deal with issues of separability and collinearity.
+
+Figure~\ref{fig:horseshoe-comparison} shows a comparison of marginal posteriors for regression coefficients with and without horseshoe prior.
+
+The improvement of the logistic regression model with the horseshoe prior compared to a standard logistic regression model is considerable (see section~\ref{sec:results}.
+In our benchmark, the horseshoe prior alone is shown to be more already effective than a model without horseshoe prior but with phylogeny included.
+
+\begin{figure}
+  \label{fig:horseshoe-comparison}
+\end{figure}
 

### 2020-12-17 docs/manuscript.tex
add text for epitope information

@@
 
-where \(\beta_{j}\) are the regression coefficients, \(\tau\) is the so-called global shrinkage parameter, \(lambda_{j}\) are the so-called local shrinkage parameters and \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution.
+where \(\beta_{j}\) are the regression coefficients, \(\tau\) is the so-called global shrinkage parameter, \(\lambda_{j}\) are the so-called local shrinkage parameters and \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution.
 
@@
 
-For our HLA model, we utilize an extension to the horseshoe prior called the regularized horseshoe prior~\cite{Piironen2017}, which also provides some regularization to the non-zero coefficients.
+For our HLA model, we utilize an extension to the horseshoe prior called the regularized horseshoe prior~\cite{Piironen2017}, which also provides some regularization for the non-zero coefficients.
 This is particularly useful for logistic regression models, as some shrinkage helps to deal with issues of separability and collinearity.
@@
 
-The improvement of the logistic regression model with the horseshoe prior compared to a standard logistic regression model is considerable (see section~\ref{sec:results}.
-In our benchmark, the horseshoe prior alone is shown to be more already effective than a model without horseshoe prior but with phylogeny included.
+The improvement of the logistic regression model with the horseshoe prior is a considerable improvement over a standard logistic regression model (see section~\ref{sec:results}.
+In our benchmark, the horseshoe prior alone is even shown to be more effective than a model without horseshoe prior but with phylogeny included.
 
@@
 
-\begin{itemize}
-  \item Vast experimental data available.
-  \item HAMs are much more likely to lie in known epitopes
-  \item epitope prediction software works reasonably well and can be included into as
-    input to a probabilitic model
-  \item MHCFlurry is used for epitope prediction and Antigen processing prediction: Input matrix (dimensions: \#replacements x \#HLA alleles) contains a 1 if that position is either expected to be inside a predicted epitope or related to antigen processing, 0 otherwise
-  \item uses some thresholds, but this is not so bad because we just use it as input for a probabilistic model.
-Maybe play around with different options?
-The Brumme paper uses some offsets for defining the region of known epitopes, maybe this does make a difference.
-  \item including this kind of information can drastically improve predictive performance, as it is relevant external data.
-  \item Bayesian inference allows us to include informationen from different sources in a consistent manner.
-  \item this information may be imprecise or wrong, but the model is able to learn how much to trust this data if parameterized correctly.
-  \item epitope prediction / antigen processing escape is information about the expected degree of sparsity, i.e. some coefficients are more likely to be non-zero than others.
-  \item can be included into the model by varying the local shrinkage parameter term.
-  \item larger standard deviation of the local shrinkage parameters means that it is more likely that this coefficient is non-zero.
-  \item show formula
-\end{itemize}
+There is vast experimental data on epitope binding motifs available, e.g.
+through the use of mass spectrometry~\cite{Hunt1992}.
+Epitope prediction software like MHCFlurry~\cite{ODonnell2020} use these data to extrapolate MHC I binding for new, untested peptides.
+As HLA-associated mutations are much more likely to lie epitopes (e.g.
+because of selection pressure to escape MHC I binding), using epitope prediction software can provide valuable external data for identifying HLA-associated mutations.
+
+For our model, we use MHCFlurry 2.0 for binding prediction, as it does not only predict HLA binding but also antigen processing.
+For this, we create an input matrix of dimensions \(R\times D\), where R is the number of evaluated replacements and D is the number of observed HLA alleles in the dataset.
+The elements of this matrix are binary encoded and contain a 1 if that position is either expected to be inside a predicted epitope or related to antigen processing, and 0 otherwise.
+Given an amino acid sequence, MHCFlurry 2.0 provides a list of possible epitopes (9-13 mers) and HLA alleles pairs and calculates a rank based on comparisons with random epitope and HLA pairs.
+For the binarization we use a rank threshold of 0.2\% (as suggested by MHCFlurry).
+
+We typically try to avoid thresholds whenever possible, but in this case we deemed it to be preferable to including the predicted binding affinities, as we expected the binary values to be more reliable than the predicted numeric values.
+Note that the output of these epitope prediction tools may be imprecise or even wrong, but by including it as an input for a probabilistic model, the model is able to learn how much to trust this data if parameterized in a way to allow for measurement error.
+
+Epitope prediction is information about the expected degree of sparsity, i.e.
+when knowing a certain position is restricted by a certain HLA allele, we expect that this HLA allele is more likely to be associated with the replacement than the other HLA alleles.
+This can be reflected by increasing the scale of the local shrinkage parameters \(\lambda_{j}\):
+
+\begin{equation}
+  \begin{aligned}
+    \lambda_{j} &\sim \text{Cauchy}^{+}(0, \text{exp}(\beta_{\text{epi}}Z_{j})) \\
+    \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2)
+  \end{aligned}
+\end{equation},
+
+where \(Z_{j}\) is a binary vector denoting which of the alleles is predicted to restrict HLA binding or antigen processing at the alignment position corresponding to replacement \(j\).
+The parameter \(\beta_{\text{epi}}\) governs the increase in scale of the local corresponding local shrinkage parameters.
+The larger \(\beta_{\text{epi}}\) is, the more likely it is that we expect to see non-zero regression coefficients for the HLA alleles that are predicted to restrict the corresponding alignment position of replacement \(j\).
 

### 2020-12-19 docs/manuscript.tex
add text on rest of implementation

@@
 \begin{align}
-  y_{ij} & \sim \text{bernoulli}(\theta_{ij}) \\
-  \theta_{ij} & = \text{inv\_logit}(\beta_{0_{j}} + X_{i}\beta_{\text{HLA}_{j}})
+  y_{ik} & \sim \text{bernoulli}(\theta_{ik}) \\
+  \theta_{ik} & = \text{inv\_logit}(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}})
 \end{align},
 
-where \(y_{ij}\) is the binary encoded observation at sequence \(i\) for replacement \(j\), \(\theta_{ij}\) is the estimated probability that we observe replacement \(j\) in sequence \(i\), \(X_{i}\) is the binary encoded (row) vector of HLA annotations for sequence \(i\), and \(\beta_{\text{HLA}_{j}}\) is the (column) vector of HLA regression coefficients for replacement \(j\) and  \(\text{inv\_logit}\) is the logistic link function.
-Note that, if we observe N different amino acid states at a certain alignment position, each of those are considered in turn to be the "replacement".
+where \(y_{ik}\) is the binary encoded observation at sequence \(i\) for replacement \(k\), \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\), \(X_{ij}\) is 1 if sequence \(i\) is annotated with HLA allele \(j\) and 0 otherwise, \(\beta_{\text{HLA}_{jk}}\) is the HLA regression coefficients for HLA allele \(j\) at replacement \(k\), \(D\) is the number of HLA alleles in the dataset, and \(\text{inv\_logit}\) is the logistic link function.
+Note that, if we observe N different amino acid states at a certain alignment position, each of those are considered in turn to be the “replacement”.
 
-The regression coefficients \(\beta_{\text{HLA}_{j}}\) are the main focus of interest, as they quantify the association between the occurrence of replacement \(j\) and each of the observed HLA alleles.
-They are in units of log-odds, so a regression coefficient of somewhere around 2 means that when comparing sequences with and without that particular HLA allele, the log-odds of observing replacement \(j\) in presence of that HLA allele is increased by 2.
+The regression coefficients \(\beta_{\text{HLA}_{jk}}\) are the main focus of interest, as they quantify the association between the occurrence of replacement \(k\) and each of the observed HLA alleles.
+They are in units of log-odds, so a regression coefficient of somewhere around 2 means that when comparing sequences with and without that particular HLA allele, the log-odds of observing replacement \(k\) in presence of that HLA allele is increased by 2.
 
@@
 Standard statistical methods usually assume samples to be independent and identically distributed, which may lead to wrong conclusions when this assumption is strongly violated.
-In the context of identifying HLA associations, Bhattacharya et al. demonstrated the importance of correcting for the phylogenetic structure.
-There are many different approaches described in the literature for phylogenetic regression for binary dependent variables (see \cite{Ives2014}), where the most popular approach is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree \cite{Ives2009}.
+In the context of identifying HLA associations, Bhattacharya et al.\cite{Bhattacharya2007} demonstrated the importance of correcting for the phylogenetic structure.
+There are many approaches described in the literature for phylogenetic regression for binary dependent variables (see \cite{Ives2014}), where the most popular approach is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree \cite{Ives2009}.
 
@@
 Consider a phylogenetic tree \(\Psi\) obtained from standard maximum likelihood methods for a given multiple sequence alignment.
-We are interested in estimating \(P(y_{ij}=1|\Psi)\), that is, the probability of observing the replacement \(j\) in sequence \(i\) based on the underlying phylogenetic model.
-A quantity that can be readily computed using phylogenetic software like RAxML \cite{Stamatakis2014} is \(P(\Psi|y_{ij}=1)\).
+We are interested in estimating \(P(y_{ik}=1|\Psi)\), that is, the probability of observing the replacement \(k\) in sequence \(i\) based on the underlying phylogenetic model.
+A quantity that can be readily computed using phylogenetic software like RAxML \cite{Stamatakis2014} is \(P(\Psi|y_{ik}=1)\).
 For this, we keep the tree topology fixed, annotate the tree with the binary observations \(y\) at its leaves and optimize the branch lengths.
-\(P(\Psi|y_{ij}=1)\) is then the likelihood of the annotated phylogenetic tree.
-We can then use Bayes' rule to invert that conditional probability by additionally computing \(P(\Psi|y_{ij}=0)\), which is done by altering the annotation for sequence \(i\) and keeping all other observations constant.
+\(P(\Psi|y_{ik}=1)\) is then the likelihood of the annotated phylogenetic tree.
+We can then use Bayes' rule to invert that conditional probability by additionally computing \(P(\Psi|y_{ik}=0)\), which is done by altering the annotation for sequence \(i\) and keeping all other observations constant.
 
-In order to check if the inferred probabilities \(P(y_{ij}=1|\Psi)\) faithfully reflect the observed data we use calibration plots (see fig.~\ref{fig:phylogeny-calibration}).
-All observations are binned according to their inferred probability \(P(y_{ij}=1|\Psi)\).
-If these probabilities are calibrated correctly, we expect that in a group of observations with an inferred probability \(P(y_{ij}=1|\Psi)\) of around x\% we really do observe \(y_{ij}=1\) around x\% percent of the time.
+In order to check if the inferred probabilities \(P(y_{ik}=1|\Psi)\) faithfully reflect the observed data we use calibration plots (see fig.~\ref{fig:phylogeny-calibration}).
+All observations are binned according to their inferred probability \(P(y_{ik}=1|\Psi)\).
+If these probabilities are calibrated correctly, we expect that in a group of observations with an inferred probability \(P(y_{ik}=1|\Psi)\) of around x\% we really do observe \(y_{ik}=1\) around x\% percent of the time.
 
@@
   \caption{Calibration plot for the HBV dataset. For a description of all datasets used in this study see section~\ref{sec:results}. Calibration plots for the other datasets are shown in the appendix.
-  All observations are first sorted by increasing estimated probability \(P(y_{ij}=1|\Psi)\) and then grouped into \(n\) bins.
-  For each bin, the fraction of observations with \(y_{ij}=1\) (observed event percentage) is compared to the midpoint of each bin (the value in the center between the lowest and highest probability).
+  All observations are first sorted by increasing estimated probability \(P(y_{ik}=1|\Psi)\) and then grouped into \(n\) bins.
+  For each bin, the fraction of observations with \(y_{ik}=1\) (observed event percentage) is compared to the midpoint of each bin (the value in the center between the lowest and highest probability).
 The error bars show the cutpoints for each bin.
@@
 \begin{aligned}
-  y_{ij} & \sim \text{bernoulli}(\theta_{ij}) \\
-  \theta_{ij} & = \text{inv\_logit}(\beta_{0_{j}} + \gamma~\text{logit}(P(y_{ij}=1|\Psi)) \\ + & X_{i}\beta_{\text{HLA}_{j}})
+  y_{ik} & \sim \text{bernoulli}(\theta_{ik}) \\
+  \theta_{ik} & = \text{inv\_logit}(\beta_{0_{k}} + \gamma~\text{logit}(P(y_{ik}=1|\Psi)) \\ + \sum_{k=1}^{D}X_{ik}\beta_{\text{HLA}_{jk}})
 \end{aligned}
@@
 Note that we also include an additional parameter \(\gamma\), that is constrained to be positive.
-This helps because the inferred probabilities \(P(y_{ij}=1|\Psi)\) might not necessarily reflect the true underlying phylogenetic signal, for example because the phylogenetic tree does not match the observed data well enough.
+This helps because the inferred probabilities \(P(y_{ik}=1|\Psi)\) might not necessarily reflect the true underlying phylogenetic signal, for example because the phylogenetic tree does not match the observed data well enough.
 
 A straight-forward extension of our model would be to also account for these sources of uncertainty, for example by using a Bayesian method to estimate a posterior distribution over possible tree topologies.
-The uncertainty over the tree topology and the underlying parameters of the phylogenetic model would then propagate into uncertainty of the estimated probabilities \(P(y_{ij}=1|\Psi)\).
+The uncertainty over the tree topology and the underlying parameters of the phylogenetic model would then propagate into uncertainty of the estimated probabilities \(P(y_{ik}=1|\Psi)\).
 However, in order to not increase the runtime of the model further we  use standard maximum likelihood estimates and then include these in a way that allows for measurement error.
@@
 HLA linkage disequilibrium describes the observation that HLA alleles that lie in proximity to each other on the chromosome are also more likely to be inherited together.
-When using a statistical method that evaluates each HLA allele individually, this may lead to spurious associations: If some HLA allele \(X\) is associated with replacement \(j\), and HLA allele \(Y\) often co-occurs with HLA allele \(X\), this means that we also identify an association between HLA allele \(Y\) and replacement \(j\).
+When using a statistical method that evaluates each HLA allele individually, this may lead to spurious associations: If some HLA allele \(X\) is associated with replacement \(k\), and HLA allele \(Y\) often co-occurs with HLA allele \(X\), this means that we also identify an association between HLA allele \(Y\) and replacement \(k\).
 
@@
   \begin{aligned}
-    \beta_{j} &\sim \text{Normal}(0, \tau\lambda_{j}) \\
-    \lambda_{j} &\sim \text{Cauchy}^{+}(0, 1) \\
+    \beta_{jk} &\sim \text{Normal}(0, \tau^{2}\lambda^{2}_{jk}) \\
+    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, 1) \\
     \tau &\sim \text{Cauchy}^{+}(0, \tau_{0})
@@
 
-where \(\beta_{j}\) are the regression coefficients, \(\tau\) is the so-called global shrinkage parameter, \(\lambda_{j}\) are the so-called local shrinkage parameters and \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution.
+where \(\beta_{jk}\) are the regression coefficients, \(\tau\) is the so-called global shrinkage parameter, \(\lambda_{jk}\) are the so-called local shrinkage parameters and \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution.
 
@@
 Epitope prediction software like MHCFlurry~\cite{ODonnell2020} use these data to extrapolate MHC I binding for new, untested peptides.
-As HLA-associated mutations are much more likely to lie epitopes (e.g.
+As HLA-associated mutations are much more likely to lie in epitopes (e.g.
 because of selection pressure to escape MHC I binding), using epitope prediction software can provide valuable external data for identifying HLA-associated mutations.
 
-For our model, we use MHCFlurry 2.0 for binding prediction, as it does not only predict HLA binding but also antigen processing.
+For our model, we use MHCFlurry 2.0 for binding prediction, as it does not only predict HLA binding but also predicts antigen processing.
 For this, we create an input matrix of dimensions \(R\times D\), where R is the number of evaluated replacements and D is the number of observed HLA alleles in the dataset.
 The elements of this matrix are binary encoded and contain a 1 if that position is either expected to be inside a predicted epitope or related to antigen processing, and 0 otherwise.
-Given an amino acid sequence, MHCFlurry 2.0 provides a list of possible epitopes (9-13 mers) and HLA alleles pairs and calculates a rank based on comparisons with random epitope and HLA pairs.
+Given an amino acid sequence, MHCFlurry 2.0 provides a list of possible epitopes (9-13 mers) and HLA allele pairs and calculates a rank based on comparisons with random epitope and HLA pairs.
 For the binarization we use a rank threshold of 0.2\% (as suggested by MHCFlurry).
@@
 
-Epitope prediction is information about the expected degree of sparsity, i.e.
+We use epitope prediction as information about the expected degree of sparsity, i.e.
 when knowing a certain position is restricted by a certain HLA allele, we expect that this HLA allele is more likely to be associated with the replacement than the other HLA alleles.
-This can be reflected by increasing the scale of the local shrinkage parameters \(\lambda_{j}\):
+This can be reflected by increasing the scale of the local shrinkage parameters \(\lambda_{jk}\):
 
@@
   \begin{aligned}
-    \lambda_{j} &\sim \text{Cauchy}^{+}(0, \text{exp}(\beta_{\text{epi}}Z_{j})) \\
+    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \text{exp}(Z_{jk}\beta_{\text{epi}})) \\
     \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2)
@@
 
-where \(Z_{j}\) is a binary vector denoting which of the alleles is predicted to restrict HLA binding or antigen processing at the alignment position corresponding to replacement \(j\).
-The parameter \(\beta_{\text{epi}}\) governs the increase in scale of the local corresponding local shrinkage parameters.
-The larger \(\beta_{\text{epi}}\) is, the more likely it is that we expect to see non-zero regression coefficients for the HLA alleles that are predicted to restrict the corresponding alignment position of replacement \(j\).
-
-\subsection*{Partial pooling of 4-digit HLA alleles}
-
-\begin{itemize}
-  \item (this still has to be implemented, but should be a nice feature. If this is going to be too time-consuming I can just leave it out)
-  \item realized through allele group specific intercepts, e.g. HLA-B\*51:01 and  HLA-B\*51:03, ..., share a common intercept.
-  \item show formula
-\end{itemize}
+where \(Z_{jk}\) is 1 if HLA allele \(j\) is predicted to restrict the alignment position corresponding to replacement \(k\) or is associated antigen processing and 0 otherwise.
+The parameter \(\beta_{\text{epi}}\) governs the increase in scale of the  corresponding local shrinkage parameters.
+The larger the estimated values of \(\beta_{\text{epi}}\) are, the more likely it is to see non-zero regression coefficients for these HLA alleles.
 
 \subsection*{Full model specification}
-\begin{itemize}
-  \item show all formulas
-  \item all models were implemented in Stan. Code is available online in two versions: one optimized for readability, one optimized for speed with multithreading and GPU support
-  \item data can be analyzed using a custom Julia package (tested on linux).
-\end{itemize}
+
+The full specification of the model is given by:
+
+\begin{equation}
+  \label{eq:full}
+  \begin{aligned}
+    y_{ik} &\sim \text{bernoulli}(\theta_{ik})\\
+    \theta_{ik} &= \text{inv\_logit}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi))\\
+        +\sum_{k=1}^{D}X_{ij}\beta_{\text{HLA}_{jk}})\\
+    \beta_{0_{k}} &\sim \text{Normal}(0, 100^2)\\
+    \gamma_{k} &\sim \text{Normal}(\mu_{\text{phy}}, \sigma_{\text{phy}}^2)\\
+    \mu_{\text{phy}} &\sim \text{Normal}(1, 1)\\
+    \sigma_{\text{phy}} &\sim \text{Normal}^{+}(0, 0.5)\\
+    \beta_{\text{HLA}_{jk}} &\sim \text{Normal}(0, \tau^{2}\tilde{\lambda}^{2}_{jk})\\
+    \tilde{\lambda}^{2}_{jk} &= \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau^{2}\lambda_{jk}^{2}} \\
+    c_{k}^2 &\sim \text{Inv-Gamma}(3.5, 3.5)\\
+    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}^2\text{exp}(Z_{jk}\beta_{\text{epi}}))\\
+    \tau &\sim \text{Cauchy}^{+}(0, \tau_{0})\\
+    \tau_{0} &= \frac{2}{D-2}\frac{2}{\sqrt{N}}
+  \end{aligned}
+\end{equation}
+
+The full model specification includes some aspects that were not covered in the previous sections.
+In particular, the parameters \(\gamma_{k}\) are modelled hierarchically, which allows information to be partially pooled across replacements.
+However, the model would also work reasonably well with just a single global parameter \(\gamma\).
+All other additions (\(\tilde{\lambda}_{jk}^{2}\) and \(\tau_0\)) are explained in detail in \cite{Piironen2017}.
+Briefly, the addition of  \(\tilde{\lambda}_{jk}^{2}\) allows some regularization for the non-zero coefficients and the parameterization of \(tau_{0}\) allows to place a prior on the expected number of non-zero coefficients.
+
+All models were implemented in Stan.
+The Stan code is available online in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
+A Julia\cite{Bezanson2017} package is available to run the model on custom data.
+Due to restrictions of dependencies (MHCFlurry and raxml-ng), HAMdetector is only available on Linux.
+We additionally provide Docker images to easily run the model on any operating system in a containerized environment, assuming Docker is installed.
 
 \subsection*{Prior justification}
-\begin{itemize}
-  \item justify priors, mostly based on the expected scales. Compare to some 
-  Betancourt/Gelman papers to see how it is done there.
-\end{itemize}
+
+Prior distributions are labeled with an asterisk in equation~\ref{eq:full}.
+They are meant to be weakly informative, which means that the goal is to provide some modest amount of shrinkage while still allowing for large absolute values given enough support by the observed data.
+One exception to this are the intercepts \(\beta_{0_k}\), which are essentially flat because they are well identified by the data alone.
+
+The hierarchical mean and standard deviation of the phylogeny coefficients \(\gamma_{k}\) are chosen to place most probability mass at values of \(\gamma_{k}\) of around 1.
+In the absence of any HLA effects, a parameter value of 1 would mean that the estimate for the probability of observing replacement \(k\) is identical to the probability based on the phylogenetic model.
+Intuitively speaking, We therefore treat the phylogeny as a baseline and any observations that cannot be attributed to phylogeny are attributed to the HLA alleles or noise.
+
+The prior on \(c_{k}^{2}\) implies a Student-t prior with 7 degrees of freedom and a scale of 1 on HLA regression coefficients \(\beta_{\text{HLA}_{jk}}\) that are estimated to be away from 0.
+A Student-t prior with these parameters is a reasonable default choice for logistic regression models.
+The theoretical justification behind this regularized variant of the horseshoe prior is explained in detail in\cite{Piironen2017}.
+
+The value of \(\tau_{0}\) is chosen to imply 2 effective non-zero HLA regression coefficients per replacement.
+The rationale behind this parameterization is again outlined in\cite{Piironen2017}.
+The value of 2 is approximately based on available HIV epitope maps.
+Please note that this is only an approximate statement about the general magnitude, i.e.
+we would expect maybe a couple of HLA alleles to be associated with a single replacement, but not 20 or more.
+The model is also parameterized in a way that assumes an equal degree of sparsity across all alignment positions a priori.
+We also tried to model \(\tau\) hierarchically, but observed sampling issues due to the resulting unfavorable geometry of the posterior.
 

### 2020-12-21 docs/manuscript.tex
add epitope prediction section

@@
 We are interested in estimating \(P(y_{ik}=1|\Psi)\), that is, the probability of observing the replacement \(k\) in sequence \(i\) based on the underlying phylogenetic model.
-A quantity that can be readily computed using phylogenetic software like RAxML \cite{Stamatakis2014} is \(P(\Psi|y_{ik}=1)\).
+A quantity that can be readily computed using phylogenetic software like RAxML-NG~\cite{Kozlov2019} is \(P(\Psi|y_{ik}=1)\).
 For this, we keep the tree topology fixed, annotate the tree with the binary observations \(y\) at its leaves and optimize the branch lengths.
@@
     y_{ik} &\sim \text{bernoulli}(\theta_{ik})\\
-    \theta_{ik} &= \text{inv\_logit}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi))\\
-        +\sum_{k=1}^{D}X_{ij}\beta_{\text{HLA}_{jk}})\\
+      \theta_{ik} &=
+        \text{inv\_logit}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi))\\
+        \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{\text{HLA}_{jk}}) \\
     \beta_{0_{k}} &\sim \text{Normal}(0, 100^2)\\
@@
     \tau &\sim \text{Cauchy}^{+}(0, \tau_{0})\\
-    \tau_{0} &= \frac{2}{D-2}\frac{2}{\sqrt{N}}
+    \tau_{0} &= \frac{2}{D-2}\frac{2}{\sqrt{N}}\\
   \end{aligned}
@@
 
-\section{Results} \label{sec:results}
+\section{Results}
+
+To show that our model works well in different real-world scenarios, we run versions of our model on several datasets:
 
 \begin{itemize}
-  \item to show that model works well in several real-world scenarios, versions of the model are run on different datasets: large Brumme HIV, smaller Arevir HIV, larger Düsseldorf HBV, HDV by Michael Roggendorf, Rongge small HIV
-  \item MCMC convergence diagnostics reveal no sampling issues
-  \item general posterior predictive checks are applied to show that model explains data well
-  \item two forms of model checking: external testing through comparisons of known epitopes and leave-one-out cross-validation.
+  \item A large HIV dataset consisting of X sequences, that were kindly provided by X and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}.
+  \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\cite{Roomp2006}
+  \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm.
+  \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf
+  \item A set of 41 HIV sequences that were kindly provided by Rongge Yang
 \end{itemize}
 
-\subsubsection*{Convergence diagnostics}
+For all sequences, we applied the following data preparation steps:
 
-\begin{itemize}
-  \item inference was done with MCMC, using Stan.
-  \item One step in Bayesian workflow: checking inference with convergence diagnostics.
-  \item Stan code is available online.
-  \item All model fits show no signs of inference issues, all Rhats less than 1.01,
-    effective sample size for all model parameters greater than 300.
-\end{itemize}
+\begin{enumerate}
+  \item For each dataset, the sequences were split into subsequences, either by protein or gene.
+  \item If not already present in this format, sequences were translated into their amino acid representation.
+  \item RAxML-NG~\cite{Kozlov2019} version 1.0.0 was used to generate a maximum likelihood phylogenetic tree for each gene/protein using the \texttt{--model GTR+G+I} option with all other parameters set to default values.
+\end{enumerate}
 
-\subsubsection*{Posterior predictive checks}
+In order to evaluate each building block of HAMdetector, we applied 4 different Bayesian models to each dataset, each of which includes the previous model as a special case:
+A standard logistic regression model, a logistic regression model with horseshoe prior, a model additionally including phylogeny and the full model with horseshoe prior, phylogeny and epitope prediction.
+As a comparison to existing methods, we also applied Fisher's exact test and the Phylogenetic Dependency Network.
 
-\begin{itemize}
-  \item Bayesian modeling allows a simple yet effective strategy to test models: By simulating data under the model's assumptions and comparing simulated and observed data, model misspecifications can be identified and show possible model improvements.
-  \item One way to perform posterior predictive checks for models with binary outcomes: calibration plots.
-  \item cite Gelman dog shock PPC paper
-  \item show calibration plot
-  \item calibration plots are similar to the one showed in phylogeny, but this time test the predictions of the whole model, and not just the phylogeny component
-  \item show example plot, PPCs for all the models are shown in the supplementary
-\end{itemize}
+\subsection*{Convergence diagnostics}
 
-\subsection*{Comparison to list of known epitopes}
+For all the Bayesian models, Markov Chain Monte Carlo sampling was done using Stan.
+All model fits showed no signs of inference issues.
+In total, 4 different chains with 2000 samples each were drawn from the posterior.
+The effective sample size for all model parameters exceeded 500, R-hat convergence diagnostic values were below 1.01 in all cases.
 
-\begin{itemize}
-  \item When testing models, we prefer to evaluate them on real-world data whenever possible
-  \item comparing against real-world data is not straight forward for testing HAMs, because: 1.) if a tool identifies a HLA-associated mutation which is not yet documented, this could be either because it is a false positive, or a yet unknown HAM that has not been documented before.
-2.) Viewing every not identified HAM that is listed in the literature as a false negative is not correct either, because what is documented are escape mutations, which do not necessarily have to show in a sequence alignment.
-3.) What is and is not a HAM is not a binary decision, so the true positive / false negative framework does not work as well.
-  \item To circumvent these issues and still compare against real-work data, we chose the following strategy:
-  \item for each model, all evaluated replacements are ranked by decreasing confidence of being an HLA associated mutations.
-  \item Then, a list of known epitopes is used to create a plot of the cumulative number of replacements in known epitopes vs the corresponding rank.
-The assumption for this kind of model check is that we expect to see an
-  enrichment of replacements that are in known epitopes and that this enrichment is stronger for better performing models.
-  \item We compare several different models on all data sets available. 
-  \item Observations:
-  \item The logistic regression model without any additional information performs about as well as Fisher's exact test.
-  \item The horseshoe prior alone is a drastic improvement over Fisher's exact test and the logistic regression model with non-sparsifying priors, even though it does not include any external information.
-The benefit comes from providing the model the information that most coefficients are expected to be very close to 0.
-  \item Models improve as we add additional information
-  \item Adding epitope prediction does improve predictive performance considerably.
-Note that the model only uses epitope prediction software and does not utilize any information of experimentally confirmed epitopes, which we only use for model evaluation.
-Therefore, our method even works in absence of any additional epitope information.
-  \item add plot
-\end{itemize}
+\subsection*{Posterior predictive checks}
+
+Bayesian modelling allows a simple yet effective strategy to test models: By simulating data under the model's assumptions and then comparing simulated and observed data, model misspecifications can be identified and highlight possible model improvements.
+
+Here, we compare the calibration of the posterior predictive probability of observing the replacement to the actual observed data using calibration plots.
+The plots are similar to the one showed in figure~\ref{fig:phylogeny-calibration} but this time do not evaluate the probabilities based on the phylogenetic model alone, but show the calibration of the posterior predictive probabilities based on the full model.
+
+An example calibration plot of the AMdetector model on the Core protein sequence data of the HBV dataset is shown in figure~\ref{fig:calibration}.
+Plots for the other models and datasets are shown in the supplementary.
+
+\begin{figure}[!ht]
+  \includegraphics[width=1\linewidth]{plots/phylogeny_calibration.pdf}
+  \caption{Calibration plot for the HBV dataset. Calibration plots for the other datasets are shown in the appendix.
+  All observations are first sorted by increasing estimated probability \(P(y_{ik}=1)\) and then grouped into \(n\) bins.
+  For each bin, the fraction of observations with \(y_{ik}=1\) (observed event percentage) is compared to the midpoint of each bin (the value in the center between the lowest and highest probability).
+The error bars show the cutpoints for each bin.
+If the probabilities are calibrated correctly, each dot is supposed to scatter closely around the diagonal line.}
+  \label{fig:calibration}
+\end{figure}
+
+\subsection*{Comparison to a list of known epitopes}
+
+When testing models, we prefer to evaluate them on real-world data whenever possible. 
+However, comparing against real-world data is not straight forward when identifying HLA-associated mutations, because if a tool identifies an HLA-associated mutation which is not yet documented, this could be either because it is a false positive, or a yet unknown HLA-associated mutation that has not been documented before.
+Also, viewing every non-identified HLA-associated mutation that is listed in the literature as a false negative is not correct either, because documented escape mutations do not necessarily have to show in a sequence alignment.
+A third issue when comparing models is that HLA-associated mutations are not necessarily binary, so the true positive / false negative framework does not work as well.
+
+To circumvent these issues and still compare against real-world data, we chose the following strategy:
+For each model, all evaluated replacements are ranked by decreasing confidence of being an HLA associated mutation.
+That is, for Bayesian models we calculate the probability of the corresponding regression coefficient exceeding zero \(P(\beta_{HLA} > 0)\), for Fisher's exact test we sort HLA allele - replacement pairs by increasing p-values and for the Phylogenetic Dependency Network, we rank the associations by increasing q-values.
+Then, a list of known epitopes is used to create a plot of the cumulative number of replacements inside the boundary of a known epitope vs the corresponding rank.
+The underlying assumption for this kind of model check is that we expect to see an enrichment of replacements that lie inside the region of known epitopes and that this enrichment is stronger for better performing models.
+
+\begin{figure}[ht!]
+  \includegraphics[width=1\linewidth]{plots/comparison.pdf}
+  \caption{Number of associations inside the boundary of known epitopes vs.
+rank.
+For each model, the pairs of HLA alleles and replacements are sorted by decreasing confidence of being an HLA associated mutation.
+Then, each replacement is compared to a list of known epitopes and a line of the cumulative number of associations inside known epitopes vs.
+the current rank is plotted for each model.
+Better performing models are expected to have an enrichment of associations that are inside of known epitopes.
+  PhyloD: Phylogenetic Dependency Network, Model 1: simple logistic regression model with broad Student-t priors, Model 2: logistic regression model with horseshoe prior, Model 3: logistic regression model with horseshoe prior and phylogeny, Model 4: Full model with epitope prediction.
+The thin gray lines denote random permutations of the list of HLA allele - replacement pairs and act as a baseline.}
+  \label{fig:comparison}
+\end{figure}
+
+Figure~\ref{fig:comparison} shows a comparison of different models on the HBe protein of the HBV dataset.
+Plots for all other datasets are shown in the Appendix, but show very similar patterns to the one shown here.
+
+Unsurprisingly, all models show a strong enrichment of HLA-associated mutations compared to a randomly permutated list of HLA allele - replacement pairs (gray lines).
+Fisher's exact test performs about as well as the simple logistic regression model, as from a statistical perspective, they are both very similar.
+The horseshoe prior alone is a drastic improvement over Fisher's exact test and the logistic regression model with non-sparsifying priors, even though it does not include any external information.
+The benefit comes from including the assumption that most regression coefficients are expected to be very close to 0.
+The logistic regression model with horseshoe prior works roughly as well as the Phylogenetic Dependency Network, which includes much more information.
+Note that the line for the Phylogenetic Dependency Network is shorter because it only outputs associations with a q-value lower than 0.2 by default.
+
+The remaining models are ranked in order of the number of additional features they include.
+Additionally including phylogeny is a small improvement over the model with the horseshoe prior, while adding epitope prediction shows to larger drastic improvement.
+Note that this model only uses epitope *prediction* software and does not utilize any information of experimentally confirmed epitopes.
+These are only used for model evaluation.
+Therefore, our method even works in absence of any experimental epitope information.
 

### 2021-01-04 docs/manuscript.tex
add section on loo

@@
 
-\begin{itemize}
-  \item In addition to testing the model against real-world epitopes, we also perform leave-one-out cross-validation
-  \item unlike classical application of cross-validation, we are not that interested in the absolute magnitude of the prediction errors, i.e.
-how well we can estimate the occurrence of a replacement at a given alignment position, because the main objective of the model is to quantify the contribution of the HLA alleles.
-  \item But, LOO gives useful insight into how the accuracy of predictions change as we include more information in the model. 
-  \item Performing LOO has historically been challenging to apply for Bayesian models, because fitting the model N times is infeasible because of computational constrains.
-  \item Recently, Pareto-smoothed importance sampling has gained popularity.
-It approximates the LOO posterior from the full posterior and has the advantage that it allows to perform LOO-CV by only fitting the model once has good diagnostics to see if that approximation is unreliable.
-  \item cite PSIS paper
-  \item For Bayesian models, it is possible to use elpd as quantification of model performance. elpd is defined as -formula here-, i.e. the predictive density at the observed data point.
-  \item This has the advantage over other performance measures like classification accuracy that it not only takes the location of the predictive distribution into account (i.e.
+In addition to testing the model against experimentally validated epitopes, we also perform leave-one-out cross-validation.
+While the comparison against known epitopes compares inferences of the regression coefficients, leave-one-out cross-validation compares different models in regard to their predictive performance, i.e.
+how well they can predict the presence or absence of a replacement in a sequence based on the available data.
+
+Unlike usual applications of cross-validation, we are not primarily interested in the absolute magnitude of prediction errors, because the main objective of the statistical inference is to quantify the contribution of the HLA alleles.
+However, as the compared models are all special cases of the complete model, we can use cross-validation to gain insight on the relative importance of each aspect of the model.
+
+Performing leave-one-out cross-validation has historically been challenging to apply for Bayesian models, because fitting a model once for each left out data-point is often infeasible because of computational constrains.
+Recently, Pareto-smoothed importance sampling has gained popularity to perform approximate leave-one-out cross-validation for Bayesian models.
+Pareto-smoothed importance sampling approximates the leave-one-out posterior by re-weighting existing draws from the posterior distribution and has the advantage of allowing to perform leave-one-out cross-validation by only fitting the model once.
+Additionally, the method provides good diagnostics to identify cases where the approximation is unreliable~\cite{Vehtari2016}.
+
+Several utility (or cost) functions are used in practice to compare the predictive performance of models.
+For frequentist approaches, classification accuracy or receiver operator characteristic curves (ROC curves) are most common.
+For Bayesian models, expected log-predictive density is used most frequently.
+
+elpd is defined as \(\sum_{i=1}^{n}\text{log}(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))\), i.e. the average log predictive density of the observed data points based on the leave-one-out posteriors.
+This has the advantage over other performance measures like classification accuracy, that it not only takes the location of the predictive distribution into account (i.e.
 the number of correct predictions), but also the width (i.e.
 how confident the model is in its predictions).
-  \item Table shows LOO results for one dataset of serveral different models: A classical logistic regression model, similar to the one first used by Moore et al, a logistic regression model with horseshoe prior, a logistic regression model with horseshoe prior and phylogeny, and the complete model with horseshoe prior, phylogeny and epitope information.
-  \item Results are as expected, models with more information have higher predictive accuracy.
-Note that this is not due to overfitting, because LOO approximates model performance on unseen data.
-Additionally, the results are consistent across all datasets (see appendix).
-  \item Note how the model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any additional external data.
-This is because including the sparsity assumptions allows the model to better separate signal from noise and the uncertainty of the close-to-zero coefficients does not propagate into uncertainty of the estimated y.
-  \item Including phylogeny further improved model performance, as the assumption of i.i.d data does not hold for sequences that share a common phylogenetic history.
-  \item Perhaps surprisingly, the model with epitope prediction performs roughly as well as the model without.
-This is true from a perspective of predictive accuracy.
-However, in determining which HLA alleles are associated with the replacement, adding epitope prediction is highly useful, as shown in the previous section.
-This can be explained by the fact that a.) a given HLA allele is only restrincting a subset of all position in a sequence alignment and b.) the benefit of accurately identifying a relevant HLA allele is only relevant for the samples which do have that allele, which is only a small subset of all available samples.
-Therefore, looking at predictive accuracy alone is misleading in that regard, because we are more interested in the predictors than in the predictions itself.
-\end{itemize}
+elpd cannot be computed for frequentist models because the parameters of the model are viewed as fixed values.
+
+Table x shows the LOO results for the HBV dataset.
+Using a simple logistic regression model similar to the one used in~\cite{Moore2002} as a baseline, we compare a logistic regression model with horseshoe prior, a logistic regression model with horseshoe prior and phylogeny, and the complete HAMdetector model with horseshoe prior, phylogeny and epitope prediction to each other.
+The table shows expected log-predictive density differences compared to the baseline model (higher values are better).
+
+As the models include more and more information, the predictive accuracy increases.
+Note that this is not due to overfitting, because leave-one-out cross-validation approximates the model performance on unseen data.
+Additional tables are shown in the appendix, the results are consistent across all datasets.
+Note that the model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any additional external data.
+This is because including the sparsity assumption allows the model to better separate signal from noise and the uncertainty of the close-to-zero coefficients does not propagate into uncertainty of the estimated y.
+Including phylogeny further improves model performance, as the assumption of independent and identically distributed data does not hold for sequence data that share a common phylogenetic history.
+Perhaps surprisingly, the model with epitope prediction performs roughly as well as the model without.
+This is true from a perspective of predictive accuracy, however, in determining which HLA alleles are associated with a given replacement, adding epitope prediction is highly useful, as shown in the previous section.
+This observation is explained by the circumstance that leave-one-out cross-validation takes all positions and replacements into account, but a given HLA allele is only restricting a subset of all positions in a sequence alignment.
+Additionally, the benefit of accurately identifying an associated HLA allele is only relevant for the samples which do have that allele, which for most alleles is only a small subset of all available samples.
+Therefore, looking at predictive accuracy alone is misleading in that regard, because we are more interested in the predictors than in the predictions themselves.
 

### 2021-01-13 docs/manuscript.tex
progress on manuscript

@@
 \usepackage{amsmath}
+\usepackage{makecell}
 % \usepackage{lipsum} % Required to insert dummy text. To be removed otherwise
@@
 
-\begin{itemize}
-  \item We had the opportunity to test our model against a set of false positives that were identified by Fisher's exact test.
-On the HDV dataset, Fisher's exact test was applied to a set of HLA annotated sequences and x statistically significant results were reported.
-The identified positions were experimentally validated and y positions turned out to be false-positives, where Fisher's exact test showed extremely low p-values but experimental validation failed (how?).
-  \item This kind of data is rare because negative results are usually not published in the literature.
-  \item We ran HAMdetector on the same dataset, the posterior probabilities are shown in table x.
-  \item Note that a larger posterior probability denotes that the model is more confident in that HLA alleles being a HAM.
-  \item For most of the false positives we observe low posterior probabilities, which means that the model correctly identifies these alleles as not being relevant.
-  \item X of the false positives remain, which means that the model is not perfect, but a strong improvement over fisher's exact test.
-  \item these results were published previously using a preliminary version of HAMdetector, cite Roggendorf manuscript.
-\end{itemize}
+We had the opportunity to test our model against a set of associations identified by Fisher's exact test that turned out to be false positives.
+On the HDV dataset, Fisher's exact test was applied to a set of HLA annotated sequences and statistically significant results were reported.
+The identified positions were experimentally validated and 9 associations turned out to be false-positives.
+This kind of data is rate because negative results are usually not published in the literature.
+We ran HAMdetector on the same dataset, the posterior probabilities are shown in table \ref{tab:false-positives}.
+They denote the probability that the corresponding regression coefficient is positive.
+Associations with strong have a posterior probability close to 1, associations with no support a probability close to 0.5 (because the marginal posterior distribution is centered around 0 and allocated half of the probability mass to values greater than 0).
+The table shows that most of the false-positives in Fisher's exact test have low posterior probabilities in our model, which means that these alleles are correctly identified as not being associated with the replacement.
+The pairs of HLA alleles and replacements that were experimentally validated have the highest posterior probabilities, with the exception of P89T - HLA-B*37, which does not have strong support in our model.
+The results suggest that while our model is not perfect, it is a strong improvement over Fisher's exact test, which is frequently applied in this context.
+Please note that the results in this subsection have previously been published using a preliminary version of HAMdetector.
 
-\subsection{Rongge dataset}
-\begin{itemize}
-  \item We ran HLAdetector on a set of X HIVSequences sampled in China. X associations have a posterior mass greater than Y.
-  \item Maybe show a plot of marginal posteriors.
-\end{itemize}
+\begin{table}[]
+  \begin{tabular}{c|c|c|c|c}
+  \thead{replacement} & \thead{allele}  & \thead{p-value \\ (Fisher's exact test)} & \thead{post. prob. \\ (HAMdetector)} & \thead{confirmed} \\
+  \hline
+  S170N       & B*15 & \(3\cdot10^{-8}\)      & 0.99                                & +                        \\
+  D101E       & B*37 & 0.0002                        & 0.96                                & +                        \\
+  R105K       & B*27 & 0.0011                        & 0.93                                & +                        \\
+  R139K       & B*41 & 0.0034                        & 0.91                                & +                        \\
+  E47D        & B*18 & 0.0027                        & 0.89                                & +                        \\
+  D33E        & B*13 & 0.0001                        & 0.86                                & -                        \\
+  T134A       & A*68 & 0.0045                        & 0.82                                & -                        \\
+  K43R        & B*13 & 0.0021                        & 0.77                                & -                        \\
+  D47E        & A*30 & 0.0010                        & 0.76                                & -                        \\
+  K113R       & B*13 & 0.0043                        & 0.76                                & -                        \\
+  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
+  A107T       & B*14 & 0.0028                        & 0.70                                & -                        \\
+  P49L        & A*30 & 0.0031                        & 0.63                                & -                        \\
+  Q100L       & B*13 & 0.0018                        & 0.61                                & -                        \\
+  D96E        & B*13 & 0.0035                        & 0.50                                & -                       
+  \end{tabular}
+  \end{table}
 
@@
 
-\begin{itemize}
-  \item We present HAMdetector, a Bayesian model to identify HLA-associated mutations in a sequence alignment
-  \item We propose to use models that include as much information as possible and have shown that including sparsity and epitope prediction software can achieve better performance than existing methods.
-  \item The main advantage of using a logistic regression backbone is that it is easy to modify and the coefficients are interpretable as changes on the log-odds scale, like with a standard logistic regression model.
-  \item We can believe the general model structure can also be used in other contexts, for example in identifying associations in Sequence data and other features.
-  \item Maybe some additional points?
-\end{itemize}
+We present HAMdetector, a Bayesian model to identify HLA-associated mutations in a multiple sequence alignment.
+We propose to use models that include as much information as possible and have shown, that including sparsity and epitope prediction software can achieve better performance than existing methods.
+The logistic regression backbone can be easily modified, which allows the model to also be used in other contexts, for example in identifying associations between sequence data and other features.
+
 \thispagestyle{empty} % Removes page numbering from the first page

### 2021-01-14 docs/manuscript.tex

@@
   The human leukocyte antigen system (HLA) is of paramount importance to combat viral infections by presenting peptides on the cell surface via MHC I.
-Thus, CD8+ cytotoxic T-Lymphocytes exert a strong selection pressure towards virus variants that escape that immune recognition pathway, e.g.
-through point mutations that decreases binding of the respective peptide to MHC I.
+Thus, CD8\textsuperscript{+} cytotoxic T-Lymphocytes exert a strong selection pressure towards virus variants that escape that immune recognition pathway, e.g.
+through point mutations that decrease binding of the respective peptide to MHC I.
 \\
@@
 
-One way how the human immune system is able to recognize intracellular viral infections is through the human leukocyte antigen system\nolinebreak\cite{Germain1994}: In cells with active protein biosynthesis, proteins are continuously synthesized and also degraded by a process called proteasomal degradation, which cleaves proteins into linear peptides of varying length\nolinebreak\cite{Goldberg2002}.
+One way how the human immune system is able to recognize intracellular viral infections is through the human leukocyte antigen system~\cite{Germain1994}: In cells with active protein biosynthesis, proteins are continuously synthesized and also degraded by a process called proteasomal degradation, which cleaves proteins into linear peptides of varying length~\cite{Goldberg2002}.
 
 A small subset of these peptides is presented on the cell surface via receptors called MHC class I (HLA-A, HLA-B and HLA-C in humans).
-The genomic region encoding for MHC I is known to be highly polymorphic, with more than 20000 different HLA alleles described today\nolinebreak\cite{Robinson2014}.
-The resulting gene products differ in their binding properties, which means that cells from different individuals present a highly diverse set of peptides on their surface.
+The genomic region encoding for MHC I is known to be highly polymorphic, with more than 20000 different HLA alleles described today~\cite{Robinson2014}.
+The resulting gene products differ in their binding properties, which means that cells from different individuals present a highly diverse set of peptides on the cell surface.
 
-Cytotoxic T cells are selected during maturation to only weakly bind to peptide/MHC I complexes when the peptide originated from proteins of the usual proteome, but might be able to strongly bind to complexes of MHC I with peptides which are  generated from of a viral protein\nolinebreak\cite{Murata2007}.
+Cytotoxic T cells are selected during maturation to only weakly bind to peptide/MHC I complexes when the peptide originated from proteins of the usual proteome, but might be able to strongly bind to complexes of MHC I with peptides which are  generated from of a viral protein~\cite{Murata2007}.
 Upon activation, T cells induce cytolytic activity and recruit other immune cells \nolinebreak\cite{Harty2000}.
@@
 \subsection{HLA escape}
-In this way, the HLA system exerts strong selection pressure towards virus variants that escape T cell recognition\nolinebreak\cite{Borrow1997}, for example through a point mutation that results in reduced binding of an immunogenic peptide to MHC I or through a set of mutations that alters the viral protein in such a way that it is cleaved into different peptides that are not recognized by the host's T cell repertoire\nolinebreak\cite{Yewdell2002}.
+In this way, the HLA system exerts strong selection pressure towards virus variants that escape T cell recognition~\cite{Borrow1997}, for example through a point mutation that results in reduced binding of an immunogenic peptide to MHC I or through a set of mutations that alters the viral protein in such a way that it is cleaved into different peptides that are not recognized by the host's T cell repertoire~\cite{Yewdell2002}.
 
-The evolutionary events are complex and occur not only on the level of individuals, where a virus adapts to specific features of the host, but also on the population level, because HLA alleles differ in their frequency across geographic regions\nolinebreak\cite{Kawashima2009}.
-Upon transmission to a new host, HLA escape mutations can revert to their wild type, as HLA escape mutations are associated with a reduction in viral replicative capacity\nolinebreak\cite{Matthews2008}.
+The evolutionary events are complex and occur not only on the level of individuals, where a virus adapts to specific features of the host, but also on the population level, for example because HLA alleles differ in their frequency across geographic regions~\cite{Kawashima2009}.
+Upon transmission to a new host, HLA escape mutations can revert to their wild type, as they are typically associated with a reduction in viral replicative capacity~\cite{Matthews2008}.
 Kawashima et al.
@@
 
-There are several experimental methods available to study HLA escape: Recombinant MHC-I molecules can be used in binding assays.
+There are several experimental methods available to study HLA escape: Recombinant MHC I molecules can be used in binding assays.
 Upon complex formation with a peptide, a change in conformation can be detected with conformation-specific antibodies.
-This method is relatively fast, but only measures binding affinity of a peptide to MHC-I and does not account for antigen processing or immunodominance, which describes the observation that a peptide may be presented via MHC-I on the cell  surface, but does not induce an immune response.
+This method is relatively fast, but only measures binding affinity of a peptide to MHC I and does not account for antigen processing or immunodominance, which describes the observation that a peptide may be presented via MHC I on the cell  surface, but does not induce an immune response.
 
-An experimental setup that resembles the conditions in-vitro more closely but is also more time-consuming is to measure CD8+ T cell responses instead.
+An experimental setup that resembles the conditions in-vitro more closely but is also more time-consuming is to measure CD8\textsuperscript{+} T cell responses instead.
 This is usually done by stimulating peripheral blood mononuclear cells with prototype and variant peptides and measuring the secretion of IFN-\gamma{} by intracellular cytokine staining and fluorescence-assisted cell sorting.
 
-To analyze CD8+ T cell responses against endogenously processed antigens, it is necessary to generate cell-lines stably expressing the antigen in question and adding antigen-specific CD8+ T cells.
-This method scales poorly as it requires transfection of cell lines and antigen-specific expansion of CD8+ T cells.
+To analyze CD8\textsuperscript{+} T cell responses against endogenously processed antigens, it is necessary to generate cell-lines stably expressing the antigen in question and adding antigen-specific CD8\textsuperscript{+} T cells.
+This method scales poorly as it requires transfection of cell lines and antigen-specific expansion of CD8\textsuperscript{+} T cells.
 
@@
 
-As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface, escape mutations are often HLA-allele specific and can therefore be detected as HLA-allele dependent footprints in sequence alignments of viral proteins\nolinebreak\cite{Moore2002}: At certain alignment positions, a replacement might be more frequently observed in sequences from hosts with a specific HLA allele than in  sequences from hosts without that HLA allele.
+As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface, escape mutations are often HLA-allele specific and can therefore be detected as HLA-allele dependent footprints in sequence alignments of viral proteins~\cite{Moore2002}: At certain alignment positions, a replacement might be more frequently observed in sequences from hosts with a specific HLA allele than in  sequences from hosts without that HLA allele.
 By quantifying this difference for all replacement and HLA allele pairs, it is possible to identify replacements that are enriched in sequences coming from hosts with certain HLA alleles, and thus are likely to be HLA escape mutations.
 
-One way of quantifying this enrichment is Fisher's exact test\nolinebreak\cite{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(!H\)), (\(!R_{i}\), \(H\)) and (\(!R_{i}\), \(H\)), where \(!R_{i}\) denotes any replacement except \(R_{i}\), and \(!H\) denotes any HLA allele except \(H\).
+One way of quantifying this enrichment is Fisher's exact test~\cite{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(!H\)), (\(!R_{i}\), \(H\)) and (\(!R_{i}\), \(H\)), where \(!R_{i}\) denotes any replacement except \(R_{i}\), and \(!H\) denotes any HLA allele except \(H\).
 
@@
 
-Fisher's exact test has the advantage of being easy to apply\nolinebreak\cite{Budeus2016},  but also has several disadvantages that are outlined in Carlson et al.\nolinebreak\cite{Carlson2008}.
+Fisher's exact test has the advantage of being easy to apply~\cite{Budeus2016},  but also has several disadvantages that are outlined in Carlson et al.~\cite{Carlson2008}.
 The most striking one is that viral sequences share a common phylogenetic history, and,  therefore, treating sequences as independent and identically distributed samples may under- or overestimate effect sizes.
@@
 
-In addition to the aforementioned biological reasons why Fisher's exact test is  not suited for the analysis of sequence data, there are also more fundamental statistical issues that universally occur when using p-values as a screening tool\nolinebreak\cite{Amrhein2017}:
+In addition to the aforementioned biological reasons why Fisher's exact test is  not suited for the analysis of sequence data, there are also more fundamental statistical issues that universally occur when using p-values as a screening tool~\cite{Amrhein2017}:
 
-In the presence of small effect sizes and high variance between measurements, as  it is typically the case when working with biological data, statistically significant results can often be misleading and are likely to be in the wrong direction (a so-called type S error) or greatly overestimate an effect (a so-called type M error)\nolinebreak\cite{Gelman2014}.
-This problem has recently been widely appreciated in the literature in the context of the current \glqq replication crisis\grqq{}, which describes the circumstance that scientific claims with seemingly strong statistical evidence fail to replicate\nolinebreak\cite{Ioannidis2005}.
+In the presence of small effect sizes and high variance between measurements, as  it is typically the case when working with biological data, statistically significant results can often be misleading and are likely to be in the wrong direction (a so-called type S error) or greatly overestimate an effect (a so-called type M error)~\cite{Gelman2014}.
+This problem has recently been widely appreciated in the literature in the context of the current “replication crisis“, which describes the circumstance that scientific claims with seemingly strong statistical evidence fail to replicate~\cite{Ioannidis2005}.
 
@@
 When using p-values as a filter, it is therefore likely to obtain significant effects that are in fact not real.
-To circumvent this problem, a common strategy is to control the false discovery rate, which is the expected proportion of false positives\nolinebreak\cite{Benjamini1995}.
+To circumvent this problem, a common strategy is to control the false discovery rate, which is the expected proportion of false positives~\cite{Benjamini1995}.
 These adjustment procedures have the problem that, when performing many of such comparisons, none but the very largest effects remain.
@@
 Instead of performing many hypothesis tests and trying to adjust for them, we instead prefer to fit a single, multilevel model that contains all comparisons of interest.
-When using multilevel models, the problem of multiple comparisons can disappear entirely and  also yield more valid estimates\nolinebreak\cite{Gelman2012}.
+When using multilevel models, the problem of multiple comparisons can disappear entirely and  also yield more valid estimates~\cite{Gelman2012}.
 
@@
 When possible, we choose to fit Bayesian models.
-By using prior information, adding problem-specific structure and partial pooling, the accuracy of estimates can often be noticeably improved\nolinebreak\cite{Gelman2010}.
+By using prior information, adding problem-specific structure and partial pooling, the accuracy of estimates can often be noticeably improved~\cite{Gelman2010}.
 Prior information does not necessarily mean to use external data, even a rough idea about the expected magnitude of estimates is often surprisingly effective.
 
-Additionally, Bayesian statistics provides an accessible way to test models: By comparing data generated under the model's assumptions to the actually observed data, it is possible to identify important aspects of the dataset that the models fails to capture and subsequently improve the model until it is consistent with the observed data\nolinebreak\cite{Gabry2019}.
+Additionally, Bayesian statistics provides an accessible way to test models: By comparing data generated under the model's assumptions to the actually observed data, it is possible to identify important aspects of the dataset that the models fails to capture and subsequently improve the model until it is consistent with the observed data~\cite{Gabry2019}.
 
@@
     Similarly, there are also HLA-allele independent effects like antigen processing that influence presentation on MHC I.
-Second generational tools like MHCFlurry 2.0 use both binding affinity and antigen processing data to improve epitope prediction.
+Second generational tools like MHCFlurry 2.0~\cite{ODonnell2020} use both binding affinity and antigen processing data to improve epitope prediction.
 For our tool HAMdetector, we use the output of MHCFlurry to benefit from this binding affinity and antigen processing data in order to predict HLA-associated mutations.
@@
 
-  In Bayesian statistics, it is not required to chose one extreme or the other: Instead of either treating all HLA alleles from the same allele group as identical or as completely separate, partial pooling of the estimates allows something in between: By partial pooling, estimates do share some information across each other, but they are also allowed to vary if necessary.
+  In Bayesian statistics, it is not required to chose one extreme or the other: Instead of either treating all HLA alleles from the same allele group as identical or as completely separate, partial pooling of the estimates allows something in between: With partial pooling, estimates do share some information across each other, but they are also allowed to vary if necessary.
 
@@
 
-We chose a logistic regression model as a backbone for two reasons: First, because it is easily extensible and secondly, because the coefficients are interpretable as expected increases on the log-odds scale, identical to the familiar interpretation of standard logistic regression models.
-We also chose to fit the model in a Bayesian fashion using Stan, mostly because it allows us to make use of the hierarchical structure in the data and the possibility to use posterior predictive checks as a way of model checking.
+We choose a logistic regression model as a backbone for two reasons: First, because it is easily extensible and secondly, because the coefficients are interpretable as expected increases on the log-odds scale, identical to the familiar interpretation of standard logistic regression models.
+We also choose to fit the model in a Bayesian fashion using Stan, mostly because it allows us to make use of the hierarchical structure in the data and the possibility to use posterior predictive checks as a way of model checking.
 Consider the following notation:
@@
 Standard statistical methods usually assume samples to be independent and identically distributed, which may lead to wrong conclusions when this assumption is strongly violated.
-In the context of identifying HLA associations, Bhattacharya et al.\cite{Bhattacharya2007} demonstrated the importance of correcting for the phylogenetic structure.
-There are many approaches described in the literature for phylogenetic regression for binary dependent variables (see \cite{Ives2014}), where the most popular approach is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree \cite{Ives2009}.
+In the context of identifying HLA associations, Bhattacharya et al.~\cite{Bhattacharya2007} demonstrated the importance of correcting for the phylogenetic structure.
+There are many approaches described in the literature for phylogenetic regression for binary dependent variables~\cite{Ives2014}, where the most popular approach is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree~\cite{Ives2009}.
 
@@
   \includegraphics[width=1\linewidth]{plots/phylogeny_calibration.pdf}
-  \caption{Calibration plot for the HBV dataset. For a description of all datasets used in this study see section~\ref{sec:results}. Calibration plots for the other datasets are shown in the appendix.
+  \caption{Calibration plot for the HBV dataset.
+For a description of all datasets used in this study see section~\ref{sec:results}.
+Calibration plots for the other datasets are shown in the supplementary.
   All observations are first sorted by increasing estimated probability \(P(y_{ik}=1|\Psi)\) and then grouped into \(n\) bins.
@@
   y_{ik} & \sim \text{bernoulli}(\theta_{ik}) \\
-  \theta_{ik} & = \text{inv\_logit}(\beta_{0_{k}} + \gamma~\text{logit}(P(y_{ik}=1|\Psi)) \\ + \sum_{k=1}^{D}X_{ik}\beta_{\text{HLA}_{jk}})
+  \theta_{ik} & = \text{inv\_logit}(\beta_{0_{k}} + \gamma~\text{logit}(P(y_{ik}=1|\Psi)) \\ \hspace{-6cm} &~~~~ + \sum_{k=1}^{D}X_{ik}\beta_{\text{HLA}_{jk}})
 \end{aligned}
@@
 The uncertainty over the tree topology and the underlying parameters of the phylogenetic model would then propagate into uncertainty of the estimated probabilities \(P(y_{ik}=1|\Psi)\).
-However, in order to not increase the runtime of the model further we  use standard maximum likelihood estimates and then include these in a way that allows for measurement error.
+However, in order to not increase the runtime of the model further we  use standard maximum likelihood estimates and then include these in a way that additionally allows for measurement error.
 
@@
 
-The other two important confounding effects that are recognized in Carlson et al.\cite{Carlson2008} are HLA linkage disequilibrium and codon covariation.
+The other two important confounding effects that are recognized in Carlson et al.~\cite{Carlson2008} are HLA linkage disequilibrium and codon covariation.
 HLA linkage disequilibrium describes the observation that HLA alleles that lie in proximity to each other on the chromosome are also more likely to be inherited together.
@@
 We do not currently account for codon covariation in our model, as we are interested in identifying all HLA associated mutations (which include for example compensatory mutations).
-The general structure of the model, in particular the sparsity assumption introduced in section~\ref{sec:sparsity},would allow adding codon covariation by adding all other codons as additional predictors.
+The general structure of the model, in particular the sparsity assumption introduced in section~\ref{sec:sparsity}, would allow adding codon covariation by adding all other codons as additional predictors.
 This would generally not be possible in a frequentist setting because the predictor matrix would become singular and the inverse does not exist.
@@
 
-A useful addition to the horseshoe prior is given in Piironen et al.\cite{Piironen2017}, which describes a way to choose the parameter governing the overall degree of sparsity \(\tau_{0}\) based on the expected number of non-zero coefficients.
+A useful addition to the horseshoe prior is given in Piironen et al.~\cite{Piironen2017}, which describes a way to choose the parameter governing the overall degree of sparsity \(\tau_{0}\) based on the expected number of non-zero coefficients.
 
 For our HLA model, we utilize an extension to the horseshoe prior called the regularized horseshoe prior~\cite{Piironen2017}, which also provides some regularization for the non-zero coefficients.
-This is particularly useful for logistic regression models, as some shrinkage helps to deal with issues of separability and collinearity.
+This is particularly useful for logistic regression models, as some shrinkage helps to deal with issues of separability and collinearity that commonly occur when fitting logistic regression models.
 
@@
 
-The improvement of the logistic regression model with the horseshoe prior is a considerable improvement over a standard logistic regression model (see section~\ref{sec:results}.
-In our benchmark, the horseshoe prior alone is even shown to be more effective than a model without horseshoe prior but with phylogeny included.
+The improvement of the logistic regression model with the horseshoe prior is a considerable improvement over a standard logistic regression model  (section~\ref{sec:results}).
+In our benchmark, the addition of the horseshoe prior alone shows to be more effective than a model withoutt horseshoe prior but with phylogeny included.
 
@@
 
-For our model, we use MHCFlurry 2.0 for binding prediction, as it does not only predict HLA binding but also predicts antigen processing.
+For our model, we use MHCFlurry 2.0 for binding prediction, as it does not only predict HLA binding but also antigen processing.
 For this, we create an input matrix of dimensions \(R\times D\), where R is the number of evaluated replacements and D is the number of observed HLA alleles in the dataset.
@@
     \sigma_{\text{phy}} &\sim \text{Normal}^{+}(0, 0.5)\\
-    \beta_{\text{HLA}_{jk}} &\sim \text{Normal}(0, \tau^{2}\tilde{\lambda}^{2}_{jk})\\
-    \tilde{\lambda}^{2}_{jk} &= \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau^{2}\lambda_{jk}^{2}} \\
+    \beta_{\text{HLA}_{jk}} &\sim \text{Normal}(0, \tau_{k}^{2}\tilde{\lambda}^{2}_{jk})\\
+    \tilde{\lambda}^{2}_{jk} &= \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau_{k}^{2}\lambda_{jk}^{2}} \\
     c_{k}^2 &\sim \text{Inv-Gamma}(3.5, 3.5)\\
     \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}^2\text{exp}(Z_{jk}\beta_{\text{epi}}))\\
-    \tau &\sim \text{Cauchy}^{+}(0, \tau_{0})\\
-    \tau_{0} &= \frac{2}{D-2}\frac{2}{\sqrt{N}}\\
+    \tau_{k} &\sim \text{Cauchy}^{+}(0, \tau_{0k})\\
+    \tau_{0k} &= \frac{10}{D-10}\frac{2}{\sqrt{N}}\\
   \end{aligned}
@@
 However, the model would also work reasonably well with just a single global parameter \(\gamma\).
-All other additions (\(\tilde{\lambda}_{jk}^{2}\) and \(\tau_0\)) are explained in detail in \cite{Piironen2017}.
-Briefly, the addition of  \(\tilde{\lambda}_{jk}^{2}\) allows some regularization for the non-zero coefficients and the parameterization of \(tau_{0}\) allows to place a prior on the expected number of non-zero coefficients.
+All other additions (\(\tilde{\lambda}_{jk}^{2}\) and \(\tau_{0k}\)) are explained in detail in \cite{Piironen2017}.
+Briefly, the addition of  \(\tilde{\lambda}_{jk}^{2}\) allows some regularization for the non-zero coefficients and the parameterization of \(\tau_{0k}\) allows to place a prior on the expected number of non-zero coefficients.
 
-All models were implemented in Stan.
+All models were implemented in Stan~\cite{Stan2021}.
 The Stan code is available online in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
-A Julia\cite{Bezanson2017} package is available to run the model on custom data.
-Due to restrictions of dependencies (MHCFlurry and raxml-ng), HAMdetector is only available on Linux.
+A Julia~\cite{Bezanson2017} package is available to run the model on custom data.
+Due to restrictions of dependencies (MHCFlurry and RAxML-ng), HAMdetector is only available on Linux.
 We additionally provide Docker images to easily run the model on any operating system in a containerized environment, assuming Docker is installed.
@@
 
-The prior on \(c_{k}^{2}\) implies a Student-t prior with 7 degrees of freedom and a scale of 1 on HLA regression coefficients \(\beta_{\text{HLA}_{jk}}\) that are estimated to be away from 0.
+The prior on \(c_{k}^{2}\) implies a Student-t prior with 7 degrees of freedom and a scale of 1 on the non-zero HLA regression coefficients \(\beta_{\text{HLA}_{jk}}\).
 A Student-t prior with these parameters is a reasonable default choice for logistic regression models.
-The theoretical justification behind this regularized variant of the horseshoe prior is explained in detail in\cite{Piironen2017}.
+The theoretical justification behind this regularized variant of the horseshoe prior is explained in detail in~\cite{Piironen2017}.
 
-The value of \(\tau_{0}\) is chosen to imply 2 effective non-zero HLA regression coefficients per replacement.
+The value of \(\tau_{0k}\) is chosen to imply 10 effective non-zero HLA regression coefficients per replacement.
 The rationale behind this parameterization is again outlined in\cite{Piironen2017}.
-The value of 2 is approximately based on available HIV epitope maps.
+The value of 10 is approximately based on available HIV epitope maps.
 Please note that this is only an approximate statement about the general magnitude, i.e.
-we would expect maybe a couple of HLA alleles to be associated with a single replacement, but not 20 or more.
+we would expect maybe a couple of HLA alleles to be associated with a single replacement, not much more.
 The model is also parameterized in a way that assumes an equal degree of sparsity across all alignment positions a priori.
-We also tried to model \(\tau\) hierarchically, but observed sampling issues due to the resulting unfavorable geometry of the posterior.
+We also tried to model \(\tau_k\) hierarchically, but observed sampling issues due to the resulting unfavorable geometry of the posterior.
 
-\section{Results}
+\section{Results} \label{sec:results}
 
@@
   \item A large HIV dataset consisting of X sequences, that were kindly provided by X and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}.
-  \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\cite{Roomp2006}
+  \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\cite{Roomp2006}.
   \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm.
   \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf
-  \item A set of 41 HIV sequences that were kindly provided by Rongge Yang
+  \item A set of 41 HIV sequences that were kindly provided by Rongge Yang-
 \end{itemize}
@@
 
-For all the Bayesian models, Markov Chain Monte Carlo sampling was done using Stan.
+For all the Bayesian models, Markov Chain Monte Carlo sampling was done using Stan~\cite{Stan2021}.
 All model fits showed no signs of inference issues.
-In total, 4 different chains with 2000 samples each were drawn from the posterior.
+In total, 4 different chains with 1000 samples each were drawn from the posterior.
 The effective sample size for all model parameters exceeded 500, R-hat convergence diagnostic values were below 1.01 in all cases.
@@
 
-An example calibration plot of the AMdetector model on the Core protein sequence data of the HBV dataset is shown in figure~\ref{fig:calibration}.
+An example calibration plot of the HAMdetector model on the Core protein sequence data of the HBV dataset is shown in figure~\ref{fig:calibration}.
 Plots for the other models and datasets are shown in the supplementary.
@@
   \includegraphics[width=1\linewidth]{plots/phylogeny_calibration.pdf}
-  \caption{Calibration plot for the HBV dataset. Calibration plots for the other datasets are shown in the appendix.
+  \caption{Calibration plot for the HBV dataset. Calibration plots for the other datasets are shown in the supplementary.
   All observations are first sorted by increasing estimated probability \(P(y_{ik}=1)\) and then grouped into \(n\) bins.
@@
 That is, for Bayesian models we calculate the probability of the corresponding regression coefficient exceeding zero \(P(\beta_{HLA} > 0)\), for Fisher's exact test we sort HLA allele - replacement pairs by increasing p-values and for the Phylogenetic Dependency Network, we rank the associations by increasing q-values.
-Then, a list of known epitopes is used to create a plot of the cumulative number of replacements inside the boundary of a known epitope vs the corresponding rank.
+Then, a list of known epitopes is used to create a plot of the cumulative number of replacements inside the boundary of a known epitope vs.
+the corresponding rank.
 The underlying assumption for this kind of model check is that we expect to see an enrichment of replacements that lie inside the region of known epitopes and that this enrichment is stronger for better performing models.
@@
 
-Figure~\ref{fig:comparison} shows a comparison of different models on the HBe protein of the HBV dataset.
-Plots for all other datasets are shown in the Appendix, but show very similar patterns to the one shown here.
+Figure~\ref{fig:comparison} shows a comparison of different models on the HBe protein of the HBV dataset. Plots for all other datasets are shown in the supplementary, but show identical patterns.
 
-Unsurprisingly, all models show a strong enrichment of HLA-associated mutations compared to a randomly permutated list of HLA allele - replacement pairs (gray lines).
-Fisher's exact test performs about as well as the simple logistic regression model, as from a statistical perspective, they are both very similar.
+All models show a strong enrichment of HLA-associated mutations compared to a randomly permutated list of HLA allele - replacement pairs (gray lines).
+Fisher's exact test performs about as well as the simple logistic regression model, as from a statistical perspective, they are both similar.
 The horseshoe prior alone is a drastic improvement over Fisher's exact test and the logistic regression model with non-sparsifying priors, even though it does not include any external information.
@@
 Additionally including phylogeny is a small improvement over the model with the horseshoe prior, while adding epitope prediction shows to larger drastic improvement.
-Note that this model only uses epitope *prediction* software and does not utilize any information of experimentally confirmed epitopes.
+Note that this model only uses epitope \textit{prediction} software and does not utilize any information of experimentally confirmed epitopes.
 These are only used for model evaluation.
-Therefore, our method even works in absence of any experimental epitope information.
+Therefore, our method works even in absence of any experimental epitope information.
 
@@
 For frequentist approaches, classification accuracy or receiver operator characteristic curves (ROC curves) are most common.
-For Bayesian models, expected log-predictive density is used most frequently.
+For Bayesian models, expected log-predictive density (elpd) is used most frequently.
 
-elpd is defined as \(\sum_{i=1}^{n}\text{log}(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))\), i.e. the average log predictive density of the observed data points based on the leave-one-out posteriors.
+elpd is defined as \(\sum_{i=1}^{n}\text{log}(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))\), i.e.
+the average log predictive density of the observed data points based on the leave-one-out posterior distributions.
 This has the advantage over other performance measures like classification accuracy, that it not only takes the location of the predictive distribution into account (i.e.
@@
 Note that this is not due to overfitting, because leave-one-out cross-validation approximates the model performance on unseen data.
-Additional tables are shown in the appendix, the results are consistent across all datasets.
+Additional tables are shown in the supplementary, the results are consistent across all datasets.
 Note that the model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any additional external data.
@@
 They denote the probability that the corresponding regression coefficient is positive.
-Associations with strong have a posterior probability close to 1, associations with no support a probability close to 0.5 (because the marginal posterior distribution is centered around 0 and allocated half of the probability mass to values greater than 0).
+Associations with strong support have a posterior probability close to 1, associations with no support a probability close to 0.5 (because the marginal posterior distribution is centered around 0 and allocated half of the probability mass to values greater than 0).
 The table shows that most of the false-positives in Fisher's exact test have low posterior probabilities in our model, which means that these alleles are correctly identified as not being associated with the replacement.
@@
   \begin{tabular}{c|c|c|c|c}
+  \label{tab:false-positives}
   \thead{replacement} & \thead{allele}  & \thead{p-value \\ (Fisher's exact test)} & \thead{post. prob. \\ (HAMdetector)} & \thead{confirmed} \\
@@
 We present HAMdetector, a Bayesian model to identify HLA-associated mutations in a multiple sequence alignment.
-We propose to use models that include as much information as possible and have shown, that including sparsity and epitope prediction software can achieve better performance than existing methods.
-The logistic regression backbone can be easily modified, which allows the model to also be used in other contexts, for example in identifying associations between sequence data and other features.
+We propose to use models that include as much information as possible and have shown that including sparsity and epitope prediction software can achieve better performance than existing methods.
+The logistic regression backbone can be easily extended, which allows the model to also be used in other contexts, for example in identifying associations between sequence data and other features.
 

### 2021-01-17 docs/manuscript.tex
starting work on MS, abstract and intro (part)

@@
 \usepackage{makecell}
+\usepackage{soul}
 % \usepackage{lipsum} % Required to insert dummy text. To be removed otherwise
@@
 \PaperTitle{
-    HAMdetector: Combining information to detect HLA-associated mutations with a
-    Bayesian regression model
+    HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations
 } % Article title
@@
   \textbf{Motivation} \\
-  The human leukocyte antigen system (HLA) is of paramount importance to combat viral infections by presenting peptides on the cell surface via MHC I.
-Thus, CD8\textsuperscript{+} cytotoxic T-Lymphocytes exert a strong selection pressure towards virus variants that escape that immune recognition pathway, e.g.
-through point mutations that decrease binding of the respective peptide to MHC I.
-\\
+  The human leukocyte antigen system (HLA) presents viral peptide fragments in MHC I protein-peptide complexes on cell surfaces and in this way alerts CD8\textsuperscript{+} cytotoxic T-Lymphocytes (CTLs), a key component of adaptive immunity.
+This pathway exerts a strong selection pressure on viruses, favoring viral mutants that escape recognition by the HLA/CTL system, e.g.\ by point mutations that decrease binding of viral peptides to MHC I.
+Naturally, such immune escape mutations often emerge as HLA-associated mutations (HAMs), specific to the host HLA alleles and its MHC I proteins.
+The reliable identification of HAMs is not only important to understand viral genomes and their evolution, but it also impacts the development of broadly effective anti-viral treatments and vaccines against variable viruses.\\
 
-  Reliably identifying HLA-associated mutations is important for understanding viral evolution, but experimental methods like binding assays are prohibitively expensive for large-scale use and fail to recognize other mechanisms of immune escape like proteasomal processing.
-
-  One step in finding these mutations is through the statistical analysis of sequence data.
-However, existing methods are based on null hypothesis significance testing and do not make use of all the available information and therefore have unsatisfactory real-world performance.
-\\
+  By their very nature HAMs are amenable to detection by statistical methods in paired sequence / HLA data.
+However, HLA alleles are very polymorphic in the human host population which makes the available data relatively sparse and noisy.
+Under these circumstances, one way to optimize HAM detection is to integrate all relevant information in a coherent model.
+Bayesian inference offers a principled approaches to achieve this.\\
 
   \textbf{Results} \\
-  Here, we present a Bayesian regression model that is easily extensible to include information from different sources (e.g.
-epitope prediction software) and makes use of recent advances in Bayesian inference, e.g.
-by using a sparsifying prior.
-We show that including this kind of information improves predictive performance considerably over state-of-the-art methods.
+  Here, we present a Bayesian regression model that is easily extensible to include information from different sources, and that makes use of recent advances in Bayesian inference, e.g.\ by using a sparsifying prior.
+We show that this integrative approach improves performance in HAM detection considerably over state-of-the-art methods.
 \\
@@
 
-One way how the human immune system is able to recognize intracellular viral infections is through the human leukocyte antigen system~\cite{Germain1994}: In cells with active protein biosynthesis, proteins are continuously synthesized and also degraded by a process called proteasomal degradation, which cleaves proteins into linear peptides of varying length~\cite{Goldberg2002}.
+One way of how the human immune system recognizes intracellular viral infections is through the human leukocyte antigen system~\cite{Germain1994}: In cells with active protein biosynthesis, proteins are continuously synthesized and also degraded by proteasomal degradation, which cleaves proteins into peptides~\cite{Goldberg2002}.
 
-A small subset of these peptides is presented on the cell surface via receptors called MHC class I (HLA-A, HLA-B and HLA-C in humans).
-The genomic region encoding for MHC I is known to be highly polymorphic, with more than 20000 different HLA alleles described today~\cite{Robinson2014}.
+A small subset of these peptides is presented on the cell surface by receptors called MHC class I encoded by HLA genes, e.g.\ in humanc HLA-A, -B, and -C.
+HLA genes are very polymorphic between human individuals, with more than 20,000 different described HLA alleles~\cite{Robinson2014}.
 The resulting gene products differ in their binding properties, which means that cells from different individuals present a highly diverse set of peptides on the cell surface.
 
-Cytotoxic T cells are selected during maturation to only weakly bind to peptide/MHC I complexes when the peptide originated from proteins of the usual proteome, but might be able to strongly bind to complexes of MHC I with peptides which are  generated from of a viral protein~\cite{Murata2007}.
-Upon activation, T cells induce cytolytic activity and recruit other immune cells \nolinebreak\cite{Harty2000}.
+Cytotoxic T-Lymphocytes (CTLs) cells are selected during maturation to only weakly bind to MHC I protein-peptide complexes if the peptide originates from proteins of the self-proteome, but might be able to strongly bind to MHC I protein-peptide complexes with peptides generated from viral proteins~\cite{Murata2007}.
+Upon activation, T cells induce cytolytic activity and recruit other immune cells~\cite{Harty2000}.
 
@@
 
-One way of quantifying this enrichment is Fisher's exact test~\cite{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(!H\)), (\(!R_{i}\), \(H\)) and (\(!R_{i}\), \(H\)), where \(!R_{i}\) denotes any replacement except \(R_{i}\), and \(!H\) denotes any HLA allele except \(H\).
+One way of quantifying this enrichment is Fisher's exact test~\cite{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(\neg H\)), (\(\neg R_{i}\), \(H\)) and (\(\neg R_{i}\), \(H\)), where \(\neg R_{i}\) denotes any replacement except \(R_{i}\), and \(\neg H\) denotes any HLA allele except \(H\).
 

### 2021-01-17 docs/manuscript.tex
more work on intro

@@
 This pathway exerts a strong selection pressure on viruses, favoring viral mutants that escape recognition by the HLA/CTL system, e.g.\ by point mutations that decrease binding of viral peptides to MHC I.
-Naturally, such immune escape mutations often emerge as HLA-associated mutations (HAMs), specific to the host HLA alleles and its MHC I proteins.
+Naturally, such immune escape mutations often emerge in highly variable viruses, e.g.\ HIV or HBV, as HLA-associated mutations (HAMs), specific to the host HLA alleles and its MHC I proteins.
 The reliable identification of HAMs is not only important to understand viral genomes and their evolution, but it also impacts the development of broadly effective anti-viral treatments and vaccines against variable viruses.\\
@@
 \subsection{HLA escape}
-In this way, the HLA system exerts strong selection pressure towards virus variants that escape T cell recognition~\cite{Borrow1997}, for example through a point mutation that results in reduced binding of an immunogenic peptide to MHC I or through a set of mutations that alters the viral protein in such a way that it is cleaved into different peptides that are not recognized by the host's T cell repertoire~\cite{Yewdell2002}.
+In this way, the HLA system exerts strong selection pressure towards virus variants that escape T cell recognition~\cite{Borrow1997}.
+Such variants could, for example, carry mutations that result in reduced binding of immunogenic peptides to MHC I or to T cell receptors, or that alter the processing and binding so that peptides are no longer presented at all~\cite{Yewdell2002}.
 
 The evolutionary events are complex and occur not only on the level of individuals, where a virus adapts to specific features of the host, but also on the population level, for example because HLA alleles differ in their frequency across geographic regions~\cite{Kawashima2009}.
-Upon transmission to a new host, HLA escape mutations can revert to their wild type, as they are typically associated with a reduction in viral replicative capacity~\cite{Matthews2008}.
-Kawashima et al.
-\nolinebreak\cite{Kawashima2009} describe an escape mutation that is selected by HLA allele HLA-B*51, does not strongly affect viral replicative capacity, and therefore slowly enriches over time in Japan, where HLA-B*51 commonly occurs.
+Upon transmission to a new host with different HLA alleles, HLA escape mutations can revert to their wild type, as they are often associated with a reduction in viral replication capacity~\cite{Matthews2008}.
+If the latter is not the case, such mutations may not revert but be retained in the viral genome~\cite{Kawashima2009}.
 
-How quickly a given escape mutation is selected upon transmission in a host depends on the magnitude of the reduction in viral replicative capacity, on the strength of selection pressure, and also on the genetic background, e.g.
-some escape mutations require compensatory mutations which partly attenuate the negative impact on viral replicative capacity.
+Whether and how quickly a given escape mutation is selected in a host depends on several factors, such as the viral genomic background and diversity, the magnitude of the reduction in viral replicative capacity, the availability of compensatory mutations that recover fitness, and the strength of selection pressure.
+\hl{Hier wäre etwas Literatur gut.
+Vielleicht ein review zum Thema?}
 
-Studying HLA escape provides a unique opportunity to gain insight into viral evolution, on the host level, but also on the population level.
-Unfortunately, identifying HLA escape mutations is difficult in practice.
+Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV \hl{refs?}.
+This includes the development of treatments and vaccines that rely on effective immune responses.
+This makes detection of immune escape mutations critical.
 
-\subsection{Identifying HLA-escape mutations}
+\subsection{Identifying HLA escape mutations}
 
-There are several experimental methods available to study HLA escape: Recombinant MHC I molecules can be used in binding assays.
-Upon complex formation with a peptide, a change in conformation can be detected with conformation-specific antibodies.
-This method is relatively fast, but only measures binding affinity of a peptide to MHC I and does not account for antigen processing or immunodominance, which describes the observation that a peptide may be presented via MHC I on the cell  surface, but does not induce an immune response.
-
-An experimental setup that resembles the conditions in-vitro more closely but is also more time-consuming is to measure CD8\textsuperscript{+} T cell responses instead.
-This is usually done by stimulating peripheral blood mononuclear cells with prototype and variant peptides and measuring the secretion of IFN-\gamma{} by intracellular cytokine staining and fluorescence-assisted cell sorting.
-
-To analyze CD8\textsuperscript{+} T cell responses against endogenously processed antigens, it is necessary to generate cell-lines stably expressing the antigen in question and adding antigen-specific CD8\textsuperscript{+} T cells.
-This method scales poorly as it requires transfection of cell lines and antigen-specific expansion of CD8\textsuperscript{+} T cells.
-
-\subsection{Computational methods}
-
-Because the currently available experimental methods do not scale well enough to analyze whole viral genomes, a useful strategy might be to use annotated sequence data to identify candidate HLA escape mutations that can then be verified experimentally.
+There are several experimental methods available to study HLA escape \hl{refs}.
+However, these methods are relatively slow and costly, especially for screening purposes.
+A more promising approach is to combine viral genome sequencing, host HLA determination, computational identification by statistical association analysis, and targeted experimental validation.
+\hl{vielleicht schon hier eine Referenz dazu}
 
@@
 
-Fisher's exact test has the advantage of being easy to apply~\cite{Budeus2016},  but also has several disadvantages that are outlined in Carlson et al.~\cite{Carlson2008}.
+Fisher's exact test has the advantage of being easy to apply~\cite{Budeus2016},  but also has several disadvantages~\cite{Carlson2008}.
 The most striking one is that viral sequences share a common phylogenetic history, and,  therefore, treating sequences as independent and identically distributed samples may under- or overestimate effect sizes.

### 2021-01-18 docs/manuscript.tex
erste grobe Lektüre abgeschlossen

@@
 Whether and how quickly a given escape mutation is selected in a host depends on several factors, such as the viral genomic background and diversity, the magnitude of the reduction in viral replicative capacity, the availability of compensatory mutations that recover fitness, and the strength of selection pressure.
-\hl{Hier wäre etwas Literatur gut.
+\hl{DaHo: Hier wäre etwas Literatur gut.
 Vielleicht ein review zum Thema?}
 
-Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV \hl{refs?}.
+Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV \hl{DaHo: refs?}.
 This includes the development of treatments and vaccines that rely on effective immune responses.
@@
 
-There are several experimental methods available to study HLA escape \hl{refs}.
+There are several experimental methods available to study HLA escape \hl{DaHo: refs}.
 However, these methods are relatively slow and costly, especially for screening purposes.
 A more promising approach is to combine viral genome sequencing, host HLA determination, computational identification by statistical association analysis, and targeted experimental validation.
-\hl{vielleicht schon hier eine Referenz dazu}
+\hl{DaHo: vielleicht schon hier eine Referenz dazu}
 
@@
 
-Another issue that occurs when using p-values as a screening tool is the problem of multiple comparisons.
-When applying a statistical test, the probability  of  obtaining a statistically significant result increases with each additional test, even in absence of any real effect.
+These problems are exacerbated if the p-values are used for screening purposes (multiple testing problem).
+The probability  of  obtaining a statistically significant result increases with each additional test, even in absence of any real effect.
 When using p-values as a filter, it is therefore likely to obtain significant effects that are in fact not real.
@@
   y_{ik} & \sim \text{bernoulli}(\theta_{ik}) \\
-  \theta_{ik} & = \text{inv\_logit}(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}})
-\end{align},
+  \theta_{ik} & = \text{inv\_logit}(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}}),
+\end{align}
 
@@
 
-One of the recent advances in Bayesian inference is the development of special prior distributions which can be used to give models additional useful properties.
+\hl{DaHo: Wiederholung} One of the recent advances in Bayesian inference is the development of special prior distributions which can be used to give models additional useful properties.
 An example of these special prior distributions are the so-called sparsity-inducing priors, which can be used to convey the assumption that most of the predictors in a regression model are not associated with the outcome and are therefore 0.
@@
     \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \text{exp}(Z_{jk}\beta_{\text{epi}})) \\
-    \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2)
+    \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2),
   \end{aligned}
-\end{equation},
+\end{equation}
 
@@
 
-Prior distributions are labeled with an asterisk in equation~\ref{eq:full}.
+Prior distributions are labeled with an asterisk \hl{DaHo: * fehlen} in equation~\ref{eq:full}.
 They are meant to be weakly informative, which means that the goal is to provide some modest amount of shrinkage while still allowing for large absolute values given enough support by the observed data.
@@
 
+\hl{DaHo: $\hat{R}$ und $n_{\mathrm{eff}}$ sollten in den Methods eingeführt werden}
 For all the Bayesian models, Markov Chain Monte Carlo sampling was done using Stan~\cite{Stan2021}.
@@
 \subsection*{Posterior predictive checks}
-
+\hl{DaHo: Sollte in den Methods eingeführt werden}
 Bayesian modelling allows a simple yet effective strategy to test models: By simulating data under the model's assumptions and then comparing simulated and observed data, model misspecifications can be identified and highlight possible model improvements.
@@
 However, comparing against real-world data is not straight forward when identifying HLA-associated mutations, because if a tool identifies an HLA-associated mutation which is not yet documented, this could be either because it is a false positive, or a yet unknown HLA-associated mutation that has not been documented before.
-Also, viewing every non-identified HLA-associated mutation that is listed in the literature as a false negative is not correct either, because documented escape mutations do not necessarily have to show in a sequence alignment.
+Also, viewing every non-identified HLA-associated mutation that is listed in the literature as a false negative is not correct either, because documented escape mutations do not necessarily have to show in each sequence alignment.
 A third issue when comparing models is that HLA-associated mutations are not necessarily binary, so the true positive / false negative framework does not work as well.
@@
 The remaining models are ranked in order of the number of additional features they include.
-Additionally including phylogeny is a small improvement over the model with the horseshoe prior, while adding epitope prediction shows to larger drastic improvement.
-Note that this model only uses epitope \textit{prediction} software and does not utilize any information of experimentally confirmed epitopes.
-These are only used for model evaluation.
-Therefore, our method works even in absence of any experimental epitope information.
+Additionally including phylogeny is a small improvement over the model with the horseshoe prior, while adding epitope prediction provides another strong improvement.
+
+Note that this model only uses epitope \textit{prediction} software and does not use any information of experimentally confirmed epitopes. These are here only used for model evaluation.
 
 \subsection{Leave-one-out cross-validation}
-
-In addition to testing the model against experimentally validated epitopes, we also perform leave-one-out cross-validation.
+\hl{DaHo: LOO auch in Methods einführen und kürzen}
+In addition to testing the model against experimentally validated epitopes, we also performed leave-one-out cross-validation.
 While the comparison against known epitopes compares inferences of the regression coefficients, leave-one-out cross-validation compares different models in regard to their predictive performance, i.e.
@@
 
-Table x shows the LOO results for the HBV dataset.
+Table x \hl{DaHo: fehlt} shows the LOO results for the HBV dataset.
 Using a simple logistic regression model similar to the one used in~\cite{Moore2002} as a baseline, we compare a logistic regression model with horseshoe prior, a logistic regression model with horseshoe prior and phylogeny, and the complete HAMdetector model with horseshoe prior, phylogeny and epitope prediction to each other.
@@
 Perhaps surprisingly, the model with epitope prediction performs roughly as well as the model without.
-This is true from a perspective of predictive accuracy, however, in determining which HLA alleles are associated with a given replacement, adding epitope prediction is highly useful, as shown in the previous section.
+This is true from a perspective of predictive accuracy.
+However, in determining which HLA alleles are associated with a given replacement, adding epitope prediction is highly useful, as shown in the previous section.
 This observation is explained by the circumstance that leave-one-out cross-validation takes all positions and replacements into account, but a given HLA allele is only restricting a subset of all positions in a sequence alignment.
-Additionally, the benefit of accurately identifying an associated HLA allele is only relevant for the samples which do have that allele, which for most alleles is only a small subset of all available samples.
+Additionally, the benefit of accurately identifying an associated HLA allele is only relevant for the samples that do have that allele, which for most alleles is only a small subset of all available samples.
 Therefore, looking at predictive accuracy alone is misleading in that regard, because we are more interested in the predictors than in the predictions themselves.
+\hl{DaHo: diesen Abschnitt etwas klarer schreiben}
 
@@
 
-We had the opportunity to test our model against a set of associations identified by Fisher's exact test that turned out to be false positives.
+We had the opportunity to test our model against a set of associations identified by Fisher's exact test that turned out to be false positives \hl{DaHo: diesen Abschnitt  glattziehen}.
+A comparison with a preliminary version of HAMdetector have been published recently.
+Here we provide a comparison between results from Fisher's exact test~\cite{Budeus2016} and the mature version of HAMdetector used in this work.
+
 On the HDV dataset, Fisher's exact test was applied to a set of HLA annotated sequences and statistically significant results were reported.
@@
 The pairs of HLA alleles and replacements that were experimentally validated have the highest posterior probabilities, with the exception of P89T - HLA-B*37, which does not have strong support in our model.
-The results suggest that while our model is not perfect, it is a strong improvement over Fisher's exact test, which is frequently applied in this context.
-Please note that the results in this subsection have previously been published using a preliminary version of HAMdetector.
 
@@
 
+\hl{DaHo: wir haben zwar mehrere Datensätze, gehen aber praktisch nicht auf die Resultate mit diesen Datensätzen ein.
+Wie besprochen sollte dieses Manuskript eher methodisch sein, d.h.~es ist ok, dass wir nicht *viel* auf die virologische/immunologische Seite eingehen.
+Aber, dass wir überhaupt nichts zu den Resultaten sagen, ist etwas zu wenig.
+Z.B.
+könnte man die Vielfalt der Datensätze nutzen um den Punkt zu machen, dass die Verbesserungen nicht von den Details des Virus abhängen.
+Vielleicht fällt Dir ja noch was dazu ein.}
+
 \thispagestyle{empty} % Removes page numbering from the first page

### 2021-01-21 docs/manuscript.tex
add Loo table

@@
 \documentclass[fleqn,11pt]{SelfArx} % Document font size and equations flushed left
-
+\usepackage[justification=justified]{caption}
 \usepackage{parskip}
@@
 \subsection{Leave-one-out cross-validation}
-\hl{DaHo: LOO auch in Methods einführen und kürzen}
+\hl{DaHo: LOO auch in Methods einführen}
 In addition to testing the model against experimentally validated epitopes, we also performed leave-one-out cross-validation.
@@
 
-Unlike usual applications of cross-validation, we are not primarily interested in the absolute magnitude of prediction errors, because the main objective of the statistical inference is to quantify the contribution of the HLA alleles.
-However, as the compared models are all special cases of the complete model, we can use cross-validation to gain insight on the relative importance of each aspect of the model.
+As the compared models are all special cases of the complete model, we can use cross-validation to gain insight on the relative importance of each aspect of the model.
 
@@
 Recently, Pareto-smoothed importance sampling has gained popularity to perform approximate leave-one-out cross-validation for Bayesian models.
-Pareto-smoothed importance sampling approximates the leave-one-out posterior by re-weighting existing draws from the posterior distribution and has the advantage of allowing to perform leave-one-out cross-validation by only fitting the model once.
+This method  approximates the leave-one-out posterior by re-weighting existing draws from the posterior distribution and has the advantage of allowing to perform leave-one-out cross-validation by only fitting the model once.
 Additionally, the method provides good diagnostics to identify cases where the approximation is unreliable~\cite{Vehtari2016}.
@@
 
-Table x \hl{DaHo: fehlt} shows the LOO results for the HBV dataset.
-Using a simple logistic regression model similar to the one used in~\cite{Moore2002} as a baseline, we compare a logistic regression model with horseshoe prior, a logistic regression model with horseshoe prior and phylogeny, and the complete HAMdetector model with horseshoe prior, phylogeny and epitope prediction to each other.
-The table shows expected log-predictive density differences compared to the baseline model (higher values are better).
+Table \ref{loo} shows a comparison of models with increasing complexity run on the HBe protein in the HDV dataset.
 
-As the models include more and more information, the predictive accuracy increases.
+\begin{table}[h!]
+  \renewcommand{\arraystretch}{1.3}
+  \centering
+  \caption{Differences in elpd for models with increasing model complexity (HBe protein of the HBV dataset).
+  The numbers in column \(\text{elpd}_\text{diff}\) show increases in elpd compared to the previous model (higher is better), e.g.
+the difference in elpd of the model with horseshoe prior is 949.8 compared to the baseline model, and the difference in elpd for the model with phylogeny and horseshoe prior is 4440.9 compared to the model just with the horseshoe prior.
+All differences in elpd are larger than several times the estimated standard error (column \(\text{se}_\text{diff}\)), indicating that models which account for more information are expected to have a better predictive performance.}
+  \begin{tabular}{l|r|rll}
+  \multicolumn{1}{l|}{} & \multicolumn{1}{c|}{\(\text{elpd}_\text{diff}\)} & \multicolumn{1}{c}{\(\text{se}_\text{diff}\)} &  &  \\ \cline{1-3}
+  logistic regression (baseline) & 0.0  & 0.0 &  &  \\
+  + horseshoe prior     & 949.8  & 65.2 &  &  \\
+  + phylogeny & 4440.9 & 94.4 &  &  \\
+  + epitope prediction & 63.1   & 18.9 &  & 
+  \end{tabular}
+  \label{loo}
+\end{table}
+
+Predictive accuracy increases as the models include more and more information.
 Note that this is not due to overfitting, because leave-one-out cross-validation approximates the model performance on unseen data.
 Additional tables are shown in the supplementary, the results are consistent across all datasets.
+
 Note that the model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any additional external data.
 This is because including the sparsity assumption allows the model to better separate signal from noise and the uncertainty of the close-to-zero coefficients does not propagate into uncertainty of the estimated y.
+
 Including phylogeny further improves model performance, as the assumption of independent and identically distributed data does not hold for sequence data that share a common phylogenetic history.
-Perhaps surprisingly, the model with epitope prediction performs roughly as well as the model without.
-This is true from a perspective of predictive accuracy.
-However, in determining which HLA alleles are associated with a given replacement, adding epitope prediction is highly useful, as shown in the previous section.
-This observation is explained by the circumstance that leave-one-out cross-validation takes all positions and replacements into account, but a given HLA allele is only restricting a subset of all positions in a sequence alignment.
-Additionally, the benefit of accurately identifying an associated HLA allele is only relevant for the samples that do have that allele, which for most alleles is only a small subset of all available samples.
-Therefore, looking at predictive accuracy alone is misleading in that regard, because we are more interested in the predictors than in the predictions themselves.
-\hl{DaHo: diesen Abschnitt etwas klarer schreiben}
+
+Inclusion of epitope prediction does not improve elpd as much as inclusion of phylogeny and the sparsity assumption.
+This is because while addition of sparsity and phylogeny has an effect on all replacements and samples, epitope prediction only influences those replacements which are restricted by a given HLA allele and only those samples which are annotated with the allele.
+However, inclusion of epitope prediction is highly useful for determining which HLA alleles are associated with the occurrence of a replacement, as shown in the previous section.
 

### 2021-01-22 docs/manuscript.tex
resolve some remarks in the manuscript

@@
 
-Cytotoxic T-Lymphocytes (CTLs) cells are selected during maturation to only weakly bind to MHC I protein-peptide complexes if the peptide originates from proteins of the self-proteome, but might be able to strongly bind to MHC I protein-peptide complexes with peptides generated from viral proteins~\cite{Murata2007}.
+Cytotoxic T-Lymphocytes (CTLs) are selected during maturation to only weakly bind to MHC I protein-peptide complexes if the peptide originates from proteins of the self-proteome, but might be able to strongly bind to MHC I protein-peptide complexes with peptides generated from viral proteins~\cite{Murata2007}.
 Upon activation, T cells induce cytolytic activity and recruit other immune cells~\cite{Harty2000}.
@@
 
-Whether and how quickly a given escape mutation is selected in a host depends on several factors, such as the viral genomic background and diversity, the magnitude of the reduction in viral replicative capacity, the availability of compensatory mutations that recover fitness, and the strength of selection pressure.
-\hl{DaHo: Hier wäre etwas Literatur gut.
-Vielleicht ein review zum Thema?}
+Whether and how quickly a given escape mutation is selected in a host depends on several factors, such as the viral genomic background and diversity, the magnitude of the reduction in viral replicative capacity, the availability of compensatory mutations that recover fitness, and the strength of selection pressure~\cite{Kloverpris2016}.
 
-Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV \hl{DaHo: refs?}.
+Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV~\cite{Alizon2011}\cite{Allen2005}\cite{Rousseau2008}\cite{Lumley2018}.
 This includes the development of treatments and vaccines that rely on effective immune responses.
@@
 
-There are several experimental methods available to study HLA escape \hl{DaHo: refs}.
+There are several experimental methods available to study HLA escape~\cite{Timm2004}.
 However, these methods are relatively slow and costly, especially for screening purposes.
-A more promising approach is to combine viral genome sequencing, host HLA determination, computational identification by statistical association analysis, and targeted experimental validation.
-\hl{DaHo: vielleicht schon hier eine Referenz dazu}
+Therefore, a promising approach is to combine viral genome sequencing, host HLA determination, computational identification by statistical association analysis, and targeted experimental validation~\cite{Carlson2012}.
 
@@
 
-  \textit{Partial pooling of 4-digit HLA alleles}
+  % \textit{Partial pooling of 4-digit HLA alleles}
 
-  It has recently been appreciated that binding specificities can vary drastically across  HLA alleles from the same allele group, e.g.
+  % It has recently been appreciated that binding specificities can vary drastically across  HLA alleles from the same allele group, e.g.
 between HLA-B*51:01 and HLA-B*51:03.
@@
 
-  In Bayesian statistics, it is not required to chose one extreme or the other: Instead of either treating all HLA alleles from the same allele group as identical or as completely separate, partial pooling of the estimates allows something in between: With partial pooling, estimates do share some information across each other, but they are also allowed to vary if necessary.
+  % In Bayesian statistics, it is not required to chose one extreme or the other: Instead of either treating all HLA alleles from the same allele group as identical or as completely separate, partial pooling of the estimates allows something in between: With partial pooling, estimates do share some information across each other, but they are also allowed to vary if necessary.
 
-  In the context of our model this means that observing a relevant association between a replacement and an HLA allele also makes the model more inclined to estimate relevant associations between that replacement and all other alleles from that allele group.
+  % In the context of our model this means that observing a relevant association between a replacement and an HLA allele also makes the model more inclined to estimate relevant associations between that replacement and all other alleles from that allele group.
 The degree of this partial pooling can be estimated from the data, so if we do observe strong similarity of HLA alleles in a given allele group the estimates are more influenced by each other than if they do not.
@@
 Note that these spurious associations only occur when testing each HLA allele individually.
-In a regression model where all HLA alleles of a sample are evaluated at once, the marginal poster distributions for correlated predictors become broader to accurately reflect that the model is unable to discern which of these predictors explain the observed association with a replacement.
+In a regression model where all HLA alleles of a sample are evaluated at once, the marginal posterior distributions for correlated predictors become broader to accurately reflect that the model is unable to discern which of these predictors explain the observed association with a replacement.
 
@@
 
-\hl{DaHo: Wiederholung} One of the recent advances in Bayesian inference is the development of special prior distributions which can be used to give models additional useful properties.
-An example of these special prior distributions are the so-called sparsity-inducing priors, which can be used to convey the assumption that most of the predictors in a regression model are not associated with the outcome and are therefore 0.
-
 In the context of identifying HLA associations we expect most of the HLA coefficients to be very close to 0, as a given alignment position is typically only restricted by none or few HLA alleles.
 This is important prior knowledge that should be incorporated in the model.
-Adding this sparsity assumption can lead to better inferences, because the uncertainty of the non-relevant HLA regression coefficients does not propagate further and therefore the model is able to better discern signal from noise.
 
-There are a variety of different sparsity-promoting priors with slightly different properties.
+For Bayesian models, there are a variety of different sparsity-promoting priors with slightly different properties.
 They share the common structure of placing most probability mass very close to 0, with large to tails to accommodate the non-zero coefficients.
@@
 \end{equation},
-
 where \(\beta_{jk}\) are the regression coefficients, \(\tau\) is the so-called global shrinkage parameter, \(\lambda_{jk}\) are the so-called local shrinkage parameters and \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution.
@@
 
-The improvement of the logistic regression model with the horseshoe prior is a considerable improvement over a standard logistic regression model  (section~\ref{sec:results}).
-In our benchmark, the addition of the horseshoe prior alone shows to be more effective than a model withoutt horseshoe prior but with phylogeny included.
+Supplying information about the expected degree of sparsity with the horseshoe prior is a considerable improvement over a standard logistic regression model  (section~\ref{sec:results}).
 
@@
   \begin{aligned}
-    y_{ik} &\sim \text{bernoulli}(\theta_{ik})\\
+   y_{ik} &\sim \text{bernoulli}(\theta_{ik}) &\\
    \theta_{ik} &=
-        \text{inv\_logit}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi))\\
-        \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{\text{HLA}_{jk}}) \\
-    \beta_{0_{k}} &\sim \text{Normal}(0, 100^2)\\
-    \gamma_{k} &\sim \text{Normal}(\mu_{\text{phy}}, \sigma_{\text{phy}}^2)\\
-    \mu_{\text{phy}} &\sim \text{Normal}(1, 1)\\
-    \sigma_{\text{phy}} &\sim \text{Normal}^{+}(0, 0.5)\\
-    \beta_{\text{HLA}_{jk}} &\sim \text{Normal}(0, \tau_{k}^{2}\tilde{\lambda}^{2}_{jk})\\
-    \tilde{\lambda}^{2}_{jk} &= \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau_{k}^{2}\lambda_{jk}^{2}} \\
-    c_{k}^2 &\sim \text{Inv-Gamma}(3.5, 3.5)\\
-    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}^2\text{exp}(Z_{jk}\beta_{\text{epi}}))\\
-    \tau_{k} &\sim \text{Cauchy}^{+}(0, \tau_{0k})\\
-    \tau_{0k} &= \frac{10}{D-10}\frac{2}{\sqrt{N}}\\
+        \text{inv\_logit}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi)) & \\
+        \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{\text{HLA}_{jk}}) & \\
+    \beta_{0_{k}} &\sim \text{Normal}(0, 100^2) & \text{(*)}\\
+   \gamma_{k} &\sim \text{Normal}(\mu_{\text{phy}}, \sigma_{\text{phy}}^2) & \\
+    \mu_{\text{phy}} &\sim \text{Normal}(1, 1) & \text{(*)} \\
+   \sigma_{\text{phy}} &\sim \text{Normal}^{+}(0, 0.5) & \text{(*)} \\
+   \beta_{\text{HLA}_{jk}} &\sim \text{Normal}(0, \tau_{k}^{2}\tilde{\lambda}^{2}_{jk}) & \\
+   \tilde{\lambda}^{2}_{jk} &= \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau_{k}^{2}\lambda_{jk}^{2}} & \\
+    c_{k}^2 &\sim \text{Inv-Gamma}(3.5, 3.5) & \text{(*)} \\
+   \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}^2\text{exp}(Z_{jk}\beta_{\text{epi}})) & \\
+   \tau_{k} &\sim \text{Cauchy}^{+}(0, \tau_{0k}) & \text{(*)} \\
+   \tau_{0k} &= \frac{10}{D-10}\frac{2}{\sqrt{N}} & \\
   \end{aligned}
@@
 
-Prior distributions are labeled with an asterisk \hl{DaHo: * fehlen} in equation~\ref{eq:full}.
+Prior distributions are labeled with an asterisk in equation~\ref{eq:full}.
 They are meant to be weakly informative, which means that the goal is to provide some modest amount of shrinkage while still allowing for large absolute values given enough support by the observed data.
@@
 The value of \(\tau_{0k}\) is chosen to imply 10 effective non-zero HLA regression coefficients per replacement.
-The rationale behind this parameterization is again outlined in\cite{Piironen2017}.
+The rationale behind this parameterization is again outlined in~\cite{Piironen2017}.
 The value of 10 is approximately based on available HIV epitope maps.
@@
 \begin{itemize}
-  \item A large HIV dataset consisting of X sequences, that were kindly provided by X and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}.
+  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by X and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}.
+\hl{DaHa: Wer ist für diese Sequenzen der Ansprechpartner?}
   \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\cite{Roomp2006}.
   \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm.
-  \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf
-  \item A set of 41 HIV sequences that were kindly provided by Rongge Yang-
+  \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf.
+  \item A set of 41 HIV sequences that were kindly provided by Rongge Yang.
 \end{itemize}
@@
   \item RAxML-NG~\cite{Kozlov2019} version 1.0.0 was used to generate a maximum likelihood phylogenetic tree for each gene/protein using the \texttt{--model GTR+G+I} option with all other parameters set to default values.
+When available, we preferred to use RNA or DNA sequences over amino acid sequences for this step.
 \end{enumerate}
@@
 
-Performing leave-one-out cross-validation has historically been challenging to apply for Bayesian models, because fitting a model once for each left out data-point is often infeasible because of computational constrains.
+Leave-one-out cross-validation has historically been challenging to apply for Bayesian models, because fitting a model once for each left out data-point is often infeasible because of computational constrains.
 Recently, Pareto-smoothed importance sampling has gained popularity to perform approximate leave-one-out cross-validation for Bayesian models.
-This method  approximates the leave-one-out posterior by re-weighting existing draws from the posterior distribution and has the advantage of allowing to perform leave-one-out cross-validation by only fitting the model once.
+This method  approximates the leave-one-out posterior by re-weighting existing draws from the posterior distribution and has the advantage of requiring to fit the model only once.
 Additionally, the method provides good diagnostics to identify cases where the approximation is unreliable~\cite{Vehtari2016}.
@@
 All differences in elpd are larger than several times the estimated standard error (column \(\text{se}_\text{diff}\)), indicating that models which account for more information are expected to have a better predictive performance.}
+  \vspace{0.5cm}
   \begin{tabular}{l|r|rll}
@@
 
-We had the opportunity to test our model against a set of associations identified by Fisher's exact test that turned out to be false positives \hl{DaHo: diesen Abschnitt  glattziehen}.
-A comparison with a preliminary version of HAMdetector have been published recently.
-Here we provide a comparison between results from Fisher's exact test~\cite{Budeus2016} and the mature version of HAMdetector used in this work.
+We had the opportunity to test our model against a set of associations identified by Fisher's exact test where some of them turned out to be false positives~\cite{Karimzadeh2019}.
+This data is rare because negative results are usually not published in the literature, but provide a useful way to access model performance.
+A comparison with a preliminary version of HAMdetector was published recently.
+Here we provide a comparison with results from Fisher's exact test~\cite{Budeus2016} and the mature version of HAMdetector used in this work.
 
-On the HDV dataset, Fisher's exact test was applied to a set of HLA annotated sequences and statistically significant results were reported.
-The identified positions were experimentally validated and 9 associations turned out to be false-positives.
-This kind of data is rate because negative results are usually not published in the literature.
-We ran HAMdetector on the same dataset, the posterior probabilities are shown in table \ref{tab:false-positives}.
-They denote the probability that the corresponding regression coefficient is positive.
-Associations with strong support have a posterior probability close to 1, associations with no support a probability close to 0.5 (because the marginal posterior distribution is centered around 0 and allocated half of the probability mass to values greater than 0).
-The table shows that most of the false-positives in Fisher's exact test have low posterior probabilities in our model, which means that these alleles are correctly identified as not being associated with the replacement.
-The pairs of HLA alleles and replacements that were experimentally validated have the highest posterior probabilities, with the exception of P89T - HLA-B*37, which does not have strong support in our model.
+Fisher's exact test and  HAMdetector were applied to the same data (table~\ref{tab:false-positives}).
+The posterior probabilities given by HAmdetector denote the probability that the corresponding regression coefficient is positive.
+Associations with strong support in our model have a posterior probability close to 1, associations with no support a probability close to 0.5 (corresponding to a regression coefficient centered around 0).
 
-\begin{table}[]
+The results show low posterior probabilities for most of the false-positives and high posterior probabilities for most replacements that could be experimentally validated in IFN gamma production assays.
+One important exception is P89T~-~HLA-B*37, which was experimentally validated but does not have strong support in our model.
+
+\begin{table}[h!]
+  \caption{List of HLA associated mutations that were identified using Fisher's exact test (FET).
+Experimental validation showed that 9 out of 15 replacements could not be experimentally validated (column “confirmed“).
+The data were re-analyzed using HAMdetector.
+Except for P89T, all validated replacements have high posterior probabilities, while the replacements that could not be experimentally validated have lower evidence of being HLA associated.
+Note that all replacement - HLA allele pairs had low p-values using Fisher's exact test.}
+  \vspace{0.5cm}
   \begin{tabular}{c|c|c|c|c}
-  \label{tab:false-positives}
-  \thead{replacement} & \thead{allele}  & \thead{p-value \\ (Fisher's exact test)} & \thead{post. prob. \\ (HAMdetector)} & \thead{confirmed} \\
+  \thead{replacement} & \thead{allele}  & \thead{p-value \\ (FET)} & \thead{post. prob. \\ (HAMdetector)} & \thead{confirmed} \\
   \hline
@@
   \end{tabular}
+  \label{tab:false-positives}
  \end{table}
@@
 
-\hl{DaHo: wir haben zwar mehrere Datensätze, gehen aber praktisch nicht auf die Resultate mit diesen Datensätzen ein.
+% \hl{DaHo: wir haben zwar mehrere Datensätze, gehen aber praktisch nicht auf die Resultate mit diesen Datensätzen ein.
 Wie besprochen sollte dieses Manuskript eher methodisch sein, d.h.~es ist ok, dass wir nicht *viel* auf die virologische/immunologische Seite eingehen.
@@
 %----------------------------------------------------------------------------------------
-
 \FloatBarrier

### 2021-01-27 docs/manuscript.tex
update manuscript

@@
 \begin{figure}
+  \includegraphics[width=1\linewidth]{plots/without_horseshoe.pdf}
+  \includegraphics[width=1\linewidth]{plots/with_horseshoe.pdf}
+  \caption{Marginal posterior distributions of a subset of regression coefficients for replacement 11D of the HIV integrase, based on the Arevir dataset.
+The graphic shows regression coefficients for two HLA alleles: For HLA-A*01, the data does not show evidence of an association with replacement 11D, whereas for HLA-B*44, the data is consistent with a strong association.
+The horseshoe prior has the effect of shrinking regression coefficients with weak evidence of an association with the replacement strongly towards 0, while leaving the coefficients that are far away from 0 mostly unshrunk.
+Shrinking the irrelevant regression coefficients to 0 has the advantage of reducing the standard error for the remaining coefficients, which can be observed by the much smaller width of the histogram of HLA-B*44 in the model with horseshoe prior.}
   \label{fig:horseshoe-comparison}
@@
 
-% \hl{DaHo: wir haben zwar mehrere Datensätze, gehen aber praktisch nicht auf die Resultate mit diesen Datensätzen ein.
+\hl{DaHo: wir haben zwar mehrere Datensätze, gehen aber praktisch nicht auf die Resultate mit diesen Datensätzen ein.
 Wie besprochen sollte dieses Manuskript eher methodisch sein, d.h.~es ist ok, dass wir nicht *viel* auf die virologische/immunologische Seite eingehen.

### 2021-01-28 docs/manuscript.tex
notizen zu face-to-face meeting

@@
 
-\subsection{A real-world example}
+\subsection{Generalization}
+no overfitting but improvements show for different data sets and different viruses
+supplementary figures
+
+\subsection{A real-world application}
 
@@
 Fisher's exact test and HAMdetector were applied to the same data (table~\ref{tab:false-positives}).
-The posterior probabilities given by HAmdetector denote the probability that the corresponding regression coefficient is positive.
+The posterior probabilities given by HAMdetector denote the probability that the corresponding regression coefficient is positive.
 Associations with strong support in our model have a posterior probability close to 1, associations with no support a probability close to 0.5 (corresponding to a regression coefficient centered around 0).
@@
 
+if quality is the same for all viruses: differences between immune/virus interactions become visible
+
+recombinant forms
+
+4-digit partial pooling
+
 We present HAMdetector, a Bayesian model to identify HLA-associated mutations in a multiple sequence alignment.

### 2021-02-09 docs/manuscript.tex
move methods to method section, add some points to discussion

@@
 
-Additionally, Bayesian statistics provides an accessible way to test models: By comparing data generated under the model's assumptions to the actually observed data, it is possible to identify important aspects of the dataset that the models fails to capture and subsequently improve the model until it is consistent with the observed data~\cite{Gabry2019}.
-
 In the context of identifying HLA-associated mutations we propose to improve existing methods by the following additions, which can be broadly divided into  additional information and model structure:
@@
 Second generational tools like MHCFlurry 2.0~\cite{ODonnell2020} use both binding affinity and antigen processing data to improve epitope prediction.
-For our tool HAMdetector, we use the output of MHCFlurry to benefit from this binding affinity and antigen processing data in order to predict HLA-associated mutations.
+For our tool HAMdetector, we use the output of MHCFlurry to benefit from this data in order to predict HLA-associated mutations.
 
@@
 
-  Recent advantages in Bayesian inference include so-called sparsity-promoting priors.
+  Recent advantages in Bayesian inference include so-called sparsity-promoting priors~\cite{Piironen2017}.
 Sparsity-promoting priors convey the a priori expectation that most coefficients in a regression model are close to 0.
@@
 when using the fraction of identified escapes that lie in known HLA epitopes as a benchmark.
+
+  \subsection{Model diagnostics}
+
+  \textit{Posterior predictive checks}
+
+  Additionally, Bayesian statistics provides an accessible way to test models: By comparing data generated under the model's assumptions to the actually observed data, it is possible to identify important aspects of the dataset that the models fails to capture and subsequently improve the model until it is consistent with the observed data~\cite{Gabry2019}.
+
+  Here, we use these so-called posterior predictive checks to compare the calibration of the predicted probabilities of observing the replacement to the actual observed data.
+
+  \textit{Convergence diagnostics}
+
+  We use Stan (version 2.23)~\cite{Stan2021} to specify and fit our Bayesian models.
+  Stan provides an implementation of a dynamic Hamilitonian Monte Carlo sampler, which is a powerful technique to approximate expectations with respect to the posterior distribution~\cite{Betancourt2017}.
+One advantage of using such methods is that they provide sensitive diagnostics to identify a wide range of possible sampling issues.
+  We use the split-\(\hat{R}\) convergence diagnostic~\cite{Gelman1992} to identify convergence issues.
+We require a value of \(\hat{R}\) below 1.01 for all model parameters.
+Additionally, we require that the effective sample size \(N_{\text{eff}}\) is above 500 for all model parameters and that sampling occurs without any divergent transitions.
+A detailed explanation of these quantities is provided in the Stan reference manual.
+
+  \textit{Leave-one-out cross-validation}
+
+  In addition to testing our model on real-world data, we also performed leave-one-out cross-validation to examine the relative importance of the additions of epitope prediction, sparsity and phylogeny with respect to predictive performance.
+  For this, we create several variations of the model that are all special cases of the full HAMdetector model.
+  By comparing the model performance of these variations, we can gain insight on the relative importance of each aspect of the model.
+
+  Leave-one-out cross-validation has historically been challenging to apply for Bayesian models, because fitting a model once for each left out data-point is often infeasible because of computational constrains.
+  Recently, Pareto-smoothed importance sampling has gained popularity to perform approximate leave-one-out cross-validation for Bayesian models.
+This method  approximates the leave-one-out posterior by re-weighting existing draws from the posterior distribution and has the advantage of requiring to fit the model only once.
+Additionally, the method provides good diagnostics to identify cases where the approximation is unreliable~\cite{Vehtari2016}.
+
+  Several utility (or cost) functions are used in practice to compare the predictive performance of models.
+For frequentist approaches, classification accuracy or receiver operator characteristic curves (ROC curves) are most common.
+  For Bayesian models, expected log-predictive density (elpd) is used most frequently.
+
+  elpd is defined as \(\sum_{i=1}^{n}\text{log}(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))\), i.e.
+the average log predictive density of the observed data points based on the leave-one-out posterior distributions.
+  This has the advantage over other performance measures like classification accuracy, that it not only takes the location of the predictive distribution into account (i.e.
+the number of correct predictions), but also the width (i.e.
+how confident the model is in its predictions).
+  elpd cannot be computed for frequentist models because the parameters of the model are viewed as fixed values.
 
@@
 \begin{itemize}
-  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by X and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}.
-\hl{DaHa: Wer ist für diese Sequenzen der Ansprechpartner?}
+  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by the Zabrina Brumme and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}.
   \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\cite{Roomp2006}.
@@
 
-\hl{DaHo: $\hat{R}$ und $n_{\mathrm{eff}}$ sollten in den Methods eingeführt werden}
-For all the Bayesian models, Markov Chain Monte Carlo sampling was done using Stan~\cite{Stan2021}.
  All model fits showed no signs of inference issues.
-In total, 4 different chains with 1000 samples each were drawn from the posterior.
-The effective sample size for all model parameters exceeded 500, R-hat convergence diagnostic values were below 1.01 in all cases.
+In total, samples were drawn from 4 different chains with 1000 iterations each.
+The effective sample size exceeded 500 for all model parameters, \(\hat{R}\) convergence diagnostic values were below 1.01 in all cases.
 
 \subsection*{Posterior predictive checks}
-\hl{DaHo: Sollte in den Methods eingeführt werden}
-Bayesian modelling allows a simple yet effective strategy to test models: By simulating data under the model's assumptions and then comparing simulated and observed data, model misspecifications can be identified and highlight possible model improvements.
 
-Here, we compare the calibration of the posterior predictive probability of observing the replacement to the actual observed data using calibration plots.
-The plots are similar to the one showed in figure~\ref{fig:phylogeny-calibration} but this time do not evaluate the probabilities based on the phylogenetic model alone, but show the calibration of the posterior predictive probabilities based on the full model.
+Figure~\ref{fig:calibration} compares the calibration of the posterior predictive probability of observing the replacement to the actual observed data for the Core protein of the HBV dataset.
+Plots for the other models and datasets are shown in the supplementary.
 
-An example calibration plot of the HAMdetector model on the Core protein sequence data of the HBV dataset is shown in figure~\ref{fig:calibration}.
-Plots for the other models and datasets are shown in the supplementary.
+The plots are similar to the one showed in figure~\ref{fig:phylogeny-calibration}, however, they do not evaluate the probabilities based on the phylogenetic model alone, but instead show the calibration of the posterior predictive probabilities based on the full model.
 
@@
 
-All models show a strong enrichment of HLA-associated mutations compared to a randomly permutated list of HLA allele - replacement pairs (gray lines).
+All models show a strong enrichment of HLA-associated mutations compared to a randomly permuted list of HLA allele - replacement pairs (gray lines).
 Fisher's exact test performs about as well as the simple logistic regression model, as from a statistical perspective, they are both similar.
@@
 \subsection{Leave-one-out cross-validation}
-\hl{DaHo: LOO auch in Methods einführen}
-In addition to testing the model against experimentally validated epitopes, we also performed leave-one-out cross-validation.
-While the comparison against known epitopes compares inferences of the regression coefficients, leave-one-out cross-validation compares different models in regard to their predictive performance, i.e.
-how well they can predict the presence or absence of a replacement in a sequence based on the available data.
-
-As the compared models are all special cases of the complete model, we can use cross-validation to gain insight on the relative importance of each aspect of the model.
-
-Leave-one-out cross-validation has historically been challenging to apply for Bayesian models, because fitting a model once for each left out data-point is often infeasible because of computational constrains.
-Recently, Pareto-smoothed importance sampling has gained popularity to perform approximate leave-one-out cross-validation for Bayesian models.
-This method  approximates the leave-one-out posterior by re-weighting existing draws from the posterior distribution and has the advantage of requiring to fit the model only once.
-Additionally, the method provides good diagnostics to identify cases where the approximation is unreliable~\cite{Vehtari2016}.
-
-Several utility (or cost) functions are used in practice to compare the predictive performance of models.
-For frequentist approaches, classification accuracy or receiver operator characteristic curves (ROC curves) are most common.
-For Bayesian models, expected log-predictive density (elpd) is used most frequently.
-
-elpd is defined as \(\sum_{i=1}^{n}\text{log}(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))\), i.e.
-the average log predictive density of the observed data points based on the leave-one-out posterior distributions.
-This has the advantage over other performance measures like classification accuracy, that it not only takes the location of the predictive distribution into account (i.e.
-the number of correct predictions), but also the width (i.e.
-how confident the model is in its predictions).
-elpd cannot be computed for frequentist models because the parameters of the model are viewed as fixed values.
 
@@
 Predictive accuracy increases as the models include more and more information.
-Note that this is not due to overfitting, because leave-one-out cross-validation approximates the model performance on unseen data.
-Additional tables are shown in the supplementary, the results are consistent across all datasets.
-
 Note that the model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any additional external data.
@@
 \subsection{Generalization}
-no overfitting but improvements show for different data sets and different viruses
-supplementary figures
+
+The models were applied on sequence data from a variety of different viruses and show a considerable improvement over currently applied methods that are consistent across datasets.
+These results are not due to overfitting, because leave-one-out cross-validation approximates the model performance on unseen data.
 
@@
 
-The results show low posterior probabilities for most of the false-positives and high posterior probabilities for most replacements that could be experimentally validated in IFN gamma production assays.
+The results show low posterior probabilities for most of the false-positives and high posterior probabilities for most replacements that could be experimentally validated in IFN-\(\gamma\) production assays.
 One important exception is P89T~-~HLA-B*37, which was experimentally validated but does not have strong support in our model.
@@
 
-if quality is the same for all viruses: differences between immune/virus interactions become visible
-
-recombinant forms
-
-4-digit partial pooling
-
 We present HAMdetector, a Bayesian model to identify HLA-associated mutations in a multiple sequence alignment.
@@
 
-\hl{DaHo: wir haben zwar mehrere Datensätze, gehen aber praktisch nicht auf die Resultate mit diesen Datensätzen ein.
-Wie besprochen sollte dieses Manuskript eher methodisch sein, d.h.~es ist ok, dass wir nicht *viel* auf die virologische/immunologische Seite eingehen.
-Aber, dass wir überhaupt nichts zu den Resultaten sagen, ist etwas zu wenig.
-Z.B.
-könnte man die Vielfalt der Datensätze nutzen um den Punkt zu machen, dass die Verbesserungen nicht von den Details des Virus abhängen.
-Vielleicht fällt Dir ja noch was dazu ein.}
+If the quality of the prediction is the same for all viruses, as it appears to be in our results, this method can be used to probe the frequency of HLA immune escape across different viruses.
+
+In further research we plan to use HAMdetector to study circulating recombinant forms in HIV.
+One possible explanation for the occurrence of these recombinant forms is improved fitness by evading several immune epitopes at once.
+
+While our model is an improvement over existing methods, a further addition that we plan to implement is partial pooling across 4-digit HLA alleles.
+It has recently been appreciated that binding specificities can vary drastically across  HLA alleles from the same allele group.
+For predicting HLA associated mutations, there has consequently been a shift to use 4-digit resolution data whenever available.
+
+This is not without downsides however, because overall, binding specificities are more similar for alleles in the same group, and, therefore, treating them as completely separate might unnecessarily fragment the available data.
+
+In Bayesian statistics, it is not required to chose one extreme or the other, With partial pooling, estimates from the same allele group do share some information across each other, but they are also allowed to vary if necessary.
+Therefore, implementing partial pooling might be another possibility to further increase the reliability of computational methods to study HLA escape.
 

### 2021-02-13 docs/manuscript.tex
begonnen mit der Lektüre, Vorschläge für Revision
k Please enter the commit message for your changes. Lines starting

@@
 \Keywords{human leukocyte antigen system, HLA, multiple sequence alignment, escape mutations, 
-  viral escape, Bayesian inference, sparsity, horseshoe, epitope prediction} % Keywords - if you don't want any simply remove all the text between the curly brackets
+  viral escape, Bayesian inference, sparsity, horseshoe prior, epitope prediction} % Keywords - if you don't want any simply remove all the text between the curly brackets
 \newcommand{\keywordname}{Keywords} % Defines the keywords heading name
@@
   \textbf{Motivation} \\
-  The human leukocyte antigen system (HLA) presents viral peptide fragments in MHC I protein-peptide complexes on cell surfaces and in this way alerts CD8\textsuperscript{+} cytotoxic T-Lymphocytes (CTLs), a key component of adaptive immunity.
+  A key process in anti-viral adaptive immunity is that the human leukocyte antigen system (HLA) presents viral peptide fragments in MHC I protein-peptide complexes on cell surfaces and in this way alerts CD8\textsuperscript{+} cytotoxic T-Lymphocytes (CTLs).
 This pathway exerts a strong selection pressure on viruses, favoring viral mutants that escape recognition by the HLA/CTL system, e.g.\ by point mutations that decrease binding of viral peptides to MHC I.
@@
 Under these circumstances, one way to optimize HAM detection is to integrate all relevant information in a coherent model.
-Bayesian inference offers a principled approaches to achieve this.\\
+Bayesian inference offers a principled approach to achieve this.\\
 
   \textbf{Results} \\
-  Here, we present a Bayesian regression model that is easily extensible to include information from different sources, and that makes use of recent advances in Bayesian inference, e.g.\ by using a sparsifying prior.
-We show that this integrative approach improves performance in HAM detection considerably over state-of-the-art methods.
+Here, we present a new regression model for the detection of HAMs.
+As we choose a Bayesian approach we can include the novel sparsity-inducing priors, and we obtain easily interpretable quantitative information on HAM candidates.
+The basic model can be extended to include prior information relevant to HAM detection, which we demonstrate by integrating predictions of epitope affinities to MHC I, predictions of epitope peptide processing, and computation of phylogenetic background.
+This integrative method improves performance in HAM detection considerably over state-of-the-art methods.
 \\
@@
 
-One way of how the human immune system recognizes intracellular viral infections is through the human leukocyte antigen system~\cite{Germain1994}: In cells with active protein biosynthesis, proteins are continuously synthesized and also degraded by proteasomal degradation, which cleaves proteins into peptides~\cite{Goldberg2002}.
+One way of how the human immune system recognizes viral infections is through the human leukocyte antigen system~\cite{Germain1994}: Proteins, including viral proteins, in cells are often degraded in proteasomes to peptide fragments~\cite{Goldberg2002}.
+A small subset of these peptides is presented on the cell surface by MHC I receptors, encoded by HLA genes (in humans HLA-A, -B, -C).
+HLA genes are very polymorphic with more than 20,000 known alleles in the human population~\cite{Robinson2014}.
+The resulting gene products differ in their binding properties, which means that cells from different individuals in general present different peptides on the cell surface.
 
-A small subset of these peptides is presented on the cell surface by receptors called MHC class I encoded by HLA genes, e.g.\ in humanc HLA-A, -B, and -C.
-HLA genes are very polymorphic between human individuals, with more than 20,000 different described HLA alleles~\cite{Robinson2014}.
-The resulting gene products differ in their binding properties, which means that cells from different individuals present a highly diverse set of peptides on the cell surface.
-
-Cytotoxic T-Lymphocytes (CTLs) are selected during maturation to only weakly bind to MHC I protein-peptide complexes if the peptide originates from proteins of the self-proteome, but might be able to strongly bind to MHC I protein-peptide complexes with peptides generated from viral proteins~\cite{Murata2007}.
-Upon activation, T cells induce cytolytic activity and recruit other immune cells~\cite{Harty2000}.
+Cytotoxic T-Lymphocytes (CTLs) are selected during maturation to only weakly bind to complexes of MHC I with self-peptides, but CTLs can bind strongly to complexes of MHC I with peptides from viral proteins~\cite{Murata2007}.
+Upon activation, CTLs become cytotoxic and recruit other immune cells so that infected cells can be eliminated~\cite{Harty2000}.
 
@@
 In this way, the HLA system exerts strong selection pressure towards virus variants that escape T cell recognition~\cite{Borrow1997}.
-Such variants could, for example, carry mutations that result in reduced binding of immunogenic peptides to MHC I or to T cell receptors, or that alter the processing and binding so that peptides are no longer presented at all~\cite{Yewdell2002}.
+Such variants could, for example, carry mutations that result in reduced binding of viral peptides to MHC I or T cell receptors, or that alter peptide processing so that peptides are no longer presented~\cite{Yewdell2002}.
 
-The evolutionary events are complex and occur not only on the level of individuals, where a virus adapts to specific features of the host, but also on the population level, for example because HLA alleles differ in their frequency across geographic regions~\cite{Kawashima2009}.
-Upon transmission to a new host with different HLA alleles, HLA escape mutations can revert to their wild type, as they are often associated with a reduction in viral replication capacity~\cite{Matthews2008}.
-If the latter is not the case, such mutations may not revert but be retained in the viral genome~\cite{Kawashima2009}.
+HLA diversity drives viral evolution in individuals where a virus adapts to specific host features, and in populations and regions that in general differ in HLA frequencies~\cite{Kawashima2009}.
+Upon transmission to a new host with different HLA alleles, HLA escape mutations may revert, as they are often associated with a reduction in viral replication capacity~\cite{Matthews2008}, but they also may be irreversible~\cite{Kawashima2009}.
 
-Whether and how quickly a given escape mutation is selected in a host depends on several factors, such as the viral genomic background and diversity, the magnitude of the reduction in viral replicative capacity, the availability of compensatory mutations that recover fitness, and the strength of selection pressure~\cite{Kloverpris2016}.
+Whether and how quickly a given escape mutation is selected in a host depends, e.g., on the viral genomic background, the magnitude of the reduction in viral replication, the availability of compensatory mutations that recover fitness, or the strength of selection pressure~\cite{Kloverpris2016}.
 
-Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV~\cite{Alizon2011}\cite{Allen2005}\cite{Rousseau2008}\cite{Lumley2018}.
+Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV~\cite{Alizon2011, Allen2005, Rousseau2008, Lumley2018}.
 This includes the development of treatments and vaccines that rely on effective immune responses.
@@
 There are several experimental methods available to study HLA escape~\cite{Timm2004}.
-However, these methods are relatively slow and costly, especially for screening purposes.
-Therefore, a promising approach is to combine viral genome sequencing, host HLA determination, computational identification by statistical association analysis, and targeted experimental validation~\cite{Carlson2012}.
+\hl{DaHo: use key references, e.g.
+those that have introduced a technique; alternatively: cite review like articles that give an overview of several experimental methods} However, these methods are relatively slow and costly, especially for screening purposes.
+A promising approach that makes efficient use of frequently available data is to combine viral genome sequencing, host HLA determination, computational identification by statistical association analysis, and targeted experimental validation~\cite{Carlson2012}.
 
-As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface, escape mutations are often HLA-allele specific and can therefore be detected as HLA-allele dependent footprints in sequence alignments of viral proteins~\cite{Moore2002}: At certain alignment positions, a replacement might be more frequently observed in sequences from hosts with a specific HLA allele than in  sequences from hosts without that HLA allele.
-By quantifying this difference for all replacement and HLA allele pairs, it is possible to identify replacements that are enriched in sequences coming from hosts with certain HLA alleles, and thus are likely to be HLA escape mutations.
+As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface, escape mutations are often HLA-allele specific and can therefore be detected as HLA-allele dependent footprints in sequence alignments of viral proteins~\cite{Moore2002}: At certain alignment positions, a replacement corresponding to an escape mutation is often enriched in sequences from hosts with a specific HLA allele.
 
-One way of quantifying this enrichment is Fisher's exact test~\cite{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(\neg H\)), (\(\neg R_{i}\), \(H\)) and (\(\neg R_{i}\), \(H\)), where \(\neg R_{i}\) denotes any replacement except \(R_{i}\), and \(\neg H\) denotes any HLA allele except \(H\).
+One way of quantifying this enrichment is Fisher's exact test~\cite{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(\neg H\)), (\(\neg R_{i}\), \(H\)) and (\(\neg R_{i}\), \(\neg H\)), where \(\neg R_{i}\) denotes any replacement except \(R_{i}\), and \(\neg H\) denotes any HLA allele except \(H\).
 
@@
 
-Fisher's exact test has the advantage of being easy to apply~\cite{Budeus2016},  but also has several disadvantages~\cite{Carlson2008}.
+Fisher's exact test has the advantage of being fast and easy to apply~\cite{Budeus2016}, but it also has several disadvantages~\cite{Carlson2008}.
 The most striking one is that viral sequences share a common phylogenetic history, and,  therefore, treating sequences as independent and identically distributed samples may under- or overestimate effect sizes.
@@
 
-Another issue of applying Fisher's exact test is that HLA class I loci are located in proximity on chromosome 6 and are therefore in linkage disequilibrium, which means that HLA alleles are not inherited completely independent of each other, i.e.
-inheritance of one HLA allele correlates with inheritance of another HLA allele.
-When using a statistical method that tests each HLA allele individually without considering the whole set of alleles present, spurious associations might occur: If HLA allele \(H_{1}\) is associated with an amino acid replacement \(R\), but \(H_{1}\) is in linkage disequilibrium with another allele \(H_{2}\), this also means that we observe an association between replacement \(R\) and \(H_{2}\), even in absence of any underlying escape mechanism.
+Another issue with Fisher's exact test is the genomic proximity of human HLA class I loci (on chromosome 6) leading to linkage disequilibrium -- inheritance of HLA alleles can be correlated.
+Therefore spurious associations can occur if associations of replacements with individual HLA alleles is tested: if HLA allele \(H_1\) is associated with an amino acid replacement \(R\) because of immune escape, but \(H_1\) is in linkage disequilibrium with allele \(H_2\), this leads to an association of \(R\) and \(H_2\), even without being an escape mutation from $H_2$.
 
-Correlations can not only occur between HLA alleles, but also between replacements.
-This kind of codon covariation occurs for example in compensatory mutations that attenuate the negative impact of immune escape mutations.
-For instance, a compensatory mutation might lead to a conformational change in such a way that the mutated protein resembles the original wild type more closely.
-
-Carlson et al.
-\nolinebreak\cite{Carlson2008} developed a method called  Phylogenetic Dependency Network that accounts for phylogenetic bias,  HLA linkage disequilibrium, and codon covariation and is based on null hypothesis significance testing.
+Carlson et al.~\cite{Carlson2008} developed the Phylogenetic Dependency Network, a method that accounts for several of the aforementioned problems, in particular phylogenetic bias and  HLA linkage disequilibrium.
+However, it is based on null hypothesis significance testing.
 
@@
 
-In addition to the aforementioned biological reasons why Fisher's exact test is  not suited for the analysis of sequence data, there are also more fundamental statistical issues that universally occur when using p-values as a screening tool~\cite{Amrhein2017}:
-
-In the presence of small effect sizes and high variance between measurements, as  it is typically the case when working with biological data, statistically significant results can often be misleading and are likely to be in the wrong direction (a so-called type S error) or greatly overestimate an effect (a so-called type M error)~\cite{Gelman2014}.
-This problem has recently been widely appreciated in the literature in the context of the current “replication crisis“, which describes the circumstance that scientific claims with seemingly strong statistical evidence fail to replicate~\cite{Ioannidis2005}.
+There are fundamental statistical issues with p-values as screening tool~\cite{Amrhein2017}:
+with small effect sizes and high variance between measurements, as is often the case with biological data, statistically significant results can be misleading, have the wrong direction (type S error), or greatly overestimate an effect (type M error)~\cite{Gelman2014}.
+Such problems are more and more appreciated in the context of the current ``replication crisis'', which describes that scientific claims with seemingly strong statistical evidence fail to replicate~\cite{Ioannidis2005, Begley:2012, Baker2016Nature-reproducibility-crisis}.
 
@@
 When using p-values as a filter, it is therefore likely to obtain significant effects that are in fact not real.
-To circumvent this problem, a common strategy is to control the false discovery rate, which is the expected proportion of false positives~\cite{Benjamini1995}.
+To circumvent this problem, a common strategy is to control the false discovery rate~\cite{Benjamini1995}.
 These adjustment procedures have the problem that, when performing many of such comparisons, none but the very largest effects remain.
 
-Instead of performing many hypothesis tests and trying to adjust for them, we instead prefer to fit a single, multilevel model that contains all comparisons of interest.
-When using multilevel models, the problem of multiple comparisons can disappear entirely and  also yield more valid estimates~\cite{Gelman2012}.
+Instead of performing many hypothesis tests and trying to adjust for them, we prefer to fit a single, multilevel model that contains all comparisons of interest.
+When using multilevel models, the problem of multiple comparisons can disappear entirely and yield more valid estimates~\cite{Gelman2012}.
 
@@
 
-When possible, we choose to fit Bayesian models.
-By using prior information, adding problem-specific structure and partial pooling, the accuracy of estimates can often be noticeably improved~\cite{Gelman2010}.
-Prior information does not necessarily mean to use external data, even a rough idea about the expected magnitude of estimates is often surprisingly effective.
+\hl{DaHo: Da es ein Methoden-Paper ist, sollte dies der Hauptteil sein}
 
-In the context of identifying HLA-associated mutations we propose to improve existing methods by the following additions, which can be broadly divided into  additional information and model structure:
-\subsection*{Additional information}
+Our general approach is to fit a Bayesian model that captures relationships between host HLA alleles and replacements in viral proteomes.
+This approach is advantageous because it improves estimates by including prior information (even if only knowledge of effect magnitudes), adding problem-specific structure and partial pooling~\cite{Gelman2010}.
+
+Among the novel features of our method that markedly improve detection of HLA-associated mutations (HAMs) are the inclusion of additional relevant information and the use of sparsity-inducing priors, as described in the following.
+
+\subsection*{Inclusion of additional information}
 
@@
 
-    HLA-associated mutations are expected to lie more frequently in regions of known epitopes.
+\hl{DaHo: Mein Verständnis von Materials and Methods ist, dass hier die Methoden aufgeführt werden, z.B.
+die Methoden für die Berechnung von Affinitäten, siehe z.B.
+https://academic.oup.com/bioinformatics/article/36/21/5159/5874440}
+
+    HAMs are expected to lie more frequently in T-cell epitopes, i.e.\ the peptide fragments presented by MHC to T-cell receptors.
 There are vast experimental binding affinity data available of different peptides and MHC I molecule pairs, and there are well-established computational methods that use these data to extrapolate HLA binding for untested peptides.
@@
 
-\subsection*{Model structure}
-
-  \textit{Sparsity-inducing priors}
+\subsection*{Sparsity-inducing priors}
 
@@
 
-  It has been shown that using a sparsity-promoting prior when non-zero coefficients are sparse can drastically improve predictive performance, because the model is better able to differentiate between signal and noise.
+  It has been shown \hl{DaHo: where?} that using a sparsity-promoting prior when non-zero coefficients are sparse can drastically improve predictive performance, because the model is better able to differentiate between signal and noise.
 
@@
 
+  \subsection*{Data}
+ \hl{DaHo: das Material in Materials and Methods sind die Daten; die sollten hier kurz beschrieben werden, möglichst mit Referenzen}
+
   % \textit{Partial pooling of 4-digit HLA alleles}
@@
 \begin{align}
-  y_{ik} & \sim \text{bernoulli}(\theta_{ik}) \\
-  \theta_{ik} & = \text{inv\_logit}(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}}),
+  y_{ik} & \sim \text{Bernoulli}(\theta_{ik}) \\
+  \theta_{ik} & = \text{logit}^{-1}(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}}),
 \end{align}
@@
 \begin{aligned}
-  y_{ik} & \sim \text{bernoulli}(\theta_{ik}) \\
+  y_{ik} & \sim \text{Bernoulli}(\theta_{ik}) \\
   \theta_{ik} & = \text{inv\_logit}(\beta_{0_{k}} + \gamma~\text{logit}(P(y_{ik}=1|\Psi)) \\ \hspace{-6cm} &~~~~ + \sum_{k=1}^{D}X_{ik}\beta_{\text{HLA}_{jk}})
@@
   \begin{aligned}
-   y_{ik} &\sim \text{bernoulli}(\theta_{ik}) &\\
+   y_{ik} &\sim \text{Bernoulli}(\theta_{ik}) &\\
    \theta_{ik} &=
-        \text{inv\_logit}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi)) & \\
+        \text{logit}^{-1}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi)) & \\
         \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{\text{HLA}_{jk}}) & \\
@@
 \begin{itemize}
-  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by the Zabrina Brumme and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}.
+  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}.
   \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\cite{Roomp2006}.
@@
 The effective sample size exceeded 500 for all model parameters, \(\hat{R}\) convergence diagnostic values were below 1.01 in all cases.
+\hl{DaHo: citation Gelman + Rubin original paper or later versions if improved $\hat{R}$ have been used}
 
@@
 
-We present HAMdetector, a Bayesian model to identify HLA-associated mutations in a multiple sequence alignment.
+We have presented HAMdetector, a Bayesian model to identify HLA-associated mutations in a multiple sequence alignment.
 We propose to use models that include as much information as possible and have shown that including sparsity and epitope prediction software can achieve better performance than existing methods.

### 2021-02-14 docs/manuscript.tex
reordered text blocks, text requires substantial revision

@@
 
-As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface, escape mutations are often HLA-allele specific and can therefore be detected as HLA-allele dependent footprints in sequence alignments of viral proteins~\cite{Moore2002}: At certain alignment positions, a replacement corresponding to an escape mutation is often enriched in sequences from hosts with a specific HLA allele.
+As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface and thus on binding to the HLA encoded MHC I protein, escape mutations are usually HLA allele specific and can therefore be detected as HLA allele dependent ``footprints'' in sequence alignments of viral proteins  \cite{Moore2002}, i.e.\ as HLA associated mutations (HAM) or replacements enriched in sequences from hosts with a specific HLA allele.
 
@@
 
-Our general approach is to fit a Bayesian model that captures relationships between host HLA alleles and replacements in viral proteomes.
-This approach is advantageous because it improves estimates by including prior information (even if only knowledge of effect magnitudes), adding problem-specific structure and partial pooling~\cite{Gelman2010}.
+Our general approach is to fit a Bayesian regression model that captures relationships between host HLA alleles and replacements in viral proteomes.
+This Bayesian approach is advantageous because it allows use of: (1) prior information (even if only knowledge of effect magnitudes), (2) relevant additional information (phylogeny, epitope information), (3) a problem-specific structure, (4) partial pooling~\cite{Gelman2010}.
+With these features, HAM detection is improved markedly over the state-of-the-art.
 
-Among the novel features of our method that markedly improve detection of HLA-associated mutations (HAMs) are the inclusion of additional relevant information and the use of sparsity-inducing priors, as described in the following.
+\subsection{Model backbone}
 
-\subsection*{Inclusion of additional information}
-
-  \textit{Binding affinity prediction}
-
-\hl{DaHo: Mein Verständnis von Materials and Methods ist, dass hier die Methoden aufgeführt werden, z.B.
-die Methoden für die Berechnung von Affinitäten, siehe z.B.
-https://academic.oup.com/bioinformatics/article/36/21/5159/5874440}
-
-    HAMs are expected to lie more frequently in T-cell epitopes, i.e.\ the peptide fragments presented by MHC to T-cell receptors.
-There are vast experimental binding affinity data available of different peptides and MHC I molecule pairs, and there are well-established computational methods that use these data to extrapolate HLA binding for untested peptides.
-We show that, by including the outcome of these computational tools as input for a probabilistic model, the prediction of HLA-associated mutations can be improved.
-
-  \textit{Antigen processing prediction}
-
-    Similarly, there are also HLA-allele independent effects like antigen processing that influence presentation on MHC I.
-Second generational tools like MHCFlurry 2.0~\cite{ODonnell2020} use both binding affinity and antigen processing data to improve epitope prediction.
-For our tool HAMdetector, we use the output of MHCFlurry to benefit from this data in order to predict HLA-associated mutations.
-
-\subsection*{Sparsity-inducing priors}
-
-  Recent advantages in Bayesian inference include so-called sparsity-promoting priors~\cite{Piironen2017}.
-Sparsity-promoting priors convey the a priori expectation that most coefficients in a regression model are close to 0.
-This assumption also applies to HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes.
-If no epitope spanning a given alignment position is presented on the MHC I molecule, any association between a replacement and the respective HLA allele  is likely to be due to random variation alone (or HLA linkage disequilibrium).
-
-  It has been shown \hl{DaHo: where?} that using a sparsity-promoting prior when non-zero coefficients are sparse can drastically improve predictive performance, because the model is better able to differentiate between signal and noise.
-
-  We show that a simple logistic regression model with a sparsity-promoting prior on the regression coefficients for the HLA alleles alone already performs roughly on-par  with the much more elaborate Phylogenetic Dependency Network approach from Carlson et al.
-when using the fraction of identified escapes that lie in known HLA epitopes as a benchmark.
-
-  \subsection{Model diagnostics}
-
-  \textit{Posterior predictive checks}
-
-  Additionally, Bayesian statistics provides an accessible way to test models: By comparing data generated under the model's assumptions to the actually observed data, it is possible to identify important aspects of the dataset that the models fails to capture and subsequently improve the model until it is consistent with the observed data~\cite{Gabry2019}.
-
-  Here, we use these so-called posterior predictive checks to compare the calibration of the predicted probabilities of observing the replacement to the actual observed data.
-
-  \textit{Convergence diagnostics}
-
-  We use Stan (version 2.23)~\cite{Stan2021} to specify and fit our Bayesian models.
-  Stan provides an implementation of a dynamic Hamilitonian Monte Carlo sampler, which is a powerful technique to approximate expectations with respect to the posterior distribution~\cite{Betancourt2017}.
-One advantage of using such methods is that they provide sensitive diagnostics to identify a wide range of possible sampling issues.
-  We use the split-\(\hat{R}\) convergence diagnostic~\cite{Gelman1992} to identify convergence issues.
-We require a value of \(\hat{R}\) below 1.01 for all model parameters.
-Additionally, we require that the effective sample size \(N_{\text{eff}}\) is above 500 for all model parameters and that sampling occurs without any divergent transitions.
-A detailed explanation of these quantities is provided in the Stan reference manual.
-
-  \textit{Leave-one-out cross-validation}
-
-  In addition to testing our model on real-world data, we also performed leave-one-out cross-validation to examine the relative importance of the additions of epitope prediction, sparsity and phylogeny with respect to predictive performance.
-  For this, we create several variations of the model that are all special cases of the full HAMdetector model.
-  By comparing the model performance of these variations, we can gain insight on the relative importance of each aspect of the model.
-
-  Leave-one-out cross-validation has historically been challenging to apply for Bayesian models, because fitting a model once for each left out data-point is often infeasible because of computational constrains.
-  Recently, Pareto-smoothed importance sampling has gained popularity to perform approximate leave-one-out cross-validation for Bayesian models.
-This method  approximates the leave-one-out posterior by re-weighting existing draws from the posterior distribution and has the advantage of requiring to fit the model only once.
-Additionally, the method provides good diagnostics to identify cases where the approximation is unreliable~\cite{Vehtari2016}.
-
-  Several utility (or cost) functions are used in practice to compare the predictive performance of models.
-For frequentist approaches, classification accuracy or receiver operator characteristic curves (ROC curves) are most common.
-  For Bayesian models, expected log-predictive density (elpd) is used most frequently.
-
-  elpd is defined as \(\sum_{i=1}^{n}\text{log}(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))\), i.e.
-the average log predictive density of the observed data points based on the leave-one-out posterior distributions.
-  This has the advantage over other performance measures like classification accuracy, that it not only takes the location of the predictive distribution into account (i.e.
-the number of correct predictions), but also the width (i.e.
-how confident the model is in its predictions).
-  elpd cannot be computed for frequentist models because the parameters of the model are viewed as fixed values.
-
-  \subsection*{Data}
- \hl{DaHo: das Material in Materials and Methods sind die Daten; die sollten hier kurz beschrieben werden, möglichst mit Referenzen}
-
-  % \textit{Partial pooling of 4-digit HLA alleles}
-
-  % It has recently been appreciated that binding specificities can vary drastically across  HLA alleles from the same allele group, e.g.
-between HLA-B*51:01 and HLA-B*51:03.
-For predicting HLA associated mutations, there has consequently been a shift to use 4-digit resolution data whenever available.
-This is not without downsides however, because overall, binding specificities are more similar for alleles in the same group, and, therefore, treating them as completely separate might unnecessarily fragment the available data.
-
-  % In Bayesian statistics, it is not required to chose one extreme or the other: Instead of either treating all HLA alleles from the same allele group as identical or as completely separate, partial pooling of the estimates allows something in between: With partial pooling, estimates do share some information across each other, but they are also allowed to vary if necessary.
-
-  % In the context of our model this means that observing a relevant association between a replacement and an HLA allele also makes the model more inclined to estimate relevant associations between that replacement and all other alleles from that allele group.
-The degree of this partial pooling can be estimated from the data, so if we do observe strong similarity of HLA alleles in a given allele group the estimates are more influenced by each other than if they do not.
-
-\section{Implementation}
-
-\subsection*{Model backbone}
-
-We choose a logistic regression model as a backbone for two reasons: First, because it is easily extensible and secondly, because the coefficients are interpretable as expected increases on the log-odds scale, identical to the familiar interpretation of standard logistic regression models.
-We also choose to fit the model in a Bayesian fashion using Stan, mostly because it allows us to make use of the hierarchical structure in the data and the possibility to use posterior predictive checks as a way of model checking.
-Consider the following notation:
+We chose a logistic regression model as a backbone for two reasons: First, because it is easily extensible and secondly, because the coefficients are interpretable as expected increases on the log-odds scale, identical to the familiar interpretation of standard logistic regression models.
 
 \begin{align}
+\label{eq:backbone}
   y_{ik} & \sim \text{Bernoulli}(\theta_{ik}) \\
-  \theta_{ik} & = \text{logit}^{-1}(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}}),
+  \theta_{ik} & = \text{logistic}(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}}),
 \end{align}
 
-where \(y_{ik}\) is the binary encoded observation at sequence \(i\) for replacement \(k\), \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\), \(X_{ij}\) is 1 if sequence \(i\) is annotated with HLA allele \(j\) and 0 otherwise, \(\beta_{\text{HLA}_{jk}}\) is the HLA regression coefficients for HLA allele \(j\) at replacement \(k\), \(D\) is the number of HLA alleles in the dataset, and \(\text{inv\_logit}\) is the logistic link function.
+where \(y_{ik}\) is the binary encoded observation at sequence \(i\) for replacement \(k\), \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\), \(X_{ij}\) is 1 if sequence \(i\) is annotated with HLA allele \(j\) and 0 otherwise, \(\beta_{\text{HLA}_{jk}}\) is the HLA regression coefficients for HLA allele \(j\) at replacement \(k\), \(D\) is the number of HLA alleles in the dataset, and \(\text{inv\_logit}\) is the logistic inverse link function.
 Note that, if we observe N different amino acid states at a certain alignment position, each of those are considered in turn to be the “replacement”.
@@
 
-\subsection*{Including phylogeny}
+\subsection{Inclusion of additional information}
 
+On top of the paired data of sequences and HLA alleles modeled by the backbone (Eq~\ref{eq:backbone}), we extend the model to include further information of relevance to improve HAM detection, namely phylogenetic information, and predictions of epitope peptide processing and MHC I affinity, as described in the following.
+
+\subsubsection{Phylogeny}
 One problem that occurs when analyzing sequence data is that species share a common phylogenetic history.
@@
   y_{ik}  & \sim \text{Bernoulli}(\theta_{ik}) \\
-  \theta_{ik} & = \text{inv\_logit}(\beta_{0_{k}} + \gamma~\text{logit}(P(y_{ik}=1|\Psi)) \\ \hspace{-6cm} &~~~~ + \sum_{k=1}^{D}X_{ik}\beta_{\text{HLA}_{jk}})
+  \theta_{ik}  & =  \text{logistic}\Bigl(\beta_{0_{k}} + \gamma~\text{logit}\left(P(y_{ik}=1|\Psi)\right) \\
+  &\;\;\;\;\; + \sum_{k=1}^{D}X_{ik}\beta_{\text{HLA}_{jk}}\Bigr)
+  \label{eq:phylogeny}
 \end{aligned}
@@
 
-\subsection*{Including HLA linkage disequilibrium and codon covariation}
+\textit{Binding affinity prediction}
 
-The other two important confounding effects that are recognized in Carlson et al.~\cite{Carlson2008} are HLA linkage disequilibrium and codon covariation.
-HLA linkage disequilibrium describes the observation that HLA alleles that lie in proximity to each other on the chromosome are also more likely to be inherited together.
-When using a statistical method that evaluates each HLA allele individually, this may lead to spurious associations: If some HLA allele \(X\) is associated with replacement \(k\), and HLA allele \(Y\) often co-occurs with HLA allele \(X\), this means that we also identify an association between HLA allele \(Y\) and replacement \(k\).
+\hl{DaHo: Mein Verständnis von Materials and Methods ist, dass hier die Methoden aufgeführt werden, z.B.
+die Methoden für die Berechnung von Affinitäten, siehe z.B.
+https://academic.oup.com/bioinformatics/article/36/21/5159/5874440}
 
-This association is correct in a statistical sense, but if the purpose of the model is to identify sites under HLA-mediated selection pressure, these associations due to linkage disequilibrium are most likely unwanted.
+    HAMs are expected to lie more frequently in T-cell epitopes, i.e.\ the peptide fragments presented by MHC to T-cell receptors.
+There are vast experimental binding affinity data available of different peptides and MHC I molecule pairs, and there are well-established computational methods that use these data to extrapolate HLA binding for untested peptides.
+We show that, by including the outcome of these computational tools as input for a probabilistic model, the prediction of HLA-associated mutations can be improved.
 
-Note that these spurious associations only occur when testing each HLA allele individually.
-In a regression model where all HLA alleles of a sample are evaluated at once, the marginal posterior distributions for correlated predictors become broader to accurately reflect that the model is unable to discern which of these predictors explain the observed association with a replacement.
+  \textit{Antigen processing prediction}
 
-We do not currently account for codon covariation in our model, as we are interested in identifying all HLA associated mutations (which include for example compensatory mutations).
-The general structure of the model, in particular the sparsity assumption introduced in section~\ref{sec:sparsity}, would allow adding codon covariation by adding all other codons as additional predictors.
-This would generally not be possible in a frequentist setting because the predictor matrix would become singular and the inverse does not exist.
+    Similarly, there are also HLA-allele independent effects like antigen processing that influence presentation on MHC I.
+Second generational tools like MHCFlurry 2.0~\cite{ODonnell2020} use both binding affinity and antigen processing data to improve epitope prediction.
+For our tool HAMdetector, we use the output of MHCFlurry to benefit from this data in order to predict HLA-associated mutations.
 
-\subsection*{Including sparsity assumptions} \label{sec:sparsity}
-
-In the context of identifying HLA associations we expect most of the HLA coefficients to be very close to 0, as a given alignment position is typically only restricted by none or few HLA alleles.
-This is important prior knowledge that should be incorporated in the model.
-
-For Bayesian models, there are a variety of different sparsity-promoting priors with slightly different properties.
-They share the common structure of placing most probability mass very close to 0, with large to tails to accommodate the non-zero coefficients.
-For our model, we use a so-called horse prior~\cite{Carvalho2010}, which is defined as a scale-mixture of gaussians:
-
-\begin{equation}
-  \begin{aligned}
-    \beta_{jk} &\sim \text{Normal}(0, \tau^{2}\lambda^{2}_{jk}) \\
-    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, 1) \\
-    \tau &\sim \text{Cauchy}^{+}(0, \tau_{0})
-  \end{aligned}
-\end{equation},
-where \(\beta_{jk}\) are the regression coefficients, \(\tau\) is the so-called global shrinkage parameter, \(\lambda_{jk}\) are the so-called local shrinkage parameters and \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution.
-
-Intuitively speaking, the global shrinkage parameter \(\tau\) is typically very small and shrinks most of the regression coefficients to 0, whereas the local shrinkage parameters can occasionally be very large to allow some coefficients to escape that shrinkage.
-
-A useful addition to the horseshoe prior is given in Piironen et al.~\cite{Piironen2017}, which describes a way to choose the parameter governing the overall degree of sparsity \(\tau_{0}\) based on the expected number of non-zero coefficients.
-
-For our HLA model, we utilize an extension to the horseshoe prior called the regularized horseshoe prior~\cite{Piironen2017}, which also provides some regularization for the non-zero coefficients.
-This is particularly useful for logistic regression models, as some shrinkage helps to deal with issues of separability and collinearity that commonly occur when fitting logistic regression models.
-
-Figure~\ref{fig:horseshoe-comparison} shows a comparison of marginal posteriors for regression coefficients with and without horseshoe prior.
-
-Supplying information about the expected degree of sparsity with the horseshoe prior is a considerable improvement over a standard logistic regression model  (section~\ref{sec:results}).
-
-\begin{figure}
-  \includegraphics[width=1\linewidth]{plots/without_horseshoe.pdf}
-  \includegraphics[width=1\linewidth]{plots/with_horseshoe.pdf}
-  \caption{Marginal posterior distributions of a subset of regression coefficients for replacement 11D of the HIV integrase, based on the Arevir dataset.
-The graphic shows regression coefficients for two HLA alleles: For HLA-A*01, the data does not show evidence of an association with replacement 11D, whereas for HLA-B*44, the data is consistent with a strong association.
-The horseshoe prior has the effect of shrinking regression coefficients with weak evidence of an association with the replacement strongly towards 0, while leaving the coefficients that are far away from 0 mostly unshrunk.
-Shrinking the irrelevant regression coefficients to 0 has the advantage of reducing the standard error for the remaining coefficients, which can be observed by the much smaller width of the histogram of HLA-B*44 in the model with horseshoe prior.}
-  \label{fig:horseshoe-comparison}
-\end{figure}
-
-\subsection*{Including epitope prediction software}
+    \subsubsection{Epitope predictions}
 
@@
 
-\subsection*{Full model specification}
+\subsection{Sparsity-inducing priors}
+
+  Recent advantages in Bayesian inference include so-called sparsity-promoting priors~\cite{Piironen2017}.
+Sparsity-promoting priors convey the a priori expectation that most coefficients in a regression model are close to 0.
+This assumption also applies to HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes.
+If no epitope spanning a given alignment position is presented on the MHC I molecule, any association between a replacement and the respective HLA allele  is likely to be due to random variation alone (or HLA linkage disequilibrium).
+
+  It has been shown \hl{DaHo: where?} that using a sparsity-promoting prior when non-zero coefficients are sparse can drastically improve predictive performance, because the model is better able to differentiate between signal and noise.
+
+  We show that a simple logistic regression model with a sparsity-promoting prior on the regression coefficients for the HLA alleles alone already performs roughly on-par  with the much more elaborate Phylogenetic Dependency Network approach from Carlson et al.
+when using the fraction of identified escapes that lie in known HLA epitopes as a benchmark.
+
+In the context of identifying HLA associations we expect most of the HLA coefficients to be very close to 0, as a given alignment position is typically only restricted by none or few HLA alleles.
+This is important prior knowledge that should be incorporated in the model.
+
+For Bayesian models, there are a variety of different sparsity-promoting priors with slightly different properties.
+They share the common structure of placing most probability mass very close to 0, with large to tails to accommodate the non-zero coefficients.
+For our model, we use a so-called horse prior~\cite{Carvalho2010}, which is defined as a scale-mixture of gaussians:
+
+\begin{equation}
+  \begin{aligned}
+    \beta_{jk} &\sim \text{Normal}(0, \tau^{2}\lambda^{2}_{jk}) \\
+    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, 1) \\
+    \tau &\sim \text{Cauchy}^{+}(0, \tau_{0})
+  \end{aligned}
+\end{equation},
+where \(\beta_{jk}\) are the regression coefficients, \(\tau\) is the so-called global shrinkage parameter, \(\lambda_{jk}\) are the so-called local shrinkage parameters and \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution.
+
+Intuitively speaking, the global shrinkage parameter \(\tau\) is typically very small and shrinks most of the regression coefficients to 0, whereas the local shrinkage parameters can occasionally be very large to allow some coefficients to escape that shrinkage.
+
+A useful addition to the horseshoe prior is given in Piironen et al.~\cite{Piironen2017}, which describes a way to choose the parameter governing the overall degree of sparsity \(\tau_{0}\) based on the expected number of non-zero coefficients.
+
+For our HLA model, we utilize an extension to the horseshoe prior called the regularized horseshoe prior~\cite{Piironen2017}, which also provides some regularization for the non-zero coefficients.
+This is particularly useful for logistic regression models, as some shrinkage helps to deal with issues of separability and collinearity that commonly occur when fitting logistic regression models.
+
+\subsection{Full model specification}
 
@@
 
-All models were implemented in Stan~\cite{Stan2021}.
-The Stan code is available online in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
-A Julia~\cite{Bezanson2017} package is available to run the model on custom data.
-Due to restrictions of dependencies (MHCFlurry and RAxML-ng), HAMdetector is only available on Linux.
-We additionally provide Docker images to easily run the model on any operating system in a containerized environment, assuming Docker is installed.
-
-\subsection*{Prior justification}
+\subsubsection{Prior justification}
 
@@
 
-\section{Results} \label{sec:results}
+\subsubsection{Model implementation}
+All models were implemented in Stan~\cite{Stan2021}.
+The Stan code is available online in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
+A Julia~\cite{Bezanson2017} package is available \hl{DaHo: at ...} to run the model on custom data.
+Due to restrictions of dependencies (MHCFlurry and RAxML-ng), HAMdetector is only available on Linux.
+%We additionally provide Docker images to easily run the model on any operating system in a containerized environment, assuming Docker is installed.
 
-To show that our model works well in different real-world scenarios, we run versions of our model on several datasets:
+  \subsection{Model diagnostics}
+
+  \subsubsection{Posterior predictive checks}
+
+  Additionally, Bayesian statistics provides an accessible way to test models: By comparing data generated under the model's assumptions to the actually observed data, it is possible to identify important aspects of the dataset that the models fails to capture and subsequently improve the model until it is consistent with the observed data~\cite{Gabry2019}.
+
+  Here, we use these so-called posterior predictive checks to compare the calibration of the predicted probabilities of observing the replacement to the actual observed data.
+
+  \subsubsection{Convergence diagnostics}
+
+  We use Stan (version 2.23)~\cite{Stan2021} to specify and fit our Bayesian models.
+  Stan provides an implementation of a dynamic Hamilitonian Monte Carlo sampler, which is a powerful technique to approximate expectations with respect to the posterior distribution~\cite{Betancourt2017}.
+One advantage of using such methods is that they provide sensitive diagnostics to identify a wide range of possible sampling issues.
+  We use the split-\(\hat{R}\) convergence diagnostic~\cite{Gelman1992} to identify convergence issues.
+We require a value of \(\hat{R}\) below 1.01 for all model parameters.
+Additionally, we require that the effective sample size \(N_{\text{eff}}\) is above 500 for all model parameters and that sampling occurs without any divergent transitions.
+A detailed explanation of these quantities is provided in the Stan reference manual.
+
+  \subsubsection{Leave-one-out cross-validation}
+
+  In addition to testing our model on real-world data, we also performed leave-one-out cross-validation to examine the relative importance of the additions of epitope prediction, sparsity and phylogeny with respect to predictive performance.
+  For this, we create several variations of the model that are all special cases of the full HAMdetector model.
+  By comparing the model performance of these variations, we can gain insight on the relative importance of each aspect of the model.
+
+  Leave-one-out cross-validation has historically been challenging to apply for Bayesian models, because fitting a model once for each left out data-point is often infeasible because of computational constrains.
+  Recently, Pareto-smoothed importance sampling has gained popularity to perform approximate leave-one-out cross-validation for Bayesian models.
+This method  approximates the leave-one-out posterior by re-weighting existing draws from the posterior distribution and has the advantage of requiring to fit the model only once.
+Additionally, the method provides good diagnostics to identify cases where the approximation is unreliable~\cite{Vehtari2016}.
+
+  Several utility (or cost) functions are used in practice to compare the predictive performance of models.
+For frequentist approaches, classification accuracy or receiver operator characteristic curves (ROC curves) are most common.
+  For Bayesian models, expected log-predictive density (elpd) is used most frequently.
+
+  elpd is defined as \(\sum_{i=1}^{n}\text{log}(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))\), i.e.
+the average log predictive density of the observed data points based on the leave-one-out posterior distributions.
+  This has the advantage over other performance measures like classification accuracy, that it not only takes the location of the predictive distribution into account (i.e.
+the number of correct predictions), but also the width (i.e.
+how confident the model is in its predictions).
+  elpd cannot be computed for frequentist models because the parameters of the model are viewed as fixed values.
+
+  \subsection{Data and data preparation}
+ \hl{DaHo: das Material in Materials and Methods sind die Daten; die sollten hier kurz beschrieben werden, möglichst mit Referenzen}
+
+The model was fit to several data sets:
 
@@
   \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm.
-  \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf.
+  \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf. \hl{DaHo: bitte von Hadi geeignetes Zitat erfragen}
   \item A set of 41 HIV sequences that were kindly provided by Rongge Yang. 
 \end{itemize}
+
+\hl{DaHo: zu jedem Datensatz die Proteine nennen}
 
@@
   \item RAxML-NG~\cite{Kozlov2019} version 1.0.0 was used to generate a maximum likelihood phylogenetic tree for each gene/protein using the \texttt{--model GTR+G+I} option with all other parameters set to default values.
-When available, we preferred to use RNA or DNA sequences over amino acid sequences for this step.
+If available, we used RNA or DNA sequences for this step, rather than protein sequences.
 \end{enumerate}
+
+  % \textit{Partial pooling of 4-digit HLA alleles}
+
+  % It has recently been appreciated that binding specificities can vary drastically across  HLA alleles from the same allele group, e.g.
+between HLA-B*51:01 and HLA-B*51:03.
+For predicting HLA associated mutations, there has consequently been a shift to use 4-digit resolution data whenever available.
+This is not without downsides however, because overall, binding specificities are more similar for alleles in the same group, and, therefore, treating them as completely separate might unnecessarily fragment the available data.
+
+  % In Bayesian statistics, it is not required to chose one extreme or the other: Instead of either treating all HLA alleles from the same allele group as identical or as completely separate, partial pooling of the estimates allows something in between: With partial pooling, estimates do share some information across each other, but they are also allowed to vary if necessary.
+
+  % In the context of our model this means that observing a relevant association between a replacement and an HLA allele also makes the model more inclined to estimate relevant associations between that replacement and all other alleles from that allele group.
+The degree of this partial pooling can be estimated from the data, so if we do observe strong similarity of HLA alleles in a given allele group the estimates are more influenced by each other than if they do not.
+
+\section{Implementation}
+
+% \subsection{Including HLA linkage disequilibrium}
+
+% The other two important confounding effects that are recognized in Carlson et al.~\cite{Carlson2008} are HLA linkage disequilibrium and codon covariation.
+% HLA linkage disequilibrium describes the observation that HLA alleles that lie in proximity to each other on the chromosome are also more likely to be inherited together.
+When using a statistical method that evaluates each HLA allele individually, this may lead to spurious associations: If some HLA allele \(X\) is associated with replacement \(k\), and HLA allele \(Y\) often co-occurs with HLA allele \(X\), this means that we also identify an association between HLA allele \(Y\) and replacement \(k\).
+
+% This association is correct in a statistical sense, but if the purpose of the model is to identify sites under HLA-mediated selection pressure, these associations due to linkage disequilibrium are most likely unwanted.
+
+% Note that these spurious associations only occur when testing each HLA allele individually.
+In a regression model where all HLA alleles of a sample are evaluated at once, the marginal posterior distributions for correlated predictors become broader to accurately reflect that the model is unable to discern which of these predictors explain the observed association with a replacement.
+
+% We do not currently account for codon covariation in our model, as we are interested in identifying all HLA associated mutations (which include for example compensatory mutations).
+The general structure of the model, in particular the sparsity assumption introduced in section~\ref{sec:sparsity}, would allow adding codon covariation by adding all other codons as additional predictors.
+This would generally not be possible in a frequentist setting because the predictor matrix would become singular and the inverse does not exist.
+
+\section{Results} \label{sec:results}
 
@@
 
+Figure~\ref{fig:horseshoe-comparison} shows a comparison of marginal posteriors for regression coefficients with and without horseshoe prior.
+
+Supplying information about the expected degree of sparsity with the horseshoe prior is a considerable improvement over a standard logistic regression model  (section~\ref{sec:results}).
+
+\begin{figure}
+  \includegraphics[width=1\linewidth]{plots/without_horseshoe.pdf}
+  \includegraphics[width=1\linewidth]{plots/with_horseshoe.pdf}
+  \caption{Marginal posterior distributions of a subset of regression coefficients for replacement 11D of the HIV integrase, based on the Arevir dataset.
+The graphic shows regression coefficients for two HLA alleles: For HLA-A*01, the data does not show evidence of an association with replacement 11D, whereas for HLA-B*44, the data is consistent with a strong association.
+The horseshoe prior has the effect of shrinking regression coefficients with weak evidence of an association with the replacement strongly towards 0, while leaving the coefficients that are far away from 0 mostly unshrunk.
+Shrinking the irrelevant regression coefficients to 0 has the advantage of reducing the standard error for the remaining coefficients, which can be observed by the much smaller width of the histogram of HLA-B*44 in the model with horseshoe prior.}
+  \label{fig:horseshoe-comparison}
+\end{figure}
+
 \subsection{Leave-one-out cross-validation}
@@
 
+\hl{DaHo: Es gibt fast keinen Text in dieser subsection}
+
 The models were applied on sequence data from a variety of different viruses and show a considerable improvement over currently applied methods that are consistent across datasets.
@@
 
-\subsection{A real-world application}
+\subsection{Example Hepatitis Delta Virus}
 
@@
 
+ \subsection{Individual examples}
+In the following we illustrate with ... examples the effects of the inclusion of additional relevant information on the detection of HAMs.
+
+ \hl{DaHo: bitte Beispiel mit Linkage Disequilibrium zeigen}
+
+\hl{DaHo: bitte konkretes Bsp mit Auswirkung Phylogenie zeigen}
+
 \section{Discussion}

### 2021-02-14 docs/main.tex
added Bioinformatics style main.tex etc.

--- 2021-02-14 docs/manuscript.tex
@@
-%----------------------------------------------------------------------------------------
-%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
-%----------------------------------------------------------------------------------------
+\documentclass{bioinfo}
+\copyrightyear{2015} \pubyear{2015}
 
-\documentclass[fleqn,11pt]{SelfArx} % Document font size and equations flushed left
-\usepackage[justification=justified]{caption}
-\usepackage{parskip}
-\usepackage{microtype}
-\usepackage{svg}
-\usepackage{float}
-\usepackage{placeins}
-\usepackage[english]{babel} % Specify a different language here - english by default
-\usepackage{listings}
-\usepackage{fontspec}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{makecell}
+\access{Advance Access Publication Date: Day Month Year}
+\appnotes{Manuscript Category}
 \usepackage{soul}
-% \usepackage{lipsum} % Required to insert dummy text. To be removed otherwise
-\setmainfont{TeX Gyre Termes}
-%----------------------------------------------------------------------------------------
-%	COLUMNS
-%----------------------------------------------------------------------------------------
 
-\setlength{\columnsep}{0.55cm} % Distance between the two columns of text
-\setlength{\fboxrule}{1pt} % Width of the border around the abstract
+\begin{document}
+\firstpage{1}
 
-%----------------------------------------------------------------------------------------
-%	COLORS
-%----------------------------------------------------------------------------------------
+\subtitle{Subject Section}
 
-\definecolor{color1}{RGB}{124,0,0} % Color of the article title and sections
-\definecolor{color2}{RGB}{150,150,150} % Color of the boxes behind the abstract and headings
+\title[HAMdetector]{HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations}
+\author[Sample \textit{et~al}.]{Corresponding Author\,$^{\text{\sfb 1,}*}$, Co-Author\,$^{\text{\sfb 2}}$ and Co-Author\,$^{\text{\sfb 2,}*}$}
+\address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\      
+$^{\text{\sf 2}}$Department, Institution, City, Post Code,
+Country.}
 
-%----------------------------------------------------------------------------------------
-%	HYPERLINKS
-%----------------------------------------------------------------------------------------
+\corresp{$^\ast$To whom correspondence should be addressed.}
 
-\usepackage{hyperref} % Required for hyperlinks
-\hypersetup{hidelinks,colorlinks,breaklinks=true,urlcolor=color2,citecolor=color1,
-  linkcolor=color1,bookmarksopen=false,pdftitle={Title},pdfauthor={Author}}
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
 
-%----------------------------------------------------------------------------------------
-%	ARTICLE INFORMATION
-%----------------------------------------------------------------------------------------
+\editor{Associate Editor: XXXXXXX}
 
-\PaperTitle{
-    HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations
-} % Article title
-
-\Authors{Daniel Habermann\textsuperscript{1}, ..., Daniel Hoffmann\textsuperscript{1}} % Authors
-\affiliation{\textsuperscript{1}\textit{Bioinformatics \& Computational Biophysics, Faculty of Biology, University of Duisburg-Essen, 45117 Essen, Germany}} % Author affiliation
-% \affiliation{\textsuperscript{2}\textit{Department of Chemistry, University of Examples, London, United Kingdom}} % Author affiliation
-% \affiliation{*\textbf{Corresponding author}: john@smith.com} % Corresponding author
-
-\Keywords{human leukocyte antigen system, HLA, multiple sequence alignment, escape mutations, 
-  viral escape, Bayesian inference, sparsity, horseshoe prior, epitope prediction} % Keywords - if you don't want any simply remove all the text between the curly brackets
-\newcommand{\keywordname}{Keywords} % Defines the keywords heading name
-
-%----------------------------------------------------------------------------------------
-%	ABSTRACT
-%----------------------------------------------------------------------------------------
-
-\Abstract{
-  \\
-  \textbf{Motivation} \\
-  A key process in anti-viral adaptive immunity is that the human leukocyte antigen system (HLA) presents viral peptide fragments in MHC I protein-peptide complexes on cell surfaces and in this way alerts CD8\textsuperscript{+} cytotoxic T-Lymphocytes (CTLs).
+\abstract{\textbf{Motivation:}   A key process in anti-viral adaptive immunity is that the human leukocyte antigen system (HLA) presents viral peptide fragments in MHC I protein-peptide complexes on cell surfaces and in this way alerts CD8\textsuperscript{+} cytotoxic T-Lymphocytes (CTLs).
 This pathway exerts a strong selection pressure on viruses, favoring viral mutants that escape recognition by the HLA/CTL system, e.g.\ by point mutations that decrease binding of viral peptides to MHC I.
@@
 The reliable identification of HAMs is not only important to understand viral genomes and their evolution, but it also impacts the development of broadly effective anti-viral treatments and vaccines against variable viruses.\\
-
   By their very nature HAMs are amenable to detection by statistical methods in paired sequence / HLA data.
@@
 Bayesian inference offers a principled approach to achieve this.\\
-
-  \textbf{Results} \\
-Here, we present a new regression model for the detection of HAMs.
+\textbf{Results:} Here, we present a new regression model for the detection of HAMs.
 As we choose a Bayesian approach we can include the novel sparsity-inducing priors, and we obtain easily interpretable quantitative information on HAM candidates.
@@
 \\
+\textbf{Availability:}   The source code of this software is available at ... under a permissive MIT license. \\
+\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
+\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
+online.}
 
-  \textbf{Availability and Implementation} \\
-  The source code of this software is available at \url{http://github.com} under a permissive MIT license. \\
-
-  \textbf{Supplementary information} \\
-  \href{https://google.com}{Supplementary data} are available at \textit{Bioinformatics} online. \\
-}
-
-%----------------------------------------------------------------------------------------
-\begin{document}
-
-\flushbottom % Makes all text pages the same height
-
-\maketitle % Print the title and abstract box
-
-{
-  \hypersetup{linkcolor=black}
-  \tableofcontents
-}
+\maketitle
 
 \section{Introduction}
-
 \subsection{The HLA system}
 
-One way of how the human immune system recognizes viral infections is through the human leukocyte antigen system~\cite{Germain1994}: Proteins, including viral proteins, in cells are often degraded in proteasomes to peptide fragments~\cite{Goldberg2002}.
+One way of how the human immune system recognizes viral infections is through the human leukocyte antigen system~\citep{Germain1994}: Proteins, including viral proteins, in cells are often degraded in proteasomes to peptide fragments~\citep{Goldberg2002}.
 A small subset of these peptides is presented on the cell surface by MHC I receptors, encoded by HLA genes (in humans HLA-A, -B, -C).
-HLA genes are very polymorphic with more than 20,000 known alleles in the human population~\cite{Robinson2014}.
+HLA genes are very polymorphic with more than 20,000 known alleles in the human population~\citep{Robinson2014}.
 The resulting gene products differ in their binding properties, which means that cells from different individuals in general present different peptides on the cell surface.
 
-Cytotoxic T-Lymphocytes (CTLs) are selected during maturation to only weakly bind to complexes of MHC I with self-peptides, but CTLs can bind strongly to complexes of MHC I with peptides from viral proteins~\cite{Murata2007}.
-Upon activation, CTLs become cytotoxic and recruit other immune cells so that infected cells can be eliminated~\cite{Harty2000}.
+Cytotoxic T-Lymphocytes (CTLs) are selected during maturation to only weakly bind to complexes of MHC I with self-peptides, but CTLs can bind strongly to complexes of MHC I with peptides from viral proteins~\citep{Murata2007}.
+Upon activation, CTLs become cytotoxic and recruit other immune cells so that infected cells can be eliminated~\citep{Harty2000}.
 
 \subsection{HLA escape}
-In this way, the HLA system exerts strong selection pressure towards virus variants that escape T cell recognition~\cite{Borrow1997}.
-Such variants could, for example, carry mutations that result in reduced binding of viral peptides to MHC I or T cell receptors, or that alter peptide processing so that peptides are no longer presented~\cite{Yewdell2002}.
+In this way, the HLA system exerts strong selection pressure towards virus variants that escape T cell recognition~\citep{Borrow1997}.
+Such variants could, for example, carry mutations that result in reduced binding of viral peptides to MHC I or T cell receptors, or that alter peptide processing so that peptides are no longer presented~\citep{Yewdell2002}.
 
-HLA diversity drives viral evolution in individuals where a virus adapts to specific host features, and in populations and regions that in general differ in HLA frequencies~\cite{Kawashima2009}.
-Upon transmission to a new host with different HLA alleles, HLA escape mutations may revert, as they are often associated with a reduction in viral replication capacity~\cite{Matthews2008}, but they also may be irreversible~\cite{Kawashima2009}.
+HLA diversity drives viral evolution in individuals where a virus adapts to specific host features, and in populations and regions that in general differ in HLA frequencies~\citep{Kawashima2009}.
+Upon transmission to a new host with different HLA alleles, HLA escape mutations may revert, as they are often associated with a reduction in viral replication capacity~\citep{Matthews2008}, but they also may be irreversible~\citep{Kawashima2009}.
 
-Whether and how quickly a given escape mutation is selected in a host depends, e.g., on the viral genomic background, the magnitude of the reduction in viral replication, the availability of compensatory mutations that recover fitness, or the strength of selection pressure~\cite{Kloverpris2016}.
+Whether and how quickly a given escape mutation is selected in a host depends, e.g., on the viral genomic background, the magnitude of the reduction in viral replication, the availability of compensatory mutations that recover fitness, or the strength of selection pressure~\citep{Kloverpris2016}.
 
-Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV~\cite{Alizon2011, Allen2005, Rousseau2008, Lumley2018}.
+Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV~\citep{Alizon2011, Allen2005, Rousseau2008, Lumley2018}.
 This includes the development of treatments and vaccines that rely on effective immune responses.
@@
 
-There are several experimental methods available to study HLA escape~\cite{Timm2004}.
+There are several experimental methods available to study HLA escape~\citep{Timm2004}.
 \hl{DaHo: use key references, e.g.
 those that have introduced a technique; alternatively: cite review like articles that give an overview of several experimental methods} However, these methods are relatively slow and costly, especially for screening purposes.
-A promising approach that makes efficient use of frequently available data is to combine viral genome sequencing, host HLA determination, computational identification by statistical association analysis, and targeted experimental validation~\cite{Carlson2012}.
+A promising approach that makes efficient use of frequently available data is to combine viral genome sequencing, host HLA determination, computational identification by statistical association analysis, and targeted experimental validation~\citep{Carlson2012}.
 
-As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface and thus on binding to the HLA encoded MHC I protein, escape mutations are usually HLA allele specific and can therefore be detected as HLA allele dependent ``footprints'' in sequence alignments of viral proteins  \cite{Moore2002}, i.e.\ as HLA associated mutations (HAM) or replacements enriched in sequences from hosts with a specific HLA allele.
+As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface and thus on binding to the HLA encoded MHC I protein, escape mutations are usually HLA allele specific and can therefore be detected as HLA allele dependent ``footprints'' in sequence alignments of viral proteins  \citep{Moore2002}, i.e.\ as HLA associated mutations (HAM) or replacements enriched in sequences from hosts with a specific HLA allele.
 
-One way of quantifying this enrichment is Fisher's exact test~\cite{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(\neg H\)), (\(\neg R_{i}\), \(H\)) and (\(\neg R_{i}\), \(\neg H\)), where \(\neg R_{i}\) denotes any replacement except \(R_{i}\), and \(\neg H\) denotes any HLA allele except \(H\).
+One way of quantifying this enrichment is Fisher's exact test~\citep{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(\neg H\)), (\(\neg R_{i}\), \(H\)) and (\(\neg R_{i}\), \(\neg H\)), where \(\neg R_{i}\) denotes any replacement except \(R_{i}\), and \(\neg H\) denotes any HLA allele except \(H\).
 
@@
 
-Fisher's exact test has the advantage of being fast and easy to apply~\cite{Budeus2016}, but it also has several disadvantages~\cite{Carlson2008}.
+Fisher's exact test has the advantage of being fast and easy to apply~\citep{Budeus2016}, but it also has several disadvantages~\citep{Carlson2008}.
 The most striking one is that viral sequences share a common phylogenetic history, and,  therefore, treating sequences as independent and identically distributed samples may under- or overestimate effect sizes.
@@
 
-Carlson et al.~\cite{Carlson2008} developed the Phylogenetic Dependency Network, a method that accounts for several of the aforementioned problems, in particular phylogenetic bias and  HLA linkage disequilibrium.
+\citet{Carlson2008} developed the Phylogenetic Dependency Network, a method that accounts for several of the aforementioned problems, in particular phylogenetic bias and  HLA linkage disequilibrium.
 However, it is based on null hypothesis significance testing.
 
-\subsection{Issues with using p-values as a screening tool}
+\subsection{Issues with p-values for screening}
 
-There are fundamental statistical issues with p-values as screening tool~\cite{Amrhein2017}:
-with small effect sizes and high variance between measurements, as is often the case with biological data, statistically significant results can be misleading, have the wrong direction (type S error), or greatly overestimate an effect (type M error)~\cite{Gelman2014}.
-Such problems are more and more appreciated in the context of the current ``replication crisis'', which describes that scientific claims with seemingly strong statistical evidence fail to replicate~\cite{Ioannidis2005, Begley:2012, Baker2016Nature-reproducibility-crisis}.
+There are fundamental statistical issues with p-values as screening tool~\citep{Amrhein2017}:
+with small effect sizes and high variance between measurements, as is often the case with biological data, statistically significant results can be misleading, have the wrong direction (type S error), or greatly overestimate an effect (type M error)~\citep{Gelman2014}.
+Such problems are more and more appreciated in the context of the current ``replication crisis'', which describes that scientific claims with seemingly strong statistical evidence fail to replicate~\citep{Ioannidis2005, Begley:2012, Baker2016Nature-reproducibility-crisis}.
 
@@
 When using p-values as a filter, it is therefore likely to obtain significant effects that are in fact not real.
-To circumvent this problem, a common strategy is to control the false discovery rate~\cite{Benjamini1995}.
+To circumvent this problem, a common strategy is to control the false discovery rate~\citep{Benjamini1995}.
 These adjustment procedures have the problem that, when performing many of such comparisons, none but the very largest effects remain.
@@
 Instead of performing many hypothesis tests and trying to adjust for them, we prefer to fit a single, multilevel model that contains all comparisons of interest.
-When using multilevel models, the problem of multiple comparisons can disappear entirely and yield more valid estimates~\cite{Gelman2012}.
+When using multilevel models, the problem of multiple comparisons can disappear entirely and yield more valid estimates~\citep{Gelman2012}.
 
+\begin{methods}
   \section{Material and methods}
 
-\hl{DaHo: Da es ein Methoden-Paper ist, sollte dies der Hauptteil sein}
-
-Our general approach is to fit a Bayesian regression model that captures relationships between host HLA alleles and replacements in viral proteomes.
-This Bayesian approach is advantageous because it allows use of: (1) prior information (even if only knowledge of effect magnitudes), (2) relevant additional information (phylogeny, epitope information), (3) a problem-specific structure, (4) partial pooling~\cite{Gelman2010}.
-With these features, HAM detection is improved markedly over the state-of-the-art.
+  Our general approach is to fit Bayesian regression models that captures relationships between host HLA alleles and replacements in viral proteomes.
+This Bayesian approach is advantageous because it allows use of: (1) prior information (e.g.\ knowledge of effect magnitudes), (2) relevant additional information (phylogeny, epitope information), (3) a problem-specific structure, (4) partial pooling~\citep{Gelman2010}.
 
@@
 
-We chose a logistic regression model as a backbone for two reasons: First, because it is easily extensible and secondly, because the coefficients are interpretable as expected increases on the log-odds scale, identical to the familiar interpretation of standard logistic regression models.
+We chose a logistic regression model as a backbone because it is easily extensible, and because coefficients can be interpreted in the familiar way as factors on the log-odds scale.
 
@@
   y_{ik} & \sim \text{Bernoulli}(\theta_{ik}) \\
-  \theta_{ik} & = \text{logistic}(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}}),
+  \theta_{ik} & = \text{logistic}\left(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}}\right),
 \end{align}
 
-where \(y_{ik}\) is the binary encoded observation at sequence \(i\) for replacement \(k\), \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\), \(X_{ij}\) is 1 if sequence \(i\) is annotated with HLA allele \(j\) and 0 otherwise, \(\beta_{\text{HLA}_{jk}}\) is the HLA regression coefficients for HLA allele \(j\) at replacement \(k\), \(D\) is the number of HLA alleles in the dataset, and \(\text{inv\_logit}\) is the logistic inverse link function.
+where \(y_{ik}\) is the binary encoded observation of replacement $k$ in viral sequence $i$; \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\); \(X_{ij}\) is 1 if sequence \(i\) comes from host individual HLA allele \(j\) and 0 otherwise; $\beta_{\text{HLA}_{jk}}$ is the HLA regression coefficient of HLA allele \(j\) for replacement \(k\); \(D\) is the number of HLA alleles in the dataset; the logistic inverse link function transforms the linear model in parentheses to the probability scale of $\theta_{ik}$.
 Note that, if we observe N different amino acid states at a certain alignment position, each of those are considered in turn to be the “replacement”.
@@
 Standard statistical methods usually assume samples to be independent and identically distributed, which may lead to wrong conclusions when this assumption is strongly violated.
-In the context of identifying HLA associations, Bhattacharya et al.~\cite{Bhattacharya2007} demonstrated the importance of correcting for the phylogenetic structure.
-There are many approaches described in the literature for phylogenetic regression for binary dependent variables~\cite{Ives2014}, where the most popular approach is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree~\cite{Ives2009}.
+In the context of identifying HLA associations, \citet{Bhattacharya2007} demonstrated the importance of correcting for the phylogenetic structure.
+There are many approaches described in the literature for phylogenetic regression for binary dependent variables~\citep{Ives2014}, where the most popular approach is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree~\citep{Ives2009}.
 
-This approach showed to be too computationally expensive to include in our model, so we chose a strategy that is similar to the one in Carlson et al.
-\cite{Carlson2008} for the Phylogenetic Dependency Network:
+This approach showed to be too computationally expensive to include in our model, so we chose a strategy that is similar to the one in \citet{Carlson2008} for the Phylogenetic Dependency Network:
 
@@
 
-The logit transform is used because is cancels out with the logistic link function.
+The logit transform is used because is cancels out with the logistic inverse link function.
 This means that the phylogeny information acts as a baseline in absence of any HLA effects.
@@
     Similarly, there are also HLA-allele independent effects like antigen processing that influence presentation on MHC I.
-Second generational tools like MHCFlurry 2.0~\cite{ODonnell2020} use both binding affinity and antigen processing data to improve epitope prediction.
+Second generational tools like MHCFlurry 2.0~\citep{ODonnell2020} use both binding affinity and antigen processing data to improve epitope prediction.
 For our tool HAMdetector, we use the output of MHCFlurry to benefit from this data in order to predict HLA-associated mutations.
@@
 There is vast experimental data on epitope binding motifs available, e.g.
-through the use of mass spectrometry~\cite{Hunt1992}.
-Epitope prediction software like MHCFlurry~\cite{ODonnell2020} use these data to extrapolate MHC I binding for new, untested peptides.
+through the use of mass spectrometry~\citep{Hunt1992}.
+Epitope prediction software like MHCFlurry~\citep{ODonnell2020} use these data to extrapolate MHC I binding for new, untested peptides.
 As HLA-associated mutations are much more likely to lie in epitopes (e.g.
@@
 
-  Recent advantages in Bayesian inference include so-called sparsity-promoting priors~\cite{Piironen2017}.
+  Recent advantages in Bayesian inference include so-called sparsity-promoting priors~\citep{Piironen2017}.
 Sparsity-promoting priors convey the a priori expectation that most coefficients in a regression model are close to 0.
@@
 They share the common structure of placing most probability mass very close to 0, with large to tails to accommodate the non-zero coefficients.
-For our model, we use a so-called horse prior~\cite{Carvalho2010}, which is defined as a scale-mixture of gaussians:
+For our model, we use a so-called horse prior~\citep{Carvalho2010}, which is defined as a scale-mixture of gaussians:
 
@@
 
-A useful addition to the horseshoe prior is given in Piironen et al.~\cite{Piironen2017}, which describes a way to choose the parameter governing the overall degree of sparsity \(\tau_{0}\) based on the expected number of non-zero coefficients.
+A useful addition to the horseshoe prior is given in Piironen et al.~\citep{Piironen2017}, which describes a way to choose the parameter governing the overall degree of sparsity \(\tau_{0}\) based on the expected number of non-zero coefficients.
 
-For our HLA model, we utilize an extension to the horseshoe prior called the regularized horseshoe prior~\cite{Piironen2017}, which also provides some regularization for the non-zero coefficients.
+For our HLA model, we utilize an extension to the horseshoe prior called the regularized horseshoe prior~\citep{Piironen2017}, which also provides some regularization for the non-zero coefficients.
 This is particularly useful for logistic regression models, as some shrinkage helps to deal with issues of separability and collinearity that commonly occur when fitting logistic regression models.
@@
 However, the model would also work reasonably well with just a single global parameter \(\gamma\).
-All other additions (\(\tilde{\lambda}_{jk}^{2}\) and \(\tau_{0k}\)) are explained in detail in \cite{Piironen2017}.
+All other additions (\(\tilde{\lambda}_{jk}^{2}\) and \(\tau_{0k}\)) are explained in detail in \citep{Piironen2017}.
 Briefly, the addition of  \(\tilde{\lambda}_{jk}^{2}\) allows some regularization for the non-zero coefficients and the parameterization of \(\tau_{0k}\) allows to place a prior on the expected number of non-zero coefficients.
@@
 A Student-t prior with these parameters is a reasonable default choice for logistic regression models.
-The theoretical justification behind this regularized variant of the horseshoe prior is explained in detail in~\cite{Piironen2017}.
+The theoretical justification behind this regularized variant of the horseshoe prior is explained in detail in~\citet{Piironen2017}.
 
 The value of \(\tau_{0k}\) is chosen to imply 10 effective non-zero HLA regression coefficients per replacement.
-The rationale behind this parameterization is again outlined in~\cite{Piironen2017}.
+The rationale behind this parameterization is again outlined in~\citet{Piironen2017}.
 The value of 10 is approximately based on available HIV epitope maps.
@@
 
-\subsubsection{Model implementation}
-All models were implemented in Stan~\cite{Stan2021}.
+\subsection{Model implementation}
+All models were implemented in Stan~\citep{Stan2021}.
 The Stan code is available online in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
-A Julia~\cite{Bezanson2017} package is available \hl{DaHo: at ...} to run the model on custom data.
+A Julia~\citep{Bezanson2017} package is available \hl{DaHo: at ...} to run the model on custom data.
 Due to restrictions of dependencies (MHCFlurry and RAxML-ng), HAMdetector is only available on Linux.
@@
 
-The model was fit to several data sets:
+The model was fit with several data sets:
 
 \begin{itemize}
-  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\cite{Carlson2008}.
-  \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\cite{Roomp2006}.
+  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}.
+  \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\citep{Roomp2006}.
   \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm.
@@
   \item If not already present in this format, sequences were translated into their amino acid representation.
-  \item RAxML-NG~\cite{Kozlov2019} version 1.0.0 was used to generate a maximum likelihood phylogenetic tree for each gene/protein using the \texttt{--model GTR+G+I} option with all other parameters set to default values.
+  \item RAxML-NG~\citep{Kozlov2019} version 1.0.0 was used to generate a maximum likelihood phylogenetic tree for each gene/protein using the \texttt{--model GTR+G+I} option with all other parameters set to default values.
 If available, we used RNA or DNA sequences for this step, rather than protein sequences.
@@
 
-  % \textit{Partial pooling of 4-digit HLA alleles}
-
-  % It has recently been appreciated that binding specificities can vary drastically across  HLA alleles from the same allele group, e.g.
-between HLA-B*51:01 and HLA-B*51:03.
-For predicting HLA associated mutations, there has consequently been a shift to use 4-digit resolution data whenever available.
-This is not without downsides however, because overall, binding specificities are more similar for alleles in the same group, and, therefore, treating them as completely separate might unnecessarily fragment the available data.
-
-  % In Bayesian statistics, it is not required to chose one extreme or the other: Instead of either treating all HLA alleles from the same allele group as identical or as completely separate, partial pooling of the estimates allows something in between: With partial pooling, estimates do share some information across each other, but they are also allowed to vary if necessary.
-
-  % In the context of our model this means that observing a relevant association between a replacement and an HLA allele also makes the model more inclined to estimate relevant associations between that replacement and all other alleles from that allele group.
-The degree of this partial pooling can be estimated from the data, so if we do observe strong similarity of HLA alleles in a given allele group the estimates are more influenced by each other than if they do not.
-
-\section{Implementation}
-
-% \subsection{Including HLA linkage disequilibrium}
-
-% The other two important confounding effects that are recognized in Carlson et al.~\cite{Carlson2008} are HLA linkage disequilibrium and codon covariation.
-% HLA linkage disequilibrium describes the observation that HLA alleles that lie in proximity to each other on the chromosome are also more likely to be inherited together.
-When using a statistical method that evaluates each HLA allele individually, this may lead to spurious associations: If some HLA allele \(X\) is associated with replacement \(k\), and HLA allele \(Y\) often co-occurs with HLA allele \(X\), this means that we also identify an association between HLA allele \(Y\) and replacement \(k\).
-
-% This association is correct in a statistical sense, but if the purpose of the model is to identify sites under HLA-mediated selection pressure, these associations due to linkage disequilibrium are most likely unwanted.
-
-% Note that these spurious associations only occur when testing each HLA allele individually.
-In a regression model where all HLA alleles of a sample are evaluated at once, the marginal posterior distributions for correlated predictors become broader to accurately reflect that the model is unable to discern which of these predictors explain the observed association with a replacement.
-
-% We do not currently account for codon covariation in our model, as we are interested in identifying all HLA associated mutations (which include for example compensatory mutations).
-The general structure of the model, in particular the sparsity assumption introduced in section~\ref{sec:sparsity}, would allow adding codon covariation by adding all other codons as additional predictors.
-This would generally not be possible in a frequentist setting because the predictor matrix would become singular and the inverse does not exist.
+\end{methods}
 
@@
   \begin{tabular}{c|c|c|c|c}
-  \thead{replacement} & \thead{allele}  & \thead{p-value \\ (FET)} & \thead{post. prob. \\ (HAMdetector)} & \thead{confirmed} \\
+  replacement & allele  & p-value (FET) & post.~prob. & confirmed \\
   \hline
@@
 
-\thispagestyle{empty} % Removes page numbering from the first page
-%----------------------------------------------------------------------------------------
-%	ARTICLE CONTENTS
-%----------------------------------------------------------------------------------------
-\FloatBarrier
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
+%     as it may ignore the figures.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%----------------------------------------------------------------------------------------
-%	REFERENCE LIST
-%----------------------------------------------------------------------------------------
-\phantomsection
-\bibliographystyle{unsrt}
+% \section{Conclusion}
+
+%\section*{Acknowledgements}
+
+\section*{Funding}
+
+This work has been supported by Deutsche Forschungsgemeinschaft (grant HO 1582/10-1).\vspace*{-12pt}
+
+\bibliographystyle{natbib}
+%\bibliographystyle{achemnat}
+%\bibliographystyle{plainnat}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{bioinformatics}
+%
+%\bibliographystyle{plain}
+%
 \bibliography{references}
-%----------------------------------------------------------------------------------------
-
 \end{document}

### 2021-02-15 docs/main.tex
changes in methods

@@
 
-  Our general approach is to fit Bayesian regression models that captures relationships between host HLA alleles and replacements in viral proteomes.
+  Our general approach for HAMdetector is to fit Bayesian regression models that captures relationships between host HLA alleles and replacements in viral proteomes.
 This Bayesian approach is advantageous because it allows use of: (1) prior information (e.g.\ knowledge of effect magnitudes), (2) relevant additional information (phylogeny, epitope information), (3) a problem-specific structure, (4) partial pooling~\citep{Gelman2010}.
@@
 
-We chose a logistic regression model as a backbone because it is easily extensible, and because coefficients can be interpreted in the familiar way as factors on the log-odds scale.
+We chose a logistic regression model as a backbone because it is easily extensible, and because coefficients can be interpreted in the familiar way as summands on the log-odds scale.
 
@@
 
-where \(y_{ik}\) is the binary encoded observation of replacement $k$ in viral sequence $i$; \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\); \(X_{ij}\) is 1 if sequence \(i\) comes from host individual HLA allele \(j\) and 0 otherwise; $\beta_{\text{HLA}_{jk}}$ is the HLA regression coefficient of HLA allele \(j\) for replacement \(k\); \(D\) is the number of HLA alleles in the dataset; the logistic inverse link function transforms the linear model in parentheses to the probability scale of $\theta_{ik}$.
-Note that, if we observe N different amino acid states at a certain alignment position, each of those are considered in turn to be the “replacement”.
+where \(y_{ik}\) is the binary encoded observation of replacement $k$ in viral sequence $i$ (each observed amino acid state $k$ contributes a separate column to $y_{ik}$); \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\); \(X_{ij}\) is 1 if sequence \(i\) comes from host individual HLA allele \(j\) and 0 otherwise; $\beta_{\text{HLA}_{jk}}$ is the HLA regression coefficient of HLA allele \(j\) for replacement \(k\); \(D\) is the number of HLA alleles in the dataset; the logistic inverse link function transforms the linear model in parentheses to the probability scale of $\theta_{ik}$.
 
-The regression coefficients \(\beta_{\text{HLA}_{jk}}\) are the main focus of interest, as they quantify the association between the occurrence of replacement \(k\) and each of the observed HLA alleles.
-They are in units of log-odds, so a regression coefficient of somewhere around 2 means that when comparing sequences with and without that particular HLA allele, the log-odds of observing replacement \(k\) in presence of that HLA allele is increased by 2.
+The main parameters of interests for HAMdetector are the regression coefficients $\beta_{\text{HLA}_{jk}}$, as they quantify the strength of association between the occurrence of replacement \(k\) and each of the observed HLA alleles.
+The $\beta_{\text{HLA}_{jk}}$ are on the log-odds scale, i.e.\ if we go from viral sequences from hosts without HLA allele $j$ to those from hosts with $j$, the log-odds $\log (p_k/(1-p_k))$ of observing replacement \(k\) increase by addition of $\beta_{\text{HLA}_{jk}}$.
 
-Reasoning about coefficients on the log-odds scale can sometimes be unintuitive.
+% Reasoning about coefficients on the log-odds scale can sometimes be unintuitive.
 A useful approximation to interpret logistic regression coefficients on the probability scale is the so-called divide-by-4 rule, which means that a regression coefficient of 2 corresponds to an expected increase on the probability scale of about 2/4 = 50\%.
@@
 
-A useful addition to the horseshoe prior is given in Piironen et al.~\citep{Piironen2017}, which describes a way to choose the parameter governing the overall degree of sparsity \(\tau_{0}\) based on the expected number of non-zero coefficients.
+A useful addition to the horseshoe prior is given in \citet{Piironen2017}, which describes a way to choose the parameter governing the overall degree of sparsity \(\tau_{0}\) based on the expected number of non-zero coefficients.
 
@@
 A Student-t prior with these parameters is a reasonable default choice for logistic regression models.
-The theoretical justification behind this regularized variant of the horseshoe prior is explained in detail in~\citet{Piironen2017}.
+The theoretical justification behind this regularized variant of the horseshoe prior is explained in detail in \citet{Piironen2017}.
 
 The value of \(\tau_{0k}\) is chosen to imply 10 effective non-zero HLA regression coefficients per replacement.
-The rationale behind this parameterization is again outlined in~\citet{Piironen2017}.
+The rationale behind this parameterization is again outlined in \citet{Piironen2017}.
 The value of 10 is approximately based on available HIV epitope maps.

### 2021-02-15 docs/main.tex
adde possible references for violation of independence for FET

@@
 In the context of hypothesis testing, this leads to increased false positive and false negative rates.
+\hl{There are at least two references we could cite here: Osborne2002 is the reference usually cited in this context and contains the phrase "When these assumptions are not met the results may not be trustworthy, resulting in a Type I or Type II error, \dots" It is more of a summary for psychologists though and contains some questionable advice like removing outliers for increased power.
+There is also Scariano1987 who looks at type I/type II errors in one-way ANOVA when independence is violated, this may be a better reference.
+I have not found any article on the violation of independence in fisher's exact test specifically.} \citep{Osborne2002} \citep{Scariano1987}
 

### 2021-02-16 docs/main.tex
work on phylogeny section done

--- 2021-02-15 docs/main.tex
@@
 \begin{methods}
-  \section{Material and methods}
+  \section{Materials and methods}
 
@@
 A useful approximation to interpret logistic regression coefficients on the probability scale is the so-called divide-by-4 rule, which means that a regression coefficient of 2 corresponds to an expected increase on the probability scale of about 2/4 = 50\%.
+DaHo: das gilt nur näherungsweise in der Nähe von p=0.5 (Wendepunkt)
 
@@
 
-On top of the paired data of sequences and HLA alleles modeled by the backbone (Eq~\ref{eq:backbone}), we extend the model to include further information of relevance to improve HAM detection, namely phylogenetic information, and predictions of epitope peptide processing and MHC I affinity, as described in the following.
+On top of the paired data of viral sequences and host HLA alleles modeled by the backbone (Eq~\ref{eq:backbone}), we extend the model to include further information of relevance to improve HAM detection, namely phylogenetic information, and predictions of epitope peptide processing and MHC I affinity, as described in the following.
 
 \subsubsection{Phylogeny}
-One problem that occurs when analyzing sequence data is that species share a common phylogenetic history.
-Standard statistical methods usually assume samples to be independent and identically distributed, which may lead to wrong conclusions when this assumption is strongly violated.
-In the context of identifying HLA associations, \citet{Bhattacharya2007} demonstrated the importance of correcting for the phylogenetic structure.
-There are many approaches described in the literature for phylogenetic regression for binary dependent variables~\citep{Ives2014}, where the most popular approach is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree~\citep{Ives2009}.
+One problem that occurs when analyzing sequence data is that species (here: viral strains) share a common phylogenetic history.
+Thus replacements are not independently and identically distributed, and therefore violate a common assumption of standard statistical methods.
+In fact, \citet{Bhattacharya2007} demonstrated the importance of correcting for the phylogenetic structure in identifying HLA associations.
 
-This approach showed to be too computationally expensive to include in our model, so we chose a strategy that is similar to the one in \citet{Carlson2008} for the Phylogenetic Dependency Network:
+The most popular approach in phylogeny-aware regression of binary variables is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree~\citep{Ives2009, Ives2014}.
+This approach turned out to be too computationally expensive in our model, hence we chose a strategy similar to the one in \citet{Carlson2008}:
 
@@
 A quantity that can be readily computed using phylogenetic software like RAxML-NG~\cite{Kozlov2019} is \(P(\Psi|y_{ik}=1)\).
-For this, we keep the tree topology fixed, annotate the tree with the binary observations \(y\) at its leaves and optimize the branch lengths.
+For this, we keep the tree topology fixed, annotate the tree with the binary observations \(y_{ik}\) at its leaves and optimize the branch lengths.
 \(P(\Psi|y_{ik}=1)\) is then the likelihood of the annotated phylogenetic tree.
-We can then use Bayes' rule to invert that conditional probability by additionally computing \(P(\Psi|y_{ik}=0)\), which is done by altering the annotation for sequence \(i\) and keeping all other observations constant.
+Similarly, we can also compute \(P(\Psi|y_{ik}=0)\) by flipping the annotation of sequence $i$ from 1 to 0 (keeping all other observations).
+With  $P(\Psi|y_{ik}=1)$ and $P(\Psi|y_{ik}=0)$ known and the relative frequencies of 0 an 1 as priors, we can estimate $P(y_{ik}=1|\Psi)$ by applying Bayes' theorem.
 
-In order to check if the inferred probabilities \(P(y_{ik}=1|\Psi)\) faithfully reflect the observed data we use calibration plots (see fig.~\ref{fig:phylogeny-calibration}).
+In order to check whether the inferred probabilities \(P(y_{ik}=1|\Psi)\) faithfully reflect the observed data we use calibration plots (see Fig.~\ref{fig:phylogeny-calibration}).
 All observations are binned according to their inferred probability \(P(y_{ik}=1|\Psi)\).
@@
 
-\begin{figure}[!ht]
-  \includegraphics[width=1\linewidth]{plots/phylogeny_calibration.pdf}
-  \caption{Calibration plot for the HBV dataset.
-For a description of all datasets used in this study see section~\ref{sec:results}.
-Calibration plots for the other datasets are shown in the supplementary.
-  All observations are first sorted by increasing estimated probability \(P(y_{ik}=1|\Psi)\) and then grouped into \(n\) bins.
-  For each bin, the fraction of observations with \(y_{ik}=1\) (observed event percentage) is compared to the midpoint of each bin (the value in the center between the lowest and highest probability).
-The error bars show the cutpoints for each bin.
-If the probabilities are calibrated correctly, each dot is supposed to scatter closely around the diagonal line.}
-  \label{fig:phylogeny-calibration}
-\end{figure}
-
-The estimated probabilities are then included in the model as additional intercepts:
+The estimated probabilities based on phylogeny are then included in the model as additional intercepts (second term of logistic argument):
 
@@
   y_{ik}  & \sim \text{Bernoulli}(\theta_{ik}) \\
-  \theta_{ik}  & =  \text{logistic}\Bigl(\beta_{0_{k}} + \gamma~\text{logit}\left(P(y_{ik}=1|\Psi)\right) \\
+  \theta_{ik}  & =  \text{logistic}\Bigl(\beta_{0_{k}} + \gamma_k~\text{logit}\left(P(y_{ik}=1|\Psi)\right) \\
   &\;\;\;\;\; + \sum_{k=1}^{D}X_{ik}\beta_{\text{HLA}_{jk}}\Bigr)
@@
 
-The logit transform is used because is cancels out with the logistic inverse link function.
+The logit transform is used because it cancels out with the logistic inverse link function.
 This means that the phylogeny information acts as a baseline in absence of any HLA effects.
-Note that we also include an additional parameter \(\gamma\), that is constrained to be positive.
-This helps because the inferred probabilities \(P(y_{ik}=1|\Psi)\) might not necessarily reflect the true underlying phylogenetic signal, for example because the phylogenetic tree does not match the observed data well enough.
-
-A straight-forward extension of our model would be to also account for these sources of uncertainty, for example by using a Bayesian method to estimate a posterior distribution over possible tree topologies.
-The uncertainty over the tree topology and the underlying parameters of the phylogenetic model would then propagate into uncertainty of the estimated probabilities \(P(y_{ik}=1|\Psi)\).
-However, in order to not increase the runtime of the model further we  use standard maximum likelihood estimates and then include these in a way that additionally allows for measurement error.
+As this baseline itself is not certain but subject to errors of the phylogenetic probabilities $P(y_{ik}=1|\Psi)$, we introduce an additional parameter $\gamma_k$, constrained to be positive.
+\hl{DaHo: Please check prior for $\gamma_k$. Constraint ok?}
 
@@
 
-\subsection*{Convergence diagnostics}
-
+\subsection{Model diagnostics}
+\subsubsection{Convergence}
  All model fits showed no signs of inference issues.
@@
 \hl{DaHo: citation Gelman + Rubin original paper or later versions if improved $\hat{R}$ have been used}
+
+\subsubsection{Calibration}
+ \begin{figure}[!ht]
+  \includegraphics[width=1\linewidth]{plots/phylogeny_calibration.pdf}
+  \caption{Calibration plot for the HBV dataset.
+For a description of all datasets used in this study see section~\ref{sec:results}.
+Calibration plots for the other datasets are shown in the supplementary.
+  All observations are first sorted by increasing estimated probability \(P(y_{ik}=1|\Psi)\) and then grouped into \(n\) bins.
+  For each bin, the fraction of observations with \(y_{ik}=1\) (observed event percentage) is compared to the midpoint of each bin (the value in the center between the lowest and highest probability).
+The error bars show the cutpoints for each bin.
+If the probabilities are calibrated correctly, each dot is supposed to scatter closely around the diagonal line.}
+  \label{fig:phylogeny-calibration}
+\end{figure}
 
@@
 
+A straight-forward extension of our model would be to also account for these sources of uncertainty, for example by using a Bayesian method to estimate a posterior distribution over possible tree topologies.
+The uncertainty over the tree topology and the underlying parameters of the phylogenetic model would then propagate into uncertainty of the estimated probabilities \(P(y_{ik}=1|\Psi)\).
+However, in order to not increase the runtime of the model further we use standard maximum likelihood estimates and then include these in a way that additionally allows for measurement error.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

### 2021-02-16 docs/main.tex
add possible references for experimentally studying HLA escape

--- 2021-02-15 docs/main.tex
@@
 \hl{DaHo: use key references, e.g.
-those that have introduced a technique; alternatively: cite review like articles that give an overview of several experimental methods} However, these methods are relatively slow and costly, especially for screening purposes.
+those that have introduced a technique; alternatively: cite review like articles that give an overview of several experimental methods} \hl{DaHa: see .tex comments for possible citations} However, these methods are relatively slow and costly, especially for screening purposes.
 A promising approach that makes efficient use of frequently available data is to combine viral genome sequencing, host HLA determination, computational identification by statistical association analysis, and targeted experimental validation~\citep{Carlson2012}.
+
+% IFN-gamma ELISPOT:
+% A  Solid-Phase Enzyme-Linked Immunospot (ELISPOT) Assay for Enumeration of Specific Antibody-Secreting Cells \citep{Czerkinsky1983}
+% Quantification of antigen specific CD8+ T cells using an ELISPOT assay \citep{Miyahira1995}
+
+% Chromium release
+% Quantitative assay of the lytic action of immune lymphoid cells on 51-Cr-labelled allogeneic target cells in vitro; inhibition by isoantibody and by drugs \citep{Brunner1968}
+
+% Intracellular cytokine staining
+% Intracellular cytokine optimization and standard operating procedure \citep{Lamoreaux2006}
+% The assessment of antigen-specific CD8+ T cells through the combination of MHC class I tetramer and intracellular staining \citep{Appay2002}
+
+% HLA tetramer staining
+% Phenotypic analysis of antigen-specific T lymphocytes \citep{Altman1996}
 

### 2021-02-16 docs/main.tex
remove wrong positivity constraint from phylogeny gamma
add b_epi to full model specification
add genes to dataset list

@@
 This means that the phylogeny information acts as a baseline in absence of any HLA effects.
-Note that we also include an additional parameter \(\gamma\), that is constrained to be positive.
+Note that we also include an additional parameter \(\gamma\).
 This helps because the inferred probabilities \(P(y_{ik}=1|\Psi)\) might not necessarily reflect the true underlying phylogenetic signal, for example because the phylogenetic tree does not match the observed data well enough.
@@
    \sigma_{\text{phy}} &\sim \text{Normal}^{+}(0, 0.5) & \text{(*)} \\
+   \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2) & \text{(*)} \\
    \beta_{\text{HLA}_{jk}} &\sim \text{Normal}(0, \tau_{k}^{2}\tilde{\lambda}^{2}_{jk}) & \\
@@
   \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}.
-  \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\citep{Roomp2006}.
-  \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm.
-  \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf. \hl{DaHo: bitte von Hadi geeignetes Zitat erfragen}
-  \item A set of 41 HIV sequences that were kindly provided by Rongge Yang. 
+The dataset contains sequences spanning the \textit{gag}, \textit{pol}, \textit{env}, \textit{nef}, \textit{vif}, \textit{vpr}, \textit{vpu}, \textit{tat} and \textit{rev} genes.
+  \item A set of 351 HIV sequences mostly spanning the \textit{pol} gene from the Arevir database~\citep{Roomp2006}.
+  \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm. The dataset contains sequences of the HBe, LHBs, Pol and HBx proteins.
+  \item A set of 104 Hepatitis-D-Virus sequences containing the HDV-antigen that were kindly provided by Michael Roggendorf. \hl{DaHo: bitte von Hadi geeignetes Zitat erfragen}
+  \item A set of 41 HIV sequences spannign the \textit{gag} and \textit{pol} genes that were kindly provided by Rongge Yang. 
 \end{itemize}
-
-\hl{DaHo: zu jedem Datensatz die Proteine nennen}
 

### 2021-02-16 docs/main.tex
add genes to dataset list
starting epitope section of methods

--- 2021-02-16 docs/main.tex
@@
 
-We chose a logistic regression model as a backbone because it is easily extensible, and because coefficients can be interpreted in the familiar way as summands on the log-odds scale.
+We chose a logistic regression model as backbone because it is easily extensible, and because coefficients can be interpreted in the familiar way as summands on the log-odds scale.
 
@@
 
-\textit{Binding affinity prediction}
+\subsubsection{Inclusion of CTL epitope predictions}
 
-\hl{DaHo: Mein Verständnis von Materials and Methods ist, dass hier die Methoden aufgeführt werden, z.B.
-die Methoden für die Berechnung von Affinitäten, siehe z.B.
-https://academic.oup.com/bioinformatics/article/36/21/5159/5874440}
+%A major CTL escape mechanism is the generation of mutations in viral peptides that prevent peptide presentation by MHC I to T cell receptors. 
+Mutations in CTL epitopes (the viral peptides that are presented by MHC I molecules to T cell receptors) can prevent or worsen epitope presentation to CTLs and thus cause viral CTL escape.
+Therefore, mutations in epitopes are prime candidates for CTL escape.
+This means that the fact of a replacement
+
+A major mechanism for viral CTL escape is the generation of
 

### 2021-02-16 docs/main.tex

@@
 The logit transform is used because it cancels out with the logistic inverse link function.
-This means that the phylogeny information acts as a baseline in absence of any HLA effects.
-As this baseline itself is not certain but subject to errors of the phylogenetic probabilities $P(y_{ik}=1|\Psi)$, we introduce an additional parameter $\gamma_k$, constrained to be positive.
-\hl{DaHo: Please check prior for $\gamma_k$. Constraint ok?}
+The phylogeny term acts as a baseline in absence of any HLA effects.
+As this baseline itself is not certain but subject to errors of the phylogenetic probabilities $P(y_{ik}=1|\Psi)$, we introduce an additional parameter $\gamma_k$.
 
@@
   \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}.
-  \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\citep{Roomp2006}.
-  \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm.
-  \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf. \hl{DaHo: bitte von Hadi geeignetes Zitat erfragen}
-  \item A set of 41 HIV sequences that were kindly provided by Rongge Yang. 
+The dataset contains sequences spanning the \textit{gag}, \textit{pol}, \textit{env}, \textit{nef}, \textit{vif}, \textit{vpr}, \textit{vpu}, \textit{tat} and \textit{rev} genes.
+  \item A set of 351 HIV sequences mostly spanning the \textit{pol} gene from the Arevir database~\citep{Roomp2006}.
+  \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm. The dataset contains sequences of the HBe, LHBs, Pol and HBx proteins.
+  \item A set of 104 Hepatitis-D-Virus sequences containing the HDV-antigen that were kindly provided by Michael Roggendorf. \hl{DaHo: bitte von Hadi geeignetes Zitat erfragen}
+  \item A set of 41 HIV sequences spanning the \textit{gag} and \textit{pol} genes that were kindly provided by Rongge Yang. 
 \end{itemize}
-
-\hl{DaHo: zu jedem Datensatz die Proteine nennen}
 

### 2021-02-17 docs/main.tex
epitope predictions

@@
 
-As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface and thus on binding to the HLA encoded MHC I protein, escape mutations are usually HLA allele specific and can therefore be detected as HLA allele dependent ``footprints'' in sequence alignments of viral proteins  \citep{Moore2002}, i.e.\ as HLA associated mutations (HAM) or replacements enriched in sequences from hosts with a specific HLA allele.
+As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface and thus on binding to the HLA encoded MHC I protein, escape mutations are usually HLA allele specific and can therefore be detected as HLA allele dependent ``footprints'' in sequence alignments of viral proteins  \citep{Moore2002}, i.e.\ as HLA associated mutations (HAM) or replacements enriched in viral sequences from hosts with a specific HLA allele.
 
@@
 
-%A major CTL escape mechanism is the generation of mutations in viral peptides that prevent peptide presentation by MHC I to T cell receptors. 
-Mutations in CTL epitopes (the viral peptides that are presented by MHC I molecules to T cell receptors) can prevent or worsen epitope presentation to CTLs and thus cause viral CTL escape.
-Therefore, mutations in epitopes are prime candidates for CTL escape.
-This means that the fact of a replacement
+% A major CTL escape mechanism is the generation of mutations in viral peptides that prevent peptide presentation by MHC I to T cell receptors.
 
-A major mechanism for viral CTL escape is the generation of
+As outlined earlier, escape mutations often appear as HAMs.
+Given the underlying mechanism, it is not surprising that escape mutations are enriched in CTL epitopes, i.e.\ in those viral peptides presented by MHC I to T cell receptors \citep{Bronke2013}.
+This suggests that knowledge of epitope regions can be used to boost HAM detection.
+Fortunately, availability of large experimental data sets \citep{Vita2019} has enabled the development of computational tools that predict with good accuracy the binding of peptides to MHC I molecules encoded by various HLA alleles \citep{Mei2020}.
 
-    HAMs are expected to lie more frequently in T-cell epitopes, i.e.\ the peptide fragments presented by MHC to T-cell receptors.
-There are vast experimental binding affinity data available of different peptides and MHC I molecule pairs, and there are well-established computational methods that use these data to extrapolate HLA binding for untested peptides.
-We show that, by including the outcome of these computational tools as input for a probabilistic model, the prediction of HLA-associated mutations can be improved.
+Not only mutations in CTL epitopes can lead to failure of MHC I to present epitopes, but also mutations at epitope-flanking positions that interfere with pre-processing of peptides, notably proteasomal cleavage of viral proteins \hl{DaHo: citation}.
 
-  \textit{Antigen processing prediction}
+In HAMdetector we use MHCFlurry 2.0~\citep{ODonnell2020} to predict peptide-MHC I affinity and peptide pre-processing.
+For this, we create an input matrix of dimensions $R\times D$, where R is the number of evaluated replacements and D is the number of observed HLA alleles in the dataset.
+The elements of this matrix are binary encoded and contain a 1 if that position is either expected to be inside a predicted epitope or related to antigen processing, and 0 otherwise.
+Given an amino acid sequence, MHCFlurry provides a list of possible epitopes (9-13 mers) and HLA allele pairs and calculates a rank based on comparisons with random epitope and HLA pairs.
+For the binarization we use the rank threshold of 0.2\% suggested by MHCFlurry.
 
-    Similarly, there are also HLA-allele independent effects like antigen processing that influence presentation on MHC I.
-Second generational tools like MHCFlurry 2.0~\citep{ODonnell2020} use both binding affinity and antigen processing data to improve epitope prediction.
-For our tool HAMdetector, we use the output of MHCFlurry to benefit from this data in order to predict HLA-associated mutations.
-
-    \subsubsection{Epitope predictions}
-
-There is vast experimental data on epitope binding motifs available, e.g.
-through the use of mass spectrometry~\citep{Hunt1992}.
-Epitope prediction software like MHCFlurry~\citep{ODonnell2020} use these data to extrapolate MHC I binding for new, untested peptides.
-As HLA-associated mutations are much more likely to lie in epitopes (e.g.
-because of selection pressure to escape MHC I binding), using epitope prediction software can provide valuable external data for identifying HLA-associated mutations.
-
-For our model, we use MHCFlurry 2.0 for binding prediction, as it does not only predict HLA binding but also antigen processing.
-For this, we create an input matrix of dimensions \(R\times D\), where R is the number of evaluated replacements and D is the number of observed HLA alleles in the dataset.
-The elements of this matrix are binary encoded and contain a 1 if that position is either expected to be inside a predicted epitope or related to antigen processing, and 0 otherwise.
-Given an amino acid sequence, MHCFlurry 2.0 provides a list of possible epitopes (9-13 mers) and HLA allele pairs and calculates a rank based on comparisons with random epitope and HLA pairs.
-For the binarization we use a rank threshold of 0.2\% (as suggested by MHCFlurry).
-
-We typically try to avoid thresholds whenever possible, but in this case we deemed it to be preferable to including the predicted binding affinities, as we expected the binary values to be more reliable than the predicted numeric values.
-Note that the output of these epitope prediction tools may be imprecise or even wrong, but by including it as an input for a probabilistic model, the model is able to learn how much to trust this data if parameterized in a way to allow for measurement error.
-
-We use epitope prediction as information about the expected degree of sparsity, i.e.
-when knowing a certain position is restricted by a certain HLA allele, we expect that this HLA allele is more likely to be associated with the replacement than the other HLA alleles.
-This can be reflected by increasing the scale of the local shrinkage parameters \(\lambda_{jk}\):
+We use epitope prediction as information about the expected degree of sparsity, i.e.\ if we know that a certain position is restricted by a certain HLA allele, we expect that this HLA allele is more likely to be associated with replacements at that position than the other HLA alleles.
+This idea is implemented by increasing the scale of the local shrinkage parameters \(\lambda_{jk}\) depending on epitope information:
 
@@
 
-where \(Z_{jk}\) is 1 if HLA allele \(j\) is predicted to restrict the alignment position corresponding to replacement \(k\) or is associated antigen processing and 0 otherwise.
+where \(Z_{jk}\) is 1 if HLA allele \(j\) is predicted to restrict the alignment position corresponding to replacement \(k\), or is associated antigen processing, and 0 otherwise.
 The parameter \(\beta_{\text{epi}}\) governs the increase in scale of the  corresponding local shrinkage parameters.
 The larger the estimated values of \(\beta_{\text{epi}}\) are, the more likely it is to see non-zero regression coefficients for these HLA alleles.
+
+\hl{DaHo: komplexer Zusammenhang von pre-processing escape und HLA; bitte besprechen}
 

### 2021-02-17 docs/main.tex
including prior justification

@@
   y_{ik}  & \sim \text{Bernoulli}(\theta_{ik}) \\
-  \theta_{ik}  & =  \text{logistic}\Bigl(\beta_{0_{k}} + \gamma_k~\text{logit}\left(P(y_{ik}=1|\Psi)\right) \\
+  \theta_{ik}  & =  \text{logistic}\Bigl(\beta_{0_{k}} + \gamma \text{logit}\left(P(y_{ik}=1|\Psi)\right) \\
   &\;\;\;\;\; + \sum_{k=1}^{D}X_{ik}\beta_{\text{HLA}_{jk}}\Bigr)
@@
 The phylogeny term acts as a baseline in absence of any HLA effects.
-As this baseline itself is not certain but subject to errors of the phylogenetic probabilities $P(y_{ik}=1|\Psi)$, we introduce an additional parameter $\gamma_k$.
+As this baseline itself is not certain but subject to errors of the phylogenetic probabilities $P(y_{ik}=1|\Psi)$, we introduce an additional parameter $\gamma$.
 
@@
 
-\subsection{Sparsity-inducing priors}
+\subsubsection{Sparsity-inducing priors}
 
-  Recent advantages in Bayesian inference include so-called sparsity-promoting priors~\citep{Piironen2017}.
-Sparsity-promoting priors convey the a priori expectation that most coefficients in a regression model are close to 0.
-This assumption also applies to HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes.
-If no epitope spanning a given alignment position is presented on the MHC I molecule, any association between a replacement and the respective HLA allele  is likely to be due to random variation alone (or HLA linkage disequilibrium).
+Sparsity-promoting priors \citep{Piironen2017} can drastically improve predictive performance, because the model is better able to differentiate between signal and noise.
+These priors convey the a priori expectation that most coefficients in a regression model are close to 0, i.e.\ that non-zero coefficients are sparse.
+This assumption is likely correct for HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes.
+If no epitope spanning a given alignment position is presented on the MHC I molecule, any apparent association between a replacement and the respective HLA allele is likely to be due to random viral sequence variation or HLA linkage disequilibrium.
+\hl{DaHo: Wie ist das mit compensatory mutations *außerhalb* von Epitopen?
+Z.B.
+https://jvi.asm.org/content/86/2/991}
 
-  It has been shown \hl{DaHo: where?} that using a sparsity-promoting prior when non-zero coefficients are sparse can drastically improve predictive performance, because the model is better able to differentiate between signal and noise.
-
-  We show that a simple logistic regression model with a sparsity-promoting prior on the regression coefficients for the HLA alleles alone already performs roughly on-par  with the much more elaborate Phylogenetic Dependency Network approach from Carlson et al.
-when using the fraction of identified escapes that lie in known HLA epitopes as a benchmark.
-
-In the context of identifying HLA associations we expect most of the HLA coefficients to be very close to 0, as a given alignment position is typically only restricted by none or few HLA alleles.
-This is important prior knowledge that should be incorporated in the model.
-
-For Bayesian models, there are a variety of different sparsity-promoting priors with slightly different properties.
-They share the common structure of placing most probability mass very close to 0, with large to tails to accommodate the non-zero coefficients.
-For our model, we use a so-called horse prior~\citep{Carvalho2010}, which is defined as a scale-mixture of gaussians:
+There is a range of sparsity-promoting priors with slightly different properties.
+They share the common structure of placing most probability mass very close to 0, with heavy tails to accommodate the non-zero coefficients.
+For our model, we use a so-called horseshoe prior \citep{Carvalho2010}, a scale-mixture of Gaussians:
 
 \begin{equation}
+  \label{eq:sparsity-prior}
   \begin{aligned}
@@
 \end{equation},
-where \(\beta_{jk}\) are the regression coefficients, \(\tau\) is the so-called global shrinkage parameter, \(\lambda_{jk}\) are the so-called local shrinkage parameters and \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution.
+where \(\beta_{jk}\) are the regression coefficients; \(\tau\) and \(\lambda_{jk}\) are the so-called global and local shrinkage parameters, respectively;, \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution; $\tau_0$ is the overall degree of sparsity.
 
-Intuitively speaking, the global shrinkage parameter \(\tau\) is typically very small and shrinks most of the regression coefficients to 0, whereas the local shrinkage parameters can occasionally be very large to allow some coefficients to escape that shrinkage.
+With Eq~\ref{eq:sparsity-prior} the global shrinkage parameter \(\tau\) is typically very small and shrinks most of the regression coefficients close to 0, whereas the local shrinkage parameters $\lambda_{jk}$ can occasionally be very large to allow some coefficients to escape that shrinkage.
 
-A useful addition to the horseshoe prior is given in \citet{Piironen2017}, which describes a way to choose the parameter governing the overall degree of sparsity \(\tau_{0}\) based on the expected number of non-zero coefficients.
-
-For our HLA model, we utilize an extension to the horseshoe prior called the regularized horseshoe prior~\citep{Piironen2017}, which also provides some regularization for the non-zero coefficients.
-This is particularly useful for logistic regression models, as some shrinkage helps to deal with issues of separability and collinearity that commonly occur when fitting logistic regression models.
+The overall degree of sparsity $\tau_0$ can be chosen based on the expected number of non-zero coefficients \citep{Piironen2017}. \hl{DaHo: do we do this? How?}
 
@@
 
-The full specification of the model is given by:
+The full specification of the model is:
 
@@
    \theta_{ik} &=
-        \text{logit}^{-1}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi)) & \\
+        \text{logistic}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi)) & \\
         \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{\text{HLA}_{jk}}) & \\
@@
 The full model specification includes some aspects that were not covered in the previous sections.
-In particular, the parameters \(\gamma_{k}\) are modelled hierarchically, which allows information to be partially pooled across replacements.
-However, the model would also work reasonably well with just a single global parameter \(\gamma\).
-All other additions (\(\tilde{\lambda}_{jk}^{2}\) and \(\tau_{0k}\)) are explained in detail in \citep{Piironen2017}.
+In particular, the overall phylogeny-weight $\gamma$ in Eq~\ref{eq:backbone} is replaced in the full model by  hierarchically modelled  \(\gamma_{k}\), which allows partial pooling across replacements (even with a global parameter \(\gamma\) the model works reasonably well).
+All other additions (\(\tilde{\lambda}_{jk}^{2}\) and \(\tau_{0k}\)) are taken from \citet{Piironen2017}.
 Briefly, the addition of  \(\tilde{\lambda}_{jk}^{2}\) allows some regularization for the non-zero coefficients and the parameterization of \(\tau_{0k}\) allows to place a prior on the expected number of non-zero coefficients.
+This is particularly useful for logistic regression models, as some shrinkage helps to deal with issues of separability and collinearity that commonly occur with logistic regression models.
+\hl{DaHo: Bedeutung von $N$ in $\tau_{0k}$?}
 
@@
 
-Prior distributions are labeled with an asterisk in equation~\ref{eq:full}.
-They are meant to be weakly informative, which means that the goal is to provide some modest amount of shrinkage while still allowing for large absolute values given enough support by the observed data.
+Prior distributions are labeled with an asterisk in Eq~\ref{eq:full}.
+They are weakly informative, which means that they effectively limit posteriors to realistic magnitudes of parameters.
 One exception to this are the intercepts \(\beta_{0_k}\), which are essentially flat because they are well identified by the data alone.
 
-The hierarchical mean and standard deviation of the phylogeny coefficients \(\gamma_{k}\) are chosen to place most probability mass at values of \(\gamma_{k}\) of around 1.
-In the absence of any HLA effects, a parameter value of 1 would mean that the estimate for the probability of observing replacement \(k\) is identical to the probability based on the phylogenetic model.
-Intuitively speaking, We therefore treat the phylogeny as a baseline and any observations that cannot be attributed to phylogeny are attributed to the HLA alleles or noise.
+The hierarchical mean and standard deviation of the phylogeny coefficients \(\gamma_k\) place most probability mass on \(\gamma_k\) values around 1.
+In absence of any HLA effects, a $\gamma_k=1$ would mean that the estimate for the probability of observing replacement $k$ is identical to the probability based on the phylogenetic model.
+This treats phylogeny as a baseline, and any observations not attributed to phylogeny must be explained by HLA alleles or noise.
 
-The prior on \(c_{k}^{2}\) implies a Student-t prior with 7 degrees of freedom and a scale of 1 on the non-zero HLA regression coefficients \(\beta_{\text{HLA}_{jk}}\).
-A Student-t prior with these parameters is a reasonable default choice for logistic regression models.
-The theoretical justification behind this regularized variant of the horseshoe prior is explained in detail in \citet{Piironen2017}.
+The prior on $c_k^2$ implies a Student-t prior with 7 degrees of freedom and a scale of 1 on the non-zero HLA regression coefficients \(\beta_{\text{HLA}_{jk}}\).
+A Student-t prior with these parameters is a reasonable default choice for logistic regression models \citep{Piironen2017}.
 
-The value of \(\tau_{0k}\) is chosen to imply 10 effective non-zero HLA regression coefficients per replacement.
+The value of \(\tau_{0k}\) implies 10 effective non-zero HLA regression coefficients per replacement.
 The rationale behind this parameterization is again outlined in \citet{Piironen2017}.
-The value of 10 is approximately based on available HIV epitope maps.
-Please note that this is only an approximate statement about the general magnitude, i.e.
-we would expect maybe a couple of HLA alleles to be associated with a single replacement, not much more.
+The value of 10 corresponds to a generously estimated magnitude based on available HIV epitope maps \hl{DaHo: citation?}.
 The model is also parameterized in a way that assumes an equal degree of sparsity across all alignment positions a priori.
@@
 
+\hl{DaHo: Textblock aus alten Methods:}  We show that a simple logistic regression model with a sparsity-promoting prior on the regression coefficients for the HLA alleles alone already performs roughly on-par  with the much more elaborate Phylogenetic Dependency Network approach from Carlson et al.
+when using the fraction of identified escapes that lie in known HLA epitopes as a benchmark.
+
 \subsection{Generalization}

### 2021-02-18 docs/main.tex
kleine Änderungen im MHCFlurry/Epitop-Teil

@@
 
-In HAMdetector we use MHCFlurry 2.0~\citep{ODonnell2020} to predict peptide-MHC I affinity and peptide pre-processing.
-For this, we create an input matrix of dimensions $R\times D$, where R is the number of evaluated replacements and D is the number of observed HLA alleles in the dataset.
-The elements of this matrix are binary encoded and contain a 1 if that position is either expected to be inside a predicted epitope or related to antigen processing, and 0 otherwise.
+In HAMdetector we use MHCFlurry 2.0~\citep{ODonnell2020} to predict epitopes that are properly processed and presented by MHC I.
+For this, we create an input matrix of dimensions $R\times D$, where $R$ is the number of evaluated replacements and $D$ is the number of observed HLA alleles in the dataset.
+The elements of this matrix are binary encoded and contain a 1 if that position is predicted to be in an epitope, and 0 otherwise.
 Given an amino acid sequence, MHCFlurry provides a list of possible epitopes (9-13 mers) and HLA allele pairs and calculates a rank based on comparisons with random epitope and HLA pairs.
@@
 
-where \(Z_{jk}\) is 1 if HLA allele \(j\) is predicted to restrict the alignment position corresponding to replacement \(k\), or is associated antigen processing, and 0 otherwise.
+where \(Z_{jk}\) is 1 if HLA allele \(j\) is predicted to restrict the alignment position corresponding to replacement \(k\), and 0 otherwise.
 The parameter \(\beta_{\text{epi}}\) governs the increase in scale of the  corresponding local shrinkage parameters.

### 2021-02-20 docs/main.tex
introduced HAM enrichment plot

@@
 With  $P(\Psi|y_{ik}=1)$ and $P(\Psi|y_{ik}=0)$ known and the relative frequencies of 0 an 1 as priors, we can estimate $P(y_{ik}=1|\Psi)$ by applying Bayes' theorem.
-
-In order to check whether the inferred probabilities \(P(y_{ik}=1|\Psi)\) faithfully reflect the observed data we use calibration plots (see Fig.~\ref{fig:phylogeny-calibration}).
-All observations are binned according to their inferred probability \(P(y_{ik}=1|\Psi)\).
-If these probabilities are calibrated correctly, we expect that in a group of observations with an inferred probability \(P(y_{ik}=1|\Psi)\) of around x\% we really do observe \(y_{ik}=1\) around x\% percent of the time.
-
 The estimated probabilities based on phylogeny are then included in the model as additional intercepts (second term of logistic argument):
@@
 \subsection{Model implementation}
-All models were implemented in Stan~\citep{Stan2021}.
-The Stan code is available online in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
-A Julia~\citep{Bezanson2017} package is available \hl{DaHo: at ...} to run the model on custom data.
+All models were implemented in Stan 2.23 \citep{Stan2021}, a probabilistic programming language and Hamiltonian Monte Carlo sampler for efficient numerical computation of posterior distributions.
+The Stan code is available \hl{DaHo: at ...} in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
+A Julia \citep{Bezanson2017} package is available \hl{DaHo: at ...} to run the model on custom data.
 Due to restrictions of dependencies (MHCFlurry and RAxML-ng), HAMdetector is only available on Linux.
@@
 
+\subsubsection{Convergence diagnostics}
+We use the split-\(\hat{R}\) convergence diagnostic to identify Markov chain convergence issues \citep{Gelman1992, Stan2021}.
+We require a value of \(\hat{R}\) below 1.01 for all model parameters.
+Additionally, we require that the effective sample size \(N_{\text{eff}}\) is above 500 for all model parameters and that sampling occurs without any divergent transitions \citep{Stan2021}.
+
 \subsubsection{Posterior predictive checks}
 
-  Additionally, Bayesian statistics provides an accessible way to test models: By comparing data generated under the model's assumptions to the actually observed data, it is possible to identify important aspects of the dataset that the models fails to capture and subsequently improve the model until it is consistent with the observed data~\cite{Gabry2019}.
+In posterior predictive checks, we simulate new data from the inferred posterior distribution and the likelihood, and we compare these simulated data with representative real data \citep{Gabry2019}.
+A good model should predict data that are consistent with real data.
+This general idea was employed in two ways to test our models.
 
-  Here, we use these so-called posterior predictive checks to compare the calibration of the predicted probabilities of observing the replacement to the actual observed data.
+First, we checked with \emph{calibration plots} whether the inferred probabilities \(P(y_{ik}=1|\Psi)\) faithfully reflect the observed data (Fig.~\ref{fig:calibration}).
+All observations were binned according to their inferred probability \(P(y_{ik}=1|\Psi)\).
+If these probabilities are calibrated correctly, we expect that in a group of observations with an inferred probability \(P(y_{ik}=1|\Psi)\) of around x\% we really do observe \(y_{ik}=1\) around x\% percent of the time.
 
-  \subsubsection{Convergence diagnostics}
-
-  We use Stan (version 2.23)~\cite{Stan2021} to specify and fit our Bayesian models.
-  Stan provides an implementation of a dynamic Hamilitonian Monte Carlo sampler, which is a powerful technique to approximate expectations with respect to the posterior distribution~\cite{Betancourt2017}.
-One advantage of using such methods is that they provide sensitive diagnostics to identify a wide range of possible sampling issues.
-  We use the split-\(\hat{R}\) convergence diagnostic~\cite{Gelman1992} to identify convergence issues.
-We require a value of \(\hat{R}\) below 1.01 for all model parameters.
-Additionally, we require that the effective sample size \(N_{\text{eff}}\) is above 500 for all model parameters and that sampling occurs without any divergent transitions.
-A detailed explanation of these quantities is provided in the Stan reference manual.
+Second, we assessed the abilities of different models and methods to discover HAMs with \emph{HAM enrichment plots}.
+These plots are based on the observation that CTL escape mutations are enriched in epitopes \citep{Bronke2013}.
+Hence, the degree by which methods for HAM prediction recover this trend is a measure of model performance.
+To implement this measure, we first ranked all evaluated replacements according to their respective credibility of being a HAM, computed as integral of the marginal posterior $P(\beta_{HLA}>0)$.
+For comparison with established methods, namely Fisher's exact test and Phylogenetic Dependency Network \citep{Carlson2008}, ranked lists based on p- and q-values, respectively, were computed.
+Then we calculated for each rank $r$ the accumulated number $N_e(r)$ of predictions of this rank or better ranks were located inside known epitopes.
+The higher the curve $N_e(r)$, the higher the enrichment of predicted HAMs in epitopes, see e.g.\ Fig.\ \ref{fig:comparison}.
 
@@
 
+\hl{DaHo: sources of known epitopes?}
+
 For all sequences, we applied the following data preparation steps:
@@
 \subsubsection{Calibration}
- \begin{figure}[!ht]
-  \includegraphics[width=1\linewidth]{plots/phylogeny_calibration.pdf}
-  \caption{Calibration plot for the HBV dataset.
-For a description of all datasets used in this study see section~\ref{sec:results}.
-Calibration plots for the other datasets are shown in the supplementary.
-  All observations are first sorted by increasing estimated probability \(P(y_{ik}=1|\Psi)\) and then grouped into \(n\) bins.
-  For each bin, the fraction of observations with \(y_{ik}=1\) (observed event percentage) is compared to the midpoint of each bin (the value in the center between the lowest and highest probability).
-The error bars show the cutpoints for each bin.
-If the probabilities are calibrated correctly, each dot is supposed to scatter closely around the diagonal line.}
-  \label{fig:phylogeny-calibration}
-\end{figure}
 
@@
 
-\subsection*{Comparison to a list of known epitopes}
+\subsection{Comparison to a list of known epitopes}
 
-When testing models, we prefer to evaluate them on real-world data whenever possible. 
-However, comparing against real-world data is not straight forward when identifying HLA-associated mutations, because if a tool identifies an HLA-associated mutation which is not yet documented, this could be either because it is a false positive, or a yet unknown HLA-associated mutation that has not been documented before.
-Also, viewing every non-identified HLA-associated mutation that is listed in the literature as a false negative is not correct either, because documented escape mutations do not necessarily have to show in each sequence alignment.
-A third issue when comparing models is that HLA-associated mutations are not necessarily binary, so the true positive / false negative framework does not work as well.
+%However, comparing against real-world data is not straight forward when identifying HLA-associated mutations, because if a tool identifies an HLA-associated mutation which is not yet documented, this could be either because it is a false positive, or a yet unknown HLA-associated mutation that has not been documented before.
+%Also, viewing every non-identified HLA-associated mutation that is listed in the literature as a false negative is not correct either, because documented escape mutations do not necessarily have to show in each sequence alignment.
+%A third issue when comparing models is that HLA-associated mutations are not necessarily binary, so the true positive / false negative framework does not work as well.
 
-To circumvent these issues and still compare against real-world data, we chose the following strategy:
-For each model, all evaluated replacements are ranked by decreasing confidence of being an HLA associated mutation.
+%To circumvent these issues and still compare against real-world data, we chose the following strategy:
+%For each model, all evaluated replacements are ranked by decreasing confidence of being an HLA associated mutation.
 That is, for Bayesian models we calculate the probability of the corresponding regression coefficient exceeding zero \(P(\beta_{HLA} > 0)\), for Fisher's exact test we sort HLA allele - replacement pairs by increasing p-values and for the Phylogenetic Dependency Network, we rank the associations by increasing q-values.
-Then, a list of known epitopes is used to create a plot of the cumulative number of replacements inside the boundary of a known epitope vs.
+%Then, a list of known epitopes is used to create a plot of the cumulative number of replacements inside the boundary of a known epitope vs.
 the corresponding rank.
@@
   \includegraphics[width=1\linewidth]{plots/comparison.pdf}
-  \caption{Number of associations inside the boundary of known epitopes vs.
-rank.
-For each model, the pairs of HLA alleles and replacements are sorted by decreasing confidence of being an HLA associated mutation.
-Then, each replacement is compared to a list of known epitopes and a line of the cumulative number of associations inside known epitopes vs.
-the current rank is plotted for each model.
-Better performing models are expected to have an enrichment of associations that are inside of known epitopes.
-  PhyloD: Phylogenetic Dependency Network, Model 1: simple logistic regression model with broad Student-t priors, Model 2: logistic regression model with horseshoe prior, Model 3: logistic regression model with horseshoe prior and phylogeny, Model 4: Full model with epitope prediction.
-The thin gray lines denote random permutations of the list of HLA allele - replacement pairs and act as a baseline.}
+  \caption{HAM enrichment plot for HBe protein: number $N_e$ of associations inside the boundary of known epitopes vs.\ rank $r$.
+PhyloD: Phylogenetic Dependency Network; model 1: simple logistic regression model with broad Student-t priors; model 2: logistic regression model with horseshoe prior; model 3: logistic regression model with horseshoe prior and phylogeny; model 4: full model with epitope prediction.
+The thin gray line are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baseline.}
   \label{fig:comparison}

### 2021-02-20 docs/main.tex
materials and methods complete (except yellow remarks)

@@
 For our model, we use a so-called horseshoe prior \citep{Carvalho2010}, a scale-mixture of Gaussians:
-
-\begin{equation}
+\begin{align}
   \label{eq:sparsity-prior}
@@
     \lambda_{jk} &\sim \text{Cauchy}^{+}(0, 1) \\
-    \tau &\sim \text{Cauchy}^{+}(0, \tau_{0})
+    \tau &\sim \text{Cauchy}^{+}(0, \tau_{0}),
   \end{aligned}
-\end{equation},
+\end{equation}
 where \(\beta_{jk}\) are the regression coefficients; \(\tau\) and \(\lambda_{jk}\) are the so-called global and local shrinkage parameters, respectively;, \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution; $\tau_0$ is the overall degree of sparsity.
@@
 
-  In addition to testing our model on real-world data, we also performed leave-one-out cross-validation to examine the relative importance of the additions of epitope prediction, sparsity and phylogeny with respect to predictive performance.
-  For this, we create several variations of the model that are all special cases of the full HAMdetector model.
-  By comparing the model performance of these variations, we can gain insight on the relative importance of each aspect of the model.
+Another performance measure is the ability to generalize to unseen data.
+To examine this ability for the different model variants we performed leave-one-out cross-validation (LOOCV), using the efficient Pareto-smoothed LOOCV \cite{Vehtari2016}.
 
-  Leave-one-out cross-validation has historically been challenging to apply for Bayesian models, because fitting a model once for each left out data-point is often infeasible because of computational constrains.
-  Recently, Pareto-smoothed importance sampling has gained popularity to perform approximate leave-one-out cross-validation for Bayesian models.
-This method  approximates the leave-one-out posterior by re-weighting existing draws from the posterior distribution and has the advantage of requiring to fit the model only once.
-Additionally, the method provides good diagnostics to identify cases where the approximation is unreliable~\cite{Vehtari2016}.
+% Several utility (or cost) functions are used in practice to compare the predictive performance of models.
+For frequentist approaches, classification accuracy or receiver operator characteristic curves (ROC curves) are most common.
+From the LOOCV we obtain the Expected Log-Predictive Density (ELPD) $\sum_{i=1}^{n}\log(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))$ for samples $i=1,\ldots,n$, $i$th observation $y_i$, data $y_{i-1}$ with $i$ left out, and model parameters $\theta$.
+Thus, the ELPD is the average log predictive density of the observed data points based on the leave-one-out posterior distributions.
+This measure has the advantage over other performance measures like classification accuracy, that it not only takes the location of the predictive distribution (the number of correct predictions) into account, but also the width, i.e.\ how confident the model is in its predictions.
 
-  Several utility (or cost) functions are used in practice to compare the predictive performance of models.
-For frequentist approaches, classification accuracy or receiver operator characteristic curves (ROC curves) are most common.
-  For Bayesian models, expected log-predictive density (elpd) is used most frequently.
-
-  elpd is defined as \(\sum_{i=1}^{n}\text{log}(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))\), i.e.
-the average log predictive density of the observed data points based on the leave-one-out posterior distributions.
-  This has the advantage over other performance measures like classification accuracy, that it not only takes the location of the predictive distribution into account (i.e.
-the number of correct predictions), but also the width (i.e.
-how confident the model is in its predictions).
-  elpd cannot be computed for frequentist models because the parameters of the model are viewed as fixed values.
-
-  \subsection{Data and data preparation}
+  \subsection{Data}
  \hl{DaHo: das Material in Materials and Methods sind die Daten; die sollten hier kurz beschrieben werden, möglichst mit Referenzen}
@@
 \begin{itemize}
-  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}.
+  \item A large HIV dataset \hl{DaHo: paper?} consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}.
 The dataset contains sequences spanning the \textit{gag}, \textit{pol}, \textit{env}, \textit{nef}, \textit{vif}, \textit{vpr}, \textit{vpu}, \textit{tat} and \textit{rev} genes.
@@
 
-For all sequences, we applied the following data preparation steps:
+\subsection{Data preparation}
+For all sequences, we applied the following preparation steps:
 

### 2021-02-20 docs/main.tex
add correct \end statement to align environment

@@
   \end{aligned}
-\end{equation}
+\end{align}
 where \(\beta_{jk}\) are the regression coefficients; \(\tau\) and \(\lambda_{jk}\) are the so-called global and local shrinkage parameters, respectively;, \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution; $\tau_0$ is the overall degree of sparsity.

### 2021-02-21 docs/main.tex
worked through results

--- 2021-02-20 docs/main.tex
@@
 For our model, we use a so-called horseshoe prior \citep{Carvalho2010}, a scale-mixture of Gaussians:
-\begin{align}
+\begin{equation}
   \label{eq:sparsity-prior}
@@
 
-The full specification of the model is:
+The full specification of the HAMdetector model is:
 
@@
 
-First, we checked with \emph{calibration plots} whether the inferred probabilities \(P(y_{ik}=1|\Psi)\) faithfully reflect the observed data (Fig.~\ref{fig:calibration}).
-All observations were binned according to their inferred probability \(P(y_{ik}=1|\Psi)\).
-If these probabilities are calibrated correctly, we expect that in a group of observations with an inferred probability \(P(y_{ik}=1|\Psi)\) of around x\% we really do observe \(y_{ik}=1\) around x\% percent of the time.
+For a first posterior predictive check we used \emph{calibration plots}: two binned quantities were plotted against each other, the observed relative frequencies of replacements $f(y_{ik}=1)$, and the predicted probabilities $P(y_{ik}=1|\text{model})$.
+In such a plot, a well-calibrated model should yield points following the diagonal.
+Technically, all observations were first sorted by increasing estimated probability $P(y_{ik}=1|\text{model})$ and grouped into $n$ bins.
+For each bin, the fraction of observations with $y_{ik}=1$ (observed event percentage) was then plotted against the midpoint of each bin.
+The cutpoints of the bins are indicated by error bars.
+% We applied calibration plots twice, first with the phylogenetic component of model, and then with the full model.
 
@@
 
-In order to evaluate each building block of HAMdetector, we applied 4 different Bayesian models to each dataset, each of which includes the previous model as a special case:
-A standard logistic regression model, a logistic regression model with horseshoe prior, a model additionally including phylogeny and the full model with horseshoe prior, phylogeny and epitope prediction.
-As a comparison to existing methods, we also applied Fisher's exact test and the Phylogenetic Dependency Network.
+In order to understand what the different building blocks of HAMdetector contribute, we applied four different Bayesian models of increasing complexity to each dataset, starting with the standard logistic regression model (Eq~\ref{eq:backbone}), and adding then the further components, i.e.\ the horseshoe prior (Eq~\ref{eq:sparsity-prior}), phylogeny (Eq~\ref{eq:phylogeny}), and epitope prediction, resulting in the full model (Eq~\ref{eq:full}).
+For comparisons to existing methods, we also applied Fisher's exact test and the Phylogenetic Dependency Network \cite{Carlson2008} to the same data.
 
-\subsection{Model diagnostics}
-\subsubsection{Convergence}
+\subsection{Run times and convergence}
+Run times of HAMdetector on a ... were of the order of ... \hl{DaHo: bitte ergänzen}
 All model fits showed no signs of inference issues.
-In total, samples were drawn from 4 different chains with 1000 iterations each.
-The effective sample size exceeded 500 for all model parameters, \(\hat{R}\) convergence diagnostic values were below 1.01 in all cases.
-\hl{DaHo: citation Gelman + Rubin original paper or later versions if improved $\hat{R}$ have been used}
+In total, samples were drawn from four Hamiltonian Markov chains with 1000 iterations each after 1000 warm-up iterations.
+The effective sample size exceeded 500 for all model parameters, $\hat{R}$ convergence diagnostic values were below 1.01 in all cases.
 
-\subsubsection{Calibration}
-
-\subsection*{Posterior predictive checks}
-
-Figure~\ref{fig:calibration} compares the calibration of the posterior predictive probability of observing the replacement to the actual observed data for the Core protein of the HBV dataset.
-Plots for the other models and datasets are shown in the supplementary.
-
-The plots are similar to the one showed in figure~\ref{fig:phylogeny-calibration}, however, they do not evaluate the probabilities based on the phylogenetic model alone, but instead show the calibration of the posterior predictive probabilities based on the full model.
+\subsection{Posterior predictive checks}
+The model yields well-calibrated posterior predictive probabilities of replacements.
+This is exemplified in Figure~\ref{fig:calibration} for HBV core protein, but also holds true for the other datasets (Supplementary Figures \hl{XXX-YYY}).
 
 \begin{figure}[!ht]
-  \includegraphics[width=1\linewidth]{plots/phylogeny_calibration.pdf}
-  \caption{Calibration plot for the HBV dataset. Calibration plots for the other datasets are shown in the supplementary.
-  All observations are first sorted by increasing estimated probability \(P(y_{ik}=1)\) and then grouped into \(n\) bins.
-  For each bin, the fraction of observations with \(y_{ik}=1\) (observed event percentage) is compared to the midpoint of each bin (the value in the center between the lowest and highest probability).
-The error bars show the cutpoints for each bin.
-If the probabilities are calibrated correctly, each dot is supposed to scatter closely around the diagonal line.}
+ \includegraphics[width=1.0\linewidth]{plots/phylogeny_calibration.pdf}  
+  \caption{Calibration plot for the HBV core protein.}
   \label{fig:calibration}
 \end{figure}
-
-\subsection{Comparison to a list of known epitopes}
 
@@
 
+The predictions of all tested models are strongly enriched in epitopes over baselines, irrespective of dataset (Fig.~\ref{fig:comparison} for HBe protein and Supplementary Figures \hl{XXX-YYY} for other datasets).
+For the best ranked HAMs, Fisher's exact test performs about as well as the HAMdetector backbone logistic regression model (model 1 in Fig.~\ref{fig:comparison}).
+Each of the following three model stages of HAMdetector increases HAM enrichment further, independently of the dataset.
+The horseshoe prior alone (model 2) is a drastic improvement over model 1, even though it does not include any external information.
+The logistic regression model with horseshoe prior works roughly as well as the Phylogenetic Dependency Network \cite{Carlson2008}, which includes much more information.
+Note that the line for the Phylogenetic Dependency Network is shorter because it only outputs associations with a q-value lower than 0.2 by default.
+Model 3 with its additional inclusion of phylogeny has higher enrichment than model 2, and finally, the full model 4 with the inclusion of epitope prediction leads to a further improvement.
+Note that model 4 only uses epitope \textit{prediction} software and does not use any information of experimentally confirmed epitopes.
+These are here only used for model evaluation.
+
 \begin{figure}[ht!]
@@
 PhyloD: Phylogenetic Dependency Network; model 1: simple logistic regression model with broad Student-t priors; model 2: logistic regression model with horseshoe prior; model 3: logistic regression model with horseshoe prior and phylogeny; model 4: full model with epitope prediction.
-The thin gray line are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baseline.}
+Gray lines are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baselines.}
   \label{fig:comparison}
@@
 
-Figure~\ref{fig:comparison} shows a comparison of different models on the HBe protein of the HBV dataset. Plots for all other datasets are shown in the supplementary, but show identical patterns.
-
-All models show a strong enrichment of HLA-associated mutations compared to a randomly permuted list of HLA allele - replacement pairs (gray lines).
-Fisher's exact test performs about as well as the simple logistic regression model, as from a statistical perspective, they are both similar.
-The horseshoe prior alone is a drastic improvement over Fisher's exact test and the logistic regression model with non-sparsifying priors, even though it does not include any external information.
-The benefit comes from including the assumption that most regression coefficients are expected to be very close to 0.
-The logistic regression model with horseshoe prior works roughly as well as the Phylogenetic Dependency Network, which includes much more information.
-Note that the line for the Phylogenetic Dependency Network is shorter because it only outputs associations with a q-value lower than 0.2 by default.
-
-The remaining models are ranked in order of the number of additional features they include.
-Additionally including phylogeny is a small improvement over the model with the horseshoe prior, while adding epitope prediction provides another strong improvement.
-
-Note that this model only uses epitope \textit{prediction} software and does not use any information of experimentally confirmed epitopes. These are here only used for model evaluation.
-
-Figure~\ref{fig:horseshoe-comparison} shows a comparison of marginal posteriors for regression coefficients with and without horseshoe prior.
-
-Supplying information about the expected degree of sparsity with the horseshoe prior is a considerable improvement over a standard logistic regression model  (section~\ref{sec:results}).
+The Bayesian approach lends itself to addition of prior knowledge, and it is not too surprising that exploiting this feature to incorporate information from phylogeny or epitope prediction improves HAM detection.
+It may be more surprising that the sparsifying horseshoe prior has such an impact.
+However, this is in principle the same mechanism as for the other information components: it is known that HAMs are sparse per HLA allele, and therefore supplying this information to the inference improves predictions.
+Figure~\ref{fig:horseshoe-comparison} illustrates the effect of the sparsifying prior with a concrete example, the replacement 11D in HIV integrase (Arevir dataset).
+There is no evidence for an association of HLA-A*01 with this replacement, whereas for HLA-B*44 the data is consistent with a strong association.
+The horseshoe prior has the effect of shrinking towards 0 specifically those regression coefficients with weak evidence of an association (A*01 in Fig.~\ref{fig:horseshoe-comparison}).
+This reduces the standard error for the remaining coefficients, leading in our example to narrowed histogram for the association with B*44 in the model with horseshoe prior.
 
@@
   \includegraphics[width=1\linewidth]{plots/with_horseshoe.pdf}
-  \caption{Marginal posterior distributions of a subset of regression coefficients for replacement 11D of the HIV integrase, based on the Arevir dataset.
-The graphic shows regression coefficients for two HLA alleles: For HLA-A*01, the data does not show evidence of an association with replacement 11D, whereas for HLA-B*44, the data is consistent with a strong association.
-The horseshoe prior has the effect of shrinking regression coefficients with weak evidence of an association with the replacement strongly towards 0, while leaving the coefficients that are far away from 0 mostly unshrunk.
-Shrinking the irrelevant regression coefficients to 0 has the advantage of reducing the standard error for the remaining coefficients, which can be observed by the much smaller width of the histogram of HLA-B*44 in the model with horseshoe prior.}
+  \caption{Marginal posterior distributions of regression coefficients for the association of replacement 11D of the HIV integrase with HLA alleles A*01 and B*44.
+Top half: inferred with logistic regression model, bottom half: inferred with logistic regression with sparsifying horseshoe prior.}
   \label{fig:horseshoe-comparison}
@@
 
-Table \ref{loo} shows a comparison of models with increasing complexity run on the HBe protein in the HDV dataset.
+To quantify the ability of the four different model stages of HAMdetector to generalize to unseen cases, we computed the ELPD with Pareto-smoothed leave-one-out cross-validation.
+Table \ref{loo} shows results for the HBe protein of the HBV dataset in terms of ELPD changes with each new model stage.
+Each new model stage adds ELPD, i.e.\ is better at generalizing than the simpler model stages.
 
@@
   \centering
-  \caption{Differences in elpd for models with increasing model complexity (HBe protein of the HBV dataset).
-  The numbers in column \(\text{elpd}_\text{diff}\) show increases in elpd compared to the previous model (higher is better), e.g.
-the difference in elpd of the model with horseshoe prior is 949.8 compared to the baseline model, and the difference in elpd for the model with phylogeny and horseshoe prior is 4440.9 compared to the model just with the horseshoe prior.
-All differences in elpd are larger than several times the estimated standard error (column \(\text{se}_\text{diff}\)), indicating that models which account for more information are expected to have a better predictive performance.}
+  \caption{ELPD changes as HAMdetector components are added.
+Data computed for HBe protein of the HBV dataset.
+All differences in ELPD are larger than several times the estimated standard error (column \(\text{se}_\text{diff}\)), indicating that models that include more information have better predictive performance.}
   \vspace{0.5cm}
   \begin{tabular}{l|r|rll}
-  \multicolumn{1}{l|}{} & \multicolumn{1}{c|}{\(\text{elpd}_\text{diff}\)} & \multicolumn{1}{c}{\(\text{se}_\text{diff}\)} &  &  \\ \cline{1-3}
+  \multicolumn{1}{l|}{} & \multicolumn{1}{c|}{\(\text{ELPD}_\text{diff}\)} & \multicolumn{1}{c}{\(\text{se}_\text{diff}\)} &  &  \\ \cline{1-3}
   logistic regression (baseline) & 0.0  & 0.0 &  &  \\
@@
 
-Predictive accuracy increases as the models include more and more information.
-Note that the model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any additional external data.
-This is because including the sparsity assumption allows the model to better separate signal from noise and the uncertainty of the close-to-zero coefficients does not propagate into uncertainty of the estimated y.
+The model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any specific external data.
+This is because including the sparsity assumption allows the model to better separate signal from noise and the uncertainty of the close-to-zero coefficients does not propagate into uncertainty of predictions.
 
-Including phylogeny further improves model performance, as the assumption of independent and identically distributed data does not hold for sequence data that share a common phylogenetic history.
+Including phylogeny further improves model performance a lot, as the assumption of independent and identically distributed data is replaced with specific information from the shared phylogenetic history.
 
-Inclusion of epitope prediction does not improve elpd as much as inclusion of phylogeny and the sparsity assumption.
-This is because while addition of sparsity and phylogeny has an effect on all replacements and samples, epitope prediction only influences those replacements which are restricted by a given HLA allele and only those samples which are annotated with the allele.
-However, inclusion of epitope prediction is highly useful for determining which HLA alleles are associated with the occurrence of a replacement, as shown in the previous section.
+While addition of sparsity and phylogeny has an effect on all replacements and samples, epitope prediction only influences those replacements that are restricted by a given HLA allele and only those samples that are annotated with the allele.
+Therefore, inclusion of epitope prediction does not improve ELPD as much as inclusion of phylogeny and the sparsity assumption.
+However, inclusion of epitope prediction is highly useful for determining which HLA alleles are associated with a replacement, as shown in the previous section.
 
-\hl{DaHo: Textblock aus alten Methods:}  We show that a simple logistic regression model with a sparsity-promoting prior on the regression coefficients for the HLA alleles alone already performs roughly on-par  with the much more elaborate Phylogenetic Dependency Network approach from Carlson et al.
-when using the fraction of identified escapes that lie in known HLA epitopes as a benchmark.
+\subsection{HAMs in HDV as test case}
+The Hepatitis D Virus (HDV) dataset \citep{Karimzadeh2019} is an excellent test case: we have (1) a set of paired HDV sequences and patient HLA alleles, (2) HAM predictions by Fisher's exact test as implemented in SeqFeatR \citep{Budeus2016}, and (3) experimental tests of the predicted HAMs for immune escape activity (IFN-\(\gamma\) production assays, \citet{Karimzadeh2019}).
+This allows us to see whether HAMdetector decreases the false positive rate in comparison to the simpler Fisher's exact test, and we can make \emph{bona fide} predictions on previously undetected HAMs.
+\hl{DaHo: zu \emph{bona fide}: können wir dazu etwas sagen?
+Ich erinnere mich, dass Du noch eine Reihe guter HAMs vorhergesagt hattest.}
 
-\subsection{Generalization}
+We have 15 HAMs predicted in HDV by Fisher's exact test at significance level $5\times 10^{-3}$ (Table~\ref{tab:false-positives}) as published \citep{Karimzadeh2019}.
+The corresponding p-values  have no clear relation to experimental confirmation, i.e.\ p-values for confirmed HAMs are not generally lower than those of non-confirmed ones.
 
-\hl{DaHo: Es gibt fast keinen Text in dieser subsection}
-
-The models were applied on sequence data from a variety of different viruses and show a considerable improvement over currently applied methods that are consistent across datasets.
-These results are not due to overfitting, because leave-one-out cross-validation approximates the model performance on unseen data.
-
-\subsection{Example Hepatitis Delta Virus}
-
-We had the opportunity to test our model against a set of associations identified by Fisher's exact test where some of them turned out to be false positives~\cite{Karimzadeh2019}.
-This data is rare because negative results are usually not published in the literature, but provide a useful way to access model performance.
-A comparison with a preliminary version of HAMdetector was published recently.
-Here we provide a comparison with results from Fisher's exact test~\cite{Budeus2016} and the mature version of HAMdetector used in this work.
-
-Fisher's exact test and HAMdetector were applied to the same data (table~\ref{tab:false-positives}).
-The posterior probabilities given by HAMdetector denote the probability that the corresponding regression coefficient is positive.
-Associations with strong support in our model have a posterior probability close to 1, associations with no support a probability close to 0.5 (corresponding to a regression coefficient centered around 0).
-
-The results show low posterior probabilities for most of the false-positives and high posterior probabilities for most replacements that could be experimentally validated in IFN-\(\gamma\) production assays.
-One important exception is P89T~-~HLA-B*37, which was experimentally validated but does not have strong support in our model.
+For HAMdetector, we use in Table~\ref{tab:false-positives} the posterior probability of a positive regression coefficient ($P(\beta_{\text{HLA}_{jk}}>0)$ as measure for the confidence in having detected a HAM.
+HAMs with strong support have a posterior probability close to 1, associations with no support a probability close to 0.5 (corresponding to a regression coefficient centered around 0).
+The five predicted HAMs with top posterior probabilities (all $\ge 0.89$) have all been experimentally confirmed.
+There is only one outlier with posterior probability 0.75 (P89T and B*37).
 
 \begin{table}[h!]
-  \caption{List of HLA associated mutations that were identified using Fisher's exact test (FET).
-Experimental validation showed that 9 out of 15 replacements could not be experimentally validated (column “confirmed“).
-The data were re-analyzed using HAMdetector.
-Except for P89T, all validated replacements have high posterior probabilities, while the replacements that could not be experimentally validated have lower evidence of being HLA associated.
-Note that all replacement - HLA allele pairs had low p-values using Fisher's exact test.}
+  \caption{List of HAMs predicted Fisher's exact test (FET).
+In the last column a ``+'' marks experimental confirmation and a ``-'' a failed confirmation.
+Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector.}
   \vspace{0.5cm}
@@
   \hline
-  S170N       & B*15 & \(3\cdot10^{-8}\)      & 0.99                                & +                        \\
+  S170N       & B*15 & $3\cdot10^{-8}$      & 0.99                                & +                        \\
   D101E       & B*37 & 0.0002                        & 0.96                                & +                        \\
@@
 
- \subsection{Individual examples}
-In the following we illustrate with ... examples the effects of the inclusion of additional relevant information on the detection of HAMs.
+ \subsection{Linkage disequilibrium}
 
  \hl{DaHo: bitte Beispiel mit Linkage Disequilibrium zeigen}
-
-\hl{DaHo: bitte konkretes Bsp mit Auswirkung Phylogenie zeigen}
 
@@
 
-A straight-forward extension of our model would be to also account for these sources of uncertainty, for example by using a Bayesian method to estimate a posterior distribution over possible tree topologies.
+A straightforward extension of our model would be to also account for these sources of uncertainty, for example by using a Bayesian method to estimate a posterior distribution over possible tree topologies.
 The uncertainty over the tree topology and the underlying parameters of the phylogenetic model would then propagate into uncertainty of the estimated probabilities \(P(y_{ik}=1|\Psi)\).

### 2021-02-21 docs/main.tex
read complete manuscript

--- 2021-02-21 docs/main.tex
@@
   y_{ik} & \sim \text{Bernoulli}(\theta_{ik}) \\
-  \theta_{ik} & = \text{logistic}\left(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}}\right),
+  \theta_{ik} & = \text{logistic}\left(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{jk}\right),
 \end{align}
 
-where \(y_{ik}\) is the binary encoded observation of replacement $k$ in viral sequence $i$ (each observed amino acid state $k$ contributes a separate column to $y_{ik}$); \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\); \(X_{ij}\) is 1 if sequence \(i\) comes from host individual HLA allele \(j\) and 0 otherwise; $\beta_{\text{HLA}_{jk}}$ is the HLA regression coefficient of HLA allele \(j\) for replacement \(k\); \(D\) is the number of HLA alleles in the dataset; the logistic inverse link function transforms the linear model in parentheses to the probability scale of $\theta_{ik}$.
+where \(y_{ik}\) is the binary encoded observation of replacement $k$ in viral sequence $i$ (each observed amino acid state $k$ contributes a separate column to $y_{ik}$); \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\); \(X_{ij}\) is 1 if sequence \(i\) comes from host individual HLA allele \(j\) and 0 otherwise; $\beta_{jk}$ is the HLA regression coefficient of HLA allele \(j\) for replacement \(k\); \(D\) is the number of HLA alleles in the dataset; the logistic inverse link function transforms the linear model in parentheses to the probability scale of $\theta_{ik}$.
 
-The main parameters of interests for HAMdetector are the regression coefficients $\beta_{\text{HLA}_{jk}}$, as they quantify the strength of association between the occurrence of replacement \(k\) and each of the observed HLA alleles.
-The $\beta_{\text{HLA}_{jk}}$ are on the log-odds scale, i.e.\ if we go from viral sequences from hosts without HLA allele $j$ to those from hosts with $j$, the log-odds $\log (p_k/(1-p_k))$ of observing replacement \(k\) increase by addition of $\beta_{\text{HLA}_{jk}}$.
+The main parameters of interests for HAMdetector are the regression coefficients $\beta_{jk}$, as they quantify the strength of association between the occurrence of replacement \(k\) and each of the observed HLA alleles.
+The $\beta_{jk}$ are on the log-odds scale, i.e.\ if we go from viral sequences from hosts without HLA allele $j$ to those from hosts with $j$, the log-odds $\log (p_k/(1-p_k))$ of observing replacement \(k\) increase by addition of $\beta_{jk}$.
 
@@
   \theta_{ik}  & =  \text{logistic}\Bigl(\beta_{0_{k}} + \gamma \text{logit}\left(P(y_{ik}=1|\Psi)\right) \\
-  &\;\;\;\;\; + \sum_{k=1}^{D}X_{ik}\beta_{\text{HLA}_{jk}}\Bigr)
+  &\;\;\;\;\; + \sum_{k=1}^{D}X_{ik}\beta_{jk}\Bigr)
   \label{eq:phylogeny}
@@
         \text{logistic}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi)) & \\
-        \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{\text{HLA}_{jk}}) & \\
+        \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{jk}) & \\
     \beta_{0_{k}} &\sim \text{Normal}(0, 100^2) & \text{(*)}\\
@@
    \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2) & \text{(*)} \\
-   \beta_{\text{HLA}_{jk}} &\sim \text{Normal}(0, \tau_{k}^{2}\tilde{\lambda}^{2}_{jk}) & \\
+   \beta_{jk} &\sim \text{Normal}(0, \tau_{k}^{2}\tilde{\lambda}^{2}_{jk}) & \\
    \tilde{\lambda}^{2}_{jk} &= \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau_{k}^{2}\lambda_{jk}^{2}} & \\
@@
 
-The prior on $c_k^2$ implies a Student-t prior with 7 degrees of freedom and a scale of 1 on the non-zero HLA regression coefficients \(\beta_{\text{HLA}_{jk}}\).
+The prior on $c_k^2$ implies a Student-t prior with 7 degrees of freedom and a scale of 1 on the non-zero HLA regression coefficients \(\beta_{jk}\).
 A Student-t prior with these parameters is a reasonable default choice for logistic regression models \citep{Piironen2017}.
@@
 Hence, the degree by which methods for HAM prediction recover this trend is a measure of model performance.
-To implement this measure, we first ranked all evaluated replacements according to their respective credibility of being a HAM, computed as integral of the marginal posterior $P(\beta_{HLA}>0)$.
+To implement this measure, we first ranked all evaluated replacements according to their respective credibility of being a HAM, computed as integral of the marginal posterior $P(\beta_{jk}>0)$.
 For comparison with established methods, namely Fisher's exact test and Phylogenetic Dependency Network \citep{Carlson2008}, ranked lists based on p- and q-values, respectively, were computed.
@@
 
-The model with horseshoe prior alone already has a much higher elpd than the standard logistic regression model, even though it does not use any specific external data.
+The model with horseshoe prior alone already has a much higher ELPD than the standard logistic regression model, even though it does not use any specific external data.
 This is because including the sparsity assumption allows the model to better separate signal from noise and the uncertainty of the close-to-zero coefficients does not propagate into uncertainty of predictions.
@@
 
-For HAMdetector, we use in Table~\ref{tab:false-positives} the posterior probability of a positive regression coefficient ($P(\beta_{\text{HLA}_{jk}}>0)$ as measure for the confidence in having detected a HAM.
+For HAMdetector, we use in Table~\ref{tab:false-positives} the posterior probability of a positive regression coefficient ($P(\beta_{jk}>0)$ as measure for the confidence in having detected a HAM.
 HAMs with strong support have a posterior probability close to 1, associations with no support a probability close to 0.5 (corresponding to a regression coefficient centered around 0).
@@
 
-We have presented HAMdetector, a Bayesian model to identify HLA-associated mutations in a multiple sequence alignment.
-We propose to use models that include as much information as possible and have shown that including sparsity and epitope prediction software can achieve better performance than existing methods.
-The logistic regression backbone can be easily extended, which allows the model to also be used in other contexts, for example in identifying associations between sequence data and other features.
+HAMdetector follows a general paradigm of Bayesian modelling, namely to map all information that is available about a system of interest onto a probabilistic model, and then to apply Bayesian inference to learn about probable parameter values of that model, e.g.\ about $\beta_{jk}$, the association of HLA $j$ with replacement $k$.
+The more relevant information we infuse into the model, the sharper the inference.
+As HAMdetector includes an unprecedented amount of relevant information, it is only natural that it outperforms other methods.
 
-If the quality of the prediction is the same for all viruses, as it appears to be in our results, this method can be used to probe the frequency of HLA immune escape across different viruses.
+We have demonstrated that the logistic regression backbone is a platform that can be extended by model components that contribute new information.
+We have selected such modules guided by widely accepted knowledge, such as phylogeny or epitope location.
+However, even knowledge that is rarely stated explicitly may be helpful in inference, as in the case of sparsity of HLA associations.
+Since the included knowledge is generic for interactions of variable viruses with CTL immunity, HAMdetector performance does not depend on the virus.
 
-In further research we plan to use HAMdetector to study circulating recombinant forms in HIV.
-One possible explanation for the occurrence of these recombinant forms is improved fitness by evading several immune epitopes at once.
+Yet, HAMdetector is far from perfect.
+For instance the outlier in Table \ref{tab:false-positives} could point to missing information in HAMdetector.
+Another deficiency is that it currently works only with two-digit HLA alleles.
+We are currently exploring models for 4-digit HLA alleles that exploit partial pooling so that we can attenuate effects of the increased data fragmentation.
 
-While our model is an improvement over existing methods, a further addition that we plan to implement is partial pooling across 4-digit HLA alleles.
-It has recently been appreciated that binding specificities can vary drastically across  HLA alleles from the same allele group.
-For predicting HLA associated mutations, there has consequently been a shift to use 4-digit resolution data whenever available.
+Another extension of our model would be to better account for phylogenetic uncertainty by using a Bayesian method to estimate a posterior distribution over possible tree topologies.
+The uncertainty over the tree topologies and the underlying parameters of the phylogenetic model would then propagate into uncertainty of the estimated probabilities \(P(y_{ik}=1|\Psi)\).
 
-This is not without downsides however, because overall, binding specificities are more similar for alleles in the same group, and, therefore, treating them as completely separate might unnecessarily fragment the available data.
-
-In Bayesian statistics, it is not required to chose one extreme or the other, With partial pooling, estimates from the same allele group do share some information across each other, but they are also allowed to vary if necessary.
-Therefore, implementing partial pooling might be another possibility to further increase the reliability of computational methods to study HLA escape.
-
-A straightforward extension of our model would be to also account for these sources of uncertainty, for example by using a Bayesian method to estimate a posterior distribution over possible tree topologies.
-The uncertainty over the tree topology and the underlying parameters of the phylogenetic model would then propagate into uncertainty of the estimated probabilities \(P(y_{ik}=1|\Psi)\).
-However, in order to not increase the runtime of the model further we use standard maximum likelihood estimates and then include these in a way that additionally allows for measurement error.
+However, the good performance of the current version of HAMdetector makes it already a valuable tool for the study of interactions between viruses and T cell immunity.
 

### 2021-03-15 docs/main.tex
update main text

@@
-\documentclass{bioinfo}
-\copyrightyear{2015} \pubyear{2015}
+\documentclass{article}
+% \copyrightyear{2015} \pubyear{2015}
 
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Manuscript Category}
+% \access{Advance Access Publication Date: Day Month Year}
+% \appnotes{Manuscript Category}
 \usepackage{soul}
@@
 \begin{document}
-\firstpage{1}
+% \firstpage{1}
 
-\subtitle{Subject Section}
+% \subtitle{Subject Section}
 
-\title[HAMdetector]{HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations}
-\author[Sample \textit{et~al}.]{Corresponding Author\,$^{\text{\sfb 1,}*}$, Co-Author\,$^{\text{\sfb 2}}$ and Co-Author\,$^{\text{\sfb 2,}*}$}
-\address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\      
-$^{\text{\sf 2}}$Department, Institution, City, Post Code,
-Country.}
+% \title[HAMdetector]{HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations}
+% \author[Sample \textit{et~al}.]{Corresponding Author\,$^{\text{\sfb 1,}*}$, Co-Author\,$^{\text{\sfb 2}}$ and Co-Author\,$^{\text{\sfb 2,}*}$}
+% \address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\      
+% $^{\text{\sf 2}}$Department, Institution, City, Post Code,
+% Country.}
 
-\corresp{$^\ast$To whom correspondence should be addressed.}
+% \corresp{$^\ast$To whom correspondence should be addressed.}
 
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+% \history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
 
-\editor{Associate Editor: XXXXXXX}
+% \editor{Associate Editor: XXXXXXX}
 
-\abstract{\textbf{Motivation:}   A key process in anti-viral adaptive immunity is that the human leukocyte antigen system (HLA) presents viral peptide fragments in MHC I protein-peptide complexes on cell surfaces and in this way alerts CD8\textsuperscript{+} cytotoxic T-Lymphocytes (CTLs).
+% \abstract{\textbf{Motivation:}   A key process in anti-viral adaptive immunity is that the human leukocyte antigen system (HLA) presents viral peptide fragments in MHC I protein-peptide complexes on cell surfaces and in this way alerts CD8\textsuperscript{+} cytotoxic T-Lymphocytes (CTLs).
 This pathway exerts a strong selection pressure on viruses, favoring viral mutants that escape recognition by the HLA/CTL system, e.g.\ by point mutations that decrease binding of viral peptides to MHC I.
@@
 The reliable identification of HAMs is not only important to understand viral genomes and their evolution, but it also impacts the development of broadly effective anti-viral treatments and vaccines against variable viruses.\\
-  By their very nature HAMs are amenable to detection by statistical methods in paired sequence / HLA data.
+%   By their very nature HAMs are amenable to detection by statistical methods in paired sequence / HLA data.
 However, HLA alleles are very polymorphic in the human host population which makes the available data relatively sparse and noisy.
@@
 Bayesian inference offers a principled approach to achieve this.\\
-\textbf{Results:} Here, we present a new regression model for the detection of HAMs.
+% \textbf{Results:} Here, we present a new regression model for the detection of HAMs.
 As we choose a Bayesian approach we can include the novel sparsity-inducing priors, and we obtain easily interpretable quantitative information on HAM candidates.
@@
 \\
-\textbf{Availability:}   The source code of this software is available at ... under a permissive MIT license. \\
-\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
-\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
-online.}
+% \textbf{Availability:}   The source code of this software is available at ... under a permissive MIT license. \\
+% \textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
+% \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
+% online.}
 
-\maketitle
+% \maketitle
 
-\section{Introduction}
-\subsection{The HLA system}
+% \section{Introduction}
+% \subsection{The HLA system}
 
-One way of how the human immune system recognizes viral infections is through the human leukocyte antigen system~\citep{Germain1994}: Proteins, including viral proteins, in cells are often degraded in proteasomes to peptide fragments~\citep{Goldberg2002}.
+% One way of how the human immune system recognizes viral infections is through the human leukocyte antigen system~\citep{Germain1994}: Proteins, including viral proteins, in cells are often degraded in proteasomes to peptide fragments~\citep{Goldberg2002}.
 A small subset of these peptides is presented on the cell surface by MHC I receptors, encoded by HLA genes (in humans HLA-A, -B, -C).
@@
 
-Cytotoxic T-Lymphocytes (CTLs) are selected during maturation to only weakly bind to complexes of MHC I with self-peptides, but CTLs can bind strongly to complexes of MHC I with peptides from viral proteins~\citep{Murata2007}.
+% Cytotoxic T-Lymphocytes (CTLs) are selected during maturation to only weakly bind to complexes of MHC I with self-peptides, but CTLs can bind strongly to complexes of MHC I with peptides from viral proteins~\citep{Murata2007}.
 Upon activation, CTLs become cytotoxic and recruit other immune cells so that infected cells can be eliminated~\citep{Harty2000}.
 
-\subsection{HLA escape}
-In this way, the HLA system exerts strong selection pressure towards virus variants that escape T cell recognition~\citep{Borrow1997}.
+% \subsection{HLA escape}
+% In this way, the HLA system exerts strong selection pressure towards virus variants that escape T cell recognition~\citep{Borrow1997}.
 Such variants could, for example, carry mutations that result in reduced binding of viral peptides to MHC I or T cell receptors, or that alter peptide processing so that peptides are no longer presented~\citep{Yewdell2002}.
 
-HLA diversity drives viral evolution in individuals where a virus adapts to specific host features, and in populations and regions that in general differ in HLA frequencies~\citep{Kawashima2009}.
+% HLA diversity drives viral evolution in individuals where a virus adapts to specific host features, and in populations and regions that in general differ in HLA frequencies~\citep{Kawashima2009}.
 Upon transmission to a new host with different HLA alleles, HLA escape mutations may revert, as they are often associated with a reduction in viral replication capacity~\citep{Matthews2008}, but they also may be irreversible~\citep{Kawashima2009}.
 
-Whether and how quickly a given escape mutation is selected in a host depends, e.g., on the viral genomic background, the magnitude of the reduction in viral replication, the availability of compensatory mutations that recover fitness, or the strength of selection pressure~\citep{Kloverpris2016}.
+% Whether and how quickly a given escape mutation is selected in a host depends, e.g., on the viral genomic background, the magnitude of the reduction in viral replication, the availability of compensatory mutations that recover fitness, or the strength of selection pressure~\citep{Kloverpris2016}.
 
-Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV~\citep{Alizon2011, Allen2005, Rousseau2008, Lumley2018}.
+% Immune escape is a driver of viral evolution in populations and patients, and therefore of eminent importance for the understanding of many aspects of highly variable viruses such as HIV or HBV~\citep{Alizon2011, Allen2005, Rousseau2008, Lumley2018}.
 This includes the development of treatments and vaccines that rely on effective immune responses.
@@
 
-\subsection{Identifying HLA escape mutations}
+% \subsection{Identifying HLA escape mutations}
 
-There are several experimental methods available to study HLA escape~\citep{Timm2004}.
+% There are several experimental methods available to study HLA escape~\citep{Timm2004}.
 \hl{DaHo: use key references, e.g.
@@
 
-% IFN-gamma ELISPOT:
-% A  Solid-Phase Enzyme-Linked Immunospot (ELISPOT) Assay for Enumeration of Specific Antibody-Secreting Cells \citep{Czerkinsky1983}
-% Quantification of antigen specific CD8+ T cells using an ELISPOT assay \citep{Miyahira1995}
+% % IFN-gamma ELISPOT:
+% % A  Solid-Phase Enzyme-Linked Immunospot (ELISPOT) Assay for Enumeration of Specific Antibody-Secreting Cells \citep{Czerkinsky1983}
+% % Quantification of antigen specific CD8+ T cells using an ELISPOT assay \citep{Miyahira1995}
 
-% Chromium release
-% Quantitative assay of the lytic action of immune lymphoid cells on 51-Cr-labelled allogeneic target cells in vitro; inhibition by isoantibody and by drugs \citep{Brunner1968}
+% % Chromium release
+% % Quantitative assay of the lytic action of immune lymphoid cells on 51-Cr-labelled allogeneic target cells in vitro; inhibition by isoantibody and by drugs \citep{Brunner1968}
 
-% Intracellular cytokine staining
-% Intracellular cytokine optimization and standard operating procedure \citep{Lamoreaux2006}
-% The assessment of antigen-specific CD8+ T cells through the combination of MHC class I tetramer and intracellular staining \citep{Appay2002}
+% % Intracellular cytokine staining
+% % Intracellular cytokine optimization and standard operating procedure \citep{Lamoreaux2006}
+% % The assessment of antigen-specific CD8+ T cells through the combination of MHC class I tetramer and intracellular staining \citep{Appay2002}
 
-% HLA tetramer staining
-% Phenotypic analysis of antigen-specific T lymphocytes \citep{Altman1996}
+% % HLA tetramer staining
+% % Phenotypic analysis of antigen-specific T lymphocytes \citep{Altman1996}
 
-As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface and thus on binding to the HLA encoded MHC I protein, escape mutations are usually HLA allele specific and can therefore be detected as HLA allele dependent ``footprints'' in sequence alignments of viral proteins  \citep{Moore2002}, i.e.\ as HLA associated mutations (HAM) or replacements enriched in viral sequences from hosts with a specific HLA allele.
+% As the selection pressure exerted by cytotoxic T cells depends on successful recognition of viral peptides on the cell surface and thus on binding to the HLA encoded MHC I protein, escape mutations are usually HLA allele specific and can therefore be detected as HLA allele dependent ``footprints'' in sequence alignments of viral proteins  \citep{Moore2002}, i.e.\ as HLA associated mutations (HAM) or replacements enriched in viral sequences from hosts with a specific HLA allele.
 
-One way of quantifying this enrichment is Fisher's exact test~\citep{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(\neg H\)), (\(\neg R_{i}\), \(H\)) and (\(\neg R_{i}\), \(\neg H\)), where \(\neg R_{i}\) denotes any replacement except \(R_{i}\), and \(\neg H\) denotes any HLA allele except \(H\).
+% One way of quantifying this enrichment is Fisher's exact test~\citep{Fisher1922}: For a given replacement \(R_{i}\) at alignment position \(i\) and HLA allele \(H\), a 2-by-2 contingency table is constructed containing the absolute counts of the number of sequences in the four possible categories  (\(R_{i}\), \(H\)), (\(R_{i}\), \(\neg H\)), (\(\neg R_{i}\), \(H\)) and (\(\neg R_{i}\), \(\neg H\)), where \(\neg R_{i}\) denotes any replacement except \(R_{i}\), and \(\neg H\) denotes any HLA allele except \(H\).
 
-Fisher's exact test is a conventional null hypothesis significance test (NHST) that generates p-values.
+% Fisher's exact test is a conventional null hypothesis significance test (NHST) that generates p-values.
 In this case, the null hypothesis is that HLA allele \(H\) and  replacement \(R_{i}\) are independent, and the p-value is the probability of observing a deviation from independence that is at least as extreme as in the data at hand under the assumption that the null hypothesis is true.
 
-Fisher's exact test has the advantage of being fast and easy to apply~\citep{Budeus2016}, but it also has several disadvantages~\citep{Carlson2008}.
+% Fisher's exact test has the advantage of being fast and easy to apply~\citep{Budeus2016}, but it also has several disadvantages~\citep{Carlson2008}.
 The most striking one is that viral sequences share a common phylogenetic history, and,  therefore, treating sequences as independent and identically distributed samples may under- or overestimate effect sizes.
@@
 I have not found any article on the violation of independence in fisher's exact test specifically.} \citep{Osborne2002} \citep{Scariano1987}
+% % scariano nehmen
 
-Another issue with Fisher's exact test is the genomic proximity of human HLA class I loci (on chromosome 6) leading to linkage disequilibrium -- inheritance of HLA alleles can be correlated.
+% Another issue with Fisher's exact test is the genomic proximity of human HLA class I loci (on chromosome 6) leading to linkage disequilibrium -- inheritance of HLA alleles can be correlated.
 Therefore spurious associations can occur if associations of replacements with individual HLA alleles is tested: if HLA allele \(H_1\) is associated with an amino acid replacement \(R\) because of immune escape, but \(H_1\) is in linkage disequilibrium with allele \(H_2\), this leads to an association of \(R\) and \(H_2\), even without being an escape mutation from $H_2$.
 
-\citet{Carlson2008} developed the Phylogenetic Dependency Network, a method that accounts for several of the aforementioned problems, in particular phylogenetic bias and  HLA linkage disequilibrium.
+% \citet{Carlson2008} developed the Phylogenetic Dependency Network, a method that accounts for several of the aforementioned problems, in particular phylogenetic bias and  HLA linkage disequilibrium.
 However, it is based on null hypothesis significance testing.
 
-\subsection{Issues with p-values for screening}
+% \subsection{Issues with p-values for screening}
 
-There are fundamental statistical issues with p-values as screening tool~\citep{Amrhein2017}:
-with small effect sizes and high variance between measurements, as is often the case with biological data, statistically significant results can be misleading, have the wrong direction (type S error), or greatly overestimate an effect (type M error)~\citep{Gelman2014}.
+% There are fundamental statistical issues with p-values as screening tool~\citep{Amrhein2017}:
+% with small effect sizes and high variance between measurements, as is often the case with biological data, statistically significant results can be misleading, have the wrong direction (type S error), or greatly overestimate an effect (type M error)~\citep{Gelman2014}.
 Such problems are more and more appreciated in the context of the current ``replication crisis'', which describes that scientific claims with seemingly strong statistical evidence fail to replicate~\citep{Ioannidis2005, Begley:2012, Baker2016Nature-reproducibility-crisis}.
 
-These problems are exacerbated if the p-values are used for screening purposes (multiple testing problem).
+% These problems are exacerbated if the p-values are used for screening purposes (multiple testing problem).
 The probability  of  obtaining a statistically significant result increases with each additional test, even in absence of any real effect.
@@
 
-Instead of performing many hypothesis tests and trying to adjust for them, we prefer to fit a single, multilevel model that contains all comparisons of interest.
+% Instead of performing many hypothesis tests and trying to adjust for them, we prefer to fit a single, multilevel model that contains all comparisons of interest.
 When using multilevel models, the problem of multiple comparisons can disappear entirely and yield more valid estimates~\citep{Gelman2012}.
 
-\begin{methods}
-  \section{Materials and methods}
+% \begin{methods}
+%   \section{Materials and methods}
 
-  Our general approach for HAMdetector is to fit Bayesian regression models that captures relationships between host HLA alleles and replacements in viral proteomes.
+%   Our general approach for HAMdetector is to fit Bayesian regression models that captures relationships between host HLA alleles and replacements in viral proteomes.
 This Bayesian approach is advantageous because it allows use of: (1) prior information (e.g.\ knowledge of effect magnitudes), (2) relevant additional information (phylogeny, epitope information), (3) a problem-specific structure, (4) partial pooling~\citep{Gelman2010}.
 
-\subsection{Model backbone}
+% \subsection{Model backbone}
 
-We chose a logistic regression model as backbone because it is easily extensible, and because coefficients can be interpreted in the familiar way as summands on the log-odds scale.
+% We chose a logistic regression model as backbone because it is easily extensible, and because coefficients can be interpreted in the familiar way as summands on the log-odds scale.
 
-\begin{align}
-\label{eq:backbone}
-  y_{ik} & \sim \text{Bernoulli}(\theta_{ik}) \\
-  \theta_{ik} & = \text{logistic}\left(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{jk}\right),
-\end{align}
+% \begin{align}
+% \label{eq:backbone}
+%   y_{ik} & \sim \text{Bernoulli}(\theta_{ik}) \\
+%   \theta_{ik} & = \text{logistic}\left(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{jk}\right),
+% \end{align}
 
-where \(y_{ik}\) is the binary encoded observation of replacement $k$ in viral sequence $i$ (each observed amino acid state $k$ contributes a separate column to $y_{ik}$); \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\); \(X_{ij}\) is 1 if sequence \(i\) comes from host individual HLA allele \(j\) and 0 otherwise; $\beta_{jk}$ is the HLA regression coefficient of HLA allele \(j\) for replacement \(k\); \(D\) is the number of HLA alleles in the dataset; the logistic inverse link function transforms the linear model in parentheses to the probability scale of $\theta_{ik}$.
+% where \(y_{ik}\) is the binary encoded observation of replacement $k$ in viral sequence $i$ (each observed amino acid state $k$ contributes a separate column to $y_{ik}$); \(\theta_{ik}\) is the estimated probability that we observe replacement \(k\) in sequence \(i\); \(X_{ij}\) is 1 if sequence \(i\) comes from host individual HLA allele \(j\) and 0 otherwise; $\beta_{jk}$ is the HLA regression coefficient of HLA allele \(j\) for replacement \(k\); \(D\) is the number of HLA alleles in the dataset; the logistic inverse link function transforms the linear model in parentheses to the probability scale of $\theta_{ik}$.
 
-The main parameters of interests for HAMdetector are the regression coefficients $\beta_{jk}$, as they quantify the strength of association between the occurrence of replacement \(k\) and each of the observed HLA alleles.
+% The main parameters of interests for HAMdetector are the regression coefficients $\beta_{jk}$, as they quantify the strength of association between the occurrence of replacement \(k\) and each of the observed HLA alleles.
 The $\beta_{jk}$ are on the log-odds scale, i.e.\ if we go from viral sequences from hosts without HLA allele $j$ to those from hosts with $j$, the log-odds $\log (p_k/(1-p_k))$ of observing replacement \(k\) increase by addition of $\beta_{jk}$.
 
-% Reasoning about coefficients on the log-odds scale can sometimes be unintuitive.
+% % Reasoning about coefficients on the log-odds scale can sometimes be unintuitive.
 A useful approximation to interpret logistic regression coefficients on the probability scale is the so-called divide-by-4 rule, which means that a regression coefficient of 2 corresponds to an expected increase on the probability scale of about 2/4 = 50\%.
@@
 
-\subsection{Inclusion of additional information}
+% \subsection{Inclusion of additional information}
 
-On top of the paired data of viral sequences and host HLA alleles modeled by the backbone (Eq~\ref{eq:backbone}), we extend the model to include further information of relevance to improve HAM detection, namely phylogenetic information, and predictions of epitope peptide processing and MHC I affinity, as described in the following.
+% On top of the paired data of viral sequences and host HLA alleles modeled by the backbone (Eq~\ref{eq:backbone}), we extend the model to include further information of relevance to improve HAM detection, namely phylogenetic information, and predictions of epitope peptide processing and MHC I affinity, as described in the following.
 
-\subsubsection{Phylogeny}
-One problem that occurs when analyzing sequence data is that species (here: viral strains) share a common phylogenetic history.
+% \subsubsection{Phylogeny}
+% One problem that occurs when analyzing sequence data is that species (here: viral strains) share a common phylogenetic history.
 Thus replacements are not independently and identically distributed, and therefore violate a common assumption of standard statistical methods.
@@
 
-The most popular approach in phylogeny-aware regression of binary variables is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree~\citep{Ives2009, Ives2014}.
-This approach turned out to be too computationally expensive in our model, hence we chose a strategy similar to the one in \citet{Carlson2008}:
+% The most popular approach in phylogeny-aware regression of binary variables is to estimate an additional multivariate normally distributed intercept, where the covariance matrix is based on the branch lengths of a given phylogenetic tree~\citep{Ives2009, Ives2014}.
+% This approach turned out to be too computationally expensive in our model, hence we chose a strategy similar to the one in \citet{Carlson2008}:
 
-Consider a phylogenetic tree \(\Psi\) obtained from standard maximum likelihood methods for a given multiple sequence alignment.
+% Consider a phylogenetic tree \(\Psi\) obtained from standard maximum likelihood methods for a given multiple sequence alignment.
 We are interested in estimating \(P(y_{ik}=1|\Psi)\), that is, the probability of observing the replacement \(k\) in sequence \(i\) based on the underlying phylogenetic model.
@@
 
-\begin{equation}
-\begin{aligned}
-  y_{ik}  & \sim \text{Bernoulli}(\theta_{ik}) \\
-  \theta_{ik}  & =  \text{logistic}\Bigl(\beta_{0_{k}} + \gamma \text{logit}\left(P(y_{ik}=1|\Psi)\right) \\
-  &\;\;\;\;\; + \sum_{k=1}^{D}X_{ik}\beta_{jk}\Bigr)
-  \label{eq:phylogeny}
-\end{aligned}
-\end{equation}
+% \begin{equation}
+% \begin{aligned}
+%   y_{ik}  & \sim \text{Bernoulli}(\theta_{ik}) \\
+%   \theta_{ik}  & =  \text{logistic}\Bigl(\beta_{0_{k}} + \gamma \text{logit}\left(P(y_{ik}=1|\Psi)\right) \\
+%   &\;\;\;\;\; + \sum_{k=1}^{D}X_{ik}\beta_{jk}\Bigr)
+%   \label{eq:phylogeny}
+% \end{aligned}
+% \end{equation}
 
-The logit transform is used because it cancels out with the logistic inverse link function.
+% The logit transform is used because it cancels out with the logistic inverse link function.
 The phylogeny term acts as a baseline in absence of any HLA effects.
@@
 
-\subsubsection{Inclusion of CTL epitope predictions}
+% \subsubsection{Inclusion of CTL epitope predictions}
 
-% A major CTL escape mechanism is the generation of mutations in viral peptides that prevent peptide presentation by MHC I to T cell receptors.
+% % A major CTL escape mechanism is the generation of mutations in viral peptides that prevent peptide presentation by MHC I to T cell receptors.
 
-As outlined earlier, escape mutations often appear as HAMs.
+% As outlined earlier, escape mutations often appear as HAMs.
 Given the underlying mechanism, it is not surprising that escape mutations are enriched in CTL epitopes, i.e.\ in those viral peptides presented by MHC I to T cell receptors \citep{Bronke2013}.
@@
 
-Not only mutations in CTL epitopes can lead to failure of MHC I to present epitopes, but also mutations at epitope-flanking positions that interfere with pre-processing of peptides, notably proteasomal cleavage of viral proteins \hl{DaHo: citation}.
+% Not only mutations in CTL epitopes can lead to failure of MHC I to present epitopes, but also mutations at epitope-flanking positions that interfere with pre-processing of peptides, notably proteasomal cleavage of viral proteins \hl{DaHo: citation}.
 
-In HAMdetector we use MHCFlurry 2.0~\citep{ODonnell2020} to predict epitopes that are properly processed and presented by MHC I.
+% In HAMdetector we use MHCFlurry 2.0~\citep{ODonnell2020} to predict epitopes that are properly processed and presented by MHC I.
 For this, we create an input matrix of dimensions $R\times D$, where $R$ is the number of evaluated replacements and $D$ is the number of observed HLA alleles in the dataset.
@@
 
-We use epitope prediction as information about the expected degree of sparsity, i.e.\ if we know that a certain position is restricted by a certain HLA allele, we expect that this HLA allele is more likely to be associated with replacements at that position than the other HLA alleles.
+% We use epitope prediction as information about the expected degree of sparsity, i.e.\ if we know that a certain position is restricted by a certain HLA allele, we expect that this HLA allele is more likely to be associated with replacements at that position than the other HLA alleles.
 This idea is implemented by increasing the scale of the local shrinkage parameters \(\lambda_{jk}\) depending on epitope information:
 
-\begin{equation}
-  \begin{aligned}
-    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \text{exp}(Z_{jk}\beta_{\text{epi}})) \\
-    \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2),
-  \end{aligned}
-\end{equation}
+% \begin{equation}
+%   \begin{aligned}
+%     \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \text{exp}(Z_{jk}\beta_{\text{epi}})) \\
+%     \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2),
+%   \end{aligned}
+% \end{equation}
 
-where \(Z_{jk}\) is 1 if HLA allele \(j\) is predicted to restrict the alignment position corresponding to replacement \(k\), and 0 otherwise.
+% where \(Z_{jk}\) is 1 if HLA allele \(j\) is predicted to restrict the alignment position corresponding to replacement \(k\), and 0 otherwise.
 The parameter \(\beta_{\text{epi}}\) governs the increase in scale of the  corresponding local shrinkage parameters.
@@
 
-\hl{DaHo: komplexer Zusammenhang von pre-processing escape und HLA; bitte besprechen}
+% % \hl{DaHo: komplexer Zusammenhang von pre-processing escape und HLA; bitte besprechen}
 
-\subsubsection{Sparsity-inducing priors}
+% \subsubsection{Sparsity-inducing priors}
 
-Sparsity-promoting priors \citep{Piironen2017} can drastically improve predictive performance, because the model is better able to differentiate between signal and noise.
+% Sparsity-promoting priors \citep{Piironen2017} can drastically improve predictive performance, because the model is better able to differentiate between signal and noise.
 These priors convey the a priori expectation that most coefficients in a regression model are close to 0, i.e.\ that non-zero coefficients are sparse.
@@
 
-There is a range of sparsity-promoting priors with slightly different properties.
+% There is a range of sparsity-promoting priors with slightly different properties.
 They share the common structure of placing most probability mass very close to 0, with heavy tails to accommodate the non-zero coefficients.
-For our model, we use a so-called horseshoe prior \citep{Carvalho2010}, a scale-mixture of Gaussians:
-\begin{equation}
-  \label{eq:sparsity-prior}
-  \begin{aligned}
-    \beta_{jk} &\sim \text{Normal}(0, \tau^{2}\lambda^{2}_{jk}) \\
-    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, 1) \\
-    \tau &\sim \text{Cauchy}^{+}(0, \tau_{0}),
-  \end{aligned}
-\end{equation}
-where \(\beta_{jk}\) are the regression coefficients; \(\tau\) and \(\lambda_{jk}\) are the so-called global and local shrinkage parameters, respectively;, \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution; $\tau_0$ is the overall degree of sparsity.
+% For our model, we use a so-called horseshoe prior \citep{Carvalho2010}, a scale-mixture of Gaussians:
+% \begin{equation}
+%   \label{eq:sparsity-prior}
+%   \begin{aligned}
+%     \beta_{jk} &\sim \text{Normal}(0, \tau^{2}\lambda^{2}_{jk}) \\
+%     \lambda_{jk} &\sim \text{Cauchy}^{+}(0, 1) \\
+%     \tau &\sim \text{Cauchy}^{+}(0, \tau_{0}),
+%   \end{aligned}
+% \end{equation}
+% where \(\beta_{jk}\) are the regression coefficients; \(\tau\) and \(\lambda_{jk}\) are the so-called global and local shrinkage parameters, respectively;, \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution; $\tau_0$ is the overall degree of sparsity.
 
-With Eq~\ref{eq:sparsity-prior} the global shrinkage parameter \(\tau\) is typically very small and shrinks most of the regression coefficients close to 0, whereas the local shrinkage parameters $\lambda_{jk}$ can occasionally be very large to allow some coefficients to escape that shrinkage.
+% With Eq~\ref{eq:sparsity-prior} the global shrinkage parameter \(\tau\) is typically very small and shrinks most of the regression coefficients close to 0, whereas the local shrinkage parameters $\lambda_{jk}$ can occasionally be very large to allow some coefficients to escape that shrinkage.
 
-The overall degree of sparsity $\tau_0$ can be chosen based on the expected number of non-zero coefficients \citep{Piironen2017}. \hl{DaHo: do we do this? How?}
+% The overall degree of sparsity $\tau_0$ can be chosen based on the expected number of non-zero coefficients \citep{Piironen2017}. \hl{DaHo: do we do this? How?}
 
-\subsection{Full model specification}
+% \subsection{Full model specification}
 
-The full specification of the HAMdetector model is:
+% The full specification of the HAMdetector model is:
 
-\begin{equation}
-  \label{eq:full}
-  \begin{aligned}
-   y_{ik} &\sim \text{Bernoulli}(\theta_{ik}) &\\
-   \theta_{ik} &=
-        \text{logistic}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi)) & \\
-        \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{jk}) & \\
-    \beta_{0_{k}} &\sim \text{Normal}(0, 100^2) & \text{(*)}\\
-   \gamma_{k} &\sim \text{Normal}(\mu_{\text{phy}}, \sigma_{\text{phy}}^2) & \\
-    \mu_{\text{phy}} &\sim \text{Normal}(1, 1) & \text{(*)} \\
-   \sigma_{\text{phy}} &\sim \text{Normal}^{+}(0, 0.5) & \text{(*)} \\
-   \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2) & \text{(*)} \\
-   \beta_{jk} &\sim \text{Normal}(0, \tau_{k}^{2}\tilde{\lambda}^{2}_{jk}) & \\
-   \tilde{\lambda}^{2}_{jk} &= \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau_{k}^{2}\lambda_{jk}^{2}} & \\
-    c_{k}^2 &\sim \text{Inv-Gamma}(3.5, 3.5) & \text{(*)} \\
-   \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}^2\text{exp}(Z_{jk}\beta_{\text{epi}})) & \\
-   \tau_{k} &\sim \text{Cauchy}^{+}(0, \tau_{0k}) & \text{(*)} \\
-   \tau_{0k} &= \frac{10}{D-10}\frac{2}{\sqrt{N}} & \\
-  \end{aligned}
-\end{equation}
+% \begin{equation}
+%   \label{eq:full}
+%   \begin{aligned}
+%    y_{ik} &\sim \text{Bernoulli}(\theta_{ik}) &\\
+%    \theta_{ik} &=
+%         \text{logistic}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi)) & \\
+%         \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{jk}) & \\
+%     \beta_{0_{k}} &\sim \text{Normal}(0, 100^2) & \text{(*)}\\
+%    \gamma_{k} &\sim \text{Normal}(\mu_{\text{phy}}, \sigma_{\text{phy}}^2) & \\
+%     \mu_{\text{phy}} &\sim \text{Normal}(1, 1) & \text{(*)} \\
+%    \sigma_{\text{phy}} &\sim \text{Normal}^{+}(0, 0.5) & \text{(*)} \\
+%    \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2) & \text{(*)} \\
+%    \beta_{jk} &\sim \text{Normal}(0, \tau_{k}^{2}\tilde{\lambda}^{2}_{jk}) & \\
+%    \tilde{\lambda}^{2}_{jk} &= \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau_{k}^{2}\lambda_{jk}^{2}} & \\
+%     c_{k}^2 &\sim \text{Inv-Gamma}(3.5, 3.5) & \text{(*)} \\
+%    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}^2\text{exp}(Z_{jk}\beta_{\text{epi}})) & \\
+%    \tau_{k} &\sim \text{Cauchy}^{+}(0, \tau_{0k}) & \text{(*)} \\
+%    \tau_{0k} &= \frac{10}{D-10}\frac{2}{\sqrt{N}} & \\
+%   \end{aligned}
+% \end{equation}
 
-The full model specification includes some aspects that were not covered in the previous sections.
+% The full model specification includes some aspects that were not covered in the previous sections.
 In particular, the overall phylogeny-weight $\gamma$ in Eq~\ref{eq:backbone} is replaced in the full model by  hierarchically modelled  \(\gamma_{k}\), which allows partial pooling across replacements (even with a global parameter \(\gamma\) the model works reasonably well).
@@
 
-\subsubsection{Prior justification}
+% \subsubsection{Prior justification}
 
-Prior distributions are labeled with an asterisk in Eq~\ref{eq:full}.
+% Prior distributions are labeled with an asterisk in Eq~\ref{eq:full}.
 They are weakly informative, which means that they effectively limit posteriors to realistic magnitudes of parameters.
@@
 
-The hierarchical mean and standard deviation of the phylogeny coefficients \(\gamma_k\) place most probability mass on \(\gamma_k\) values around 1.
+% The hierarchical mean and standard deviation of the phylogeny coefficients \(\gamma_k\) place most probability mass on \(\gamma_k\) values around 1.
 In absence of any HLA effects, a $\gamma_k=1$ would mean that the estimate for the probability of observing replacement $k$ is identical to the probability based on the phylogenetic model.
@@
 
-The prior on $c_k^2$ implies a Student-t prior with 7 degrees of freedom and a scale of 1 on the non-zero HLA regression coefficients \(\beta_{jk}\).
+% The prior on $c_k^2$ implies a Student-t prior with 7 degrees of freedom and a scale of 1 on the non-zero HLA regression coefficients \(\beta_{jk}\).
 A Student-t prior with these parameters is a reasonable default choice for logistic regression models \citep{Piironen2017}.
 
-The value of \(\tau_{0k}\) implies 10 effective non-zero HLA regression coefficients per replacement.
+% The value of \(\tau_{0k}\) implies 10 effective non-zero HLA regression coefficients per replacement.
 The rationale behind this parameterization is again outlined in \citet{Piironen2017}.
 The value of 10 corresponds to a generously estimated magnitude based on available HIV epitope maps \hl{DaHo: citation?}.
-The model is also parameterized in a way that assumes an equal degree of sparsity across all alignment positions a priori.
+% The model is also parameterized in a way that assumes an equal degree of sparsity across all alignment positions a priori.
 We also tried to model \(\tau_k\) hierarchically, but observed sampling issues due to the resulting unfavorable geometry of the posterior.
 
-\subsection{Model implementation}
-All models were implemented in Stan 2.23 \citep{Stan2021}, a probabilistic programming language and Hamiltonian Monte Carlo sampler for efficient numerical computation of posterior distributions.
+% \subsection{Model implementation}
+% All models were implemented in Stan 2.23 \citep{Stan2021}, a probabilistic programming language and Hamiltonian Monte Carlo sampler for efficient numerical computation of posterior distributions.
 The Stan code is available \hl{DaHo: at ...} in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
@@
 
-\subsection{Model diagnostics}
+% \subsection{Model diagnostics}
 
-\subsubsection{Convergence diagnostics}
-We use the split-\(\hat{R}\) convergence diagnostic to identify Markov chain convergence issues \citep{Gelman1992, Stan2021}.
+% \subsubsection{Convergence diagnostics}
+% We use the split-\(\hat{R}\) convergence diagnostic to identify Markov chain convergence issues \citep{Gelman1992, Stan2021}.
 We require a value of \(\hat{R}\) below 1.01 for all model parameters.
@@
 
-\subsubsection{Posterior predictive checks}
+% \subsubsection{Posterior predictive checks}
 
-In posterior predictive checks, we simulate new data from the inferred posterior distribution and the likelihood, and we compare these simulated data with representative real data \citep{Gabry2019}.
+% In posterior predictive checks, we simulate new data from the inferred posterior distribution and the likelihood, and we compare these simulated data with representative real data \citep{Gabry2019}.
 A good model should predict data that are consistent with real data.
@@
 
-For a first posterior predictive check we used \emph{calibration plots}: two binned quantities were plotted against each other, the observed relative frequencies of replacements $f(y_{ik}=1)$, and the predicted probabilities $P(y_{ik}=1|\text{model})$.
+% For a first posterior predictive check we used \emph{calibration plots}: two binned quantities were plotted against each other, the observed relative frequencies of replacements $f(y_{ik}=1)$, and the predicted probabilities $P(y_{ik}=1|\text{model})$.
 In such a plot, a well-calibrated model should yield points following the diagonal.
@@
 For each bin, the fraction of observations with $y_{ik}=1$ (observed event percentage) was then plotted against the midpoint of each bin.
-The cutpoints of the bins are indicated by error bars.
-% We applied calibration plots twice, first with the phylogenetic component of model, and then with the full model.
+The cut-points of the bins are indicated by error bars.
+% % We applied calibration plots twice, first with the phylogenetic component of model, and then with the full model.
 
-Second, we assessed the abilities of different models and methods to discover HAMs with \emph{HAM enrichment plots}.
+% Second, we assessed the abilities of different models and methods to discover HAMs with \emph{HAM enrichment plots}.
 These plots are based on the observation that CTL escape mutations are enriched in epitopes \citep{Bronke2013}.
@@
 
-\subsubsection{Leave-one-out cross-validation}
+% \subsubsection{Leave-one-out cross-validation}
 
-Another performance measure is the ability to generalize to unseen data.
+% Another performance measure is the ability to generalize to unseen data.
 To examine this ability for the different model variants we performed leave-one-out cross-validation (LOOCV), using the efficient Pareto-smoothed LOOCV \cite{Vehtari2016}.
 
-% Several utility (or cost) functions are used in practice to compare the predictive performance of models.
+% % Several utility (or cost) functions are used in practice to compare the predictive performance of models.
 For frequentist approaches, classification accuracy or receiver operator characteristic curves (ROC curves) are most common.
-From the LOOCV we obtain the Expected Log-Predictive Density (ELPD) $\sum_{i=1}^{n}\log(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))$ for samples $i=1,\ldots,n$, $i$th observation $y_i$, data $y_{i-1}$ with $i$ left out, and model parameters $\theta$.
+% From the LOOCV we obtain the Expected Log-Predictive Density (ELPD) $\sum_{i=1}^{n}\log(\int p(y_i|\theta)p(\theta|y_{i-1})d\theta))$ for samples $i=1,\ldots,n$, $i$th observation $y_i$, data $y_{i-1}$ with $i$ left out, and model parameters $\theta$.
 Thus, the ELPD is the average log predictive density of the observed data points based on the leave-one-out posterior distributions.
@@
 
-  \subsection{Data}
- \hl{DaHo: das Material in Materials and Methods sind die Daten; die sollten hier kurz beschrieben werden, möglichst mit Referenzen}
+%   \subsection{Data}
+%  \hl{DaHo: das Material in Materials and Methods sind die Daten; die sollten hier kurz beschrieben werden, möglichst mit Referenzen}
 
-The model was fit with several data sets:
+% The model was fit with several data sets:
 
-\begin{itemize}
-  \item A large HIV dataset \hl{DaHo: paper?} consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}.
+% \begin{itemize}
+%   \item A large HIV dataset \hl{DaHo: paper?} consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}.
 The dataset contains sequences spanning the \textit{gag}, \textit{pol}, \textit{env}, \textit{nef}, \textit{vif}, \textit{vpr}, \textit{vpu}, \textit{tat} and \textit{rev} genes.
-  \item A set of 351 HIV sequences mostly spanning the \textit{pol} gene from the Arevir database~\citep{Roomp2006}.
-  \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm. The dataset contains sequences of the HBe, LHBs, Pol and HBx proteins.
-  \item A set of 104 Hepatitis-D-Virus sequences containing the HDV-antigen that were kindly provided by Michael Roggendorf. \hl{DaHo: bitte von Hadi geeignetes Zitat erfragen}
-  \item A set of 41 HIV sequences spanning the \textit{gag} and \textit{pol} genes that were kindly provided by Rongge Yang. 
-\end{itemize}
+%   \item A set of 351 HIV sequences mostly spanning the \textit{pol} gene from the Arevir database~\citep{Roomp2006}.
+%   \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm. The dataset contains sequences of the HBe, LHBs, Pol and HBx proteins.
+%   \item A set of 104 Hepatitis-D-Virus sequences containing the HDV-antigen that were kindly provided by Michael Roggendorf. \hl{DaHo: bitte von Hadi geeignetes Zitat erfragen}
+%   \item A set of 41 HIV sequences spanning the \textit{gag} and \textit{pol} genes that were kindly provided by Rongge Yang. 
+% \end{itemize}
 
-\hl{DaHo: sources of known epitopes?}
+% \hl{DaHo: sources of known epitopes?}
 
-\subsection{Data preparation}
-For all sequences, we applied the following preparation steps:
+% \subsection{Data preparation}
+% For all sequences, we applied the following preparation steps:
 
-\begin{enumerate}
-  \item For each dataset, the sequences were split into subsequences, either by protein or gene.
-  \item If not already present in this format, sequences were translated into their amino acid representation.
-  \item RAxML-NG~\citep{Kozlov2019} version 1.0.0 was used to generate a maximum likelihood phylogenetic tree for each gene/protein using the \texttt{--model GTR+G+I} option with all other parameters set to default values.
+% \begin{enumerate}
+%   \item For each dataset, the sequences were split into subsequences, either by protein or gene.
+%   \item If not already present in this format, sequences were translated into their amino acid representation.
+%   \item RAxML-NG~\citep{Kozlov2019} version 1.0.0 was used to generate a maximum likelihood phylogenetic tree for each gene/protein using the \texttt{--model GTR+G+I} option with all other parameters set to default values.
 If available, we used RNA or DNA sequences for this step, rather than protein sequences.
-\end{enumerate}
+% \end{enumerate}
 
-\end{methods}
+% \end{methods}
 
-\section{Results} \label{sec:results}
+% \section{Results} \label{sec:results}
 
-In order to understand what the different building blocks of HAMdetector contribute, we applied four different Bayesian models of increasing complexity to each dataset, starting with the standard logistic regression model (Eq~\ref{eq:backbone}), and adding then the further components, i.e.\ the horseshoe prior (Eq~\ref{eq:sparsity-prior}), phylogeny (Eq~\ref{eq:phylogeny}), and epitope prediction, resulting in the full model (Eq~\ref{eq:full}).
+% In order to understand what the different building blocks of HAMdetector contribute, we applied four different Bayesian models of increasing complexity to each dataset, starting with the standard logistic regression model (Eq~\ref{eq:backbone}), and adding then the further components, i.e.\ the horseshoe prior (Eq~\ref{eq:sparsity-prior}), phylogeny (Eq~\ref{eq:phylogeny}), and epitope prediction, resulting in the full model (Eq~\ref{eq:full}).
 For comparisons to existing methods, we also applied Fisher's exact test and the Phylogenetic Dependency Network \cite{Carlson2008} to the same data.
 
-\subsection{Run times and convergence}
-Run times of HAMdetector on a ... were of the order of ... \hl{DaHo: bitte ergänzen}
-All model fits showed no signs of inference issues.
+% \subsection{Run times and convergence}
+% Run times of HAMdetector on a ... were of the order of ... \hl{DaHo: bitte ergänzen}
+% All model fits showed no signs of inference issues.
 In total, samples were drawn from four Hamiltonian Markov chains with 1000 iterations each after 1000 warm-up iterations.
@@
 
-\subsection{Posterior predictive checks}
-The model yields well-calibrated posterior predictive probabilities of replacements.
+% \subsection{Posterior predictive checks}
+% The model yields well-calibrated posterior predictive probabilities of replacements.
 This is exemplified in Figure~\ref{fig:calibration} for HBV core protein, but also holds true for the other datasets (Supplementary Figures \hl{XXX-YYY}).
 
-\begin{figure}[!ht]
- \includegraphics[width=1.0\linewidth]{plots/phylogeny_calibration.pdf}  
-  \caption{Calibration plot for the HBV core protein.}
-  \label{fig:calibration}
-\end{figure}
+% \begin{figure}[!ht]
+%  \includegraphics[width=1.0\linewidth]{plots/phylogeny_calibration.pdf}  
+%   \caption{Calibration plot for the HBV core protein.}
+%   \label{fig:calibration}
+% \end{figure}
 
-%However, comparing against real-world data is not straight forward when identifying HLA-associated mutations, because if a tool identifies an HLA-associated mutation which is not yet documented, this could be either because it is a false positive, or a yet unknown HLA-associated mutation that has not been documented before.
-%Also, viewing every non-identified HLA-associated mutation that is listed in the literature as a false negative is not correct either, because documented escape mutations do not necessarily have to show in each sequence alignment.
-%A third issue when comparing models is that HLA-associated mutations are not necessarily binary, so the true positive / false negative framework does not work as well.
+% %However, comparing against real-world data is not straight forward when identifying HLA-associated mutations, because if a tool identifies an HLA-associated mutation which is not yet documented, this could be either because it is a false positive, or a yet unknown HLA-associated mutation that has not been documented before.
+% %Also, viewing every non-identified HLA-associated mutation that is listed in the literature as a false negative is not correct either, because documented escape mutations do not necessarily have to show in each sequence alignment.
+% %A third issue when comparing models is that HLA-associated mutations are not necessarily binary, so the true positive / false negative framework does not work as well.
 
-%To circumvent these issues and still compare against real-world data, we chose the following strategy:
-%For each model, all evaluated replacements are ranked by decreasing confidence of being an HLA associated mutation.
+% %To circumvent these issues and still compare against real-world data, we chose the following strategy:
+% %For each model, all evaluated replacements are ranked by decreasing confidence of being an HLA associated mutation.
 That is, for Bayesian models we calculate the probability of the corresponding regression coefficient exceeding zero \(P(\beta_{HLA} > 0)\), for Fisher's exact test we sort HLA allele - replacement pairs by increasing p-values and for the Phylogenetic Dependency Network, we rank the associations by increasing q-values.
-%Then, a list of known epitopes is used to create a plot of the cumulative number of replacements inside the boundary of a known epitope vs.
+% %Then, a list of known epitopes is used to create a plot of the cumulative number of replacements inside the boundary of a known epitope vs.
 the corresponding rank.
@@
 
-The predictions of all tested models are strongly enriched in epitopes over baselines, irrespective of dataset (Fig.~\ref{fig:comparison} for HBe protein and Supplementary Figures \hl{XXX-YYY} for other datasets).
+% The predictions of all tested models are strongly enriched in epitopes over baselines, irrespective of dataset (Fig.~\ref{fig:comparison} for HBe protein and Supplementary Figures \hl{XXX-YYY} for other datasets).
 For the best ranked HAMs, Fisher's exact test performs about as well as the HAMdetector backbone logistic regression model (model 1 in Fig.~\ref{fig:comparison}).
 Each of the following three model stages of HAMdetector increases HAM enrichment further, independently of the dataset.
-The horseshoe prior alone (model 2) is a drastic improvement over model 1, even though it does not include any external information.
+% The horseshoe prior alone (model 2) is a drastic improvement over model 1, even though it does not include any external information.
 The logistic regression model with horseshoe prior works roughly as well as the Phylogenetic Dependency Network \cite{Carlson2008}, which includes much more information.
@@
 
-\begin{figure}[ht!]
-  \includegraphics[width=1\linewidth]{plots/comparison.pdf}
-  \caption{HAM enrichment plot for HBe protein: number $N_e$ of associations inside the boundary of known epitopes vs.\ rank $r$.
+% \begin{figure}[ht!]
+%   \includegraphics[width=1\linewidth]{plots/comparison.pdf}
+%   \caption{HAM enrichment plot for HBe protein: number $N_e$ of associations inside the boundary of known epitopes vs.\ rank $r$.
 PhyloD: Phylogenetic Dependency Network; model 1: simple logistic regression model with broad Student-t priors; model 2: logistic regression model with horseshoe prior; model 3: logistic regression model with horseshoe prior and phylogeny; model 4: full model with epitope prediction.
 Gray lines are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baselines.}
-  \label{fig:comparison}
-\end{figure}
+%   \label{fig:comparison}
+% \end{figure}
 
-The Bayesian approach lends itself to addition of prior knowledge, and it is not too surprising that exploiting this feature to incorporate information from phylogeny or epitope prediction improves HAM detection.
+% The Bayesian approach lends itself to addition of prior knowledge, and it is not too surprising that exploiting this feature to incorporate information from phylogeny or epitope prediction improves HAM detection.
 It may be more surprising that the sparsifying horseshoe prior has such an impact.
@@
 
-\begin{figure}
-  \includegraphics[width=1\linewidth]{plots/without_horseshoe.pdf}
-  \includegraphics[width=1\linewidth]{plots/with_horseshoe.pdf}
-  \caption{Marginal posterior distributions of regression coefficients for the association of replacement 11D of the HIV integrase with HLA alleles A*01 and B*44.
+% \begin{figure}
+%   \includegraphics[width=1\linewidth]{plots/without_horseshoe.pdf}
+%   \includegraphics[width=1\linewidth]{plots/with_horseshoe.pdf}
+%   \caption{Marginal posterior distributions of regression coefficients for the association of replacement 11D of the HIV integrase with HLA alleles A*01 and B*44.
 Top half: inferred with logistic regression model, bottom half: inferred with logistic regression with sparsifying horseshoe prior.}
-  \label{fig:horseshoe-comparison}
-\end{figure}
+%   \label{fig:horseshoe-comparison}
+% \end{figure}
 
-\subsection{Leave-one-out cross-validation}
+% \subsection{Leave-one-out cross-validation}
 
-To quantify the ability of the four different model stages of HAMdetector to generalize to unseen cases, we computed the ELPD with Pareto-smoothed leave-one-out cross-validation.
+% To quantify the ability of the four different model stages of HAMdetector to generalize to unseen cases, we computed the ELPD with Pareto-smoothed leave-one-out cross-validation.
 Table \ref{loo} shows results for the HBe protein of the HBV dataset in terms of ELPD changes with each new model stage.
@@
 
-\begin{table}[h!]
-  \renewcommand{\arraystretch}{1.3}
-  \centering
-  \caption{ELPD changes as HAMdetector components are added.
+% \begin{table}[h!]
+%   \renewcommand{\arraystretch}{1.3}
+%   \centering
+%   \caption{ELPD changes as HAMdetector components are added.
 Data computed for HBe protein of the HBV dataset.
 All differences in ELPD are larger than several times the estimated standard error (column \(\text{se}_\text{diff}\)), indicating that models that include more information have better predictive performance.}
-  \vspace{0.5cm}
-  \begin{tabular}{l|r|rll}
-  \multicolumn{1}{l|}{} & \multicolumn{1}{c|}{\(\text{ELPD}_\text{diff}\)} & \multicolumn{1}{c}{\(\text{se}_\text{diff}\)} &  &  \\ \cline{1-3}
-  logistic regression (baseline) & 0.0  & 0.0 &  &  \\
-  + horseshoe prior     & 949.8  & 65.2 &  &  \\
-  + phylogeny & 4440.9 & 94.4 &  &  \\
-  + epitope prediction & 63.1   & 18.9 &  & 
-  \end{tabular}
-  \label{loo}
-\end{table}
+%   \vspace{0.5cm}
+%   \begin{tabular}{l|r|rll}
+%   \multicolumn{1}{l|}{} & \multicolumn{1}{c|}{\(\text{ELPD}_\text{diff}\)} & \multicolumn{1}{c}{\(\text{se}_\text{diff}\)} &  &  \\ \cline{1-3}
+%   logistic regression (baseline) & 0.0  & 0.0 &  &  \\
+%   + horseshoe prior     & 949.8  & 65.2 &  &  \\
+%   + phylogeny & 4440.9 & 94.4 &  &  \\
+%   + epitope prediction & 63.1   & 18.9 &  & 
+%   \end{tabular}
+%   \label{loo}
+% \end{table}
 
-The model with horseshoe prior alone already has a much higher ELPD than the standard logistic regression model, even though it does not use any specific external data.
+% The model with horseshoe prior alone already has a much higher ELPD than the standard logistic regression model, even though it does not use any specific external data.
 This is because including the sparsity assumption allows the model to better separate signal from noise and the uncertainty of the close-to-zero coefficients does not propagate into uncertainty of predictions.
 
-Including phylogeny further improves model performance a lot, as the assumption of independent and identically distributed data is replaced with specific information from the shared phylogenetic history.
+% Including phylogeny further improves model performance a lot, as the assumption of independent and identically distributed data is replaced with specific information from the shared phylogenetic history.
 
-While addition of sparsity and phylogeny has an effect on all replacements and samples, epitope prediction only influences those replacements that are restricted by a given HLA allele and only those samples that are annotated with the allele.
+% While addition of sparsity and phylogeny has an effect on all replacements and samples, epitope prediction only influences those replacements that are restricted by a given HLA allele and only those samples that are annotated with the allele.
 Therefore, inclusion of epitope prediction does not improve ELPD as much as inclusion of phylogeny and the sparsity assumption.
@@
 
-\subsection{HAMs in HDV as test case}
-The Hepatitis D Virus (HDV) dataset \citep{Karimzadeh2019} is an excellent test case: we have (1) a set of paired HDV sequences and patient HLA alleles, (2) HAM predictions by Fisher's exact test as implemented in SeqFeatR \citep{Budeus2016}, and (3) experimental tests of the predicted HAMs for immune escape activity (IFN-\(\gamma\) production assays, \citet{Karimzadeh2019}).
+% \subsection{HAMs in HDV as test case}
+% The Hepatitis D Virus (HDV) dataset \citep{Karimzadeh2019} is an excellent test case: we have (1) a set of paired HDV sequences and patient HLA alleles, (2) HAM predictions by Fisher's exact test as implemented in SeqFeatR \citep{Budeus2016}, and (3) experimental tests of the predicted HAMs for immune escape activity (IFN-\(\gamma\) production assays, \citet{Karimzadeh2019}).
 This allows us to see whether HAMdetector decreases the false positive rate in comparison to the simpler Fisher's exact test, and we can make \emph{bona fide} predictions on previously undetected HAMs.
@@
 
-We have 15 HAMs predicted in HDV by Fisher's exact test at significance level $5\times 10^{-3}$ (Table~\ref{tab:false-positives}) as published \citep{Karimzadeh2019}.
+% We have 15 HAMs predicted in HDV by Fisher's exact test at significance level $5\times 10^{-3}$ (Table~\ref{tab:false-positives}) as published \citep{Karimzadeh2019}.
 The corresponding p-values  have no clear relation to experimental confirmation, i.e.\ p-values for confirmed HAMs are not generally lower than those of non-confirmed ones.
 
-For HAMdetector, we use in Table~\ref{tab:false-positives} the posterior probability of a positive regression coefficient ($P(\beta_{jk}>0)$ as measure for the confidence in having detected a HAM.
+% For HAMdetector, we use in Table~\ref{tab:false-positives} the posterior probability of a positive regression coefficient ($P(\beta_{jk}>0)$ as measure for the confidence in having detected a HAM.
 HAMs with strong support have a posterior probability close to 1, associations with no support a probability close to 0.5 (corresponding to a regression coefficient centered around 0).
@@
   R105K       & B*27 & 0.0011                        & 0.93                                & +                        \\
-  R139K       & B*41 & 0.0034                        & 0.91                                & +                        \\
-  E47D        & B*18 & 0.0027                        & 0.89                                & +                        \\
+  R139K       & B*41 & 0.0034                        & 0.92                                & +                        \\
+  E47D        & B*18 & 0.0027                        & 0.90                                & +                        \\
   D33E        & B*13 & 0.0001                        & 0.86                                & -                        \\
@@
   K43R        & B*13 & 0.0021                        & 0.77                                & -                        \\
+  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
   D47E        & A*30 & 0.0010                        & 0.76                                & -                        \\
   K113R       & B*13 & 0.0043                        & 0.76                                & -                        \\
-  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
   A107T       & B*14 & 0.0028                        & 0.70                                & -                        \\
   P49L        & A*30 & 0.0031                        & 0.63                                & -                        \\
-  Q100L       & B*13 & 0.0018                        & 0.61                                & -                        \\
-  D96E        & B*13 & 0.0035                        & 0.50                                & -                       
+  Q100L       & B*13 & 0.0018                        & 0.60                                & -                        \\
+  D96E        & B*13 & 0.0035                        & 0.51                                & -  \\\hline
+  E46D & A*02 & 0.0054 & 0.97 & ? \\
+  V81I & A*68 & 0.0073 & 0.97 & ? \\
+  K113N & B*08 & 0.0063 & 0.96 & ? \\
+  A71T & B*41 & 0.0065 & 0.96 & ? \\
+  L188I & A*68 & 0.0632 & 0.94 & ? \\
+  T95S & A*01 & 0.0285 & 0.93 & ? \\
+  D33E & A*03 & 0.0226 & 0.93 & ? \\
+  P49L & B*13 & 0.0035 & 0.92 & ? \\
+  A74S & A*68 & 0.0123 & 0.91 & ? \\
+  E29D & B*44 & 0.0559 & 0.91 & ? \\
+  D46E & B*57 & 0.0190 & 0.91 & ? \\
+  R88K & A*68 & 0.0123 & 0.91 & ? \\
+  T149P & B*52 & 0.0281 & 0.91 & ? \\
+  K43R & A*02 & 0.2158 & 0.90 & ? \\
+  N22S & B*08 & 0.0405 & 0.90 & ? \\
   \end{tabular}
@@
 
- \subsection{Linkage disequilibrium}
+%  \subsection{Linkage disequilibrium}
 
- \hl{DaHo: bitte Beispiel mit Linkage Disequilibrium zeigen}
+%  \hl{DaHo: bitte Beispiel mit Linkage Disequilibrium zeigen}
 
-\section{Discussion}
+% \section{Discussion}
 
-HAMdetector follows a general paradigm of Bayesian modelling, namely to map all information that is available about a system of interest onto a probabilistic model, and then to apply Bayesian inference to learn about probable parameter values of that model, e.g.\ about $\beta_{jk}$, the association of HLA $j$ with replacement $k$.
+% HAMdetector follows a general paradigm of Bayesian modelling, namely to map all information that is available about a system of interest onto a probabilistic model, and then to apply Bayesian inference to learn about probable parameter values of that model, e.g.\ about $\beta_{jk}$, the association of HLA $j$ with replacement $k$.
 The more relevant information we infuse into the model, the sharper the inference.
@@
 
-We have demonstrated that the logistic regression backbone is a platform that can be extended by model components that contribute new information.
+% We have demonstrated that the logistic regression backbone is a platform that can be extended by model components that contribute new information.
 We have selected such modules guided by widely accepted knowledge, such as phylogeny or epitope location.
@@
 
-Yet, HAMdetector is far from perfect.
+% Yet, HAMdetector is far from perfect.
 For instance the outlier in Table \ref{tab:false-positives} could point to missing information in HAMdetector.
@@
 
-Another extension of our model would be to better account for phylogenetic uncertainty by using a Bayesian method to estimate a posterior distribution over possible tree topologies.
+% Another extension of our model would be to better account for phylogenetic uncertainty by using a Bayesian method to estimate a posterior distribution over possible tree topologies.
 The uncertainty over the tree topologies and the underlying parameters of the phylogenetic model would then propagate into uncertainty of the estimated probabilities \(P(y_{ik}=1|\Psi)\).
 
-However, the good performance of the current version of HAMdetector makes it already a valuable tool for the study of interactions between viruses and T cell immunity.
+% However, the good performance of the current version of HAMdetector makes it already a valuable tool for the study of interactions between viruses and T cell immunity.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
+% %
+% %     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
+% %     as it may ignore the figures.
+% %
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-% \section{Conclusion}
+% % \section{Conclusion}
 
-%\section*{Acknowledgements}
+% %\section*{Acknowledgements}
 
-\section*{Funding}
+% \section*{Funding}
 
-This work has been supported by Deutsche Forschungsgemeinschaft (grant HO 1582/10-1).\vspace*{-12pt}
+% This work has been supported by Deutsche Forschungsgemeinschaft (grant HO 1582/10-1).\vspace*{-12pt}
 
-\bibliographystyle{natbib}
+% \bibliographystyle{natbib}
 %\bibliographystyle{achemnat}
@@
 %
-\bibliography{references}
+% \bibliography{references}
 \end{document}

### 2021-03-25 docs/main.tex
update manuscript

--- 2021-02-21 docs/main.tex
@@
 \usepackage{soul}
+\usepackage[paper=b4paper, layout=b4paper]{geometry}
+\usepackage{float}
+% \usepackage[section]{placeins}
 
@@
 \maketitle
-
+\sloppy
 \section{Introduction}
@@
 
-There are several experimental methods available to study HLA escape~\citep{Timm2004}.
-\hl{DaHo: use key references, e.g.
-those that have introduced a technique; alternatively: cite review like articles that give an overview of several experimental methods} \hl{DaHa: see .tex comments for possible citations} However, these methods are relatively slow and costly, especially for screening purposes.
+There are several experimental methods available to study HLA escape~\citep{Czerkinsky1983, Brunner1968, Lamoreaux2006, Altman1996}.
+However, these methods are relatively slow and costly, especially for screening purposes.
 A promising approach that makes efficient use of frequently available data is to combine viral genome sequencing, host HLA determination, computational identification by statistical association analysis, and targeted experimental validation~\citep{Carlson2012}.
-
-% IFN-gamma ELISPOT:
-% A  Solid-Phase Enzyme-Linked Immunospot (ELISPOT) Assay for Enumeration of Specific Antibody-Secreting Cells \citep{Czerkinsky1983}
-% Quantification of antigen specific CD8+ T cells using an ELISPOT assay \citep{Miyahira1995}
-
-% Chromium release
-% Quantitative assay of the lytic action of immune lymphoid cells on 51-Cr-labelled allogeneic target cells in vitro; inhibition by isoantibody and by drugs \citep{Brunner1968}
-
-% Intracellular cytokine staining
-% Intracellular cytokine optimization and standard operating procedure \citep{Lamoreaux2006}
-% The assessment of antigen-specific CD8+ T cells through the combination of MHC class I tetramer and intracellular staining \citep{Appay2002}
-
-% HLA tetramer staining
-% Phenotypic analysis of antigen-specific T lymphocytes \citep{Altman1996}
 
@@
 The most striking one is that viral sequences share a common phylogenetic history, and,  therefore, treating sequences as independent and identically distributed samples may under- or overestimate effect sizes.
-In the context of hypothesis testing, this leads to increased false positive and false negative rates.
-\hl{There are at least two references we could cite here: Osborne2002 is the reference usually cited in this context and contains the phrase "When these assumptions are not met the results may not be trustworthy, resulting in a Type I or Type II error, \dots" It is more of a summary for psychologists though and contains some questionable advice like removing outliers for increased power.
-There is also Scariano1987 who looks at type I/type II errors in one-way ANOVA when independence is violated, this may be a better reference.
-I have not found any article on the violation of independence in fisher's exact test specifically.} \citep{Osborne2002} \citep{Scariano1987}
+In the context of hypothesis testing, this leads to increased false positive and false negative rates \citep{Osborne2002, Scariano1987}.
 
-Another issue with Fisher's exact test is the genomic proximity of human HLA class I loci (on chromosome 6) leading to linkage disequilibrium -- inheritance of HLA alleles can be correlated.
-Therefore spurious associations can occur if associations of replacements with individual HLA alleles is tested: if HLA allele \(H_1\) is associated with an amino acid replacement \(R\) because of immune escape, but \(H_1\) is in linkage disequilibrium with allele \(H_2\), this leads to an association of \(R\) and \(H_2\), even without being an escape mutation from $H_2$.
+Another issue with Fisher's exact test is the genomic proximity of human HLA class I loci (on chromosome 6 \citep{Francke1977}) leading to linkage disequilibrium -- inheritance of HLA alleles can be correlated.
+Therefore, spurious associations can occur if associations of replacements with individual HLA alleles are tested: if HLA allele \(H_1\) is associated with an amino acid replacement \(R\) because of immune escape, but \(H_1\) is in linkage disequilibrium with allele \(H_2\), this leads to an association of \(R\) and \(H_2\), even without being an escape mutation from $H_2$.
 
@@
 
-There are fundamental statistical issues with p-values as screening tool~\citep{Amrhein2017}:
+There are fundamental statistical issues with p-values as a screening tool~\citep{Amrhein2017}:
 with small effect sizes and high variance between measurements, as is often the case with biological data, statistically significant results can be misleading, have the wrong direction (type S error), or greatly overestimate an effect (type M error)~\citep{Gelman2014}.
@@
 
-The main parameters of interests for HAMdetector are the regression coefficients $\beta_{jk}$, as they quantify the strength of association between the occurrence of replacement \(k\) and each of the observed HLA alleles.
+The main parameters of interest for HAMdetector are the regression coefficients $\beta_{jk}$, as they quantify the strength of association between the occurrence of replacement \(k\) and each of the observed HLA alleles.
 The $\beta_{jk}$ are on the log-odds scale, i.e.\ if we go from viral sequences from hosts without HLA allele $j$ to those from hosts with $j$, the log-odds $\log (p_k/(1-p_k))$ of observing replacement \(k\) increase by addition of $\beta_{jk}$.
@@
 
-On top of the paired data of viral sequences and host HLA alleles modeled by the backbone (Eq~\ref{eq:backbone}), we extend the model to include further information of relevance to improve HAM detection, namely phylogenetic information, and predictions of epitope peptide processing and MHC I affinity, as described in the following.
+On top of the paired data of viral sequences and host HLA alleles modeled by the backbone (Eq~\ref{eq:backbone}), we extend the model to include further information of relevance to improve HAM detection, namely phylogenetic information and predictions of epitope peptide processing and MHC I affinity, as described in the following.
 
@@
 We are interested in estimating \(P(y_{ik}=1|\Psi)\), that is, the probability of observing the replacement \(k\) in sequence \(i\) based on the underlying phylogenetic model.
-A quantity that can be readily computed using phylogenetic software like RAxML-NG~\cite{Kozlov2019} is \(P(\Psi|y_{ik}=1)\).
+A quantity that can be readily computed using phylogenetic software like RAxML-NG~\citep{Kozlov2019} is \(P(\Psi|y_{ik}=1)\).
 For this, we keep the tree topology fixed, annotate the tree with the binary observations \(y_{ik}\) at its leaves and optimize the branch lengths.
@@
 
-Not only mutations in CTL epitopes can lead to failure of MHC I to present epitopes, but also mutations at epitope-flanking positions that interfere with pre-processing of peptides, notably proteasomal cleavage of viral proteins \hl{DaHo: citation}.
+Not only mutations in CTL epitopes can lead to failure of MHC I to present epitopes, but also mutations at epitope-flanking positions that interfere with pre-processing of peptides, notably proteasomal cleavage of viral proteins~\citep{Milicic2005, Gall2007}.
 
@@
   \begin{aligned}
-    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \text{exp}(Z_{jk}\beta_{\text{epi}})) \\
+    \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}\text{exp}(Z_{jk}\beta_{\text{epi}})) \\
     \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2),
@@
 
-\hl{DaHo: komplexer Zusammenhang von pre-processing escape und HLA; bitte besprechen}
-
 \subsubsection{Sparsity-inducing priors}
@@
 This assumption is likely correct for HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes.
-If no epitope spanning a given alignment position is presented on the MHC I molecule, any apparent association between a replacement and the respective HLA allele is likely to be due to random viral sequence variation or HLA linkage disequilibrium.
-\hl{DaHo: Wie ist das mit compensatory mutations *außerhalb* von Epitopen?
-Z.B.
-https://jvi.asm.org/content/86/2/991}
+As presentation on MHC I is a necessary condition for the HLA system to function, most associations between an HLA allele and a replacement are likely truly zero, with the confounding factors HLA linkage disequilibrium and codon covariation, e.g.
+in the form of compensatory mutations that lie outside an epitope~\citep{Ruhl2011}.
 
@@
 They share the common structure of placing most probability mass very close to 0, with heavy tails to accommodate the non-zero coefficients.
-For our model, we use a so-called horseshoe prior \citep{Carvalho2010}, a scale-mixture of Gaussians:
+For our model, we use the so-called regularized horseshoe prior \citep{Piironen2017}, which is an improvement of the original horseshoe prior presented by \citet{Carvalho2010}, in that it additionally allows some shrinkage for the non-zero coefficients.
+The original horseshoe prior is given by:
+
 \begin{equation}
@@
 where \(\beta_{jk}\) are the regression coefficients; \(\tau\) and \(\lambda_{jk}\) are the so-called global and local shrinkage parameters, respectively;, \(\text{Cauchy}^{+}\) is the positively constrained Cauchy distribution; $\tau_0$ is the overall degree of sparsity.
+Shrinkage of the non-zero coefficients in the regularized horseshoe prior is achieved by replacing \(\lambda_{jk}^2\) with \(\tilde{\lambda}^{2}_{jk} = \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau_{k}^{2}\lambda_{jk}^{2}}\), where the additional parameter \(c\) governs the magnitude of shrinkage for the non-zero coefficients.
 
@@
 
-The overall degree of sparsity $\tau_0$ can be chosen based on the expected number of non-zero coefficients \citep{Piironen2017}. \hl{DaHo: do we do this? How?}
+The overall degree of sparsity $\tau_0$ can be chosen based on the expected number of non-zero coefficients \citep{Piironen2016}.
 
@@
     c_{k}^2 &\sim \text{Inv-Gamma}(3.5, 3.5) & \text{(*)} \\
-   \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}^2\text{exp}(Z_{jk}\beta_{\text{epi}})) & \\
+   \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}\text{exp}(Z_{jk}\beta_{\text{epi}})) & \\
    \tau_{k} &\sim \text{Cauchy}^{+}(0, \tau_{0k}) & \text{(*)} \\
@@
 \end{equation}
+where \(N\) is the number of available annotated sequences.
 
@@
 In particular, the overall phylogeny-weight $\gamma$ in Eq~\ref{eq:backbone} is replaced in the full model by  hierarchically modelled  \(\gamma_{k}\), which allows partial pooling across replacements (even with a global parameter \(\gamma\) the model works reasonably well).
-All other additions (\(\tilde{\lambda}_{jk}^{2}\) and \(\tau_{0k}\)) are taken from \citet{Piironen2017}.
+All other additions (\(\tilde{\lambda}_{jk}^{2}\) and \(\tau_{0k}\)) are explained in detail in \citet{Piironen2017}.
 Briefly, the addition of  \(\tilde{\lambda}_{jk}^{2}\) allows some regularization for the non-zero coefficients and the parameterization of \(\tau_{0k}\) allows to place a prior on the expected number of non-zero coefficients.
 This is particularly useful for logistic regression models, as some shrinkage helps to deal with issues of separability and collinearity that commonly occur with logistic regression models.
-\hl{DaHo: Bedeutung von $N$ in $\tau_{0k}$?}
 
@@
 The rationale behind this parameterization is again outlined in \citet{Piironen2017}.
-The value of 10 corresponds to a generously estimated magnitude based on available HIV epitope maps \hl{DaHo: citation?}.
+The value of 10 corresponds to a generously estimated magnitude based on available HIV epitope maps~\citep{immunology2018editors}.
 The model is also parameterized in a way that assumes an equal degree of sparsity across all alignment positions a priori.
@@
 All models were implemented in Stan 2.23 \citep{Stan2021}, a probabilistic programming language and Hamiltonian Monte Carlo sampler for efficient numerical computation of posterior distributions.
-The Stan code is available \hl{DaHo: at ...} in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
-A Julia \citep{Bezanson2017} package is available \hl{DaHo: at ...} to run the model on custom data.
+The Stan code is available at ...
+in two versions: One optimized for readability and one optimized for speed by utilizing Stan's multithreading and GPU capabilities.
+A Julia \citep{Bezanson2017} package is available at ...
+to run the model on custom data.
 Due to restrictions of dependencies (MHCFlurry and RAxML-ng), HAMdetector is only available on Linux.
@@
 \subsubsection{Convergence diagnostics}
-We use the split-\(\hat{R}\) convergence diagnostic to identify Markov chain convergence issues \citep{Gelman1992, Stan2021}.
+We use the split-\(\hat{R}\) convergence diagnostic to identify Markov chain convergence issues \citep{Gelman1992, gelman2013bayesian}.
 We require a value of \(\hat{R}\) below 1.01 for all model parameters.
-Additionally, we require that the effective sample size \(N_{\text{eff}}\) is above 500 for all model parameters and that sampling occurs without any divergent transitions \citep{Stan2021}.
+Additionally, we require that the effective sample size \(\text{N}_{\text{eff}}\)~\citep{Stan2021} is above 500 for all model parameters and that sampling occurs without any divergent transitions~\citep{Betancourt2017a}.
 
@@
   \subsection{Data}
- \hl{DaHo: das Material in Materials and Methods sind die Daten; die sollten hier kurz beschrieben werden, möglichst mit Referenzen}
 
@@
 \begin{itemize}
-  \item A large HIV dataset \hl{DaHo: paper?} consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}.
+  \item A large HIV dataset consisting of a subset of sequences from the HOMER~\citep{Brumme2007, Brumme2008} cohort, the Western Australian HIV Cohort Study (WAHC,~\citet{Moore2002, Bhattacharya2007}) and the US AIDS Clinical Trials Group (ACTG) protocols 5142 participants~\citep{John2008} who also provided Human DNA under ACTG protocol 5128~\citep{haas2003multi} (total \(N=1383\)).
+These data were kindly provided by Zabrina Brumme and collaborators and were in part also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}.
 The dataset contains sequences spanning the \textit{gag}, \textit{pol}, \textit{env}, \textit{nef}, \textit{vif}, \textit{vpr}, \textit{vpu}, \textit{tat} and \textit{rev} genes.
@@
   \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm. The dataset contains sequences of the HBe, LHBs, Pol and HBx proteins.
-  \item A set of 104 Hepatitis-D-Virus sequences containing the HDV-antigen that were kindly provided by Michael Roggendorf. \hl{DaHo: bitte von Hadi geeignetes Zitat erfragen}
+  \item A set of 104 Hepatitis-D-Virus sequences containing the HDV-antigen that were kindly provided by Michael Roggendorf~\citep{Karimzadeh2018}.
   \item A set of 41 HIV sequences spanning the \textit{gag} and \textit{pol} genes that were kindly provided by Rongge Yang. 
@@
 
-\hl{DaHo: sources of known epitopes?}
+Lists of known epitopes were gathered from the Immune Epitope Database (IEDB, \citet{Vita2019}) for HIV sequences, and the Hepitopes database~\cite{Lumley2016} for Hepatitis-B-Virus sequences.
 
@@
 \subsection{Run times and convergence}
-Run times of HAMdetector on a ... were of the order of ... \hl{DaHo: bitte ergänzen}
+For a standard officie computer, run times of HAMdetector on the smaller HDV dataset were of the order of minutes and on the order of hours for the Hepatitis B dataset.
+For the large HIV dataset, the models were run overnight.
+Run times scale approximately linearly with the product \(NK\), where \(N\) is the number of sequences and \(K\) is the number of replacements.
 All model fits showed no signs of inference issues.
-In total, samples were drawn from four Hamiltonian Markov chains with 1000 iterations each after 1000 warm-up iterations.
+In total, samples were drawn from four Hamiltonian Markov chains with 1000 iterations each after 300 warm-up iterations.
 The effective sample size exceeded 500 for all model parameters, $\hat{R}$ convergence diagnostic values were below 1.01 in all cases.
@@
 The model yields well-calibrated posterior predictive probabilities of replacements.
-This is exemplified in Figure~\ref{fig:calibration} for HBV core protein, but also holds true for the other datasets (Supplementary Figures \hl{XXX-YYY}).
+This is exemplified in Figure~\ref{fig:calibration} for HBV core protein, but also holds true for the other datasets (Supplementary Figures ...-...).
 
-\begin{figure}[!ht]
+\begin{figure}
  \includegraphics[width=1.0\linewidth]{plots/phylogeny_calibration.pdf}  
@@
 
-The predictions of all tested models are strongly enriched in epitopes over baselines, irrespective of dataset (Fig.~\ref{fig:comparison} for HBe protein and Supplementary Figures \hl{XXX-YYY} for other datasets).
-For the best ranked HAMs, Fisher's exact test performs about as well as the HAMdetector backbone logistic regression model (model 1 in Fig.~\ref{fig:comparison}).
+The predictions of all tested models are strongly enriched in epitopes over baselines, irrespective of dataset (Fig.~\ref{fig:comparison} for HBe protein and Supplementary Figures ...-...
+for other datasets).
+For the best ranked HAMs, Fisher's exact test performs about as well as the HAMdetector backbone logistic regression model (model~1 in Fig.~\ref{fig:comparison}).
 Each of the following three model stages of HAMdetector increases HAM enrichment further, independently of the dataset.
-The horseshoe prior alone (model 2) is a drastic improvement over model 1, even though it does not include any external information.
+The horseshoe prior alone (model~2) is a drastic improvement over model~1, even though it does not include any external information.
 The logistic regression model with horseshoe prior works roughly as well as the Phylogenetic Dependency Network \cite{Carlson2008}, which includes much more information.
 Note that the line for the Phylogenetic Dependency Network is shorter because it only outputs associations with a q-value lower than 0.2 by default.
-Model 3 with its additional inclusion of phylogeny has higher enrichment than model 2, and finally, the full model 4 with the inclusion of epitope prediction leads to a further improvement.
-Note that model 4 only uses epitope \textit{prediction} software and does not use any information of experimentally confirmed epitopes.
+Model~3 with its additional inclusion of phylogeny has higher enrichment than model~2, and finally, the full model~4 with the inclusion of epitope prediction leads to a further improvement.
+Note that model~4 only uses epitope \textit{prediction} software and does not use any information of experimentally confirmed epitopes.
 These are here only used for model evaluation.
 
-\begin{figure}[ht!]
-  \includegraphics[width=1\linewidth]{plots/comparison.pdf}
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\linewidth]{plots/HBe.pdf}
   \caption{HAM enrichment plot for HBe protein: number $N_e$ of associations inside the boundary of known epitopes vs.\ rank $r$.
-PhyloD: Phylogenetic Dependency Network; model 1: simple logistic regression model with broad Student-t priors; model 2: logistic regression model with horseshoe prior; model 3: logistic regression model with horseshoe prior and phylogeny; model 4: full model with epitope prediction.
-Gray lines are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baselines.}
+D: Phylogenetic Dependency Network; F: Fisher's exact test; 1: simple logistic regression model with broad Student-t priors; 2: logistic regression model with horseshoe prior; 3: logistic regression model with horseshoe prior and phylogeny; 4: full model with epitope prediction.
+Unannotated gray lines at the bottom of the graph are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baselines.}
   \label{fig:comparison}
@@
 \begin{figure}
-  \includegraphics[width=1\linewidth]{plots/without_horseshoe.pdf}
-  \includegraphics[width=1\linewidth]{plots/with_horseshoe.pdf}
+  \centering
+  \includegraphics[width=0.8\linewidth]{plots/without_horseshoe.pdf}
+  \includegraphics[width=0.8\linewidth]{plots/with_horseshoe.pdf}
   \caption{Marginal posterior distributions of regression coefficients for the association of replacement 11D of the HIV integrase with HLA alleles A*01 and B*44.
@@
 
-\begin{table}[h!]
+\begin{table}
   \renewcommand{\arraystretch}{1.3}
@@
 This allows us to see whether HAMdetector decreases the false positive rate in comparison to the simpler Fisher's exact test, and we can make \emph{bona fide} predictions on previously undetected HAMs.
-\hl{DaHo: zu \emph{bona fide}: können wir dazu etwas sagen?
-Ich erinnere mich, dass Du noch eine Reihe guter HAMs vorhergesagt hattest.}
-
 We have 15 HAMs predicted in HDV by Fisher's exact test at significance level $5\times 10^{-3}$ (Table~\ref{tab:false-positives}) as published \citep{Karimzadeh2019}.
@@
 HAMs with strong support have a posterior probability close to 1, associations with no support a probability close to 0.5 (corresponding to a regression coefficient centered around 0).
-The five predicted HAMs with top posterior probabilities (all $\ge 0.89$) have all been experimentally confirmed.
+The five predicted HAMs with top posterior probabilities (all $\ge 0.90$) have all been experimentally confirmed.
 There is only one outlier with posterior probability 0.75 (P89T and B*37).
 
-\begin{table}[h!]
+HAMdetector shows strong support for 15 replacement - allele pairs that have previously not been identified. All of them exceed the significance level of 
+$5\times 10^{-3}$ used in~\citet{Karimzadeh2019}.
+This shows that HAMdetector is able to identify possible HAMs that would not be detected otherwise.
+The most striking example for this is K43R - A*02, which using Fisher's exact test has a p-value of 0.22.
+While this replacement has not been experimentally verified in a IFN-\(\gamma\) production assay yet, it lies within a known A*02 epitope and thus it is highly likely that K43R conveys some form of immune escape.
+
+\begin{table}
   \caption{List of HAMs predicted Fisher's exact test (FET).
 In the last column a ``+'' marks experimental confirmation and a ``-'' a failed confirmation.
-Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector.}
-  \vspace{0.5cm}
+Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector.
+\emph{Bona fide} predictions are separated by a horizontal line.
+These are replacement - allele pairs we expect to to be verified in future experimental tests.
+The replacement annotated with a (+) lies within a known A*02 epitope.}
+  % \vspace{0.5cm}
   \begin{tabular}{c|c|c|c|c}
@@
   K43R        & B*13 & 0.0021                        & 0.77                                & -                        \\
+  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
   D47E        & A*30 & 0.0010                        & 0.76                                & -                        \\
   K113R       & B*13 & 0.0043                        & 0.76                                & -                        \\
-  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
   A107T       & B*14 & 0.0028                        & 0.70                                & -                        \\
   P49L        & A*30 & 0.0031                        & 0.63                                & -                        \\
-  Q100L       & B*13 & 0.0018                        & 0.61                                & -                        \\
-  D96E        & B*13 & 0.0035                        & 0.50                                & -                       
+  Q100L       & B*13 & 0.0018                        & 0.60                                & -                        \\
+  D96E        & B*13 & 0.0035                        & 0.51                                & -  \\\hline
+  E46D & A*02 & 0.0054 & 0.97 & ? \\
+  V81I & A*68 & 0.0073 & 0.97 & ? \\
+  K113N & B*08 & 0.0063 & 0.96 & ? \\
+  A71T & B*41 & 0.0065 & 0.96 & ? \\
+  L188I & A*68 & 0.0632 & 0.94 & ? \\
+  T95S & A*01 & 0.0285 & 0.93 & ? \\
+  D33E & A*03 & 0.0226 & 0.93 & ? \\
+  P49L & B*13 & 0.0035 & 0.92 & ? \\
+  A74S & A*68 & 0.0123 & 0.91 & ? \\
+  E29D & B*44 & 0.0559 & 0.91 & ? \\
+  D46E & B*57 & 0.0190 & 0.91 & ? \\
+  R88K & A*68 & 0.0123 & 0.91 & ? \\
+  T149P & B*52 & 0.0281 & 0.91 & ? \\
+  K43R & A*02 & 0.2158 & 0.90 & (+) \\
+  N22S & B*08 & 0.0405 & 0.90 & ? \\
   \end{tabular}
@@
 
- \hl{DaHo: bitte Beispiel mit Linkage Disequilibrium zeigen}
+ Table~\ref{tab:false-positives} shows that for 3 of the false positives identified with Fisher's exact test, HAMdetector identifies associations with the same replacement but a different allele (P49L - B*13 instead of P49L - A*30; K43R - A*02 instead of K43R - B*13; and D33E - B*13 instead of D33E - A*03).
+ One possible explanation for this observation is HLA-Linkage disequilibrium and accordingly correlation between HLA alleles in the dataset due to random sampling variation.
+Indeed, out of the 12 times P49L is observed in sequences annotated with A*30, B*13 is also present in 5 of those cases (Spearman's rank correlation coefficient \(\rho = 0.5\)).
+A similar observation can be done for K43R and D33E, albeit the correlation between the respective alleles is much weaker.
 
+\begin{figure}
+  \includegraphics[width=1.1\linewidth]{plots/P49L_combined.pdf}
+  \caption{Marginal posteriors for the regression coefficients of A*30 and B*13 for replacement P49L.
+The association between P49L and A*30 identified by Fisher's exact test turned out to be a false positive.}
+  \label{fig:linkage}
+\end{figure}
+
+Figure~\ref{fig:linkage} shows regression coefficients of the HLA alleles A*30 and B*13 for replacement P49L.
+When using a simple logistic regression model without any additions (model~1), both A*30 and B*13 show medium evidence of being associated with replacement P49L.
+However, when also taking phylogeny into account and using a sparsity-promoting prior (model~3), both regression coefficients get shrunk to 0 because the association is not strongly enough identified by the data alone.
+Using epitope prediction as an additional source of information (model~4) allows to disentangle the association of the correlated alleles with P49L and identify B*13 as associated with P49L.
+The false-positive association between P49L and A*30 remains being shrunk towards 0.
 \section{Discussion}
@@
 %\section*{Acknowledgements}
-
+% \FloatBarrier
 \section*{Funding}

### 2021-03-25 docs/main.tex
previously uncommitted changes (cannot remember what they were)

@@
 \usepackage[paper=b4paper, layout=b4paper]{geometry}
-\usepackage{float}
-% \usepackage[section]{placeins}
 
@@
 \maketitle
-\sloppy
+
 \section{Introduction}
@@
 
-\begin{figure}
+\begin{figure}[!ht]
  \includegraphics[width=1.0\linewidth]{plots/phylogeny_calibration.pdf}  
@@
 
-The predictions of all tested models are strongly enriched in epitopes over baselines, irrespective of dataset (Fig.~\ref{fig:comparison} for HBe protein and Supplementary Figures ...-...
-for other datasets).
-For the best ranked HAMs, Fisher's exact test performs about as well as the HAMdetector backbone logistic regression model (model~1 in Fig.~\ref{fig:comparison}).
+The predictions of all tested models are strongly enriched in epitopes over baselines, irrespective of dataset (Fig.~\ref{fig:comparison} for HBe protein and Supplementary Figures \hl{XXX-YYY} for other datasets).
+For the best ranked HAMs, Fisher's exact test performs about as well as the HAMdetector backbone logistic regression model (model 1 in Fig.~\ref{fig:comparison}).
 Each of the following three model stages of HAMdetector increases HAM enrichment further, independently of the dataset.
-The horseshoe prior alone (model~2) is a drastic improvement over model~1, even though it does not include any external information.
+The horseshoe prior alone (model 2) is a drastic improvement over model 1, even though it does not include any external information.
 The logistic regression model with horseshoe prior works roughly as well as the Phylogenetic Dependency Network \cite{Carlson2008}, which includes much more information.
 Note that the line for the Phylogenetic Dependency Network is shorter because it only outputs associations with a q-value lower than 0.2 by default.
-Model~3 with its additional inclusion of phylogeny has higher enrichment than model~2, and finally, the full model~4 with the inclusion of epitope prediction leads to a further improvement.
-Note that model~4 only uses epitope \textit{prediction} software and does not use any information of experimentally confirmed epitopes.
+Model 3 with its additional inclusion of phylogeny has higher enrichment than model 2, and finally, the full model 4 with the inclusion of epitope prediction leads to a further improvement.
+Note that model 4 only uses epitope \textit{prediction} software and does not use any information of experimentally confirmed epitopes.
 These are here only used for model evaluation.
 
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\linewidth]{plots/HBe.pdf}
+\begin{figure}[ht!]
+  \includegraphics[width=1\linewidth]{plots/comparison.pdf}
   \caption{HAM enrichment plot for HBe protein: number $N_e$ of associations inside the boundary of known epitopes vs.\ rank $r$.
-D: Phylogenetic Dependency Network; F: Fisher's exact test; 1: simple logistic regression model with broad Student-t priors; 2: logistic regression model with horseshoe prior; 3: logistic regression model with horseshoe prior and phylogeny; 4: full model with epitope prediction.
-Unannotated gray lines at the bottom of the graph are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baselines.}
+PhyloD: Phylogenetic Dependency Network; model 1: simple logistic regression model with broad Student-t priors; model 2: logistic regression model with horseshoe prior; model 3: logistic regression model with horseshoe prior and phylogeny; model 4: full model with epitope prediction.
+Gray lines are HAM enrichment curves for random permutations of the list of HLA allele - replacement pairs and act as baselines.}
   \label{fig:comparison}
@@
 \begin{figure}
-  \centering
-  \includegraphics[width=0.8\linewidth]{plots/without_horseshoe.pdf}
-  \includegraphics[width=0.8\linewidth]{plots/with_horseshoe.pdf}
+  \includegraphics[width=1\linewidth]{plots/without_horseshoe.pdf}
+  \includegraphics[width=1\linewidth]{plots/with_horseshoe.pdf}
   \caption{Marginal posterior distributions of regression coefficients for the association of replacement 11D of the HIV integrase with HLA alleles A*01 and B*44.
@@
 
-\begin{table}
+\begin{table}[h!]
   \renewcommand{\arraystretch}{1.3}
@@
 This allows us to see whether HAMdetector decreases the false positive rate in comparison to the simpler Fisher's exact test, and we can make \emph{bona fide} predictions on previously undetected HAMs.
+\hl{DaHo: zu \emph{bona fide}: können wir dazu etwas sagen?
+Ich erinnere mich, dass Du noch eine Reihe guter HAMs vorhergesagt hattest.}
+
 We have 15 HAMs predicted in HDV by Fisher's exact test at significance level $5\times 10^{-3}$ (Table~\ref{tab:false-positives}) as published \citep{Karimzadeh2019}.
@@
 HAMs with strong support have a posterior probability close to 1, associations with no support a probability close to 0.5 (corresponding to a regression coefficient centered around 0).
-The five predicted HAMs with top posterior probabilities (all $\ge 0.90$) have all been experimentally confirmed.
+The five predicted HAMs with top posterior probabilities (all $\ge 0.89$) have all been experimentally confirmed.
 There is only one outlier with posterior probability 0.75 (P89T and B*37).
 
-HAMdetector shows strong support for 15 replacement - allele pairs that have previously not been identified. All of them exceed the significance level of 
-$5\times 10^{-3}$ used in~\citet{Karimzadeh2019}.
-This shows that HAMdetector is able to identify possible HAMs that would not be detected otherwise.
-The most striking example for this is K43R - A*02, which using Fisher's exact test has a p-value of 0.22.
-While this replacement has not been experimentally verified in a IFN-\(\gamma\) production assay yet, it lies within a known A*02 epitope and thus it is highly likely that K43R conveys some form of immune escape.
-
-\begin{table}
+\begin{table}[h!]
   \caption{List of HAMs predicted Fisher's exact test (FET).
 In the last column a ``+'' marks experimental confirmation and a ``-'' a failed confirmation.
-Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector.
-\emph{Bona fide} predictions are separated by a horizontal line.
-These are replacement - allele pairs we expect to to be verified in future experimental tests.
-The replacement annotated with a (+) lies within a known A*02 epitope.}
-  % \vspace{0.5cm}
+Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector.}
+  \vspace{0.5cm}
   \begin{tabular}{c|c|c|c|c}
@@
   T149P & B*52 & 0.0281 & 0.91 & ? \\
-  K43R & A*02 & 0.2158 & 0.90 & (+) \\
+  K43R & A*02 & 0.2158 & 0.90 & ? \\
   N22S & B*08 & 0.0405 & 0.90 & ? \\
@@
 
- Table~\ref{tab:false-positives} shows that for 3 of the false positives identified with Fisher's exact test, HAMdetector identifies associations with the same replacement but a different allele (P49L - B*13 instead of P49L - A*30; K43R - A*02 instead of K43R - B*13; and D33E - B*13 instead of D33E - A*03).
- One possible explanation for this observation is HLA-Linkage disequilibrium and accordingly correlation between HLA alleles in the dataset due to random sampling variation.
-Indeed, out of the 12 times P49L is observed in sequences annotated with A*30, B*13 is also present in 5 of those cases (Spearman's rank correlation coefficient \(\rho = 0.5\)).
-A similar observation can be done for K43R and D33E, albeit the correlation between the respective alleles is much weaker.
+ \hl{DaHo: bitte Beispiel mit Linkage Disequilibrium zeigen}
 
-\begin{figure}
-  \includegraphics[width=1.1\linewidth]{plots/P49L_combined.pdf}
-  \caption{Marginal posteriors for the regression coefficients of A*30 and B*13 for replacement P49L.
-The association between P49L and A*30 identified by Fisher's exact test turned out to be a false positive.}
-  \label{fig:linkage}
-\end{figure}
-
-Figure~\ref{fig:linkage} shows regression coefficients of the HLA alleles A*30 and B*13 for replacement P49L.
-When using a simple logistic regression model without any additions (model~1), both A*30 and B*13 show medium evidence of being associated with replacement P49L.
-However, when also taking phylogeny into account and using a sparsity-promoting prior (model~3), both regression coefficients get shrunk to 0 because the association is not strongly enough identified by the data alone.
-Using epitope prediction as an additional source of information (model~4) allows to disentangle the association of the correlated alleles with P49L and identify B*13 as associated with P49L.
-The false-positive association between P49L and A*30 remains being shrunk towards 0.
 \section{Discussion}
@@
 %\section*{Acknowledgements}
-% \FloatBarrier
+
 \section*{Funding}

### 2021-03-25 docs/main.tex
a few small corrections and remarks

@@
 This assumption is likely correct for HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes.
-As presentation on MHC I is a necessary condition for the HLA system to function, most associations between an HLA allele and a replacement are likely truly zero, with the confounding factors HLA linkage disequilibrium and codon covariation, e.g.
-in the form of compensatory mutations that lie outside an epitope~\citep{Ruhl2011}.
+As presentation on MHC I is a necessary condition for the HLA system to function, most associations between an HLA allele and a replacement are likely truly zero, though sometimes confounding factors such as HLA linkage disequilibrium may lead to the appearance of sizeable associations.
+\hl{DaHo: kann man so nicht schreiben, weil die kompensatorischen Mutationen zu *echten* Assoziationen führen: ``and codon covariation,
+e.g.
+in the form of compensatory mutations that lie outside an epitope~\citep{Ruhl2011}.'' Ich finde trotzdem den Punkt der kompensatorischen Mutationen wichtig; sollte vielleicht in der Diskussion aufgegriffen werden.
+Z.B.
+ist es interessant zu sehen, ob unser Verfahren bei Positionen außerhalb von Epitopen immer noch einen Vorteil bei der Erkennung von Assoziationen liefert, oder zumindest solche Assoziationen wie in \citet{Ruhl2011} noch erkennt.}
 
@@
 \subsection{Run times and convergence}
-For a standard officie computer, run times of HAMdetector on the smaller HDV dataset were of the order of minutes and on the order of hours for the Hepatitis B dataset.
+For a standard office computer, run times of HAMdetector on the smaller HDV dataset were of the order of minutes and on the order of hours for the Hepatitis B dataset.
 For the large HIV dataset, the models were run overnight.
@@
 \begin{table}[h!]
-  \caption{List of HAMs predicted Fisher's exact test (FET).
-In the last column a ``+'' marks experimental confirmation and a ``-'' a failed confirmation.
-Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector.}
+  \caption{List of HAMs predicted by Fisher's exact test (FET).
+In the last column ``+'' and ``-'' mark successful and failed experimental confirmation, respectively.
+``?'' are untested bona fide predictions.
+``post.prob.'' are posterior probabilities for positive associations computed with HAMdetector.}
   \vspace{0.5cm}

### 2021-03-25 docs/main.tex
small changes in text

@@
 
-HAMdetector shows strong support for 15 replacement - allele pairs that have previously not been identified. All of them exceed the significance level of 
-$5\times 10^{-3}$ used in~\citet{Karimzadeh2019}.
-This shows that HAMdetector is able to identify possible HAMs that would not be detected otherwise.
-The most striking example for this is K43R - A*02, which using Fisher's exact test has a p-value of 0.22.
-While this replacement has not been experimentally verified in a IFN-\(\gamma\) production assay yet, it lies within a known A*02 epitope and thus it is highly likely that K43R conveys some form of immune escape.
+HAMdetector strongly supports 15 replacement - allele pairs that have previously not been identified (question marks in last column of Table~\ref{tab:false-positives}).
+All of them have association probabilities of 0.90 or higher, while their p-values from Fisher's exact test exceed the significance level of $5\times 10^{-3}$ used in~\citet{Karimzadeh2019}.
+Given the superior performance of HAMdetector on the experimentally tested HAMs, these 15 bona fide predictions suggest that most true HAMs may still to be discovered.
+A striking example is K43R - A*02 with a p-value of 0.22 in Fisher's exact test but a HAM-probability of 0.90 and location inside an A*02 restricted epitope.
 
-\begin{table}
-  \caption{List of HAMs predicted Fisher's exact test (FET).
-In the last column a ``+'' marks experimental confirmation and a ``-'' a failed confirmation.
-Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector.
-\emph{Bona fide} predictions are separated by a horizontal line.
-These are replacement - allele pairs we expect to to be verified in future experimental tests.
-The replacement annotated with a (+) lies within a known A*02 epitope.}
-  % \vspace{0.5cm}
+\begin{table}[h!]
+  \caption{List of HAMs predicted by Fisher's exact test (FET).
+In the last column ``+'' and ``-'' mark experimentally confirmed or rejected HAMs, respectively; ``?'' below the horizontal line indicate untested bona fide predictions.
+``post.prob.'' are posterior probabilities for positive associations computed with HAMdetector.}
+  \vspace{0.5cm}
   \begin{tabular}{c|c|c|c|c}
@@
 
- Table~\ref{tab:false-positives} shows that for 3 of the false positives identified with Fisher's exact test, HAMdetector identifies associations with the same replacement but a different allele (P49L - B*13 instead of P49L - A*30; K43R - A*02 instead of K43R - B*13; and D33E - B*13 instead of D33E - A*03).
- One possible explanation for this observation is HLA-Linkage disequilibrium and accordingly correlation between HLA alleles in the dataset due to random sampling variation.
-Indeed, out of the 12 times P49L is observed in sequences annotated with A*30, B*13 is also present in 5 of those cases (Spearman's rank correlation coefficient \(\rho = 0.5\)).
-A similar observation can be done for K43R and D33E, albeit the correlation between the respective alleles is much weaker.
+ For three of the false positives proposed by Fisher's exact test (Table~\ref{tab:false-positives}), HAMdetector identifies associations with the same replacement but a different allele (P49L - B*13 instead of P49L - A*30; K43R - A*02 instead of K43R - B*13; and D33E - B*13 instead of D33E - A*03).
+One possible explanation for this observation is HLA linkage disequilibrium \hl{DaHo: and accordingly correlation between HLA alleles in the dataset due to random sampling variation.
+(Satz etwas opak.)} Indeed, out of the 12 times P49L is observed in sequences annotated with A*30, B*13 is also present in 5 of those cases (Spearman's rank correlation coefficient \(\rho = 0.5\)).
+A similar observation can be made for K43R and D33E, although the correlation between the respective alleles is much weaker.
 
@@
   \includegraphics[width=1.1\linewidth]{plots/P49L_combined.pdf}
-  \caption{Marginal posteriors for the regression coefficients of A*30 and B*13 for replacement P49L.
-The association between P49L and A*30 identified by Fisher's exact test turned out to be a false positive.}
+  \caption{Marginal posteriors for the regression coefficients of A*30 (left column) and B*13 (right column) for replacement P49L with different model stages (rows).}
   \label{fig:linkage}
@@
 Figure~\ref{fig:linkage} shows regression coefficients of the HLA alleles A*30 and B*13 for replacement P49L.
-When using a simple logistic regression model without any additions (model~1), both A*30 and B*13 show medium evidence of being associated with replacement P49L.
-However, when also taking phylogeny into account and using a sparsity-promoting prior (model~3), both regression coefficients get shrunk to 0 because the association is not strongly enough identified by the data alone.
-Using epitope prediction as an additional source of information (model~4) allows to disentangle the association of the correlated alleles with P49L and identify B*13 as associated with P49L.
-The false-positive association between P49L and A*30 remains being shrunk towards 0.
+With the simplest logistic regression model (model~1), both A*30 and B*13 have medium evidence of being associated with replacement P49L.
+However, with phylogeny and sparsity-promoting prior (model~3) both regression coefficients shrink close to 0 -- the associations are not convincingly supported by the data.
+Using epitope prediction as additional source of information (model~4) allows to disentangle the association of the correlated alleles with P49L and identify B*13 as likely associated with P49L.
+The association between P49L and A*30 (predicted by Fisher's exact test) remains shrunk towards 0.
 \section{Discussion}

### 2021-04-06 src/alleles.jl
add src/alleles.jl

@@ -18,7 +18,7 @@ struct HLAType
         for gene in ("A", "B", "C")
             c = count(genes .== gene) 
             if c > 2
-                error("HLA type must contain at most 2 $gene alleles, got $(c)")
+                error("HLA type must contain at  2 $gene alleles, got $(c)")
             end
         end
 
@@ -130,6 +130,10 @@ function Base.isless(x::HLAAllele, y::HLAAllele)
     end
 
     return false
+end
+
+function parse_allele(::Missing)
+    return missing
 end
 
 function parse_allele(x::AbstractString)

### 2021-04-08 src/phylogenetic_tree.jl
remove raxmlHPC dependency in favor of raxml-ng

@@ -23,21 +23,21 @@ end
 function phylogenetic_tree(data::AbstractHLAData)
     !ismissing(data.tree) && return data.tree
 
-    temp_fasta = tempname() * ".fasta"
-    temp_name = basename(tempname())
+    temp_prefix = tempname()
+    temp_fasta = temp_prefix * ".fasta"
+
     write_numbered_fasta(data, temp_fasta)
 
-    model = "PROTGAMMAAUTO"
-    @suppress Base.run(`raxmlHPC -s $temp_fasta -m $model -p 53 
-                        -n $(temp_name) -w $(tempdir()) -T $(Threads.nthreads())`)
-    unrooted_tree_path = joinpath(tempdir(), "RAxML_result.$(temp_name)")
-    @suppress Base.run(`raxmlHPC -f I -m $model -t $unrooted_tree_path 
-                        -w $(tempdir()) -n $(temp_name * "_rooted") 
-                        -T $(Threads.nthreads())`)
-
-    final_tree_path = joinpath(tempdir(), "RAxML_rootedTree.$(temp_name * "_rooted")")
+    model = "LG+G+I"
+    @suppress Base.run(`raxml-ng --msa $temp_fasta --model $model --threads auto 
+        --workers auto`)
+    tree = phylogenetic_tree(readline(temp_fasta * ".raxml.bestTree"))
     
-    return phylogenetic_tree(readline(final_tree_path))
+    raxml_files = ["bestModel", "bestTree", "log", "mlTrees", "rba", "startTree"]
+    rm(temp_fasta)
+    rm.(temp_fasta .* ".raxml." .* raxml_files)
+
+    return tree
 end
 
 function write_numbered_fasta(data::AbstractHLAData, filepath::String)

### 2021-04-08 test/phylogenetic_tree.jl

@@ -74,7 +74,7 @@ end
     hla_data = HLAData("test", fasta_path, rand(HLAType, 6), missing, missing)
     tree = phylogenetic_tree(hla_data)
 
-    @test leaves(tree) == [2,4,6,8,10,11]
+    @test leaves(tree) isa Vector{Int}
 end
 
 @testset "isleaf(v::Int, tree::Phylogenetic)" begin
@@ -82,8 +82,7 @@ end
     hla_data = HLAData("test", fasta_path, rand(HLAType, 6), missing, missing)
     tree = phylogenetic_tree(hla_data)
 
-    @test isleaf(2, tree)
-    @test !isleaf(1, tree)
+    @test isleaf(leaves(tree)[1], tree)
     @test !isleaf(90, tree)
 end
 
@@ -91,10 +90,12 @@ end
     fasta_path = joinpath(@__DIR__, "data", "phylogeny.fasta")
     hla_data = HLAData("test", fasta_path, rand(HLAType, 6), missing, missing)
     tree = phylogenetic_tree(hla_data)
-
+    tree_leaves = Escape.leaves(tree)
     @test get_property(tree, 1, :name) == "root"
-    @test get_property(tree, 2, :name) == get_prop(tree.graph, 2, :name)
-    @test get_property(tree, 10, :name) == get_prop(tree.graph, 10, :name)
+    @test get_property(tree, tree_leaves[1], :name) == 
+        get_prop(tree.graph, tree_leaves[1], :name)
+    @test get_property(tree, tree_leaves[1], :name) == 
+        get_prop(tree.graph, tree_leaves[1], :name)
     @test ismissing(get_property(tree, 100, :name))
     @test ismissing(get_property(tree, 1, :nonexisting))
 end

### 2021-05-03 src/hla_model_result.jl
circumvent issue in stan input data for blockwise runs in replacement_summary(::HLAModelResult)

@@ -24,6 +24,7 @@ function replacement_summary(
         n_replacement_with_hla = Int[]
     )
 
+    # using length(rs) to circumvent wrong data for blockwise runs
     R = si["R"]
     D = si["D"]
     Z = si["Z"]
@@ -32,7 +33,7 @@ function replacement_summary(
         fisher = Escape.run(Escape.FisherTest(), data; fdr = fdr)
     end
 
-    for r in 1:R
+	for r in 1:length(rs)
         fisher_idx = fisher_p ? findfirst(x -> x == rs[r], fisher.replacements) : 0
 
         for d in 1:D
@@ -41,6 +42,12 @@ function replacement_summary(
             p_value = fisher_p ? fisher.p_values[fisher_idx][alleles[d]] : 0
             n_total = fisher_p ? sum(fisher.counts[fisher_idx][alleles[d]][1, :]) : 0
             n_with_hla = fisher_p ? fisher.counts[fisher_idx][alleles[d]][1, 1] : 0
+			
+			if r > size(Z, 1)
+				z = 0
+			else 
+				z = Z[r, d]
+			end
 
             push!(df, 
                 [ 
@@ -51,7 +58,7 @@ function replacement_summary(
                     lower,
                     upper,
                     p_value,
-                    Z[r, d],
+                    z,
                     n_total,
                     n_with_hla
                 ]

### 2021-05-03 docs/main.tex
remove latex math statement B\ast15 that prevents bibtex from compiling, comment out some highlights

@@
   Our general approach for HAMdetector is to fit Bayesian regression models that captures relationships between host HLA alleles and replacements in viral proteomes.
+
   This Bayesian approach is advantageous because it allows use of: (1) prior information (e.g.\ knowledge of effect magnitudes), (2) relevant additional information (phylogeny, epitope information), (3) a problem-specific structure, (4) partial pooling~\citep{Gelman2010}.
@@
 This assumption is likely correct for HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes.
-As presentation on MHC I is a necessary condition for the HLA system to function, most associations between an HLA allele and a replacement are likely truly zero, though sometimes confounding factors such as HLA linkage disequilibrium may lead to the appearance of sizeable associations.
-\hl{DaHo: kann man so nicht schreiben, weil die kompensatorischen Mutationen zu *echten* Assoziationen führen: ``and codon covariation,
-e.g.
+As presentation on MHC I is a necessary condition for the HLA system to function, most associations between an HLA allele and a replacement are likely truly zero, with the exception of other factors such as HLA linkage disequilibrium and codon covariation, e.g.
+in the form of compensatory mutations that lie outside and epitope~\citep{Ruhl2011}.
+% \hl{DaHo: kann man so nicht schreiben, weil die kompensatorischen Mutationen zu *echten* Assoziationen führen: ``and codon covariation,
+% e.g.
 in the form of compensatory mutations that lie outside an epitope~\citep{Ruhl2011}.'' Ich finde trotzdem den Punkt der kompensatorischen Mutationen wichtig; sollte vielleicht in der Diskussion aufgegriffen werden.
@@
  For three of the false positives proposed by Fisher's exact test (Table~\ref{tab:false-positives}), HAMdetector identifies associations with the same replacement but a different allele (P49L - B*13 instead of P49L - A*30; K43R - A*02 instead of K43R - B*13; and D33E - B*13 instead of D33E - A*03).
-One possible explanation for this observation is HLA linkage disequilibrium \hl{DaHo: and accordingly correlation between HLA alleles in the dataset due to random sampling variation.
+One possible explanation for this observation is HLA linkage disequilibrium and accordingly correlation between HLA alleles in the dataset due to random sampling variation.
+
+ %\hl{DaHo: and accordingly correlation between HLA alleles in the dataset due to random sampling variation.
 (Satz etwas opak.)} Indeed, out of the 12 times P49L is observed in sequences annotated with A*30, B*13 is also present in 5 of those cases (Spearman's rank correlation coefficient \(\rho = 0.5\)).
@@
 The association between P49L and A*30 (predicted by Fisher's exact test) remains shrunk towards 0.
+
+\subsection{HAMs outside epitopes}
+\label{sec:hams-outs-epit}
+
+imperfect epitope / processing predictions
+
+compensatory mutations
+
+we expect HAMs outside epitopes - are there such HAMs?
+
+Figure
+
 \section{Discussion}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
+% %
+% %     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
+% %     as it may ignore the figures.
+% %
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%\section*{Acknowledgements}
+\section*{Acknowledgements}
 % \FloatBarrier
@@
 \bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
+% \bibliographystyle{achemnat}
 %\bibliographystyle{plainnat}
 %\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
+% \bibliographystyle{bioinformatics}
 %
-%\bibliographystyle{plain}
+% \bibliographystyle{plain}
 %

### 2021-05-04 docs/main.tex
update manuscript with section on HAMs outside epitopes

@@
 \usepackage{soul}
+\usepackage{graphicx}
 \usepackage[paper=b4paper, layout=b4paper]{geometry}
+% \usepackage{float}
+\usepackage[section]{placeins}
 \usepackage{float}
-% \usepackage[section]{placeins}
+\usepackage[export]{adjustbox}
+% \multicols{2}
+% \usepackage{minipa}
+% \usepackage{dblfloatfix}
 
@@
 
-For a first posterior predictive check we used \emph{calibration plots}: two binned quantities were plotted against each other, the observed relative frequencies of replacements $f(y_{ik}=1)$, and the predicted probabilities $P(y_{ik}=1|\text{model})$.
+For a first posterior predictive check we used \emph{calibration plots} (Fig.
+\ref{fig:calibration}): two binned quantities were plotted against each other, the observed relative frequencies of replacements $f(y_{ik}=1)$, and the predicted probabilities $P(y_{ik}=1|\text{model})$.
 In such a plot, a well-calibrated model should yield points following the diagonal.
@@
 \begin{figure}
- \includegraphics[width=1.0\linewidth]{plots/phylogeny_calibration.pdf}  
-  \caption{Calibration plot for the HBV core protein.}
+ \includegraphics[width=1.0\linewidth]{plots/calibration_plot.pdf}  
+  \caption{Calibration plot for the HBe antigen.}
   \label{fig:calibration}
@@
 
-imperfect epitope / processing predictions
+The epitope and processing predictions that HAMdetector uses are imperfect, as the underlying tools extrapolate binding affinities for new epitopes based on experimental data.
+Bayesian statistics provides a coherent framework to make use of imperfect data.
+In HAMdetector, this is achieved by an additional parameter \(\beta_{\text{epi}}\) that governs how strongly the model takes an apparent association between a replacement and the corresponding HLA allele into account.
+By default, the regression coefficients that quantify the strength of association between allele and replacement are shrunk towards 0, and only in the presence of considerable evidence in favor of an association (e.g.
+because the replacement often co-occurs with a certain HLA allele), this shrinkage is overcome by the observed data.
 
-compensatory mutations
+If the epitope prediction happens to be reliable, i.e.
+when the presence of a predicted epitope correlates strongly with the probability observing the replacement in a host with the respective HLA allele, the parameter \(\beta_{\text{epi}}\) is estimated to be large and less evidence by the sequence data is enough to escape the shrinkage and estimate a non-zero association between allele and replacement, compared to associations that do not lie inside a predicted epitope.
+Likewise, if the epitope prediction turns out to be non-reliable, \(\beta_{\text{epi}}\) is estimated to be close to 0 and the presence of a predicted epitope does not strongly affect the conclusions drawn from the sequence data.
 
-we expect HAMs outside epitopes - are there such HAMs?
+However, it is important to consider that biologically relevant HLA escape mutations do not necessarily have to lie within or close to the boundary of an epitope.
+Compensatory mutations \citep{Kelleher2001, Ruhl2011, NeumannHaefelin2011} can confer a strong selective advantage, e.g.
+by partially restoring replicative capacity that would otherwise be impaired by the exclusive presence of a certain HLA escape mutation.
+As they also facilitate HLA escape and are associated with HLA alleles in a sequence alignment, they can also be considered as HLA escape mutations.
+In a sequence alignment, compensatory mutations can occur far away from the epitope they are associated with, as they might be the result of improved physical interactions with another amino acid in the folded, three-dimensional protein.
 
-Figure
+We therefore also expect HAMs outside epitopes and one possible concern is that the model focuses too strongly on associations with replacements that lie within the boundary of predicted epitopes.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.95\linewidth]{plots/m3_m4_HBe.pdf}
+  \caption{Integral of the marginal posterior $P(\beta_{jk}>0)$ for the HAMdetector model with epitope prediction (model 4) and without epitope prediction (model 3).}
+  \label{fig:m3m4}
+\end{figure}
+
+Figure~\ref{fig:m3m4} shows posterior probabilities \(P(\beta_{jk}>0)\) for replacement--HLA allele pairs as calculated by HAMdetector with (model 4) and without (model 3) epitope prediction.
+Each replacement--HLA allele pair is represented by a dot and colored according to whether or not that position lies within a predicted epitope.
+For replacements that do not lie within a predicted epitope, both models provide similar estimates (points across the diagonal).
+However, some replacements--HLA pairs that have only weak evidence of association in model 3 have strong support in model 4, which is explained by the additional evidence provided by epitope prediction.
+The figure shows that the model is still able to identify associations outside predicted epitopes and that epitope information augments evidence obtained from sequence data.
 
@@
 
-\section*{Acknowledgements}
-% \FloatBarrier
+% \section*{Acknowledgements}
 \section*{Funding}

### 2021-05-14 src/Escape.jl
remove run_blockwise

@@ -19,7 +19,6 @@ include("phylogenetic_background.jl")
 include("epitope_prediction.jl")
 include("epitope_map.jl")
 include("hla_model.jl")
-include("hla_model_blockwise.jl")
 include("fisher_test.jl")
 include("hla_model_result.jl")
 include("hla_model_result_io.jl")

### 2021-05-14 test/loo.jl

@@ -2,7 +2,7 @@
     if !isfile(joinpath(@__DIR__, "data", "result_loo.jls"))
         result = @suppress Escape.run(
             Escape.HLAModel{4}(), Escape.HLADataset("Test").data[1], 
-            mincount = 1, iter = 200, chains = 4
+            mincount = 1, iter = 2000, chains = 4
         )
 
         serialize(joinpath(@__DIR__, "data", "result_loo.jls"), result)

### 2021-05-14 test/runtests.jl

@@ -27,7 +27,6 @@ include("hla_model_result.jl")
 include("loo.jl")
 include("epitope_map.jl")
 include("hla_model.jl")
-include("hla_model_blockwise.jl")
 include("fisher_test.jl")
 include("hla_model_result_io.jl")
 include("posterior_predictive_checks.jl")

### 2021-05-14 manuscript/manuscript.tex
rename docs folder to manuscript

--- 2021-02-14 docs/manuscript.tex
@@
 \usepackage{listings}
-\usepackage{fontspec}
-\usepackage{hyperref}
+\usepackage[no-math]{fontspec}
 \usepackage{amsmath}
@@
 
-\hl{DaHo: Da es ein Methoden-Paper ist, sollte dies der Hauptteil sein}
-
 Our general approach is to fit a Bayesian regression model that captures relationships between host HLA alleles and replacements in viral proteomes.
@@
   y_{ik} & \sim \text{Bernoulli}(\theta_{ik}) \\
-  \theta_{ik} & = \text{logistic}(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}}),
+  \theta_{ik} & = \text{logistic}\left(\beta_{0_{k}} + \sum_{j=1}^{D} X_{ij}\beta_{\text{HLA}_{jk}}\right),
 \end{align}
@@
 
-\subsubsection{Model implementation}
+\subsection{Model implementation}
 All models were implemented in Stan~\cite{Stan2021}.
@@
 
-The model was fit to several data sets:
+The model was fit with several data sets:
 

### 2021-05-14 test/loo.jl
fix broken unit test for Loo

@@ -1,14 +1,8 @@
 @testset "Loo" begin
-    if !isfile(joinpath(@__DIR__, "data", "result_loo.jls"))
-        result = @suppress Escape.run(
-            Escape.HLAModel{4}(), Escape.HLADataset("Test").data[1], 
-            mincount = 1, iter = 2000, chains = 4
-        )
-
-        serialize(joinpath(@__DIR__, "data", "result_loo.jls"), result)
-    end
-
-    result = deserialize(joinpath(@__DIR__, "data", "result_loo.jls"))
+    result = @suppress Escape.run(
+        Escape.HLAModel{4}(), Escape.HLADataset("Test").data[1], 
+        mincount = 1, iter = 10, warmup = 10, chains = 4, seed = 123
+    )
 
     sf = Escape.stanfit(result)
     p = extract(sf)

### 2021-05-14 src/hla_model_blockwise.jl
fix merge conflict

@@ -1,92 +1,0 @@
-function run_blockwise(model::HLAModel{T}, data::AbstractHLAData;
-    mincount::Int = 1,
-    depth::Int = 1,
-    partitions = Iterators.partition(replacements(data, mincount = mincount), 100),
-    wp::AbstractWorkerPool = default_worker_pool(),
-    stan_kwargs...
-) where T
-
-    input = stan_input(model, data, depth = depth, mincount = mincount)
-    alleles = sort(unique_alleles(hla_types(data), depth = depth))
-    
-    sf_files = pmap(x -> run_blockwise_(model, data, x; 
-        mincount = mincount, depth = depth, wp = wp, stan_kwargs...), wp, partitions)
-    sf_1 = deserialize(sf_files[1]) # used to access iter, chains field etc.
-    combined_d = [Dict{String, Vector{Float64}}() for i in 1:sf_1.chains]
-
-    i = 0
-    for file in sf_files
-        sf = deserialize(file)
-        for r_idx in 1:sf.data["R"]
-            i = i + 1
-            for d in 1:sf.data["D"]
-                for (res_idx, res) in enumerate(sf.result)
-                    if "b_phy.1" in keys(res)
-                        combined_d[res_idx]["b_phy.$i"] = res["b_phy.$r_idx"]
-                    end
-                    combined_d[res_idx]["b0_hla.$i"] = res["b0_hla.$r_idx"]
-                    combined_d[res_idx]["beta_hla.$i.$d"] = res["beta_hla.$r_idx.$d"]
-                end
-            end
-        end
-    end
-
-    combined_sf = StanInterface.Stanfit(sf_1.model, sf_1.data, sf_1.iter, sf_1.chains,
-        combined_d, sf_1.diagnostics)
-
-    rm.(sf_files)
-
-    return HLAModelResult(model, data, combined_sf, vcat(partitions...), alleles)
-end
-
-function run_blockwise_(model::HLAModel{T}, data::AbstractHLAData, replacements;
-    mincount, depth, wp, stan_kwargs...
-) where T
-
-    input = stan_input(model, data, depth = depth)
-    alleles = sort(unique_alleles(hla_types(data), depth = depth))
-    all_replacements = Escape.replacements(data, mincount = mincount)
-    
-    idx_(replacement) = findfirst(x -> x == replacement, all_replacements)
-    r_idx = map(idx_, replacements)
-    
-    y = Int[]
[…]

### 2021-05-14 test/hla_model_blockwise.jl

@@ -1,21 +1,0 @@
-@testset "Escape.run_blockwise(::HLAModel, ::AbstractHLAData)" begin
-    ds = Escape.HLADataset("Test")
-    m = Escape.HLAModel{4}()
-    data = ds.data[1]
-    r = Escape.replacements(data)
-
-    partitions = Iterators.partition(r, 5)
-    res_blockwise = @suppress Escape.run_blockwise(m, data, partitions = partitions,
-        iter = 5, warmup = 5, chains = 1)
-    res = @suppress Escape.run(m, data, keep_all_parameters = true, iter = 5, warmup = 5, 
-        chains = 1)
-
-    for key in keys(res.sf.result[1])
-        if startswith(key, "beta_hla")
-            @test key in keys(res_blockwise.sf.result[1])
-        end
-    end
-
-    @test res_blockwise = @suppress Escape.run_blockwise(Escape.HLAModel{1}(), data, 
-        partitions = partitions, iter = 5, warmup = 5, chains = 1) isa Escape.HLAModelResult
-end

### 2021-05-15 docs/make.jl
add Documenter.jl folder structure

@@ -1,0 +1,15 @@
+using Documenter
+using Escape
+
+makedocs(
+    sitename = "Escape",
+    format = Documenter.HTML(),
+    modules = [Escape]
+)
+
+# Documenter can also automatically deploy documentation to gh-pages.
+# See "Hosting Documentation" and deploydocs() in the Documenter manual
+# for more information.
+#=deploydocs(
+    repo = "<repository url>"
+)=#
